### 2017-06-08 manuscript.tex
very rough partial outine

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

\usepackage[round,semicolon]{natbib}

% Use times if you have the font installed; otherwise, comment out the
% following line.

\usepackage{times}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{seqsplit}

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\usepackage{newfloat}
\usepackage{amsmath}
\usepackage[labelfont=bf]{caption}
\usepackage{nameref}
\usepackage{rotating}
\usepackage{color}
\usepackage{float}

\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}

\renewcommand{\figurename}{{}}
\renewcommand{\thefigure}{{Figure \arabic{figure}}}

\renewcommand{\tablename}{{}}
\renewcommand{\thetable}{{Table \arabic{table}}}

\newfloat{suppfig}{thp}{lofsupfig}
\renewcommand{\thesuppfig}{Supplementary figure \arabic{suppfig}}
\floatname{suppfig}{}

\newfloat{supptab}{thp}{lofsuptab}
\renewcommand{\thesupptab}{Supplementary table \arabic{supptab}}
\floatname{supptab}{}

\newfloat{suppfile}{thp}{lofsupfile}
\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
\floatname{suppfile}{}

\renewcommand{\theequation}{Equation \arabic{equation}}

\definecolor{darkblue}{rgb}{0, 0.0, 0.6}

\usepackage{hyperref}
\hypersetup{colorlinks,citecolor=blue,linkcolor=darkblue,urlcolor=blue}

\usepackage{seqsplit}

\usepackage{array}
\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\newcommand{\comment}[1]{{\color{red}(\textsl{#1})}}
\newcommand{\mikecomment}[1]{{\color{magenta}(\textsl{#1})}}

\usepackage{setspace}

\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{1}

%\renewcommand{\abstractname}{\large SUMMARY}

\title{Mapping sites of shifting and constant mutational effects on the evolutionary landscape of HIV Envelope}

\author
{Hugh K. Haddox$^{1,2}$ , Adam S. Dingens$^{1,2}$, Julie Overbaugh and Jesse D. Bloom$^{1,2,5}$\\
\footnotesize{$^1$Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}\\
\footnotesize{$^2$Molecular and Cellular Biology PhD program, University of Washington, Seattle, WA} \\
\footnotesize{E-mail: \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
HIV Envelope (Env) evolves rapidly.
The immediate the evolutionary space accessible to any viral variant is largely determined by the effects of single amino-acid mutations.
However, the effects of these mutations can shift over evolutionary time as the virus traverses through sequence space.
Here we comprehensively quantify the ways in which mutational effects shift or remain constant by experimentally measuring the effects of all amino-acid mutations to two homologs of Env with 85\% (?) sequence identity.
We find...
\end{abstract}

\section*{Introduction}

\section*{Results}

\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder (?) viruses}
Previously we've done deep mutational scanning on a lab-passaged CXCR4-tropic virus isolated late in infection.
Here we sought to examine the effects of mutations in viruses more relevant to selection on HIV in nature.
We chose...

We made an alignment of group M sequences (\ref{suppfile:grpM}) and clade A (\ref{suppfile:cladeA})
The relationship between BG505 and BF520 is shown in \ref{ref:tree}

\begin{figure}
%\centerline{\includegraphics[width=5in]{?}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship among BG505 and BF520.
{\bf(A)} Alignment of group M sequences.
{\bf(B)} Zooms in on clade A sequences, shows more.
Both use alignments for actual analyses with \texttt{phydms}.}
\end{figure}

We performed deep mutational scanning as described previously with a few modifications that are very briefly described here and detailed in the methods.
We saw strong selection as indicated in \ref{fig:mutfreqs}.
More details on DMS is shown in \ref{suppfig:dms}

\begin{figure}
%\centerline{\includegraphics[width=5in]{?}}
\caption{\label{fig:mutfreqs}
Mutation frequencies before and after selection after subtracting off stop codons.
Correlations among the 2 pairs of replicates within each homolog.
So this figure has something like 4 panels.
}
\end{figure}

\subsection*{Re-scaling the experimental measurements to optimally describe HIV evolution in nature}

Re-scaled as in \ref{tab:phydms}

\begin{table}
\caption{\label{tab:phydms}
Results of \texttt{phydms} analysis with each set of averaged homologs.
Maybe two nested table showing results for both clade A and group M.
Show results for individual replicates as a supptab.
}
\end{table}

A figure with the re-scaled average preferences for both BG505 and BF520. Wildtype under them.

A supplemental file for each of these and the results for each replicate. This is a lot of files, so you might just make a zip with all eight.

\subsection*{Differences between homologs}

\begin{itemize}
\item I like your heat map of correlation coefficients with HA included.
\item You define a distance metric with is half sum absolute corrected. A little example plot of how this is calculated
\item Also as the same figure subpanel or a supplemental figure perhaps show that this is highly correlated with the other two distance measures (just show two correlation plots), and then that's probably enough
\item As a new figure or subpanel you'd have the histogram -- I think it would be cool if you could actually overlay the histogram of Env with HA. Either overlay or align with same scales. 
\item For a summary figure showing the difference, I think you should do the following: (1) For each site, make the total height of the letter stack in each direction equal to the your corrected distance measure (the same thing in the histogram). (2) The height of each letter is equal to the difference in preferences of the average for each homolog. Plot overlays of conserved and variable, and surface versus core.
\end{itemize}

\subsection*{Validating sites of mutational shifts}
Describe and validate a few of these sites

\subsection*{Looking at how shifts relate to evolution and structure}

\section*{Discussion}

\section*{Methods}

\subsection*{Sequence alignments}

\subsection*{Acknowledgments}

\bibliographystyle{mbe}
\bibliography{references.bib}

\clearpage
\normalsize

\section*{Supplementary Material}

\begin{supptab}
\caption{\label{supptab:phydms_replicates}
phydms analysis for all replicates
}
\end{supptab}

\begin{suppfig}
\caption{\label{suppfig:dms}
some plots about DMS
}
\end{suppfig}

\begin{suppfile}
\caption{\label{suppfile:grpM}
Alignment of all group M sequences numbered and aligned as ???
}
\end{suppfile}

\begin{suppfile}
\caption{\label{suppfile:cladeA}
Alignment of clade A sequences...
}
\end{suppfile}

\end{document}

### 2017-09-25 manuscript.tex
Reformatted manuscript for submitting to eLife and adding files of figures from thesis

@@
-% Use only LaTeX2e, calling the article.cls class and 12-point type.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ELIFE ARTICLE TEMPLATE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% PREAMBLE 
+\documentclass[9pt,lineno]{elife}
+% Use the onehalfspacing option for 1.5 line spacing
+% Use the doublespacing option for 2.0 line spacing
+% Please note that these options may affect formatting.
 
-\documentclass[12pt]{article}
+\usepackage{lipsum} % Required to insert dummy text
+\usepackage[version=4]{mhchem}
+\usepackage{siunitx}
+\DeclareSIUnit\Molar{M}
 
-\usepackage[round,semicolon]{natbib}
-
-% Use times if you have the font installed; otherwise, comment out the
-% following line.
-
-\usepackage{times}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{seqsplit}
-
-% The following parameters seem to provide a reasonable page setup.
-
-\topmargin 0.0cm
-\oddsidemargin 0.2cm
-\textwidth 16cm 
-\textheight 21cm
-\footskip 1.0cm
-
-\usepackage{newfloat}
-\usepackage{amsmath}
-\usepackage[labelfont=bf]{caption}
-\usepackage{nameref}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{float}
-
-\setcounter{topnumber}{8}
-\setcounter{bottomnumber}{8}
-\setcounter{totalnumber}{8}
-
-\renewcommand{\figurename}{{}}
-\renewcommand{\thefigure}{{Figure \arabic{figure}}}
-
-\renewcommand{\tablename}{{}}
-\renewcommand{\thetable}{{Table \arabic{table}}}
-
-\newfloat{suppfig}{thp}{lofsupfig}
-\renewcommand{\thesuppfig}{Supplementary figure \arabic{suppfig}}
-\floatname{suppfig}{}
-
-\newfloat{supptab}{thp}{lofsuptab}
-\renewcommand{\thesupptab}{Supplementary table \arabic{supptab}}
-\floatname{supptab}{}
-
-\newfloat{suppfile}{thp}{lofsupfile}
-\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
-\floatname{suppfile}{}
-
-\renewcommand{\theequation}{Equation \arabic{equation}}
-
-\definecolor{darkblue}{rgb}{0, 0.0, 0.6}
-
-\usepackage{hyperref}
-\hypersetup{colorlinks,citecolor=blue,linkcolor=darkblue,urlcolor=blue}
-
-\usepackage{seqsplit}
-
-\usepackage{array}
-\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\newcommand{\comment}[1]{{\color{red}(\textsl{#1})}}
-\newcommand{\mikecomment}[1]{{\color{magenta}(\textsl{#1})}}
-
-\usepackage{setspace}
-
-\renewcommand{\topfraction}{1}
-\renewcommand{\bottomfraction}{1}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{1}
-
-%\renewcommand{\abstractname}{\large SUMMARY}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ARTICLE SETUP
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{Mapping sites of shifting and constant mutational effects on the evolutionary landscape of HIV Envelope}
 
-\author
-{Hugh K. Haddox$^{1,2}$ , Adam S. Dingens$^{1,2}$, Julie Overbaugh and Jesse D. Bloom$^{1,2,5}$\\
-\footnotesize{$^1$Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}\\
-\footnotesize{$^2$Molecular and Cellular Biology PhD program, University of Washington, Seattle, WA} \\
-\footnotesize{E-mail: \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-}
+\author[1,2\authfn{1}]{Hugh K. Haddox}
+\author[1,2\authfn{1}]{Adam S. Dingens}
+\author[3]{Julie Overbaugh}
+\author[1,2,4]{Jesse D. Bloom}
+\affil[1]{Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
+\affil[2]{Molecular and Cellular Biology PhD program, University of Washington, Seattle, WA}
+\contrib[\authfn{1}]{These authors contributed equally to this work}
 
-\date{}
+\corr{jbloom@fredhutch.org}{JDB}
+% \presentadd[\authfn{5}]{eLife Sciences editorial Office, eLife Sciences, Cambridge, United Kingdom}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ARTICLE START
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 
-\section*{Results}
+\section{Results}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
 
-\section*{Methods}
+\section{Methods and Materials}
 
@@
 
-\subsection*{Acknowledgments}
+\section{Acknowledgments}
 
-\bibliographystyle{mbe}
+\nocite{*} % This command displays all refs in the bib file
 \bibliography{references.bib}
 
-\clearpage
-\normalsize
-
 \section*{Supplementary Material}
-
+% figure out how to do the supplement for eLife
+% then remove the `\iffalse` and `\fi` commands below
+\iffalse
 \begin{supptab}
@@
 \end{suppfile}
+% remove the below line when figure out how to format the supplement correctly
+\fi
 

### 2017-09-25 manuscript.tex
Added figures from thesis

@@
 \begin{figure}
-%\centerline{\includegraphics[width=5in]{?}}
+\centerline{\includegraphics[width=5in]{figures/tree/tree}}
 \caption{\label{fig:tree}
 Phylogenetic tree showing the relationship among BG505 and BF520.
-{\bf(A)} Alignment of group M sequences.
-{\bf(B)} Zooms in on clade A sequences, shows more.
-Both use alignments for actual analyses with \texttt{phydms}.}
+{\bf(A)} A tree of group-M sequences with 20 sequences per subtype for subtypes A, B, C, D, F, and G.
+{\bf(B)} A tree of subtype-A sequences with 120 sequences total, rooted using a subtype-B sequence as an outgroup.
+BG505 and BF520 cluster within subtype A, while LAI clusters within subtype B.
+The alignments we used to make these trees are the same alignments we used in the below \texttt{phydms} analysis, excluding the outgroup sequence used to root the subtype-A tree.
+The scale bars are in units of codon substitutions per site.
+}
 \end{figure}
@@
 \begin{figure}
-%\centerline{\includegraphics[width=5in]{?}}
+\centerline{\includegraphics[width=3in]{figures/dms_schematic/dms_schematic}}
+\caption{\label{fig:dms_schematic}
+Deep mutational scanning workflow.
+{\bf (A)} For each homolog, we made a library of proviral HIV plasmids with random codon-level mutations in \textit{env}.
+We then transfected the plasmids into 293T cells to generate mutant viruses, which may lack a genotype-phenotype link since transfected cell are expected to each receive multiple plasmids.
+To establish this link and select for functional variants, we first passaged the libraries in SupT1 cells at a low MOI.
+Then, we imposed a second round of selection by infecting the passaged viruses into SupT1 cells at a high MOI and then harvesting reverse-transcribed unintegrated viral DNA at 12 hours post-infection.
+Finally, we deep sequenced the libraries before and after selection.
+We also deep sequenced wildtype controls to estimate error rates due to PCR, deep sequencing, and viral replication.
+Using these sequencing data, we then inferred each site's preference for each of the 20 amino acids.
+{\bf (B)} For each homolog, we conducted this experiment in full biological triplicate, beginning at the stage of independently creating the plasmid mutant libraries.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=5in]{figures/mutfreqs/mutfreqs}}
 \caption{\label{fig:mutfreqs}
-Mutation frequencies before and after selection after subtracting off stop codons.
-Correlations among the 2 pairs of replicates within each homolog.
-So this figure has something like 4 panels.
+The deep mutational scanning experiments imposed strong purifying selection and led to reproducible estimates of each homolog's amino-acid preferences.
+{\bf(A)} For each replicate of each homolog, we deep sequenced the starting plasmid libraries (DNA) and the selected virus libraries (virus).
+Bars show the per-codon mutation frequency averaged across all sites after subtracting mutation frequencies determined using wildtype controls (Fig \ref{suppfig:ch3_mutfreqs_supp}).
+Mutations leading to stop codons were purged during the selection step to 3-17\% their starting frequencies (see red numbers), indicating strong purifying selection.
+Nonsynonymous mutations decreased to 40-50\% their starting frequencies, consistent with a large fraction of amino-acid mutations being deleterious to Env.
+Synonymous mutations only decreased to 87-95\% their starting frequencies, as would be expected if synonymous mutations tend to have more neutral effects than nonsynonymous changes.
+Panels {\bf(B)} and {\bf(C)} show the correlation in our estimates of each homolog's site-specific amino-acid preferences between experimental replicates, with {\bf(B)} showing the correlation between BF520 replicates and {\bf(C)} showing the correlation between BG505 replicates.
+Each correlation plot reports the associated Pearson correlation coefficient, which ranged from 0.59-0.75, indicating that our estimates are largely repeatable between replicates.
 }
+%\figsupp{test text}{\includegraphics[width=5cm]{figures/mutfreqs_supp/mutfreqs_supp}}
 \end{figure}
@@
 \begin{table}
-\caption{\label{tab:phydms}
-Results of \texttt{phydms} analysis with each set of averaged homologs.
-Maybe two nested table showing results for both clade A and group M.
-Show results for individual replicates as a supptab.
+\centering
+\caption{
+{\bf Phylogenetic models that incorporate Env's preferences indicate that selection was less stringent in the lab than in nature.}}
+{\footnotesize
+\begin{tabular}{cccc}
+\multicolumn{4}{c}{\bf{Group M}}\\
+\hline
+model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
+\hline
+BG505 & 0.00 & $-$58185.53 & 7: $\beta$ = 2.07, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.49, $\kappa$ = 3.16\\
+BF520 & 70.26 & $-$58220.66 & 7: $\beta$ = 2.44, $\alpha_\omega$ = 0.71, $\beta_\omega$ = 0.47, $\kappa$ = 3.11\\
+LAI & 2994.80 & $-$59682.93 & 7: $\beta$ = 1.00, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.46, $\kappa$ = 3.09\\
+BF520 site averaged & 3913.08 & $-$60142.07 & 7: $\beta$ = 2.24, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
+BG505 site averaged & 3922.08 & $-$60146.57 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
+LAI site averaged & 4083.20 & $-$60227.13 & 7: $\beta$ = 0.83, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.2\\
+YNGKP M5 & 4423.42 & $-$60392.24 & 12: $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.55, $\kappa$ = 3.20\\
+\\
+\multicolumn{4}{c}{\bf{Subtype A}}\\
+\hline
+model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
+\hline
+BF520 & 0.00 & $-$48166.45 & 7: $\beta$ = 2.80, $\alpha_\omega$ = 0.59, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
+BG505 & 234.70 & $-$48283.8 & 7: $\beta$ = 2.19, $\alpha_\omega$ = 0.58, $\beta_\omega$ = 0.32, $\kappa$ = 3.50\\
+LAI & 2860.64 & $-$49596.77 & 7: $\beta$ = 1.06, $\alpha_\omega$ = 0.48, $\beta_\omega$ = 0.30, $\kappa$ = 3.41\\
+BG505 site averaged & 3909.68 & $-$50121.29 & 7: $\beta$ = 0.93, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BF520 site averaged & 3913.70 & $-$50123.3 & 7: $\beta$ = 1.22, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+LAI site averaged & 3961.18 & $-$50147.04 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
+YNGKP M5 & 3980.00 & $-$50151.45 & 12: $\alpha_\omega$ = 0.42, $\beta_\omega$ = 0.42, $\kappa$ = 3.62
+\end{tabular}
 }
+\begin{flushleft}
+We used \texttt{phydms} to incorporate each homolog's averaged preferences into ExpCMs.
+This table shows the results of using these models to optimize the branch lengths of the group-M and subtype-A trees shown in Fig \ref{fig:ch3_tree}.
+For both trees, the ExpCMs dramatically outperformed a standard codon-substitution model (YNGKP M5), based on differences in Akaike information criterion (AIC), with BG505's and BF520's preferences describing Env's evolution better than LAI's preferences.
+The stringency parameters ($\beta$) inferred for BG505 and BF520 were both  $>$1, indicating that selection is more stringent in nature than our experiments for these homologs.
+In contrast, the stringency parameters inferred for LAI were $\sim$1, which may be due to experimental or biological differences, as discussed in the main text.
+As a control, for each homolog, we created an ExpCMs using preferences averaged across all sites in the protein.
+This averaging substantially decreased phylogenetic fit, indicating that the preferences capture site-specific selective pressures on Env in nature.
+The results for individual experimental replicates are shown in Tables \ref{tab:phydms_groupM_supp} and \ref{tab:phydms_subtypeA_supp}.
+\end{flushleft}
+\label{tab:phydms}
 \end{table}
 
-A figure with the re-scaled average preferences for both BG505 and BF520. Wildtype under them.
+\begin{figure}
+\centerline{\includegraphics[width=6.5in]{figures/prefs_BG505/prefs_BG505.png}}
+\caption{\label{fig:bg505_prefs_logo}
+The rescaled averaged site-specific amino-acid preferences for BG505.
+This logo plot shows the site-specific amino-acid preferences for BG505 after averaging between replicates and then rescaling them using the stringency parameter from Table \ref{tab:phydms} inferred for BG505 from the group-M alignment.
+Each site has a stack of 20 letters corresponding to the 20 amino acids.
+Letter heights, which sum to one at each site, are proportional to the site's preference for each amino acid.
+The top bar (R) indicates gp120 variable loops, other regions in gp120, or gp41.
+The bottom bar (WT) shows the wildtype amino-acid sequence for BG505.
+Sites are numbered according to the HXB2 numbering scheme~\cite{korber1998numbering}).
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=6.5in]{figures/prefs_BF520/prefs_BF520.png}}
+\caption{\label{fig:bf520_prefs_logo}
+The re-scaled averaged site-specific amino-acid preferences for BF520.
+This figure is the same as Fig \ref{fig:bg505_prefs_logo}, but for BF520 instead of BG505.
+}
+\end{figure}
 
@@
 
-\begin{itemize}
-\item I like your heat map of correlation coefficients with HA included.
-\item You define a distance metric with is half sum absolute corrected. A little example plot of how this is calculated
-\item Also as the same figure subpanel or a supplemental figure perhaps show that this is highly correlated with the other two distance measures (just show two correlation plots), and then that's probably enough
-\item As a new figure or subpanel you'd have the histogram -- I think it would be cool if you could actually overlay the histogram of Env with HA. Either overlay or align with same scales. 
-\item For a summary figure showing the difference, I think you should do the following: (1) For each site, make the total height of the letter stack in each direction equal to the your corrected distance measure (the same thing in the histogram).
-(2) The height of each letter is equal to the difference in preferences of the average for each homolog.
-Plot overlays of conserved and variable, and surface versus core.
-\end{itemize}
+\begin{figure}
+\centerline{\includegraphics[width=6.5in]{figures/homologs_corr/homologs_corr.png}}
+\caption{\label{fig:homologs_corr}
+Env's preferences are well correlated between BG505 and BF520.
+This figure shows the correlation of the preferences between replicates, both within and between Env homologs.
+In this figure, the preferences for each replicate have been rescaled using the stringency parameter for the corresponding homolog from the group-M analysis in Table \ref{tab:phydms}.
+We analyzed 616 sites that are shared between homologs and were in readily alignable regions of the group-M multiple-sequence alignment.
+As a control, we also compared our estimates of Env's preferences with the preferences of a non-homologous protein -- influenza HA -- across the 480 sites where these proteins overlap in sequential numbering.
+{\bf (A)} Plots showing the correlation of preferences between two replicates from the same protein.
+Here, differences reflect experimental noise.
+{\bf (B)} Plots showing the correlation of preferences between Env homologs, or between BG505 and HA.
+Here, differences reflect a combination of experimental noise and biological differences between proteins.
+The Env homologs are well correlated, whereas BG505 and HA are not correlated.
+{\bf (C)} The heat map shows the Pearson correlation coefficient for all pairwise comparisons of all experimental replicates of each homolog (coefficients are also shown on the correlation plots).
+This heat map indicates that the trends in {\bf (A)} and {\bf (B)} hold for all replicates.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=4.5in]{figures/distances/distances.png}}
+\caption{\label{fig:distances}
+Most shifts in amino-acid preference between Env homologs are small-to-intermediate in effect size after correcting for experimental noise.
+We quantified the shift in each site's preferences using a distance metric that corrects for experimental noise, as quantified by experimental replicates.
+{\bf (A)} This panel shows how the distance metric is calculated for a subset of sites.
+{\bf (B)} The blue histogram shows the distribution of site-specific $RMSD_{corrected}$ values between Env homologs for all 616 sites being compared.
+Most sites have a distance greater than zero.
+The overlayed green histogram shows the distribution of $RMSD_{corrected}$ values between BG505 Env and influenza HA for all 480 sites that overlap in sequential numbering.
+Most site-specific distances in the Env-HA comparison are much larger than the distances in the BG505-BF520 comparison.
+}
+\end{figure}
+
+Add a supplemental figure showing that the distance metric we use (half sum absolute corrected) is correlated with other distance metrics (e.g., Jensen-Shannon distance).
+
+\begin{figure}
+\centerline{\includegraphics[width=6.5in]{figures/diffs_logo/diffs_logo.png}}
+\caption{\label{fig:diffs_logo}
+{\bf The difference in Env's site-specific amino-acid preferences between homologs, scaled by site-specific $RMSD_{corrected}$ values.}
+This figure shows the estimated preference shift for each amino acid across all 616 sites in Env being compared.
+For each site, we plot the difference in the rescaled averaged preferences between homologs ($\Delta\pi_{r,a} = \pi_{r,a}^{BG505} - \pi_{r,a}^{BF520}$), where each site has 20 letters corresponding to the 20 amino acids, and where negative and positive values are shown below and above the central black line, respectively.
+Thus, amino acids above the central line are more preferred in BG505, whereas amino acids below the central line are more preferred in BF520.
+At each site, we adjusted the total height of the letters in both directions to equal that site's $RMSD_{corrected}$ from Fig \ref{fig:distances}B (i.e., $\sum_{a}| \Delta\pi_{r,a} | = RMSD_{corrected}$).
+In effect, the sites with the largest stack heights are the sites where we observe the largest differences, after accounting for noise.
+The bar indicates which sites are variable or conserved in amino-acid sequence between homologs.
+Note that adjacent sites are not always contiguous in primary sequence, since we masked sites that are either not shared between homologs, or are not readily alignable in the group-M multiple-sequence alignment.
+}
+\end{figure}
+
+Plot overlays of conserved and variable, and surface versus core?
 
@@
 \subsection*{Looking at how shifts relate to evolution and structure}
+
+\begin{figure}
+\centerline{\includegraphics[width=2.5in]{figures/distances_at_subsets_of_sites/distances_at_subsets_of_sites.png}}
+\caption{\label{fig:shifts_at_variable_sites}
+{\bf The median shift in preferences between Env homologs is similar at variable vs. conserved sites.}
+The box plots compare show the shifts at all sites, just sites that differ in wildtype amino acids between BG505 and BF520 (variable), or sites that are conserved between the homologs.
+Shifts are quantified by site-specific $RMSD_{corrected}$ values from Fig \ref{fig:ch3_distances}.
+The median shift is roughly equivalent for each group of sites.
+}
+\end{figure}
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=5in]{figures/sanger_sequencing_supp/sanger_sequencing_supp}}
+\caption{\label{suppfig:sanger_sequencing_supp}
+{\bf Sanger sequencing of the BG505 mutant plasmids revealed that codon mutations were distributed roughly uniformly, with an average of 1.5 mutations per gene.}
+We Sanger sequenced 44 clones of BG505 Env.
+We sampled these clones roughly evenly from the three replicate mutant plasmid libraries before these libraries had undergone any functional selection. 
+{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
+{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
+{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons, as expected from random mutagenesis.
+{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in BG505 numbering).
+{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
+We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
+The observed distribution is close to the expected distribution.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=5in]{figures/mutfreqs_supp/mutfreqs_supp}}
+\caption{\label{suppfig:mutfreqs_supp}
+{\bf Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.}
+This figure is similar to Fig \ref{fig:mutfreqs}, but it reports \textit{non}-error-corrected per-codon mutation frequencies for each of the mutant libraries and wildtype controls before and after selection for {\bf (A)} BG505 and {\bf (B)} BF520.
+Specifically, for each each replicate, wt-DNA and mut-DNA refer to the pre-selection wildtype and mutant plasmids, respectively; and wt-virus and mut-virus refer to the post-selection wildtype and mutant viruses, respectively.
+Note, there is only a single replicate of the wt-DNA sample for BF520.
+The bars show frequencies of different codon-level mutations.
+The left bar for each sample groups mutations as being either synonymous, non-synonymous, or leading to the introduction of a stop codon.
+The right bar for each sample groups mutations as introducing a single-nucleotide codon mutation (e.g., aaa $\rightarrow$ aaT), double-nucleotide codon mutation (e.g., aaa $\rightarrow$ aTT), or triple-nucleotide mutation (e.g., aaa $\rightarrow$ TTT).
+Each wildtype plasmid shows a low rate of single-nucleotide codon mutations, consistent with errors from PCR and deep sequencing.
+Importantly, the wildtype plasmids have a much lower mutation rate than the mutant plasmids.
+The mutation frequency of single-nucleotide mutations substantially increases for each wildtype virus, which is expected due to errors from viral replication.
+Comparing the wildtype viruses to the mutant viruses suggests that roughly a third to a half of mutations in the mutant viruses arise due to to errors from PCR, deep sequencing, and viral replication, underlying the importance of estimating these error rates using the controls.
+When inferring Env's amino-acid preferences, we use the wildtype controls to statistically correct for such errors, as described in the Methods.
+}
+\end{figure}
+
+\begin{table}
+\centering
+\caption[Results of the \texttt{phydms} analysis with group-M sequences for individual experimental replicates.]{
+{\bf Results of the \texttt{phydms} analysis with group-M sequences for individual experimental replicates.}}
+{\footnotesize
+\begin{tabular}{cccc}
+model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
+\hline
+BG505-avg & 0.00 & $-$58185.53 & 7: $\beta$ = 2.07, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.49, $\kappa$ = 3.16\\
+BF520-avg & 70.26 & $-$58220.66 & 7: $\beta$ = 2.44, $\alpha_\omega$ = 0.71, $\beta_\omega$ = 0.47, $\kappa$ = 3.11\\
+BF520-1 & 1149.48 & $-$58760.27 & 7: $\beta$ = 1.69, $\alpha_\omega$ = 0.62, $\beta_\omega$ = 0.48, $\kappa$ = 3.18\\
+BG505-1 & 1169.24 & $-$58770.15 & 7: $\beta$ = 1.35, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.52, $\kappa$ = 3.16\\
+BG505-3 & 1188.94 & $-$58780.0 & 7: $\beta$ = 1.58, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.52, $\kappa$ = 3.21\\
+BG505-2 & 1215.68 & $-$58793.37 & 7: $\beta$ = 1.44, $\alpha_\omega$ = 0.66, $\beta_\omega$ = 0.50, $\kappa$ = 3.18\\
+BF520-3 & 1550.10 & $-$58960.58 & 7: $\beta$ = 1.69, $\alpha_\omega$ = 0.66, $\beta_\omega$ = 0.54, $\kappa$ = 3.08\\
+BF520-2 & 1613.64 & $-$58992.35 & 7: $\beta$ = 1.50, $\alpha_\omega$ = 0.65, $\beta_\omega$ = 0.50, $\kappa$ = 3.18\\
+LAI-3b & 2792.12 & $-$59581.59 & 7: $\beta$ = 0.87, $\alpha_\omega$ = 0.56, $\beta_\omega$ = 0.44, $\kappa$ = 3.17\\
+LAI-avg & 2994.80 & $-$59682.93 & 7: $\beta$ = 1.00, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.46, $\kappa$ = 3.09\\
+LAI-1 & 3402.28 & $-$59886.67 & 7: $\beta$ = 0.59, $\alpha_\omega$ = 0.52, $\beta_\omega$ = 0.49, $\kappa$ = 3.17\\
+LAI-2 & 3561.38 & $-$59966.22 & 7: $\beta$ = 0.53, $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.46, $\kappa$ = 3.15\\
+LAI-3 & 3594.92 & $-$59982.99 & 7: $\beta$ = 0.45, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.40, $\kappa$ = 3.18\\
+BF520-3 site averaged & 3904.60 & $-$60137.83 & 7: $\beta$ = 2.42, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
+BF520-avg site averaged & 3913.08 & $-$60142.07 & 7: $\beta$ = 2.24, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
+BG505-1 site averaged & 3914.92 & $-$60142.99 & 7: $\beta$ = 1.51, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
+BF520-1 site averaged & 3920.58 & $-$60145.82 & 7: $\beta$ = 2.17, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.22\\
+BF520-2 site averaged & 3921.04 & $-$60146.05 & 7: $\beta$ = 2.09, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
+BG505-3 site averaged & 3921.64 & $-$60146.35 & 7: $\beta$ = 1.61, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.20\\
+BG505-avg site averaged & 3922.08 & $-$60146.57 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
+BG505-2 site averaged & 3933.90 & $-$60152.48 & 7: $\beta$ = 1.44, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.22\\
+LAI-3b site averaged & 4056.98 & $-$60214.02 & 7: $\beta$ = 1.05, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.40, $\kappa$ = 3.26\\
+LAI-avg site averaged & 4083.20 & $-$60227.13 & 7: $\beta$ = 0.83, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.26\\
+LAI-3 site averaged & 4085.00 & $-$60228.03 & 7: $\beta$ = 0.73, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
+LAI-1 site averaged & 4093.64 & $-$60232.35 & 7: $\beta$ = 0.73, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
+LAI-2 site averaged & 4100.52 & $-$60235.79 & 7: $\beta$ = 0.61, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
+YNGKP M5 & 4423.42 & $-$60392.24 & 12: $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.55, $\kappa$ = 3.20
+\end{tabular}
+}
+\begin{flushleft}
+This table is similar to Table \ref{tab:ch3_phydms}, but shows the results of ExpCMs for \textit{all} experimental replicates of each homolog, instead of just the ExpCMs made after averaging a homolog's amino-acid preferences among its replicates.
+This table only shows the results of the analysis with group-M sequences; see \ref{tab:ch3_phydms_subtypeA_supp} for the results of the analysis with subtype-A sequences.
+With the exception of LAI replicate 3b, the ExpCMs for individual experimental replicates always perform worse (based on $\Delta$AIC) than the ExpCMs for the averaged replicates.
+Moreover, the inferred stringency parameters are always lower for individual replicates than for the averaged replicates.
+These patterns are both consistent with the idea that averaging across replicates increases the accuracy of our estimates by decreasing experimental noise.
+As in \ref{tab:ch3_phydms}, the ExpCMs for individual replicates all outperform the standard model and decrease in performance when the preferences are averaged across all sites in the protein to create model that is not site-specific.
+\end{flushleft}
+\label{tab:ch3_phydms_groupM_supp}
+\end{table}
+
+\begin{table}
+\centering
+\caption[Results of the \texttt{phydms} analysis with subtype-A sequences for individual experimental replicates.]{
+{\bf Results of the \texttt{phydms} analysis with subtype-A sequences for individual experimental replicates.}}
+{\footnotesize
+\begin{tabular}{cccc}
+model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
+\hline
+BF520-avg & 0.00 & $-$48166.45 & 7: $\beta$ = 2.80, $\alpha_\omega$ = 0.59, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
+BG505-avg & 234.70 & $-$48283.8 & 7: $\beta$ = 2.19, $\alpha_\omega$ = 0.58, $\beta_\omega$ = 0.32, $\kappa$ = 3.50\\
+BF520-1 & 1027.22 & $-$48680.06 & 7: $\beta$ = 1.90, $\alpha_\omega$ = 0.57, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BG505-3 & 1236.20 & $-$48784.55 & 7: $\beta$ = 1.63, $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.33, $\kappa$ = 3.59\\
+BG505-1 & 1270.98 & $-$48801.94 & 7: $\beta$ = 1.48, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
+BG505-2 & 1392.52 & $-$48862.71 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.39, $\kappa$ = 3.41\\
+BF520-3 & 1468.76 & $-$48900.83 & 7: $\beta$ = 1.88, $\alpha_\omega$ = 0.61, $\beta_\omega$ = 0.34, $\kappa$ = 3.47\\
+BF520-2 & 1591.72 & $-$48962.31 & 7: $\beta$ = 1.68, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.34, $\kappa$ = 3.65\\
+LAI-3b & 2813.66 & $-$49573.28 & 7: $\beta$ = 0.89, $\alpha_\omega$ = 0.49, $\beta_\omega$ = 0.30, $\kappa$ = 3.54\\
+LAI-avg & 2860.64 & $-$49596.77 & 7: $\beta$ = 1.06, $\alpha_\omega$ = 0.48, $\beta_\omega$ = 0.30, $\kappa$ = 3.41\\
+LAI-1 & 3173.34 & $-$49753.12 & 7: $\beta$ = 0.68, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.33, $\kappa$ = 3.49\\
+LAI-3 & 3397.38 & $-$49865.14 & 7: $\beta$ = 0.45, $\alpha_\omega$ = 0.49, $\beta_\omega$ = 0.30, $\kappa$ = 3.49\\
+LAI-2 & 3451.26 & $-$49892.08 & 7: $\beta$ = 0.51, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.34, $\kappa$ = 3.48\\
+BF520-3 site averaged & 3900.54 & $-$50116.72 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BG505-1 site averaged & 3908.02 & $-$50120.46 & 7: $\beta$ = 0.92, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BG505-2 site averaged & 3908.50 & $-$50120.7 & 7: $\beta$ = 0.90, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BG505-avg site averaged & 3909.68 & $-$50121.29 & 7: $\beta$ = 0.93, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BF520-avg site averaged & 3913.70 & $-$50123.3 & 7: $\beta$ = 1.22, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
+BG505-3 site averaged & 3914.34 & $-$50123.62 & 7: $\beta$ = 0.94, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.60\\
+BF520-2 site averaged & 3917.84 & $-$50125.37 & 7: $\beta$ = 1.18, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.62\\
+BF520-1 site averaged & 3922.20 & $-$50127.55 & 7: $\beta$ = 1.20, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.63\\
+LAI-3b site averaged & 3959.56 & $-$50146.23 & 7: $\beta$ = 0.20, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.65\\
+LAI-3 site averaged & 3961.16 & $-$50147.03 & 7: $\beta$ = 0.03, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
+LAI-avg site averaged & 3961.18 & $-$50147.04 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
+LAI-1 site averaged & 3961.22 & $-$50147.06 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
+LAI-2 site averaged & 3961.30 & $-$50147.1 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
+YNGKP M5 & 3980.00 & $-$50151.45 & 12: $\alpha_\omega$ = 0.42, $\beta_\omega$ = 0.42, $\kappa$ = 3.62
+\end{tabular}
+}
+\begin{flushleft}
+This table is similar to Table \ref{tab:ch3_phydms_groupM_supp}, but shows the results of the analysis with subtype-A sequences.
+The trends seen for subtype A are similar to those for group M.
+\end{flushleft}
+\label{tab:phydms_subtypeA_supp}
+\end{table}
+
 \end{document}

### 2017-10-24 paper/manuscript.tex
moved manuscript into `./paper/` subdirectory

--- 2017-09-25 manuscript.tex
@@
 }
+\figdata{The alignments used to make the group-M and subtype-A trees are in files \ref{suppfile:grpM} and \ref{suppfile:cladeA}, respectively}
 \end{figure}
@@
 }
-%\figsupp{test text}{\includegraphics[width=5cm]{figures/mutfreqs_supp/mutfreqs_supp}}
+\figsupp[Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.\label{mutfreqs_supp}]{Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.
+This figure is similar to Fig \ref{fig:mutfreqs}, but it reports \textit{non}-error-corrected per-codon mutation frequencies for each of the mutant libraries and wildtype controls before and after selection for {\bf (A)} BG505 and {\bf (B)} BF520.
+Specifically, for each each replicate, wt-DNA and mut-DNA refer to the pre-selection wildtype and mutant plasmids, respectively; and wt-virus and mut-virus refer to the post-selection wildtype and mutant viruses, respectively.
+Note, there is only a single replicate of the wt-DNA sample for BF520.
+The bars show frequencies of different codon-level mutations.
+The left bar for each sample groups mutations as being either synonymous, non-synonymous, or leading to the introduction of a stop codon.
+The right bar for each sample groups mutations as introducing a single-nucleotide codon mutation (e.g., aaa $\rightarrow$ aaT), double-nucleotide codon mutation (e.g., aaa $\rightarrow$ aTT), or triple-nucleotide mutation (e.g., aaa $\rightarrow$ TTT).
+Each wildtype plasmid shows a low rate of single-nucleotide codon mutations, consistent with errors from PCR and deep sequencing.
+Importantly, the wildtype plasmids have a much lower mutation rate than the mutant plasmids.
+The mutation frequency of single-nucleotide mutations substantially increases for each wildtype virus, which is expected due to errors from viral replication.
+Comparing the wildtype viruses to the mutant viruses suggests that roughly a third to a half of mutations in the mutant viruses arise due to to errors from PCR, deep sequencing, and viral replication, underlying the importance of estimating these error rates using the controls.
+When inferring Env's amino-acid preferences, we use the wildtype controls to statistically correct for such errors, as described in the Methods.}{\includegraphics[width=5in]{figures/mutfreqs_supp/mutfreqs_supp}}
 \end{figure}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=5in]{figures/mutfreqs_supp/mutfreqs_supp}}
-\caption{\label{suppfig:mutfreqs_supp}
-{\bf Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.}
-This figure is similar to Fig \ref{fig:mutfreqs}, but it reports \textit{non}-error-corrected per-codon mutation frequencies for each of the mutant libraries and wildtype controls before and after selection for {\bf (A)} BG505 and {\bf (B)} BF520.
-Specifically, for each each replicate, wt-DNA and mut-DNA refer to the pre-selection wildtype and mutant plasmids, respectively; and wt-virus and mut-virus refer to the post-selection wildtype and mutant viruses, respectively.
-Note, there is only a single replicate of the wt-DNA sample for BF520.
-The bars show frequencies of different codon-level mutations.
-The left bar for each sample groups mutations as being either synonymous, non-synonymous, or leading to the introduction of a stop codon.
-The right bar for each sample groups mutations as introducing a single-nucleotide codon mutation (e.g., aaa $\rightarrow$ aaT), double-nucleotide codon mutation (e.g., aaa $\rightarrow$ aTT), or triple-nucleotide mutation (e.g., aaa $\rightarrow$ TTT).
-Each wildtype plasmid shows a low rate of single-nucleotide codon mutations, consistent with errors from PCR and deep sequencing.
-Importantly, the wildtype plasmids have a much lower mutation rate than the mutant plasmids.
-The mutation frequency of single-nucleotide mutations substantially increases for each wildtype virus, which is expected due to errors from viral replication.
-Comparing the wildtype viruses to the mutant viruses suggests that roughly a third to a half of mutations in the mutant viruses arise due to to errors from PCR, deep sequencing, and viral replication, underlying the importance of estimating these error rates using the controls.
-When inferring Env's amino-acid preferences, we use the wildtype controls to statistically correct for such errors, as described in the Methods.
-}
-\end{figure}
-
 \begin{table}

### 2017-10-31 paper/manuscript.tex
first 2 figures

@@
 \author[1,2\authfn{1}]{Adam S. Dingens}
-\author[3]{Julie Overbaugh}
-\author[1,2,4]{Jesse D. Bloom}
-\affil[1]{Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
+\author[1,3]{Sarah K. Hilton}
+\author[4]{Julie Overbaugh}
+\author[1,2,3]{Jesse D. Bloom}
+\affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
 \affil[2]{Molecular and Cellular Biology PhD program, University of Washington, Seattle, WA}
+\affil[3]{Department of Genome Sciences, University of Washington, Seattle, WA}
+\affil[4]{Human Biology Division, Fred Hutchinson Cancer Research Center, Seattle, WA}
 \contrib[\authfn{1}]{These authors contributed equally to this work}
@@
 
-We made an alignment of group M sequences (\ref{suppfile:grpM}) and clade A (\ref{suppfile:cladeA})
 The relationship between BG505 and BF520 is shown in \ref{ref:tree}
@@
 \begin{figure}
-\centerline{\includegraphics[width=5in]{figures/tree/tree}}
+\centerline{\bf \Large A \hspace{0.52\textwidth} \bf B \hspace{0.35\textwidth}}
+\centerline{\includegraphics[width=0.6\textwidth]{figures/tree_plot.pdf} \includegraphics[width=0.4\textwidth]{figures/masked_alignment_identity.pdf}}
 \caption{\label{fig:tree}
-Phylogenetic tree showing the relationship among BG505 and BF520.
-{\bf(A)} A tree of group-M sequences with 20 sequences per subtype for subtypes A, B, C, D, F, and G.
-{\bf(B)} A tree of subtype-A sequences with 120 sequences total, rooted using a subtype-B sequence as an outgroup.
-BG505 and BF520 cluster within subtype A, while LAI clusters within subtype B.
-The alignments we used to make these trees are the same alignments we used in the below \texttt{phydms} analysis, excluding the outgroup sequence used to root the subtype-A tree.
-The scale bars are in units of codon substitutions per site.
+CAPTION
 }
-\figdata{The alignments used to make the group-M and subtype-A trees are in files \ref{suppfile:grpM} and \ref{suppfile:cladeA}, respectively}
+\figdata{The coding sequence alignment of clade A \textit{env} sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
+\figdata{Sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
 \end{figure}
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{figures/dms_schematic/dms_schematic}}
+\centerline{\includegraphics[width=0.9\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
 \caption{\label{fig:dms_schematic}
@@
 \section*{Supplementary Material}
-% figure out how to do the supplement for eLife
-% then remove the `\iffalse` and `\fi` commands below
-\iffalse
-\begin{supptab}
-\caption{\label{supptab:phydms_replicates}
-phydms analysis for all replicates
-}
-\end{supptab}
-
-\begin{suppfig}
-\caption{\label{suppfig:dms}
-some plots about DMS
-}
-\end{suppfig}
-
-\begin{suppfile}
-\caption{\label{suppfile:grpM}
-Alignment of all group M sequences numbered and aligned as ???
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{\label{suppfile:cladeA}
-Alignment of clade A sequences...
-}
-\end{suppfile}
-% remove the below line when figure out how to format the supplement correctly
-\fi
 

### 2017-10-31 paper/manuscript.tex
figure 3: avgmutfreqs and prefs correlations

@@
 \begin{figure}
-\centerline{\includegraphics[width=5in]{figures/mutfreqs/mutfreqs}}
+{{\bf \Large A} \hspace{0.2\textwidth} BG505 \hspace{0.51\textwidth} BF520} \\
+{\includegraphics[height=0.32\textwidth]{figures/BG505_avgmutfreqs.pdf}\includegraphics[clip=true,trim=0in 0in 1.7in 0in,height=0.32\textwidth]{figures/BF520_avgmutfreqs.pdf}} \\
+{\bf \Large B} \\
+{\includegraphics[width=0.37\textwidth]{figures/BG505_prefscorr.pdf}\hspace{0.1\textwidth}\includegraphics[width=0.37\textwidth]{figures/BF520_prefscorr.pdf}}
 \caption{\label{fig:mutfreqs}
@@
 }
-\figsupp[Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.\label{mutfreqs_supp}]{Per-codon mutation frequencies of the mutant libraries and wildtype controls before and after selection.
-This figure is similar to Fig \ref{fig:mutfreqs}, but it reports \textit{non}-error-corrected per-codon mutation frequencies for each of the mutant libraries and wildtype controls before and after selection for {\bf (A)} BG505 and {\bf (B)} BF520.
-Specifically, for each each replicate, wt-DNA and mut-DNA refer to the pre-selection wildtype and mutant plasmids, respectively; and wt-virus and mut-virus refer to the post-selection wildtype and mutant viruses, respectively.
-Note, there is only a single replicate of the wt-DNA sample for BF520.
-The bars show frequencies of different codon-level mutations.
-The left bar for each sample groups mutations as being either synonymous, non-synonymous, or leading to the introduction of a stop codon.
-The right bar for each sample groups mutations as introducing a single-nucleotide codon mutation (e.g., aaa $\rightarrow$ aaT), double-nucleotide codon mutation (e.g., aaa $\rightarrow$ aTT), or triple-nucleotide mutation (e.g., aaa $\rightarrow$ TTT).
-Each wildtype plasmid shows a low rate of single-nucleotide codon mutations, consistent with errors from PCR and deep sequencing.
-Importantly, the wildtype plasmids have a much lower mutation rate than the mutant plasmids.
-The mutation frequency of single-nucleotide mutations substantially increases for each wildtype virus, which is expected due to errors from viral replication.
-Comparing the wildtype viruses to the mutant viruses suggests that roughly a third to a half of mutations in the mutant viruses arise due to to errors from PCR, deep sequencing, and viral replication, underlying the importance of estimating these error rates using the controls.
-When inferring Env's amino-acid preferences, we use the wildtype controls to statistically correct for such errors, as described in the Methods.}{\includegraphics[width=5in]{figures/mutfreqs_supp/mutfreqs_supp}}
+\figdata{The raw numerical data plotted in panel (A) are in the file \texttt{avgmutfreqs.csv}.}
+\figdata{The amino-acid preferences for each replicate and the replicate averages for each homolog are in \texttt{all\_prefs\_unscaled.zip}.}
 \end{figure}

### 2017-10-31 paper/manuscript.tex
added phydms model comparison

@@
 \begin{table}
-\centering
-\caption{
-{\bf Phylogenetic models that incorporate Env's preferences indicate that selection was less stringent in the lab than in nature.}}
-{\footnotesize
-\begin{tabular}{cccc}
-\multicolumn{4}{c}{\bf{Group M}}\\
-\hline
-model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
-\hline
-BG505 & 0.00 & $-$58185.53 & 7: $\beta$ = 2.07, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.49, $\kappa$ = 3.16\\
-BF520 & 70.26 & $-$58220.66 & 7: $\beta$ = 2.44, $\alpha_\omega$ = 0.71, $\beta_\omega$ = 0.47, $\kappa$ = 3.11\\
-LAI & 2994.80 & $-$59682.93 & 7: $\beta$ = 1.00, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.46, $\kappa$ = 3.09\\
-BF520 site averaged & 3913.08 & $-$60142.07 & 7: $\beta$ = 2.24, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
-BG505 site averaged & 3922.08 & $-$60146.57 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
-LAI site averaged & 4083.20 & $-$60227.13 & 7: $\beta$ = 0.83, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.2\\
-YNGKP M5 & 4423.42 & $-$60392.24 & 12: $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.55, $\kappa$ = 3.20\\
-\\
-\multicolumn{4}{c}{\bf{Subtype A}}\\
-\hline
-model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
-\hline
-BF520 & 0.00 & $-$48166.45 & 7: $\beta$ = 2.80, $\alpha_\omega$ = 0.59, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
-BG505 & 234.70 & $-$48283.8 & 7: $\beta$ = 2.19, $\alpha_\omega$ = 0.58, $\beta_\omega$ = 0.32, $\kappa$ = 3.50\\
-LAI & 2860.64 & $-$49596.77 & 7: $\beta$ = 1.06, $\alpha_\omega$ = 0.48, $\beta_\omega$ = 0.30, $\kappa$ = 3.41\\
-BG505 site averaged & 3909.68 & $-$50121.29 & 7: $\beta$ = 0.93, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BF520 site averaged & 3913.70 & $-$50123.3 & 7: $\beta$ = 1.22, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-LAI site averaged & 3961.18 & $-$50147.04 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
-YNGKP M5 & 3980.00 & $-$50151.45 & 12: $\alpha_\omega$ = 0.42, $\beta_\omega$ = 0.42, $\kappa$ = 3.62
-\end{tabular}
-}
-\begin{flushleft}
-We used \texttt{phydms} to incorporate each homolog's averaged preferences into ExpCMs.
-This table shows the results of using these models to optimize the branch lengths of the group-M and subtype-A trees shown in Fig \ref{fig:ch3_tree}.
-For both trees, the ExpCMs dramatically outperformed a standard codon-substitution model (YNGKP M5), based on differences in Akaike information criterion (AIC), with BG505's and BF520's preferences describing Env's evolution better than LAI's preferences.
-The stringency parameters ($\beta$) inferred for BG505 and BF520 were both  $>$1, indicating that selection is more stringent in nature than our experiments for these homologs.
-In contrast, the stringency parameters inferred for LAI were $\sim$1, which may be due to experimental or biological differences, as discussed in the main text.
-As a control, for each homolog, we created an ExpCMs using preferences averaged across all sites in the protein.
-This averaging substantially decreased phylogenetic fit, indicating that the preferences capture site-specific selective pressures on Env in nature.
-The results for individual experimental replicates are shown in Tables \ref{tab:phydms_groupM_supp} and \ref{tab:phydms_subtypeA_supp}.
-\end{flushleft}
-\label{tab:phydms}
+\begin{fullwidth}
+{\centering \include{figures/gamma_omega_modelcomparison}}
+\caption{\label{tab:phydms}
+Phylogenetic models that incorporate Env's preferences indicate that selection was less stringent in the lab than in nature.}
+\tabledata{Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites are in \texttt{modelcomparison.md}.}
+\end{fullwidth}
 \end{table}
@@
 
-\begin{table}
-\centering
-\caption[Results of the \texttt{phydms} analysis with group-M sequences for individual experimental replicates.]{
-{\bf Results of the \texttt{phydms} analysis with group-M sequences for individual experimental replicates.}}
-{\footnotesize
-\begin{tabular}{cccc}
-model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
-\hline
-BG505-avg & 0.00 & $-$58185.53 & 7: $\beta$ = 2.07, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.49, $\kappa$ = 3.16\\
-BF520-avg & 70.26 & $-$58220.66 & 7: $\beta$ = 2.44, $\alpha_\omega$ = 0.71, $\beta_\omega$ = 0.47, $\kappa$ = 3.11\\
-BF520-1 & 1149.48 & $-$58760.27 & 7: $\beta$ = 1.69, $\alpha_\omega$ = 0.62, $\beta_\omega$ = 0.48, $\kappa$ = 3.18\\
-BG505-1 & 1169.24 & $-$58770.15 & 7: $\beta$ = 1.35, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.52, $\kappa$ = 3.16\\
-BG505-3 & 1188.94 & $-$58780.0 & 7: $\beta$ = 1.58, $\alpha_\omega$ = 0.69, $\beta_\omega$ = 0.52, $\kappa$ = 3.21\\
-BG505-2 & 1215.68 & $-$58793.37 & 7: $\beta$ = 1.44, $\alpha_\omega$ = 0.66, $\beta_\omega$ = 0.50, $\kappa$ = 3.18\\
-BF520-3 & 1550.10 & $-$58960.58 & 7: $\beta$ = 1.69, $\alpha_\omega$ = 0.66, $\beta_\omega$ = 0.54, $\kappa$ = 3.08\\
-BF520-2 & 1613.64 & $-$58992.35 & 7: $\beta$ = 1.50, $\alpha_\omega$ = 0.65, $\beta_\omega$ = 0.50, $\kappa$ = 3.18\\
-LAI-3b & 2792.12 & $-$59581.59 & 7: $\beta$ = 0.87, $\alpha_\omega$ = 0.56, $\beta_\omega$ = 0.44, $\kappa$ = 3.17\\
-LAI-avg & 2994.80 & $-$59682.93 & 7: $\beta$ = 1.00, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.46, $\kappa$ = 3.09\\
-LAI-1 & 3402.28 & $-$59886.67 & 7: $\beta$ = 0.59, $\alpha_\omega$ = 0.52, $\beta_\omega$ = 0.49, $\kappa$ = 3.17\\
-LAI-2 & 3561.38 & $-$59966.22 & 7: $\beta$ = 0.53, $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.46, $\kappa$ = 3.15\\
-LAI-3 & 3594.92 & $-$59982.99 & 7: $\beta$ = 0.45, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.40, $\kappa$ = 3.18\\
-BF520-3 site averaged & 3904.60 & $-$60137.83 & 7: $\beta$ = 2.42, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
-BF520-avg site averaged & 3913.08 & $-$60142.07 & 7: $\beta$ = 2.24, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
-BG505-1 site averaged & 3914.92 & $-$60142.99 & 7: $\beta$ = 1.51, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
-BF520-1 site averaged & 3920.58 & $-$60145.82 & 7: $\beta$ = 2.17, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.22\\
-BF520-2 site averaged & 3921.04 & $-$60146.05 & 7: $\beta$ = 2.09, $\alpha_\omega$ = 0.50, $\beta_\omega$ = 0.41, $\kappa$ = 3.21\\
-BG505-3 site averaged & 3921.64 & $-$60146.35 & 7: $\beta$ = 1.61, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.20\\
-BG505-avg site averaged & 3922.08 & $-$60146.57 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.21\\
-BG505-2 site averaged & 3933.90 & $-$60152.48 & 7: $\beta$ = 1.44, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.42, $\kappa$ = 3.22\\
-LAI-3b site averaged & 4056.98 & $-$60214.02 & 7: $\beta$ = 1.05, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.40, $\kappa$ = 3.26\\
-LAI-avg site averaged & 4083.20 & $-$60227.13 & 7: $\beta$ = 0.83, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.26\\
-LAI-3 site averaged & 4085.00 & $-$60228.03 & 7: $\beta$ = 0.73, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
-LAI-1 site averaged & 4093.64 & $-$60232.35 & 7: $\beta$ = 0.73, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
-LAI-2 site averaged & 4100.52 & $-$60235.79 & 7: $\beta$ = 0.61, $\alpha_\omega$ = 0.51, $\beta_\omega$ = 0.41, $\kappa$ = 3.27\\
-YNGKP M5 & 4423.42 & $-$60392.24 & 12: $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.55, $\kappa$ = 3.20
-\end{tabular}
-}
-\begin{flushleft}
-This table is similar to Table \ref{tab:ch3_phydms}, but shows the results of ExpCMs for \textit{all} experimental replicates of each homolog, instead of just the ExpCMs made after averaging a homolog's amino-acid preferences among its replicates.
-This table only shows the results of the analysis with group-M sequences; see \ref{tab:ch3_phydms_subtypeA_supp} for the results of the analysis with subtype-A sequences.
-With the exception of LAI replicate 3b, the ExpCMs for individual experimental replicates always perform worse (based on $\Delta$AIC) than the ExpCMs for the averaged replicates.
-Moreover, the inferred stringency parameters are always lower for individual replicates than for the averaged replicates.
-These patterns are both consistent with the idea that averaging across replicates increases the accuracy of our estimates by decreasing experimental noise.
-As in \ref{tab:ch3_phydms}, the ExpCMs for individual replicates all outperform the standard model and decrease in performance when the preferences are averaged across all sites in the protein to create model that is not site-specific.
-\end{flushleft}
-\label{tab:ch3_phydms_groupM_supp}
-\end{table}
-
-\begin{table}
-\centering
-\caption[Results of the \texttt{phydms} analysis with subtype-A sequences for individual experimental replicates.]{
-{\bf Results of the \texttt{phydms} analysis with subtype-A sequences for individual experimental replicates.}}
-{\footnotesize
-\begin{tabular}{cccc}
-model & $\Delta$AIC & log likelihood & parameters: optimized values \\ 
-\hline
-BF520-avg & 0.00 & $-$48166.45 & 7: $\beta$ = 2.80, $\alpha_\omega$ = 0.59, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
-BG505-avg & 234.70 & $-$48283.8 & 7: $\beta$ = 2.19, $\alpha_\omega$ = 0.58, $\beta_\omega$ = 0.32, $\kappa$ = 3.50\\
-BF520-1 & 1027.22 & $-$48680.06 & 7: $\beta$ = 1.90, $\alpha_\omega$ = 0.57, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BG505-3 & 1236.20 & $-$48784.55 & 7: $\beta$ = 1.63, $\alpha_\omega$ = 0.53, $\beta_\omega$ = 0.33, $\kappa$ = 3.59\\
-BG505-1 & 1270.98 & $-$48801.94 & 7: $\beta$ = 1.48, $\alpha_\omega$ = 0.55, $\beta_\omega$ = 0.33, $\kappa$ = 3.53\\
-BG505-2 & 1392.52 & $-$48862.71 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.39, $\kappa$ = 3.41\\
-BF520-3 & 1468.76 & $-$48900.83 & 7: $\beta$ = 1.88, $\alpha_\omega$ = 0.61, $\beta_\omega$ = 0.34, $\kappa$ = 3.47\\
-BF520-2 & 1591.72 & $-$48962.31 & 7: $\beta$ = 1.68, $\alpha_\omega$ = 0.54, $\beta_\omega$ = 0.34, $\kappa$ = 3.65\\
-LAI-3b & 2813.66 & $-$49573.28 & 7: $\beta$ = 0.89, $\alpha_\omega$ = 0.49, $\beta_\omega$ = 0.30, $\kappa$ = 3.54\\
-LAI-avg & 2860.64 & $-$49596.77 & 7: $\beta$ = 1.06, $\alpha_\omega$ = 0.48, $\beta_\omega$ = 0.30, $\kappa$ = 3.41\\
-LAI-1 & 3173.34 & $-$49753.12 & 7: $\beta$ = 0.68, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.33, $\kappa$ = 3.49\\
-LAI-3 & 3397.38 & $-$49865.14 & 7: $\beta$ = 0.45, $\alpha_\omega$ = 0.49, $\beta_\omega$ = 0.30, $\kappa$ = 3.49\\
-LAI-2 & 3451.26 & $-$49892.08 & 7: $\beta$ = 0.51, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.34, $\kappa$ = 3.48\\
-BF520-3 site averaged & 3900.54 & $-$50116.72 & 7: $\beta$ = 1.53, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BG505-1 site averaged & 3908.02 & $-$50120.46 & 7: $\beta$ = 0.92, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BG505-2 site averaged & 3908.50 & $-$50120.7 & 7: $\beta$ = 0.90, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BG505-avg site averaged & 3909.68 & $-$50121.29 & 7: $\beta$ = 0.93, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BF520-avg site averaged & 3913.70 & $-$50123.3 & 7: $\beta$ = 1.22, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.61\\
-BG505-3 site averaged & 3914.34 & $-$50123.62 & 7: $\beta$ = 0.94, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.60\\
-BF520-2 site averaged & 3917.84 & $-$50125.37 & 7: $\beta$ = 1.18, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.62\\
-BF520-1 site averaged & 3922.20 & $-$50127.55 & 7: $\beta$ = 1.20, $\alpha_\omega$ = 0.47, $\beta_\omega$ = 0.30, $\kappa$ = 3.63\\
-LAI-3b site averaged & 3959.56 & $-$50146.23 & 7: $\beta$ = 0.20, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.65\\
-LAI-3 site averaged & 3961.16 & $-$50147.03 & 7: $\beta$ = 0.03, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
-LAI-avg site averaged & 3961.18 & $-$50147.04 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
-LAI-1 site averaged & 3961.22 & $-$50147.06 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
-LAI-2 site averaged & 3961.30 & $-$50147.1 & 7: $\beta$ = 0.01, $\alpha_\omega$ = 0.46, $\beta_\omega$ = 0.30, $\kappa$ = 3.67\\
-YNGKP M5 & 3980.00 & $-$50151.45 & 12: $\alpha_\omega$ = 0.42, $\beta_\omega$ = 0.42, $\kappa$ = 3.62
-\end{tabular}
-}
-\begin{flushleft}
-This table is similar to Table \ref{tab:ch3_phydms_groupM_supp}, but shows the results of the analysis with subtype-A sequences.
-The trends seen for subtype A are similar to those for group M.
-\end{flushleft}
-\label{tab:phydms_subtypeA_supp}
-\end{table}
-
 \end{document}

### 2017-10-31 paper/manuscript.tex
figures placed in manuscript

@@
 
+\usepackage{float}
+\newfloat{suppfile}{thp}{lofsupfile}
+\floatname{suppfile}{Supplementary file}
+
+\usepackage{placeins}
+
+\renewcommand{\floatpagefraction}{1}
+
+\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Mapping sites of shifting and constant mutational effects on the evolutionary landscape of HIV Envelope}
+\title{Shifting mutational effects on the evolutionary landscape of HIV Envelope}
 
@@
 \subsection*{Deep mutational scanning of two Env homologs from transmitted-founder (?) viruses}
-Previously we've done deep mutational scanning on a lab-passaged CXCR4-tropic virus isolated late in infection.
-Here we sought to examine the effects of mutations in viruses more relevant to selection on HIV in nature.
-We chose...
-
-The relationship between BG505 and BF520 is shown in \ref{ref:tree}
 
@@
 \end{figure}
-
-We performed deep mutational scanning as described previously with a few modifications that are very briefly described here and detailed in the methods.
-We saw strong selection as indicated in \ref{fig:mutfreqs}.
-More details on DMS is shown in \ref{suppfig:dms}
+\FloatBarrier
 
@@
 }
+\figsupp[Sanger sequencing of the BG505 mutant plasmid libraries.]{
+We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
+{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
+{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
+{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
+{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used here).
+{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
+We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
+The observed distribution is close to the expected distribution.
+}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
 \end{figure}
+\FloatBarrier
 
@@
 \end{figure}
+\FloatBarrier
 
 \subsection*{Re-scaling the experimental measurements to optimally describe HIV evolution in nature}
-
-Re-scaled as in \ref{tab:phydms}
 
@@
 \end{table}
+\FloatBarrier
 
 \begin{figure}
-\centerline{\includegraphics[width=6.5in]{figures/prefs_BG505/prefs_BG505.png}}
+\begin{fullwidth}
+{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
 \caption{\label{fig:bg505_prefs_logo}
@@
 }
+\figdata{The re-scaled preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
+\figdata{Sequence of the BG505 Env and mapping from sequential 1, 2, ... numbering of this sequence (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
+\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\end{fullwidth}
 \end{figure}
+\FloatBarrier
 
 \begin{figure}
-\centerline{\includegraphics[width=6.5in]{figures/prefs_BF520/prefs_BF520.png}}
+\begin{fullwidth}
+{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
 \caption{\label{fig:bf520_prefs_logo}
@@
 }
+\figdata{The re-scaled preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
+\figdata{Sequence of the BF520 Env and mapping from sequential 1, 2, ... numbering of this sequence (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
+\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\end{fullwidth}
 \end{figure}
-
-A supplemental file for each of these and the results for each replicate. This is a lot of files, so you might just make a zip with all eight.
+\FloatBarrier
 
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=6.5in]{figures/homologs_corr/homologs_corr.png}}
-\caption{\label{fig:homologs_corr}
-Env's preferences are well correlated between BG505 and BF520.
-This figure shows the correlation of the preferences between replicates, both within and between Env homologs.
-In this figure, the preferences for each replicate have been rescaled using the stringency parameter for the corresponding homolog from the group-M analysis in Table \ref{tab:phydms}.
-We analyzed 616 sites that are shared between homologs and were in readily alignable regions of the group-M multiple-sequence alignment.
-As a control, we also compared our estimates of Env's preferences with the preferences of a non-homologous protein -- influenza HA -- across the 480 sites where these proteins overlap in sequential numbering.
-{\bf (A)} Plots showing the correlation of preferences between two replicates from the same protein.
-Here, differences reflect experimental noise.
-{\bf (B)} Plots showing the correlation of preferences between Env homologs, or between BG505 and HA.
-Here, differences reflect a combination of experimental noise and biological differences between proteins.
-The Env homologs are well correlated, whereas BG505 and HA are not correlated.
-{\bf (C)} The heat map shows the Pearson correlation coefficient for all pairwise comparisons of all experimental replicates of each homolog (coefficients are also shown on the correlation plots).
-This heat map indicates that the trends in {\bf (A)} and {\bf (B)} hold for all replicates.
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=4.5in]{figures/distances/distances.png}}
-\caption{\label{fig:distances}
-Most shifts in amino-acid preference between Env homologs are small-to-intermediate in effect size after correcting for experimental noise.
-We quantified the shift in each site's preferences using a distance metric that corrects for experimental noise, as quantified by experimental replicates.
-{\bf (A)} This panel shows how the distance metric is calculated for a subset of sites.
-{\bf (B)} The blue histogram shows the distribution of site-specific $RMSD_{corrected}$ values between Env homologs for all 616 sites being compared.
-Most sites have a distance greater than zero.
-The overlayed green histogram shows the distribution of $RMSD_{corrected}$ values between BG505 Env and influenza HA for all 480 sites that overlap in sequential numbering.
-Most site-specific distances in the Env-HA comparison are much larger than the distances in the BG505-BF520 comparison.
-}
-\end{figure}
-
-Add a supplemental figure showing that the distance metric we use (half sum absolute corrected) is correlated with other distance metrics (e.g., Jensen-Shannon distance).
-
-\begin{figure}
-\centerline{\includegraphics[width=6.5in]{figures/diffs_logo/diffs_logo.png}}
-\caption{\label{fig:diffs_logo}
-{\bf The difference in Env's site-specific amino-acid preferences between homologs, scaled by site-specific $RMSD_{corrected}$ values.}
-This figure shows the estimated preference shift for each amino acid across all 616 sites in Env being compared.
-For each site, we plot the difference in the rescaled averaged preferences between homologs ($\Delta\pi_{r,a} = \pi_{r,a}^{BG505} - \pi_{r,a}^{BF520}$), where each site has 20 letters corresponding to the 20 amino acids, and where negative and positive values are shown below and above the central black line, respectively.
-Thus, amino acids above the central line are more preferred in BG505, whereas amino acids below the central line are more preferred in BF520.
-At each site, we adjusted the total height of the letters in both directions to equal that site's $RMSD_{corrected}$ from Fig \ref{fig:distances}B (i.e., $\sum_{a}| \Delta\pi_{r,a} | = RMSD_{corrected}$).
-In effect, the sites with the largest stack heights are the sites where we observe the largest differences, after accounting for noise.
-The bar indicates which sites are variable or conserved in amino-acid sequence between homologs.
-Note that adjacent sites are not always contiguous in primary sequence, since we masked sites that are either not shared between homologs, or are not readily alignable in the group-M multiple-sequence alignment.
-}
-\end{figure}
-
-Plot overlays of conserved and variable, and surface versus core?
-
 \subsection*{Validating sites of mutational shifts}
-Describe and validate a few of these sites
 
 \subsection*{Looking at how shifts relate to evolution and structure}
-
-\begin{figure}
-\centerline{\includegraphics[width=2.5in]{figures/distances_at_subsets_of_sites/distances_at_subsets_of_sites.png}}
-\caption{\label{fig:shifts_at_variable_sites}
-{\bf The median shift in preferences between Env homologs is similar at variable vs. conserved sites.}
-The box plots compare show the shifts at all sites, just sites that differ in wildtype amino acids between BG505 and BF520 (variable), or sites that are conserved between the homologs.
-Shifts are quantified by site-specific $RMSD_{corrected}$ values from Fig \ref{fig:ch3_distances}.
-The median shift is roughly equivalent for each group of sites.
-}
-\end{figure}
 
@@
 
-\nocite{*} % This command displays all refs in the bib file
 \bibliography{references.bib}
 
-\section*{Supplementary Material}
-
-\begin{figure}
-\centerline{\includegraphics[width=5in]{figures/sanger_sequencing_supp/sanger_sequencing_supp}}
-\caption{\label{suppfig:sanger_sequencing_supp}
-{\bf Sanger sequencing of the BG505 mutant plasmids revealed that codon mutations were distributed roughly uniformly, with an average of 1.5 mutations per gene.}
-We Sanger sequenced 44 clones of BG505 Env.
-We sampled these clones roughly evenly from the three replicate mutant plasmid libraries before these libraries had undergone any functional selection. 
-{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
-{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
-{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons, as expected from random mutagenesis.
-{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in BG505 numbering).
-{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
-We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
-The observed distribution is close to the expected distribution.
-}
-\end{figure}
+\begin{suppfile}
+\caption{
+\label{suppfile:code}
+The code to perform all steps in the analysis beginning with downloading the FASTQ files is in \texttt{analysis\_code.zip}.}
+\end{suppfile}
 

### 2017-11-01 paper/manuscript.tex
figures made thru Fig 6

@@
 \begin{figure}
-{{\bf \Large A} \hspace{0.2\textwidth} BG505 \hspace{0.51\textwidth} BF520} \\
-{\includegraphics[height=0.32\textwidth]{figures/BG505_avgmutfreqs.pdf}\includegraphics[clip=true,trim=0in 0in 1.7in 0in,height=0.32\textwidth]{figures/BF520_avgmutfreqs.pdf}} \\
+%{{\bf \Large A} \hspace{0.2\textwidth} BG505 \hspace{0.51\textwidth} BF520} \\
+%{\includegraphics[height=0.32\textwidth]{figures/BG505_avgmutfreqs.pdf}\includegraphics[clip=true,trim=0in 0in 1.7in 0in,height=0.32\textwidth]{figures/BF520_avgmutfreqs.pdf}} \\
+%{\bf \Large B} \\
+%{\includegraphics[width=0.37\textwidth]{figures/BG505_prefscorr.pdf}\hspace{0.1\textwidth}\includegraphics[width=0.37\textwidth]{figures/BF520_prefscorr.pdf}}
+\begin{minipage}[t]{0.34\textwidth}
+{\bf \Large A} \\ 
+\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
+\centerline{\small BG505}
+\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
+\centerline{\small BF520}
+\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
+\end{minipage}
+\begin{minipage}[t]{0.01\textwidth}
+\end{minipage}
+\begin{minipage}[t]{0.65\textwidth}
 {\bf \Large B} \\
-{\includegraphics[width=0.37\textwidth]{figures/BG505_prefscorr.pdf}\hspace{0.1\textwidth}\includegraphics[width=0.37\textwidth]{figures/BF520_prefscorr.pdf}}
+\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
+\end{minipage}
 \caption{\label{fig:mutfreqs}
-The deep mutational scanning experiments imposed strong purifying selection and led to reproducible estimates of each homolog's amino-acid preferences.
-{\bf(A)} For each replicate of each homolog, we deep sequenced the starting plasmid libraries (DNA) and the selected virus libraries (virus).
-Bars show the per-codon mutation frequency averaged across all sites after subtracting mutation frequencies determined using wildtype controls (Fig \ref{suppfig:ch3_mutfreqs_supp}).
-Mutations leading to stop codons were purged during the selection step to 3-17\% their starting frequencies (see red numbers), indicating strong purifying selection.
-Nonsynonymous mutations decreased to 40-50\% their starting frequencies, consistent with a large fraction of amino-acid mutations being deleterious to Env.
-Synonymous mutations only decreased to 87-95\% their starting frequencies, as would be expected if synonymous mutations tend to have more neutral effects than nonsynonymous changes.
-Panels {\bf(B)} and {\bf(C)} show the correlation in our estimates of each homolog's site-specific amino-acid preferences between experimental replicates, with {\bf(B)} showing the correlation between BF520 replicates and {\bf(C)} showing the correlation between BG505 replicates.
-Each correlation plot reports the associated Pearson correlation coefficient, which ranged from 0.59-0.75, indicating that our estimates are largely repeatable between replicates.
+CAPTION
 }
@@
 
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures/diversifying_selection_corr.pdf}
+\caption{\label{fig:divselcorr}
+Correlation between sites of diversifying and purifying selection among homologs.
+\jdbcomment{Hugh - can you make a structure or structures that color sites by evidence for purifying and diversifying selection?
+The data are in the \texttt{merged\_omegabysite.csv} file.
+Probably you could just color by the log10 P-value.
+It may be better to show one structure for diversifying and one for purifying rather than trying to show both on the same structure?
+Also, any observations about where these sites are located would be useful...}
+}
+\figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}
+\end{figure}
+\FloatBarrier
+
 \subsection*{Differences between homologs}

### 2017-11-03 paper/figures/omegabysite_structural_analysis/analysis_code.py
Add script to visualize omega-by-site values on structure

@@ -1,0 +1,144 @@
+"""
+This script is for analyzing omega-by-site values in PyMOL
+
+It is intended to be run in the PyMOL terminal as:
+    run analysis_code.py
+
+Hugh Haddox, November-2-2017
+"""
+import pymol
+from pymol import cmd
+
+structure = '5fyl'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Remove non-Env chains
+# gp41 = chain B
+# gp120 = chain G
+# 35O22 = chains D and E
+# PGT122 = chains H and L
+cmd.remove ('c;D,E,H,L,U,V')
+cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey90')
+#cmd.color('grey20', structure)
+cmd.set('cartoon_transparency', '0.5')
+cmd.set('cartoon_transparency', '0', structure)
+#cmd.hide('hetatm')
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get omega-by-site values and determine which sites evolve significantly
+# faster or slower than expected
+omega_d = {'BG505':{}, 'BF520':{}}
+sites_sig_fast = {'BG505':[], 'BF520':[]}
+sites_sig_slow = {'BG505':[], 'BF520':[]}
+Q_cutoff = 0.05
+with open('../merged_omegabysite.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		(site, omega, P, dLnL, Q, Env, log10P, log10Pdir) = line.strip().split(',')
+		omega_d[Env][site] = float(log10Pdir)
+		if float(Q) < Q_cutoff and float(omega) < 1.0:
+			sites_sig_slow[Env].append(site)
+		if float(Q) < Q_cutoff and float(omega) > 1.0:
+			sites_sig_fast[Env].append(site)
+
+# Identify min and max omega values
+h = 'BG505'
+min_omega = min(omega_d[h].values())
+max_omega = max(omega_d[h].values())
+print "\nThe minimum and maximum omega-by-site values are:"
[…]

### 2017-11-04 paper/figures/omegabysite_structural_analysis/analysis_code.py
Add structures showing omega-by-site to figure

@@ -6,90 +6,119 @@ It is intended to be run in the PyMOL terminal as:
 
 Hugh Haddox, November-2-2017
 """
+
+# Imports
 import pymol
 from pymol import cmd
 
-structure = '5fyl'
-cmd.reinitialize()
-cmd.delete('all')
-cmd.fetch(structure) #, type='pdb1')
+homologs = ['BG505', 'BF520']
+for h in homologs:
 
-# Remove non-Env chains
-# gp41 = chain B
-# gp120 = chain G
-# 35O22 = chains D and E
-# PGT122 = chains H and L
-cmd.remove ('c;D,E,H,L,U,V')
-cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+	# Fetch structure
+	structure = '5fyl'
+	cmd.reinitialize()
+	cmd.delete('all')
+	cmd.fetch(structure) #, type='pdb1')
 
-# Tweak initial display and color of Env monomers
-cmd.hide('everything')
-cmd.bg_color('white')
-cmd.show('cartoon')
-cmd.color('grey90')
-#cmd.color('grey20', structure)
-cmd.set('cartoon_transparency', '0.5')
-cmd.set('cartoon_transparency', '0', structure)
-#cmd.hide('hetatm')
+	# Remove non-Env chains
+	# gp41 = chain B
+	# gp120 = chain G
+	# 35O22 = chains D and E
+	# PGT122 = chains H and L
+	cmd.remove ('c;D,E,H,L,U,V')
+	cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
 
-# Identify unique sites in structure
-sites_in_structure = []
-cmd.iterate("(name ca)","sites_in_structure.append(resi)")
-unique_sites_in_structure = []
-for site in sites_in_structure:
-	if site not in unique_sites_in_structure:
-		unique_sites_in_structure.append(site)
+	# Tweak initial display and color of Env monomers
+	cmd.hide('everything')
+	cmd.bg_color('white')
+	cmd.show('cartoon')
+	cmd.color('grey90')
+	#cmd.color('grey20', structure)
+	cmd.set('cartoon_transparency', '0.5')
+	cmd.set('cartoon_transparency', '0', structure)
+	#cmd.hide('hetatm')
 
-# Get omega-by-site values and determine which sites evolve significantly
-# faster or slower than expected
[…]

### 2017-11-06 paper/manuscript.tex
Add structures showing omega-by-site to figure
added supp figure showing logos for all sites of purifying selection

@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{figures/diversifying_selection_corr.pdf}
+%\includegraphics[width=0.5\textwidth]{figures/diversifying_selection_corr.pdf}
+\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
 \caption{\label{fig:divselcorr}
 Correlation between sites of diversifying and purifying selection among homologs.
-\jdbcomment{Hugh - can you make a structure or structures that color sites by evidence for purifying and diversifying selection?
-The data are in the \texttt{merged\_omegabysite.csv} file.
-Probably you could just color by the log10 P-value.
-It may be better to show one structure for diversifying and one for purifying rather than trying to show both on the same structure?
-Also, any observations about where these sites are located would be useful...}
+\jdbcomment{Hugh - can you make panels showing the RSA and structure view of sites of purifying selection.}  
 }
+\figsupp[Amino-acid preferences and alignment frequencies at sites of purifying selection.]{
+This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
+Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
+For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
+The $Q$-value indicated is the maximum of that for the two homologs.
+}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
 \figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}

### 2017-11-06 paper/manuscript.tex
added plot of RSA among sites of purifying and diversifying selection

@@
 \centerline{\bf \Large A \hspace{0.52\textwidth} \bf B \hspace{0.35\textwidth}}
-\centerline{\includegraphics[width=0.6\textwidth]{figures/tree_plot.pdf} \includegraphics[width=0.4\textwidth]{figures/masked_alignment_identity.pdf}}
+\centerline{\includegraphics[width=0.64\textwidth]{figures/tree_plot.pdf} \includegraphics[width=0.36\textwidth]{figures/masked_alignment_identity.pdf}}
 \caption{\label{fig:tree}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
 \caption{\label{fig:dms_schematic}
@@
 Correlation between sites of diversifying and purifying selection among homologs.
-\jdbcomment{Hugh - can you make panels showing the RSA and structure view of sites of purifying selection.}  
+\jdbcomment{Hugh - can you make the panel showing the relationship between purifying selection and the Andrew Ward sites.}  
 }
@@
 }{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
+\figsupp[Distribution of relative solvent accessibilities among sites of diversifying and purifying selection.]{
+Sites are grouped by whether they are under diversifying or purifying selection at $Q < 0.05$ in \emph{both} homologs, or whether they do not fall into either of these categories.
+The absolute solvent accessibility of each site in the BG505 SOSIP trimer in PDB 5FYL~\citep{?} was calculated using DSSP~\citep{?}, and normalized to a relative solvent accessibility using the absolute accessibilities provided by \citet{?}.
+As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
+}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
 \figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}

### 2017-11-07 paper/figures/omegabysite_structural_analysis/analysis_code_conformational_dynamics_network.py
Add analysis of our results compared to the sites identified by the Ward group to be involved in conformational dynamics

@@ -2,7 +2,7 @@
 This script is for analyzing omega-by-site values in PyMOL
 
 It is intended to be run in the PyMOL terminal as:
-    run analysis_code.py
+    run analysis_code_conformational_dynamics_network.py
 
 Hugh Haddox, November-2-2017
 """
@@ -11,43 +11,40 @@ Hugh Haddox, November-2-2017
 import pymol
 from pymol import cmd
 
+# Fetch structure
+structure = '5vn3'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Remove non-Env chains
+# gp41 = chain A
+# gp120 = chain G
+cmd.remove ('c;C,H,I,J,K,L,M,N')
+#cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey90')
+#cmd.color('grey20', structure)
+cmd.set('cartoon_transparency', '0.5')
+cmd.set('cartoon_transparency', '0', structure)
+#cmd.hide('hetatm')
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get omega-by-site values and determine which sites evolve significantly
+# faster or slower than expected
 homologs = ['BG505', 'BF520']
 for h in homologs:
-
-	# Fetch structure
-	structure = '5fyl'
-	cmd.reinitialize()
-	cmd.delete('all')
-	cmd.fetch(structure) #, type='pdb1')
-
-	# Remove non-Env chains
-	# gp41 = chain B
-	# gp120 = chain G
-	# 35O22 = chains D and E
-	# PGT122 = chains H and L
-	cmd.remove ('c;D,E,H,L,U,V')
-	cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
-
-	# Tweak initial display and color of Env monomers
-	cmd.hide('everything')
-	cmd.bg_color('white')
-	cmd.show('cartoon')
-	cmd.color('grey90')
-	#cmd.color('grey20', structure)
-	cmd.set('cartoon_transparency', '0.5')
[…]

### 2017-11-07 paper/figures/omegabysite_structural_analysis/analysis_code_omega_by_site.py

@@ -1,0 +1,173 @@
+"""
+This script is for analyzing omega-by-site values in PyMOL
+
+It is intended to be run in the PyMOL terminal as:
+    run analysis_code_omega_by_site.py
+
+Hugh Haddox, November-2-2017
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+homologs = ['BG505', 'BF520']
+for h in homologs:
+
+	# Fetch structure
+	structure = '5fyl'
+	cmd.reinitialize()
+	cmd.delete('all')
+	cmd.fetch(structure) #, type='pdb1')
+
+	# Remove non-Env chains
+	# gp41 = chain B
+	# gp120 = chain G
+	# 35O22 = chains D and E
+	# PGT122 = chains H and L
+	cmd.remove ('c;D,E,H,L,U,V')
+	cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+
+	# Tweak initial display and color of Env monomers
+	cmd.hide('everything')
+	cmd.bg_color('white')
+	cmd.show('cartoon')
+	cmd.color('grey90')
+	#cmd.color('grey20', structure)
+	cmd.set('cartoon_transparency', '0.5')
+	cmd.set('cartoon_transparency', '0', structure)
+	#cmd.hide('hetatm')
+
+	# Identify unique sites in structure
+	sites_in_structure = []
+	cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+	unique_sites_in_structure = []
+	for site in sites_in_structure:
+		if site not in unique_sites_in_structure:
+			unique_sites_in_structure.append(site)
+
+	# Get omega-by-site values and determine which sites evolve significantly
+	# faster or slower than expected
+	omega_d = {'BG505':{}, 'BF520':{}}
+	sites_sig_fast = {'BG505':[], 'BF520':[]}
+	sites_sig_slow = {'BG505':[], 'BF520':[]}
+	Q_cutoff = 0.05
+	with open('../merged_omegabysite.csv') as f:
+		lines = f.readlines()[1:]
+		for line in lines:
+			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir) = line.strip().split(',')
+			omega_d[Env][site] = float(log10Pdir)
+			if float(Q) < Q_cutoff and float(omega) < 1.0:
+				sites_sig_slow[Env].append(site)
+			if float(Q) < Q_cutoff and float(omega) > 1.0:
+				sites_sig_fast[Env].append(site)
[…]

### 2017-11-12 paper/manuscript.tex
resized panel C in `omegabysite_structural_analysis` figure
start work on prefs dist

@@
 \figdata{Sequence of the BG505 Env and mapping from sequential 1, 2, ... numbering of this sequence (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\figdata{The $\omega_r$ values and associated $Q$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 Correlation between sites of diversifying and purifying selection among homologs.
-\jdbcomment{Hugh - can you make the panel showing the relationship between purifying selection and the Andrew Ward sites.}  
 }

### 2017-11-18 paper/manuscript.tex
added diff prefs logoplot
joy plots of pref-to-pref differences

@@
 
+\begin{figure}
+\begin{fullwidth}
+{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
+\hspace{0.01\textwidth} 
+\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
+\hspace{0.03\textwidth}
+\includegraphics[clip=true, trim=0in 0in 0in 1.2in, width=0.68\textwidth]{figures/distance_distribution.pdf}
+\vspace{0.15in}
+\\ 
+{\bf \Large C} \\
+\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
+\caption{\label{fig:diffprefs}
+Difference in preferences.
+}
+\figdata{The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
+\end{fullwidth}
+\end{figure}
+\FloatBarrier
+
 \subsection*{Validating sites of mutational shifts}

### 2017-11-25 paper/manuscript.tex
compute nearest distance to shifted / substituted sites
put shiftfeatures figure in paper

@@
 
-\author[1,2\authfn{1}]{Hugh K. Haddox}
-\author[1,2\authfn{1}]{Adam S. Dingens}
+\author[1,2,\authfn{1}]{Hugh K. Haddox}
+\author[1,2,\authfn{1}]{Adam S. Dingens}
 \author[1,3]{Sarah K. Hilton}
@@
 
-\subsection*{Validating sites of mutational shifts}
+\subsection*{Looking at how shifts relate to evolution and structure}
 
-\subsection*{Looking at how shifts relate to evolution and structure}
+\begin{figure}
+{\bf \Large A} \\
+\includegraphics[width=0.9\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
+\\
+{\bf \Large B} \hspace{0.38\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
+\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
+\hspace{0.01\textwidth}
+\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
+\hspace{0.01\textwidth}
+\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
+\caption{\label{fig:shiftfeatures}
+CAPTION.
+Note that the analyses in panels B-D are restricted to only those sites that are resolved in the Env crystal structure.
+}
+\figdata{Statistical testing of whether significantly shifted sites are more likely to have substituted between homologs are in \texttt{shifts\_vs\_subs\_table.txt}.}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.7\textwidth]{figures/entrenchment.pdf}}
+\caption{\label{fig:entrenchment}
+CAPTION
+}
+\end{figure}
 

### 2017-11-25 paper/manuscript.tex
added text from Hugh thesis: added raw, *not* yet fixed to account for changed sections, figure labels etc

@@
 \begin{abstract}
-HIV Envelope (Env) evolves rapidly.
-The immediate the evolutionary space accessible to any viral variant is largely determined by the effects of single amino-acid mutations.
-However, the effects of these mutations can shift over evolutionary time as the virus traverses through sequence space.
-Here we comprehensively quantify the ways in which mutational effects shift or remain constant by experimentally measuring the effects of all amino-acid mutations to two homologs of Env with 85\% (?) sequence identity.
-We find...
+HIV's envelope protein (Env) evolves rapidly.
+The immediate evolutionary space accessible to any viral variant is largely determined by the effects of single amino-acid mutations on viral fitness.
+Due to epistasis, these effects can "shift" over evolutionary time as the virus traverses through sequence space (e.g., the effect of a mutation tolerated in one viral variant may shift such that it is not tolerated in another related variant).
+In principle, even single amino-acid mutations can substantially alter a protein's mutational tolerance.
+However, the prevalence of such shifts in long-term protein evolution in nature is largely unknown.
+Here, we comprehensively quantified the ways in which the effects of mutations to Env have shifted or remained constant among two divergent Env homologs with 85\% sequence identity, corresponding to $>$100 amino-acid differences.
+We did so by experimentally measuring the effects of all single amino-acid mutations to each Env homolog on the ability of that homolog to support viral replication in cell culture.
+In total, we analyzed these effects at a total of 616 homologous sites.
+We observed that a small fraction of mutations had largely shifted in their effects, where a given mutation was well tolerated in one homolog, but strongly disfavored in the other.
+In contrast, we observed that the majority of mutations had shifted in their effects by only small-to-intermediate amounts.
+Thus, our data suggest that mutational effects are substantially conserved between these divergent Env homologs despite $>$100 underlying amino-acid differences.
+Overall, our data indicate that although epistasis can cause the effects of mutations to shift over evolutionary time, large shifts have been rare in the long-term evolutionary period separating the Env homologs we considered.
 \end{abstract}
@@
 \section{Introduction}
+HIV's envelope protein (Env) rapidly evolves.
+In just a single HIV-infected individual, the infecting virus can give rise to a pool of Env variants that are as diverse as all globally circulating influenza strains in a given year~\cite{korber2001evolutionary}.
+Env's ability to rapidly evolve has dire consequences for the immune system.
+Most HIV-infected individuals generate anti-Env antibodies that neutralize the virus~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
+However, Env readily evades this response through rapid sequence evolution~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
+The immune system can adapt to target new viral variants through through the elicitation of new antibodies and refining the specificity of antibodies through somatic hypermutation.
+But, Env is able to evade new immune responses each time they arise~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
+Understanding this protein's ability to do so is thus central to understanding HIV's ability to evade immunity.
+
+The immediate evolutionary space surrounding Env is largely defined by the effects of single point mutations on viral replication.
+To better understand this space, a large number of studies have characterized the effects of mutations on Env's ability to support viral replication in cell culture and evade antibodies~\cite{olshevsky1990identification,cordonnier1989single,basmaciogullari2002identification,freed1989mutational,lu2001structural,jacobs2005alanine,zwick2005anti,pantophlet2003fine,li2011mechanism,lynch2015hiv,haddox2016experimental}.
+For practical reasons, these studies often measured these effects in just a single or a few genetic backgrounds.
+However, effects of mutations can change over evolutionary time due to epistasis.
+One form of epistasis is intra-protein epistasis, where the effect of a mutation at one site in a protein is influenced by which amino acids are present at other sites in the same protein.
+Thus, as a protein evolves, substitutions at one site may "shift" the effects of mutations at other sites in the protein, such that a mutation that is not tolerated in one background is tolerated in a closely related one.
+This phenomenon has been documented for a wide variety of proteins~\cite{weinreich2006darwinian,bloom2006protein,bershtein2006robustness,ortlund2007crystal,lunzer2010pervasive,gong2013stability,natarajan2013epistasis,harms2014historical,podgornaia2015pervasive}, including Env~\cite{freed1994evidence,wang1996single,da2010fitness,gasser2016buffering}, and is described in greater detail in the Introduction of this thesis.
+
+Considering the effects of mutations in different genetic backgrounds is important for a protein as diverse as Env.
+Since HIV's introduction into humans, several phylogenetically distinct subtypes have arisen (Fig \ref{fig:ch3_tree}).
+The level of sequence divergence in Env between subtypes is typically 20-35\% amino-acid divergence~\cite{korber2001evolutionary}.
+Even within a subtype, Env variants typically differ by 15-20\% amino-acid divergence~\cite{korber2001evolutionary}.
+Thus, given that Env is $\sim$850 amino acids in length, homologs often differ by $>$100 amino acids.
+In principle, even single amino-acid changes can cause large shifts in the effect of mutations at other sites in a protein~\cite{weinreich2006darwinian,ortlund2007crystal,gong2013stability}.
+Thus, mutational effects may substantially differ between Env homologs.
+However, the frequency of large shifts in long-term protein evolution is still largely unknown.
+Some studies suggest that such shifts may be rare~\cite{doud2015site,ashenberg2013mutational}.
+
+Here, we quantified shifts in the effects of mutations on Env's ability to support viral replication in cell culture among two divergent Env homologs that share 85\% amino-acid identity.
+A technique called deep mutational scanning can be used to measure the effects of all possible single amino-acid mutations to a protein or protein domain of interest in a single high-throughput experiment~\cite{mclaughlin2012spatial,roscoe2013analyses,firnberg2014comprehensive,olson2014comprehensive,melnikov2014comprehensive,bloom2014experimentally,qi2014quantitative,thyagarajan2014inherent,stiffler2015evolvability,doud2015site,kitzman2015massively,mishra2016systematic,doud2016accurate,mavor2015determination}.
+We applied this technique to measure the ability of each homolog to tolerate all single amino-acid amino-acids across 612 homologous sites.
+A small fraction of mutations had dramatically different effects between homologs, with the mutation being well tolerated in one homolog, but highly deleterious to the other homolog.
+However, the effects of most mutations had only shifted by small-to-intermediate amounts, suggesting that the effects of most mutations are largely conserved between the divergent homologs despite underlying amino-acid differences at 15\% of sites.
 
@@
 
-\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder (?) viruses}
+\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
+Previously, we used deep mutational scanning to estimate the effects of all single amino-acid mutations to Env from the LAI strain of HIV, which is a lab-passaged CXCR4-tropic strain isolated late in infection~\cite{peden1991changes,haddox2016experimental}.
+Here, we sought to examine the effects of mutations in viruses isolated from earlier in infection, near the time of transmission.
+We chose two Envs -- BG505.W6M.C2 and BF520.W14M.C2 (hereafter referred to as BG505 Env and BF520 Env, respectively) -- both from viruses isolated from HIV-infected infants shortly after mother-to-child transmission~\cite{goo2014early}.
+Both infants rapidly developed broad plasma responses \cite{goo2014early}, and an anti-Env broadly neutralizing antibody (bNAb) has been isolated from BF520~\cite{simonich2016hiv}.
+We have previously used deep mutational scanning to comprehensively identify all amino-acid mutations that allow BF520 Env to escape a different bNAb \cite{dingens2017comprehensive}.
+BG505 Env has been extensively studied from a structural and immunological standpoint~\cite{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,falkowska2014broadly,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\cite{sanders2013next,sanders2015hiv,de2015immunogenicity}.
+Specifically, we examined the T332N variant of BG505 Env, which has a glycosylation site that is targeted by many bNAbs, but is absent in wildtype BG505 Env~\cite{sanders2013next}.
+Thus, we considered strains that are relevant to the effort to create antibody-based immunotherapies targeting Env.
+
+The trees in Fig \ref{fig:ch3_tree} show the relationship between the Envs from BG505, BF520, and LAI in context of other HIV-1 sequences.
+Fig \ref{fig:ch3_tree}A, which includes group-M sequences from several different subtypes, shows that BG505 and BF520 cluster within subtype-A, while LAI clusters within subtype-B.
+Fig \ref{fig:ch3_tree}B, which includes an larger number of just subtype-A sequences, shows that BG505 and BF520 are separated by long branches, which is typical of any two sequences in this star-like phylogeny.
+BG505 and BF520 Env are 85\% identical at the amino-acid level, while both are 73\% identical to LAI Env.
+Since each of these Envs are $\sim$850 amino-acids in length, these levels of divergence correspond to $>100$ amino-acid differences between each pair of homologs.
 
@@
 \FloatBarrier
+
+We used the deep mutational-scanning approach schematized in Fig \ref{fig:ch3_dms_schematic} to quantify the effects of all single amino-acid changes to both BG505 and BF520 Env.
+We followed the same technique that we used for LAI~\cite{haddox2016experimental}, with a few modifications.
+For each homolog, we created a library of \textit{env} genes with random codon mutations.
+We cloned these libraries into full-length proviral plasmids encoding the Q23 strain of HIV \cite{poss1999variants}, using high-efficiency cloning to obtain $>$1 million unique plasmid clones per library.
+Sanger sequencing revealed that for both BG505 and BF520, the codon mutations were evenly distributed across the genes, with an average of 1.5 and 1.1 codon mutations per gene, respectively (see Fig~\ref{suppfig:ch3_sanger_sequencing_supp} and \cite{dingens2017comprehensive}).
+We then generated mutant viruses by transfecting the plasmid library into 293T cells.
+Since transfected cells each receive multiple plasmids, the resulting viruses are unlikely to have a genotype-phenotype link.
+To establish this link and select for functional variants, we first passaged viral libraries for four days at a low initial multiplicity of infection (MOI) of 0.01 in SupT1 cells expressing CCR5.
+We then infected SupT1 cells expressing CCR5 with the passaged libraries at a high MOI ($>$1) and harvested reverse-transcribed unintegrated viral DNA 12 hours post-infection, before additional rounds of replication could occur.
+Essentially, this second infection imposes an additional round of selection for viruses with entry-competent Env variants.
+Finally, we deep sequenced the plasmid library before selection and the unintegrated proviral DNA after selection to quantify the change in frequency of each mutation.
+We estimated error rates from PCR and deep sequencing by sequencing wildtype plasmids.
+We also estimated error rates from viral replication by sequencing wildtype viruses passaged in parallel with the mutant viruses.
+We conducted the entire experiment in biological triplicate for each homolog (Fig \ref{fig:ch3_dms_schematic}B) (except that we only sequenced a single wildtype plasmid for the BF520 experiment).
 
@@
 
+For the deep mutational-scanning experiments to succeed, the selection step must efficiently purge deleterious mutations from the library.
+As a first measure of the strength of selection, we examined changes in per-codon mutation frequencies averaged across all sites in Env.
+Codon-level mutations can be classified as being nonsynonymous, synonymous, or leading to a stop codon.
+Deep sequencing of the mutant libraries allowed us to quantify the frequency each of these mutation types.
+In addition, deep sequencing of the wildtype controls allowed us to quantify mutational errors from PCR, deep sequencing, and viral replication.
+We corrected for these errors by subtracting mutation frequencies in the wildtype plasmids from the those in the mutant plasmids, and the mutation frequencies in the wildtype viruses from those in the mutant viruses.
+Figure \ref{fig:ch3_mutfreqs}A shows the resulting error-corrected per-codon mutation frequencies for each replicate before and after selection.
+The overall mutation frequency was higher in the BG505 libraries compared to the BF520 libraries, consistent with the higher mutation frequency estimated by Sanger sequencing Fig~\ref{suppfig:ch3_sanger_sequencing_supp}.
+
 \begin{figure}
-%{{\bf \Large A} \hspace{0.2\textwidth} BG505 \hspace{0.51\textwidth} BF520} \\
-%{\includegraphics[height=0.32\textwidth]{figures/BG505_avgmutfreqs.pdf}\includegraphics[clip=true,trim=0in 0in 1.7in 0in,height=0.32\textwidth]{figures/BF520_avgmutfreqs.pdf}} \\
-%{\bf \Large B} \\
-%{\includegraphics[width=0.37\textwidth]{figures/BG505_prefscorr.pdf}\hspace{0.1\textwidth}\includegraphics[width=0.37\textwidth]{figures/BF520_prefscorr.pdf}}
 \begin{minipage}[t]{0.34\textwidth}
@@
 
+The observed changes in per-codon mutation frequencies are consistent with the experiments imposing strong purifying selection.
+Mutations leading to pre-mature stop codons are expected to be highly deleterious and thus efficiently purged by selection.
+Indeed, stop codons were purged to between 3-17\% their starting frequencies in each replicate (see the red numbers in Figure \ref{fig:ch3_mutfreqs}A).
+This purging was more efficient in the BG505 replicates than in the BF520 replicates, suggesting that the BG505 experiments imposed stronger selection.
+Nonsynonymous mutations can have more varied effects depending on the site and amino-acid change in question.
+A variety of studies indicate that many if not most amino-acid mutations tend to be deleterious to any protein~\cite{guo2004protein,shafikhani1997generation,bloom2005thermodynamic}.
+Consistent with this expectation, we found that nonsynonymous mutations were purged to 40-50\% their starting frequencies across all replicates.
+In contrast, the effects of synonymous mutations might be expected to be more neutral than the effects of nonsynonymous mutations.
+This expectation does not always hold~\cite{parmley2006evidence,cuevas2012fitness,subramaniam2013serine,zanini2013quantifying}, especially in regions where codon mutations have the potential to disrupt important RNA secondary structures, such as HIV's Rev-response element, which overlaps with the \textit{env} codon sequence~\cite{fernandes2012hiv,haddox2016experimental}.
+However, when averaging across all sites, we found that the synonymous mutations were mostly retained upon selection, only decreasing to 87-95\% their starting frequency in each replicate.
+Overall, these changes are broadly consistent with selection strongly purging deleterious variants in the libraries.
+
+Next, we used the deep-sequencing data to infer each homolog's site-specific amino-acid preferences.
+For each site ($r$), we inferred that site's relative preference ($\pi_{r,a}$) for each amino acid ($a$) as values that are proportional to that amino acid's enrichment or depletion upon selection.
+These preferences are defined to sum to one at each site (i.e., $\sum_{a} \pi_{r,a} = 1$).
+In making these inferences, we used the deep-sequencing data of the wildtype controls to statistically correct for errors.
+We mutagenized 670 sites in BG505 and and 662 sites BF520.
+Thus, we estimated 14,070 (= 670 $\times$ 21) and 13,902 (=662 $\times$ 21) site-specific amino-acid preferences for these homologs, respectively.
+Fig \ref{fig:ch3_bg505_prefs_logo} shows the preferences for BG505 averaged between replicates and rescaled to optimally reflect the strength of selection in nature (as described in the next section).
+Fig \ref{fig:ch3_bf520_prefs_logo} is the same as Fig \ref{fig:ch3_bg505_prefs_logo}, but shows the preferences for BF520 instead of BG505.
+Each homolog's preferences are highly heterogeneous among sites.
+For instance, although some sites strongly prefer just a single amino acid (e.g., site 54), other sites, such as the region of variable loop 4 between sites 396-413, prefer a large number of amino acids roughly equally.
+Of sites that tolerate multiple amino acids, some sites prefer amino acids with a wide variety of chemical properties (e.g., site 31), while other sites only prefer amino acids with similar properties (e.g., site 35 only prefers aromatic amino acids).
+Overall, these data quantify each homolog's ability to tolerate each amino acid at each site.
+
+Comparing the preferences between experimental replicates of the same homolog allowed us to quantify the level of experimental noise in our estimates (Figs \ref{fig:ch3_mutfreqs} B and C).
+Our estimates were largely reproducible between replicates for a give homolog, with Pearson correlation coefficients ranging from 0.59-0.75.
+The estimates were more reproducible for BG505 than for BF520.
+This trend might be explained by the fact that stop codons were less efficiently purged in the BF520 experiments (Fig \ref{fig:ch3_mutfreqs}A).
+Thus, a higher fraction of nonfunctional variants may have randomly survived the selection step in the BF520 experiments, which would be expected to result in increased noise.
+Overall, however, the above results indicate that our estimates were largely reproducible for both homologs.
+
 \subsection*{Re-scaling the experimental measurements to optimally describe HIV evolution in nature}
+Next, we sought to compare Env's amino-acid preferences between BG505 and BF520.
+We also sought to compare these homologs to LAI~\cite{haddox2016experimental}.
+However, a concern is that differences we see between homologs may be due to experimental differences, rather than true biological differences.
+For instance, the LAI experiments involved passaging the viral libraries for a substantially longer amount of time in cell culture, which may have imposed increased selection.
+Additionally, even though we conducted the BG505 and BF520 experiments using very similar protocols, the experiments seemed to have exerted different levels of purifying selection, as indicated by differential purging of stop codons (Fig \ref{fig:ch3_mutfreqs}A).
+The strength of selection in our experiments may also differ from the strength of selection in nature.
+Ideally, however, we would like to compare the preferences after rescaling them to reflect the intensity of selection in a natural setting.
+
+To address these concerns, we estimated differences in selection strength between our experiments and nature using a phylogenetic approach.
+A software package called \texttt{phydms} can be used to incorporate site-specific amino-acid preferences into substitution models for maximum-likelihood phylogenetics~\cite{hilton2017phydms}.
+These models, known as experimentally informed codon models (ExpCMs), define the probability of a substitution at site $r$ to amino-acid $j$ from amino-acid $i$ as being proportional to the ratio of the site's preferences for these amino acids: $\frac{\pi_{r,j}}{\pi_{r,i}}$.
+Thus, if $j$ is more preferred than $i$, the probability of substituting from $j$ to $i$ is greater than substituting from $i$ to $j$.
+These models also define a stringency parameter ($\beta$) that rescales the preference for each amino acid $a$ at a site to be: $\pi_{r,a}^{\beta} / \lbrack \sum_{a} \pi_{r,a}^{\beta} \rbrack$.
+Given an alignment of natural sequences, ExpCMs can be used to infer the value of this parameter that rescales the preferences to optimally describe the evolution of these sequences.
+If the inferred stringency parameter is greater than one, the indication is that selection in nature tends to prefer the same amino acids as the experiments, but with greater stringency.
+
+We created ExpCMs with each homolog's averaged preferences.
+We then used these ExpCMs to analyze the \textit{env} sequences used to make the trees in Fig \ref{fig:ch3_tree}.
+To do so, we first used \texttt{RAxML} to infer the topology of each tree using the GTRCAT model.
+Fixing this topology, we then optimized the tree's branch lengths using either a standard codon-substitution model (YNGKP M5) or one of the ExpCMs.
+Table \ref{tab:ch3_phydms} compares the performance of each model with respect to each tree.
+For both trees, the ExpCMs describe Env's natural evolution far better than the standard model, as gauged by differences Akaike information criterion (AIC), which quantifies the likelihood of the data given the model while taking into account the number of model parameters, with higher AIC values indicating worse model performance.
+Notably, BG505 and BF520's preferences describe Env's evolution substantially better than LAI's preferences, though the ExpCM for LAI still outperforms the standard model.
+As a control, for each homolog, we averaged that homolog's preferences across all sites and made a ExpCM that modeled each site using these averaged preferences.
+As expected, these site-averaged models perform worse than the site-specific ones, indicating that the preferences recapitulate site-specific constraints in nature.
+The site-averaged models perform about as well as the standard model, which also lacks site-specific information.
+Tables \ref{tab:ch3_phydms_groupM_supp} and \ref{tab:ch3_phydms_subtypeA_supp} show the results of ExpCMs made using the preferences of individual replicates.
 
@@
 \FloatBarrier
+
+For both alignments, the ExpCMs inferred stringency parameters of $>$1 for both BG505 and BF520, indicating that selection was weaker in our experiments for these homologs than in nature.
+The results are largely consistent between the two alignments, indicating that the rescaling is robust to the different levels of sequence diversity we examined.
+We decided to rescale the preferences for BG505 and BF520 using the stringency parameters from the group-M analysis (Table \ref{tab:ch3_phydms}).
+These resulting rescaled preferences are shown in Figs \ref{fig:ch3_bg505_prefs_logo} and \ref{fig:ch3_bf520_prefs_logo} for BG505 and BF520, respectively.
+Since both homologs were rescaled with respect to the same sequences, we expect this rescaling to also normalize the preferences across experiments, helping to correct for differences in selection strength between experiments.
+
+In contrast to BG505 and BF520, for both alignments, the ExpCMs inferred a stringency parameter of $\sim$1 for LAI (Table \ref{tab:ch3_phydms}).
+Thus, both the stringency parameter and the overall model performance was lower for LAI than for the other homologs.
+One explanation for these differences is biological.
+Perhaps the preferences for LAI -- a lab-adapted virus isolated from chronic infection -- are simply less representative of the dominant selective pressures that shape Env's evolution in nature.
+An alternative explanation is experimental.
+It is possible that if we repeated the LAI deep mutational scan using the same method we used for BG505 and BF520, our estimates would change.
+We are currently repeating these experiments.
+Thus, for the remainder of this chapter, I will only focus on comparing the preferences for BG505 and BF520.
 
@@
 \subsection*{Differences between homologs}
+Having rescaled the preferences of each homolog to optimally describe Env's evolution in nature, we next sought to compare the preferences between homologs to determine how much Env's preferences have shifted over evolutionary time.
+We estimated preferences for 670 and 662 sites in BG505 and BF520, respectively.
+We compared the subset of 616 sites that are shared between homologs and are readily alignable in the group-M multiple-sequence alignment.
+Thus, in total, we compared 12,320 (=616$\times$20) amino-acid preferences between the homologs.
+
+As an initial gene-wide measure of these shifts, we simply quantified the correlation of the preferences between homologs.
+We expected some of the observed differences to be due to experimental noise rather than true biological differences.
+We estimated this noise by correlating the preferences between replicates for the same homolog after rescaling the preferences for each replicate using the stringency parameter from Table \ref{tab:ch3_phydms} for the appropriate homolog, as inferred from the group-M alignments.
+Fig \ref{fig:ch3_homologs_corr}A shows this correlation between a single pair of replicates for both BG505 and BF520.
+These comparisons provide an approximate ceiling for the expected correlation of replicates between homologs if both homologs have identical preferences.
+Conversely, if Env's preferences have substantially shifted over evolutionary time, the correlation between homologs is expected to be much lower.
+Fig \ref{fig:ch3_homologs_corr}B shows the correlation between homologs for a single pair of replicates.
+These replicates are nearly as well correlated as two replicates from the same homolog, suggesting that Env's preferences are substantially conserved between homologs.
+As a control, we correlated Env's preferences to the preferences of a non-homologous protein -- influenza hemagglutinin (HA) -- estimated in another study ~\cite{doud2016accurate} and rescaled with the appropriate stringency parameter from Table 1 of that study.
+HA's preferences were highly repeatable between the pair of replicates shown in Fig \ref{fig:ch3_homologs_corr}A.
+However, comparing replicates between BG505 and HA across all 480 sites that overlap in sequential numbering yielded a very low correlation, as would be expected when comparing preferences for non-homologous sites.
+Fig \ref{fig:ch3_homologs_corr} C shows that these trends are consistent across all three replicates for each protein.
+Thus, the correlations between Env homologs are much higher than the expected correlation between two non-homologous proteins.
 
@@
 
+Next, we sought to quantify shifts in Env's preferences at a site-specific level.
+Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.
+Thus, we quantified shifts using an approach that corrects for noise, similar to the one used in a previous study that deep mutational-scanning data between homologs of influenza's nucleoprotein \cite{doud2015site}.
+Fig \ref{fig:ch3_distances}A shows how we quantified shifts for a few example sites.
+For a given site $r$, we define the distance in preferences between an arbitrary pair of replicates $i$ and $j$ as: $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
+We then use this metric to measure the distance between all pairwise combinations of replicates from both homologs.
+We quantified the un-corrected distance between homologs as the root mean square of all replicate-replicate distances between homologs ($RMSD_{between}$).
+Since $D_{r}^{i,j}$ can range between 0-1, $RMSD_{between}$ also ranges between 0-1, with 0 indicating no differences and 1 indicating extreme differences between the homologs.
+Next, we quantified experimental noise as the root mean square of all replicate-replicate distances where both replicates came from the \textit{same} homolog ($RMSD_{within}$).
+Finally, we computed the noise-corrected distance between homologs ($RMSD_{corrected}$) by subtracting $RMSD_{between}$ by $RMSD_{within}$.
+
+Sites where we repeatedly measured large differences between homologs are expected to have $RMSD_{corrected}$ values much greater than zero (e.g., sites 512 and 309; Fig \ref{fig:ch3_distances}A).
+In contrast, sites where we repeatedly measured small differences between homologs are expected to have $RMSD_{corrected}$ values close to zero (e.g., site 296).
+Sites where the noise completely overwhelmed the biological signal are also expected to have $RMSD_{corrected}$ values close to zero (e.g., site 607).
+Thus, a $RMSD_{corrected}$ value near zero could indicate that a site's preferences are conserved between homologs, or that the noise at a site is high.
+
+The blue histogram in Fig \ref{fig:ch3_distances}B shows the distribution of site-specific $RMSD_{corrected}$ values between BG505 and BF520 for all 616 sites being compared (Fig \ref{fig:ch3_homologs_corr}).
+Most sites have distances greater than zero.
+Yet, only a small fraction of sites have large distances as extreme as sites 512 and 309 ($RMSD_{corrected}$ = 0.30-0.47; Fig \ref{fig:ch3_distances}A), as indicated by the small tail at the positive end of the distribution.
+Thus, it appears that few sites have dramatically shifted in preference.
+However, since $RMSD_{corrected}$ is influenced by noise, and since this noise is heterogeneous between sites, it was unclear exactly what the distribution would look like if Env's preferences had actually shifted at a large number of sites.
+We estimated the expected shape of this distribution by comparing BG505's preferences with the preferences from influenza HA, comparing all 480 sites that overlap in primary sequence.
+Since these proteins are non-homologous, we expected large site-specific differences in their preferences, mimicking large shifts during evolution.
+The green histogram in Fig \ref{fig:ch3_distances}B shows the resulting distribution of $RMSD_{corrected}$ values between Env and HA.
+As expected, the values in the Env-HA comparison tend to be much larger than the values in the Env-Env comparison.
+Thus, most differences between Env homologs are small-to-intermediate in effect size relative to the typical differences we observe for non-homologous sites.
+
+To examine the preference shifts in greater detail, we made a logo plot that shows the estimated shift for each amino acid at each of the 616 sites being compared (Fig \ref{fig:ch3_diffs_logo}).
+As described in the figure legend, this logo plot shows the results of subtracting the averaged preferences between homologs and then adjusting the total height of the letters at each site to correspond to that site's $RMSD_{corrected}$ value from Fig \ref{fig:ch3_distances}B.
+Most sites have small-to-intermediate stack heights, reflecting the fact that most site have relatively small-to-intermediate $RMSD_{corrected}$ values in Fig \ref{fig:ch3_distances}B.
+Sites where we detect large shifts between homologs are distributed throughout Env's primary sequence.
+The observed shifts follow a variety of patterns.
+At some sites, the preferences are strongly shifted for just a few amino acids (e.g., sites 309 and 288).
+At other sites, the preferences are strongly shifted in terms of the overall number of tolerated mutations (e.g., sites 512, 516, 599).
+Many of the largest shifts involved shifting from one hyrdrophobic amino acid to another (e.g., sites 165, 288, 307, 309).
+Overall, the shifts in the preferences are highly heterogeneous between sites.
+We are currently validating our findings for a subset of mutations by independently cloning each of these mutations into proviral plasmids, and then testing the ability of mutant viruses to replicate in cell culture.
+
 \subsection*{Looking at how shifts relate to evolution and structure}
+
+Next, we sought to explore the evolutionary basis of these shifts.
+One prediction is that the shifts would be concentrated at sites that differ between homologs.
+For instance, at site 309, Env's preferences shift with wildtype amino acid, where the BG505 wildtype amino acid (isoleucine) is more preferred in BG505 and the BF520 wildtype amino acid (leucine) is more preferred in BF520.
+This trend was observed in another study that compared shifts in mutational effects between homologs of influenza's nucleoprotein~\cite{doud2015site}.
+However, when we considered all sites, we did not find evidence that shifts at variable sites were stronger than shifts at conserved sites (Fig \ref{fig:ch3_shifts_at_variable_sites}).
+In the future, I plan to examine other evolutionary explanations for these shifts.
+I also plan to examine these shifts in context of Env's structure, which might lend insight into their functional basis.
 
@@
 \section{Discussion}
+We experimentally estimated the effects of all single amino-acid changes to two Env homologs with 85\% amino-acid identity.
+We did so using a high-throughput technique called deep mutational scanning.
+For each homolog, this approach involved making libraries of \textit{env} with random codon mutations, selecting for mutations that supported viral replication in cell culture, and then deep sequencing the starting and ending libraries to quantify the effect of each mutation.
+These experiments imposed strong purifying selection, as indicated by the near-complete depletion of mutations leading to premature stop codons.
+For each homolog, the results allowed us to estimate each site's preference for each of the 20 amino acids.
+We found that these estimates were largely repeatable between experimental replicates of the same homolog.
+
+We then compared site-specific amino-acid preferences between homologs.
+In an initial gene-wide comparison, we simply correlated the preferences between homologs across all sites.
+We expected that the differences we observed would be due to a combination of true biological differences and experimental noise.
+We quantified experimental noise by correlating the preferences between experimental replicates of the same homolog.
+When we then correlated the preferences between replicates from different homologs, we found that the correlation was nearly as high as the correlation between replicates from the same homolog, suggesting that the true preferences of each homolog are fairly well conserved.
+As a control, we compared Env's preferences with the preferences from a non-homologous protein -- influenza HA -- among sites that overlap in primary sequence of these proteins.
+As expected, the correlation in preferences between these non-homologous proteins was very low.
+This finding provided the first indication that although Env's preferences differ to some extent between homologs, they are still much more conserved than expected for two non-homologous proteins.
+
+Next, we quantified shifts in Env's amino-acid preferences at a more site-specific level.
+We computed shifts using a metric that takes into account experimental noise.
+At each site, this metric estimates the expected shift due to the noise, as captured by experimental replicates, and then corrects for this noise when measuring the shifts between homologs.
+We found that for most sites, the shifts between homologs were small-to-intermediate in effect size, with only a few sites having large shifts.
+We used the same metric to compare Env and influenza HA at sites that overlap in primary sequence.
+As expected for two non-homologous proteins, we found that most sites had large shifts in mutational effects.
+Thus, this site-specific analysis provided additional support that the shifts observed between Env homologs were not nearly as large as shifts between non-homologous proteins.
+
+Overall, these results help elucidate the strength of epistasis in Env's long-term evolution.
+Although large shifts in mutational effects can occur during protein evolution, our results indicate that such events were rare for the Env homologs we analyzed.
+Deep mutational scanning has also been used to analyze mutational shifts between other homologs.
+One study compared two homologs of influenza's nucleoprotein with 94\% amino-acid identity~\cite{doud2015site}.
+This study found that most mutational effects were conserved.
+A lower-throughput study of nucleoprotein mutations with a variety of stability effects found that these effects were also largely conserved among homologs that ranged from 94\%-72\% amino-acid identity~\cite{ashenberg2013mutational}.
+Another deep mutational-scanning study compared three homologous TIM-barrel proteins, each with 30-40\% amino-acid identity to one another~\cite{chan2017correlation}.
+Even for these considerably divergent homologs, mutational effects were still found to be correlated between homologs at many sites.
+Thus, our findings with Env are qualitatively similar to the above studies in suggesting that although epistasis can cause mutational effects to diverge over time, it does not usually completely erase site-specific preferences.
+
+Our results also shed light on the evolutionary basis of shifts in mutational effects.
+The deep mutational scan comparing nucleoprotein homologs found that shifts were enriched at sites that differed in wildtype amino-acid sequence between homologs~\cite{doud2015site}.
+This trend might be expected if proteins quickly evolve to accommodate historical amino-acid changes, as suggested from computational modeling~\cite{pollock2012amino}.
+However, the median shift in Env's preferences was similar between sites that were the same vs. different in wildtype sequence.
+Additional comparisons will be required to determine which pattern is more typical.
+
+Future work could expand upon our findings to analyze other Env homologs and the effects of mutations on other Env phenotypes.
+In this study, we compared two homologs from subtype A.
+It is possible that mutational effects are more shifted between more divergent Envs.
+Deep mutational scanning of Env from a different subtype could address this question (our previous results for LAI are difficult to compare with BG505 and BF520, for reasons described above).
+The specific phenotype we selected for in our study was the ability of Env to support viral replication in cell culture in the \textit{absence} of immune selection (e.g., antibodies).
+However, it is of great interest to characterize the effects of mutations to Env on antibody escape, especially for antibodies being used in immunotherapies or as templates in vaccine design.
+We have previously used deep mutational scanning to comprehensively identify single amino-acid mutations that allow BF520 Env to escape a bNAb~\cite{dingens2017comprehensive}.
+This technique, when applied to two different Env homologs, could be used to comprehensively determine the extent to which antibody-escape mutations have shifted during Env evolution.
 
@@
 
-\subsection*{Sequence alignments}
+\subsubsection*{Sequence numbering}
+We use the HXB2 numbering system \cite{korber1998numbering} for each Env homolog unless otherwise noted.
+We defined Env's ``variable loops'' according to \url{http://www.hiv.lanl.gov/}, but did not consider the disulfide-bonded cysteines as part of the loops.
+
+\subsubsection*{Codon-mutant libraries}
+For BF520 \textit{env}, we used the codon-mutant libraries generated from a previous study \cite{dingens2017comprehensive}.
+For BG505 \textit{env}, we generated codon-mutant libraries using essentially the same methods as the previous study, but with a few modifications.
+We computationally generated the sequences of the codon-tiling primers for the PCR-mutagenesis step using the same approach as in~\cite{dingens2017comprehensive}.
+The sequences of these primers will be made available as a supplemental file upon publication of this work.
+The end primers for this mutagenesis step were: 5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3' and 5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'.
+As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \cite{poss1999variants} -- another subtype-A transmitted/founder virus, using the same high-efficiency cloning vector used previously \cite{dingens2017comprehensive}.
+To do so, we first digested the cloning vector with BsmBI.
+We then used PCR to elongate the amplicons to include on either end 30 basepairs that are identical in sequence to the ends of the BsmBI-digested vector.
+The primers we used for this PCR were:
+5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3' and 5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'.
+Next, we used NEBuilder HiFi DNA Assembly (NEB, E2621S) to clone the \textit{env} amplicons into the BsmBI-digested plasmids.
+We purified the assembled products using Agencourt AMPure XP beads (Beckman Coulter, A63880) using a bead-to-sample ratio of 1.5, and then transformed the purified products into Stellar electrocompetent cells (Takara, 636765).
+The transformations yielded between 1.5-3.6 million unique clones for each of the three replicate libraries, as estimated by plating 1:2,000 dilutions of the transformations.
+As before, we scraped the plated colonies and maxiprepped the plasmid DNA, but this time we did not include the 4-hour outgrowth step after the scraping step.
+For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} in the \textit{env} locus of the full-length Q23 proviral plasmid.
+
+\subsubsection*{Generation and passaging of viruses}
+For BF520, we analyzed the viruses generated and passaged in our previous study \cite{dingens2017comprehensive}.
+For BG505, we generated and passaged viral libraries using essentially the same methods.
+First, for each replicate, we generated mutant viruses by transfecting 293T cells in three 6-well plates with a mixture of 2 ug mutant plasmid DNA and 6 uL FuGENE 6 Transfection Reagent (Promega, E269A) and 100 uL DMEM per well.
+293T cells were seeded in D10 media (DMEM supplemented with 10\% FBS, 1\% 200 mM L-glutamine, and 1\% of a solution of 10,000 units/mL penicillin and 10,000 $\mu$g/mL streptomycin) the day before transfection with 0.5 million cells per well, such that they were approximately 50\% confluent the next day.
+In parallel, we generated wildtype viruses by transfecting one 6-well plate of 293T cells with wildtype plasmid, using the same amount of DNA and transfection reagent as above.
+At 2days post-transfection, we harvested the transfection supernatant, passed it through a 0.2$\mu$m filter, treated the supernatant with DNAse to digest residual plasmid DNA (as in~\cite{haddox2016experimental}), and froze aliquots at -80$^{\circ}$C.
+We thawed and titered aliquots using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran as described in~\cite{dingens2017comprehensive}.
+
+Next, we conducted the initial viral passage in SupT1.CCR5 cells (obtained from Dr. James Hoxie \cite{boyd2015mutations}).
+During this passage, cells were maintained in R10 media, which has the same composition as D10 (described above), but has RPMI-1640 (GE Healthcare Life Sciences, SH30255.01) in the place of DMEM, with 10 $\mu$g/mL DEAE-dextran to enhance viral infection.
+We infected cells with 4 million (for replicate 1) or 5 million (for replicates 2 and 3) TZM-bl infectious units of mutant virus at an MOI of 0.01, with cells at a starting concentration of 1 million cells/mL in vented tissue-culture flasks (Fisher Scientific, 14-826-80).
+At 1 day post-infection, we pelleted cells, aspirated the supernatant, and resuspended cell pellets in fresh media including DEAE-dextran.
+At 2 days post-infection, we doubled the volume of each culture with fresh media including DEAE-dextran.
+At 4 days post-infection, we pelleted cells, passed the virus-containing supernatant through a 0.2 $\mu$m filter, concentrated the virus $\sim$30 fold using ultracentrifugation as described in~\cite{dingens2017comprehensive}, and then froze aliquots at -80$^{\circ}$C.
+In parallel, for each replicate, we also passaged 0.2 million (for replicate 1) or 0.5 million (for replicates 2 and 3) infectious units of wildtype virus with and using the same conditions.
+We thawed and titered these aliquots using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran.
+
+We conducted a second, much shorter viral passage by infecting cells with the passage-1 viruses.
+For each virus, we infected 1 million TMZ-bl infectious units into 1 million SupT1.CCR5 cells in the presence of 100 $\mu$g/mL DEAE-dextran (a 10-fold higher concentration than to concentration we used in the TZM-bl assay to titer the virus, meaning the actual number of infectious units in this passage was probably higher).
+Three hours post-infection, we pelleted the cells and resuspended them in fresh media without any DEAE-dextran.
+At 12 hours post-infection, we pelleted cells, washed them once with PBS, and then used a miniprep kit to harvest reverse-transcribed unintegrated viral DNA.
+
+\subsubsection*{Generation of samples for Illumina sequencing}
+We deeply sequenced each plasmid library before selection and each viral library after the two passages, as well as the wildtype plasmids and wildtype viruses that served as controls.
+First, we generated PCR amplicons of \textit{env} using the same approach as~\cite{dingens2017comprehensive} with the primers: 5'-GAAGACAGTGGCAATGAGAGTGATGG-3' and 5'-TTCCCTAACCTCAGGCCATCC-3'.
+Next, we sequenced these amplicons using a barcoded sub-amplicon sequencing strategy to reduce the sequencing error rate, as previously described~\cite{doud2016accurate,haddox2016experimental}, dividing \textit{env} into seven amplicons.
+The primer pairs used to generate these subamplicons are as follows, where N characters represent randomized sites in the barcode region of the primer:
+
+\begin{itemize}
+{\footnotesize
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGATCTTGGGGATGATAATAATCTGTAGTGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGGTGACATTGGTACACTGTAGAGTAAC-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNCCATGTGTAAAGTTAACCCCTCTCTGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGAACTTCTTATCCTTACACTTTAGGATCGC-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNCATACATTATTGTGCCCCAGCTGG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNCAATGTGCTTGTCTTATATCCCCTATTATGTC-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGGACAAGCATTCTATGCAACAGGG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNTGCTTTATTCTGCATGGGAGAGTTATAC-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGTCAAATAGCACGGGGTCAAATGAC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNCCAAGGAAGACAGCTCCTATTCCAAC-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGTGGTGGGGAGAGAAAAAAGAGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGTTTCTATTACTCCAACTAGAGTTCCAGGG-3'
+\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGGAAAACTCATCTGCACCACTAATGTG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNTGAGTATCCCTGCCTAACTCTATGTATTACAG-3'
+}
+\end{itemize}
+
+The resulting deep-sequencing data will be made available on the Sequence Read Archive upon the publication of this work.
+
+\subsubsection*{Analysis of deep-sequencing data}
+
+We analyzed the deep-sequencing data using the \texttt{dms\_tools} software package~\cite{bloom2015software}.
+The specific code we used will be made available upon the publication of this work.
+A summary of our workflow is as follows, and involves multiple programs encoded in \texttt{dms\_tools}:
+First, we aligned sequencing reads to \textit{env} and computed site-specific codon mutation frequencies using \texttt{dms\_barcodedsubamplicons}, using \texttt{dms\_summarizealignments} to make plots summarizing these frequencies.
+Next, we inferred Env's preferences from the codon counts using \texttt{dms\_inferprefs}.
+We then used a custom script to compare these preferences between homologs.
+
+\subsubsection*{Alignments and phylogenetic analyses of HIV-1 {\it env} sequences}
+We made the trees in Fig \ref{ref:ch3_tree} using multiple-sequence alignments downloaded from the HIV sequence database (\url{http://www.hiv.lanl.gov/}) that we then manually curated in the following ways:
+First, we removed sequences that differed in length from the HXB2 reference sequence (including gap characters), or which contained a premature stop codon, ambiguous residue, or frame-shift mutation.
+Next, we removed columns in the alignment for which we lack deep mutational scanning data or are not present in HXB2, columns that have $>$5\% gap characters, or columns in variable loops that looked poorly aligned by eye.
+The sequences used to make the trees in Fig \ref{ref:ch3_tree} were randomly selected from the larger alignment, with a defined number of sequences per clade, as described in the figure.
+The topologies of these trees were inferred by \texttt{RAxML} using the GTRCAT model.
+In the \texttt{phydms}~\cite{hilton2017phydms} analyses, we fixed the topologies of these trees, and then compared the ability of the different models to optimize the branch lengths.
+The trees shown in Fig \ref{ref:ch3_tree} are the results of optimizing the branch lengths using the YNGKP M5 model.
+The code we used to generate these alignments, build the trees, and conduct the \texttt{phydms} analyses will be made available upon publication of this work.
 
 \section{Acknowledgments}
+\comment{add this}
 

### 2017-11-26 paper/manuscript.tex
revised abstract to <= 150 words

@@
 %%% PREAMBLE 
-\documentclass[9pt,lineno]{elife}
+\documentclass[9pt]{elife}
+% add lineno for line numbering
 % Use the onehalfspacing option for 1.5 line spacing
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Shifting mutational effects on the evolutionary landscape of HIV Envelope}
+\title{Mapping mutational effects along the evolutionary landscape of HIV envelope}
 
@@
 \begin{abstract}
-HIV's envelope protein (Env) evolves rapidly.
-The immediate evolutionary space accessible to any viral variant is largely determined by the effects of single amino-acid mutations on viral fitness.
-Due to epistasis, these effects can "shift" over evolutionary time as the virus traverses through sequence space (e.g., the effect of a mutation tolerated in one viral variant may shift such that it is not tolerated in another related variant).
-In principle, even single amino-acid mutations can substantially alter a protein's mutational tolerance.
-However, the prevalence of such shifts in long-term protein evolution in nature is largely unknown.
-Here, we comprehensively quantified the ways in which the effects of mutations to Env have shifted or remained constant among two divergent Env homologs with 85\% sequence identity, corresponding to $>$100 amino-acid differences.
-We did so by experimentally measuring the effects of all single amino-acid mutations to each Env homolog on the ability of that homolog to support viral replication in cell culture.
-In total, we analyzed these effects at a total of 616 homologous sites.
-We observed that a small fraction of mutations had largely shifted in their effects, where a given mutation was well tolerated in one homolog, but strongly disfavored in the other.
-In contrast, we observed that the majority of mutations had shifted in their effects by only small-to-intermediate amounts.
-Thus, our data suggest that mutational effects are substantially conserved between these divergent Env homologs despite $>$100 underlying amino-acid differences.
-Overall, our data indicate that although epistasis can cause the effects of mutations to shift over evolutionary time, large shifts have been rare in the long-term evolutionary period separating the Env homologs we considered.
+The immediate evolutionary space accessible to a virus is largely determined by how single mutations affect fitness.
+These mutational effects can shift as the virus evolves.
+However, the prevalence of such shifts remains unclear.
+Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) variants that differ at $>$100 residues.
+Most mutations similarly affect both Envs, but the amino-acid preferences of a substantial minority of sites have clearly shifted.
+These shifts usually involve sites that prefer a single amino acid in one Env but tolerate many amino acids in the other.
+Surprisingly, shifts are only slightly enriched at sites that differ between the Envs -- and many occur at residues that do not even contact substitutions.
+Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the evolutionary capacity of most sites is conserved between moderately diverged viral strains.
 \end{abstract}

### 2017-11-26 paper/manuscript.tex
edits to intro

@@
 \documentclass[9pt]{elife}
+\sloppy
 % add lineno for line numbering
@@
 \DeclareSIUnit\Molar{M}
+%\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
+\usepackage{natbib}
+\usepackage{float}
+\usepackage{tikz}
 
-\usepackage{float}
 \newfloat{suppfile}{thp}{lofsupfile}
+%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
 \floatname{suppfile}{Supplementary file}
-
 \usepackage{placeins}
@@
 \author[4]{Julie Overbaugh}
-\author[1,2,3]{Jesse D. Bloom}
+\author[1,3,*]{Jesse D. Bloom}
 \affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA} 
@@
 \begin{abstract}
-The immediate evolutionary space accessible to a virus is largely determined by how single mutations affect fitness.
+The immediate evolutionary space accessible to HIV is largely determined by how single amino-acid mutations affect fitness.
 These mutational effects can shift as the virus evolves.
 However, the prevalence of such shifts remains unclear.
-Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) variants that differ at $>$100 residues.
+Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
 Most mutations similarly affect both Envs, but the amino-acid preferences of a substantial minority of sites have clearly shifted.
-These shifts usually involve sites that prefer a single amino acid in one Env but tolerate many amino acids in the other.
-Surprisingly, shifts are only slightly enriched at sites that differ between the Envs -- and many occur at residues that do not even contact substitutions.
+These shifted sites usually prefer a specific amino acid in one Env, but can tolerate many amino acids in the other.
+Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
 Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the evolutionary capacity of most sites is conserved between moderately diverged viral strains.
@@
 \section{Introduction}
-HIV's envelope protein (Env) rapidly evolves.
-In just a single HIV-infected individual, the infecting virus can give rise to a pool of Env variants that are as diverse as all globally circulating influenza strains in a given year~\cite{korber2001evolutionary}.
-Env's ability to rapidly evolve has dire consequences for the immune system.
-Most HIV-infected individuals generate anti-Env antibodies that neutralize the virus~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
-However, Env readily evades this response through rapid sequence evolution~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
-The immune system can adapt to target new viral variants through through the elicitation of new antibodies and refining the specificity of antibodies through somatic hypermutation.
-But, Env is able to evade new immune responses each time they arise~\cite{albert1990rapid,wei2003antibody,richman2003rapid}.
-Understanding this protein's ability to do so is thus central to understanding HIV's ability to evade immunity.
+HIV's envelope (Env) protein evolves very rapidly.
+The major group of HIV-1 that is responsible for the current pandemic originated from a virus that entered the human population $\sim$100 years ago~\citep{sharp2011origins,worobey2008direct,faria2014early}.
+The descendants of this virus have evolved so rapidly that their Envs now have as little as 65\% protein identity~\citep{lynch2009appreciating}.
+For comparison, protein orthologs shared between humans and mice have only diverged to a median identity of 78\% over 90 million years~\citep{waterston2002initial,hedges2006timetree}.
 
-The immediate evolutionary space surrounding Env is largely defined by the effects of single point mutations on viral replication.
-To better understand this space, a large number of studies have characterized the effects of mutations on Env's ability to support viral replication in cell culture and evade antibodies~\cite{olshevsky1990identification,cordonnier1989single,basmaciogullari2002identification,freed1989mutational,lu2001structural,jacobs2005alanine,zwick2005anti,pantophlet2003fine,li2011mechanism,lynch2015hiv,haddox2016experimental}.
-For practical reasons, these studies often measured these effects in just a single or a few genetic backgrounds.
-However, effects of mutations can change over evolutionary time due to epistasis.
-One form of epistasis is intra-protein epistasis, where the effect of a mutation at one site in a protein is influenced by which amino acids are present at other sites in the same protein.
-Thus, as a protein evolves, substitutions at one site may "shift" the effects of mutations at other sites in the protein, such that a mutation that is not tolerated in one background is tolerated in a closely related one.
-This phenomenon has been documented for a wide variety of proteins~\cite{weinreich2006darwinian,bloom2006protein,bershtein2006robustness,ortlund2007crystal,lunzer2010pervasive,gong2013stability,natarajan2013epistasis,harms2014historical,podgornaia2015pervasive}, including Env~\cite{freed1994evidence,wang1996single,da2010fitness,gasser2016buffering}, and is described in greater detail in the Introduction of this thesis.
+Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid}.
+Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
+The immediate evolutionary space accessible to any given Env is defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
+Most studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to any given Env variant.
 
-Considering the effects of mutations in different genetic backgrounds is important for a protein as diverse as Env.
-Since HIV's introduction into humans, several phylogenetically distinct subtypes have arisen (Fig \ref{fig:ch3_tree}).
-The level of sequence divergence in Env between subtypes is typically 20-35\% amino-acid divergence~\cite{korber2001evolutionary}.
-Even within a subtype, Env variants typically differ by 15-20\% amino-acid divergence~\cite{korber2001evolutionary}.
-Thus, given that Env is $\sim$850 amino acids in length, homologs often differ by $>$100 amino acids.
-In principle, even single amino-acid changes can cause large shifts in the effect of mutations at other sites in a protein~\cite{weinreich2006darwinian,ortlund2007crystal,gong2013stability}.
-Thus, mutational effects may substantially differ between Env homologs.
-However, the frequency of large shifts in long-term protein evolution is still largely unknown.
-Some studies suggest that such shifts may be rare~\cite{doud2015site,ashenberg2013mutational}.
+But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
+During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
+Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
+The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins far simpler than Env, which forms a large heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
-Here, we quantified shifts in the effects of mutations on Env's ability to support viral replication in cell culture among two divergent Env homologs that share 85\% amino-acid identity.
-A technique called deep mutational scanning can be used to measure the effects of all possible single amino-acid mutations to a protein or protein domain of interest in a single high-throughput experiment~\cite{mclaughlin2012spatial,roscoe2013analyses,firnberg2014comprehensive,olson2014comprehensive,melnikov2014comprehensive,bloom2014experimentally,qi2014quantitative,thyagarajan2014inherent,stiffler2015evolvability,doud2015site,kitzman2015massively,mishra2016systematic,doud2016accurate,mavor2015determination}.
-We applied this technique to measure the ability of each homolog to tolerate all single amino-acid amino-acids across 612 homologous sites.
-A small fraction of mutations had dramatically different effects between homologs, with the mutation being well tolerated in one homolog, but highly deleterious to the other homolog.
-However, the effects of most mutations had only shifted by small-to-intermediate amounts, suggesting that the effects of most mutations are largely conserved between the divergent homologs despite underlying amino-acid differences at 15\% of sites.
+Here we measure the effects on viral replication of all single amino-acid mutations to the ectodomains of two transmitted-founder virus Envs that differ by $>$100 mutations.
+By comparing these complete mutational maps, we identify sites where amino-acid preferences have shifted between the Envs.
+Most sites show no detectable shifts, but 30 sites have clearly altered their amino-acid preferences.
+These shifts tend to involve sites that strongly prefer a specific amino acid in one Env but tolerate many amino acids in the other.
+The shifted sites cluster in Env's structure, but are often distant from any of the amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
+By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
+Overall, our work provides across-strain genotype-to-phenotype maps that complement existing work characterizing Env's diversity at the level of protein sequence and structure.
 
@@
 \subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
+Correlations between replicates in prior studies:
+\citet{haddox2016experimental} all sites, had $R = 0.45$ to $R = 0.50$;
+
 Previously, we used deep mutational scanning to estimate the effects of all single amino-acid mutations to Env from the LAI strain of HIV, which is a lab-passaged CXCR4-tropic strain isolated late in infection~\cite{peden1991changes,haddox2016experimental}.
@@
 \end{figure}
-\FloatBarrier
 
@@
 \end{table}
-\FloatBarrier
 
@@
 \section{Acknowledgments}
-\comment{add this}
+\jdbcomment{add this}
 
 \bibliography{references.bib}
+
+\clearpage
 

### 2017-11-28 paper/figures/shifts_on_structure/pymol_analysis_code.py
Add figure of shifts on structure

@@ -1,0 +1,152 @@
+"""
+This script is for analyzing RMSD_corrected values in PyMOL
+
+It colors the structure of Env according to site-specific
+RMSD_corrected values on a color scale from grey to orange,
+with sites with significant RMSD_corrected values shown in
+spheres.
+
+When `take_pictures` is set to true, this script generates pictures
+of Env: `Env_pymol_face1.png` and `Env_pymol_face2.png`, which are
+rotated 120 degrees relative to one another.
+
+This script intended to be run in the PyMOL terminal as:
+    run pymol_analysis_code.py
+
+Hugh Haddox, November-27-2017
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '5fyl'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Remove non-Env chains
+# gp41 = chain B
+# gp120 = chain G
+# 35O22 = chains D and E
+# PGT122 = chains H and L
+cmd.remove ('c;D,E,H,L,U,V')
+cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey90')
+#cmd.color('grey20', structure)
+cmd.set('cartoon_transparency', '0.5')
+cmd.set('cartoon_transparency', '0', structure)
+#cmd.hide('hetatm')
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+    lines = f.readlines()[1:]
+    for line in lines:
+        elements = line.split(',')
+        site = elements[0]
+        RMSDcorrected = elements[1]
+        significant_shift = elements[6]
+        RMSD_dict[site] = float(RMSDcorrected)
+        sig_dict[site] = significant_shift
[…]

### 2017-11-30 paper/manuscript.tex
finalized tree figure, legend, supplements

@@
 Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
-The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins far simpler than Env, which forms a large heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
+The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
@@
 Most sites show no detectable shifts, but 30 sites have clearly altered their amino-acid preferences.
-These shifts tend to involve sites that strongly prefer a specific amino acid in one Env but tolerate many amino acids in the other.
-The shifted sites cluster in Env's structure, but are often distant from any of the amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
+These shifted sites usually strongly prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
+The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
 By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
-Overall, our work provides across-strain genotype-to-phenotype maps that complement existing work characterizing Env's diversity at the level of protein sequence and structure.
+Overall, our work provides across-strain genotype-to-phenotype maps that inform the analysis of Env's structure and evolution.
 
@@
 
+\subsection*{Two Env homologs from clade A transmitted-founder viruses}
+
+\begin{figure}
+\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.9\textwidth]{figures/tree_plot.pdf}}
+\caption{\label{fig:tree}
+Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
+The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of substantially unique sequences clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
+Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, leaving 616 alignable sites.
+The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~data~\ref{figsupp:identity}.
+The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and then branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
+}
+\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
+{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}.
+There are 616 non-masked sites.
+The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
+{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
+\figdata{\label{figdata:alignment} The sequence alignment of clade A \textit{env} coding sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
+\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
+\end{figure}
+
 \subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
+
 Correlations between replicates in prior studies:
@@
 Since each of these Envs are $\sim$850 amino-acids in length, these levels of divergence correspond to $>100$ amino-acid differences between each pair of homologs.
-
-\begin{figure}
-\centerline{\bf \Large A \hspace{0.52\textwidth} \bf B \hspace{0.35\textwidth}}
-\centerline{\includegraphics[width=0.64\textwidth]{figures/tree_plot.pdf} \includegraphics[width=0.36\textwidth]{figures/masked_alignment_identity.pdf}}
-\caption{\label{fig:tree}
-CAPTION
-}
-\figdata{The coding sequence alignment of clade A \textit{env} sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
-\figdata{Sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
-\end{figure}
 

### 2017-12-01 paper/manuscript.tex
first section of Results

@@
 \subsection*{Two Env homologs from clade A transmitted-founder viruses}
+The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
+However, all previous deep mutational scanning studies that have measured how large numbers of mutations affect HIV replication have used lab-adapted CXCR4-tropic viruses~\citep[LAI and NL4-3;][]{haddox2016experimental,al2014high} or a CCR5-tropic virus from a late-stage infection~\citep[LN40;][]{duenas2016saturation}.
+The properties of Env can vary substantially between such lab-passaged or late-stage viruses and the transmitted-founder viruses that are most relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
+
+We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
+The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
+We used the T332N variant of BG505 Env because it has a glycosylation site that is targeted by many anti-HIV antibodies but is absent in wildtype BG505~\citep{sanders2013next}.
+The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
+We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.
 
@@
 The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of substantially unique sequences clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
-Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, leaving 616 alignable sites.
-The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~data~\ref{figsupp:identity}.
+Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}, leaving 616 alignable sites.
+The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
 The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and then branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
@@
 
+Both BG505 and BF520 are from clade A of the major group of HIV-1.
+Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
+BG505 and BF520 have a typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and at 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the entire clade A alignment (Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}).
+The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
+
 \subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
@@
 \citet{haddox2016experimental} all sites, had $R = 0.45$ to $R = 0.50$;
-
-Previously, we used deep mutational scanning to estimate the effects of all single amino-acid mutations to Env from the LAI strain of HIV, which is a lab-passaged CXCR4-tropic strain isolated late in infection~\cite{peden1991changes,haddox2016experimental}.
-Here, we sought to examine the effects of mutations in viruses isolated from earlier in infection, near the time of transmission.
-We chose two Envs -- BG505.W6M.C2 and BF520.W14M.C2 (hereafter referred to as BG505 Env and BF520 Env, respectively) -- both from viruses isolated from HIV-infected infants shortly after mother-to-child transmission~\cite{goo2014early}.
-Both infants rapidly developed broad plasma responses \cite{goo2014early}, and an anti-Env broadly neutralizing antibody (bNAb) has been isolated from BF520~\cite{simonich2016hiv}.
-We have previously used deep mutational scanning to comprehensively identify all amino-acid mutations that allow BF520 Env to escape a different bNAb \cite{dingens2017comprehensive}.
-BG505 Env has been extensively studied from a structural and immunological standpoint~\cite{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,falkowska2014broadly,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\cite{sanders2013next,sanders2015hiv,de2015immunogenicity}.
-Specifically, we examined the T332N variant of BG505 Env, which has a glycosylation site that is targeted by many bNAbs, but is absent in wildtype BG505 Env~\cite{sanders2013next}.
-Thus, we considered strains that are relevant to the effort to create antibody-based immunotherapies targeting Env.
-
-The trees in Fig \ref{fig:ch3_tree} show the relationship between the Envs from BG505, BF520, and LAI in context of other HIV-1 sequences.
-Fig \ref{fig:ch3_tree}A, which includes group-M sequences from several different subtypes, shows that BG505 and BF520 cluster within subtype-A, while LAI clusters within subtype-B.
-Fig \ref{fig:ch3_tree}B, which includes an larger number of just subtype-A sequences, shows that BG505 and BF520 are separated by long branches, which is typical of any two sequences in this star-like phylogeny.
-BG505 and BF520 Env are 85\% identical at the amino-acid level, while both are 73\% identical to LAI Env.
-Since each of these Envs are $\sim$850 amino-acids in length, these levels of divergence correspond to $>100$ amino-acid differences between each pair of homologs.
 

### 2017-12-01 paper/manuscript.tex
tweaks to intro / first section Results

@@
 During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
-Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
+Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
 Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
@@
 
-Here we measure the effects on viral replication of all single amino-acid mutations to the ectodomains of two transmitted-founder virus Envs that differ by $>$100 mutations.
+Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
 By comparing these complete mutational maps, we identify sites where amino-acid preferences have shifted between the Envs.
@@
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
-However, all previous deep mutational scanning studies that have measured how large numbers of mutations affect HIV replication have used lab-adapted CXCR4-tropic viruses~\citep[LAI and NL4-3;][]{haddox2016experimental,al2014high} or a CCR5-tropic virus from a late-stage infection~\citep[LN40;][]{duenas2016saturation}.
-The properties of Env can vary substantially between such lab-passaged or late-stage viruses and the transmitted-founder viruses that are most relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
+However, the only previous deep mutational scanning study that has measured how all Env amino-acid mutations affect HIV replication is a study by our group~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
+The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
 
@@
 Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
-The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of substantially unique sequences clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
+The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
 Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}, leaving 616 alignable sites.
 The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
-The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and then branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
+The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
 }
@@
 
-Both BG505 and BF520 are from clade A of the major group of HIV-1.
+Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 have a typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and at 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the entire clade A alignment (Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}).
+BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the clade A alignment (Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

### 2017-12-01 paper/manuscript.tex
slight tweaks to `dms_schematic`

@@
 Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
-The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
-Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}, leaving 616 alignable sites.
+The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
+Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
 The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
@@
 \figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
-{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}.
+{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
 There are 616 non-masked sites.
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the clade A alignment (Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}).
+BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the clade A alignment (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
 
-\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
+\subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
+We have previously described a deep mutational scanning strategy for
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
+\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
 \caption{\label{fig:dms_schematic}
 Deep mutational scanning workflow.
-{\bf (A)} For each homolog, we made a library of proviral HIV plasmids with random codon-level mutations in \textit{env}.
-We then transfected the plasmids into 293T cells to generate mutant viruses, which may lack a genotype-phenotype link since transfected cell are expected to each receive multiple plasmids.
-To establish this link and select for functional variants, we first passaged the libraries in SupT1 cells at a low MOI.
-Then, we imposed a second round of selection by infecting the passaged viruses into SupT1 cells at a high MOI and then harvesting reverse-transcribed unintegrated viral DNA at 12 hours post-infection.
-Finally, we deep sequenced the libraries before and after selection.
-We also deep sequenced wildtype controls to estimate error rates due to PCR, deep sequencing, and viral replication.
-Using these sequencing data, we then inferred each site's preference for each of the 20 amino acids.
-{\bf (B)} For each homolog, we conducted this experiment in full biological triplicate, beginning at the stage of independently creating the plasmid mutant libraries.
+{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
+The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
+To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
+To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
+We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce the sequencing error rate.
+We also deep sequenced identically handled wildtype controls to estimate error and mutation rates.
+Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
+These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
+{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
+These replicates therefore account for \emph{all} sources of noise and error in the experiments.
 }
-\figsupp[Sanger sequencing of the BG505 mutant plasmid libraries.]{
+\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
 We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
@@
 {\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
-{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used here).
+{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
 {\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
@@
 The observed distribution is close to the expected distribution.
+Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
 }{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}

### 2017-12-02 paper/manuscript.tex
some progress on results

@@
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
-However, the only previous deep mutational scanning study that has measured how all Env amino-acid mutations affect HIV replication is a study by our group~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
+However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by our group~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
 The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
@@
 \subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
-We have previously described a deep mutational scanning strategy for
-
-Correlations between replicates in prior studies:
-\citet{haddox2016experimental} all sites, had $R = 0.45$ to $R = 0.50$;
-
-We used the deep mutational-scanning approach schematized in Fig \ref{fig:ch3_dms_schematic} to quantify the effects of all single amino-acid changes to both BG505 and BF520 Env.
-We followed the same technique that we used for LAI~\cite{haddox2016experimental}, with a few modifications.
-For each homolog, we created a library of \textit{env} genes with random codon mutations.
-We cloned these libraries into full-length proviral plasmids encoding the Q23 strain of HIV \cite{poss1999variants}, using high-efficiency cloning to obtain $>$1 million unique plasmid clones per library.
-Sanger sequencing revealed that for both BG505 and BF520, the codon mutations were evenly distributed across the genes, with an average of 1.5 and 1.1 codon mutations per gene, respectively (see Fig~\ref{suppfig:ch3_sanger_sequencing_supp} and \cite{dingens2017comprehensive}).
-We then generated mutant viruses by transfecting the plasmid library into 293T cells.
-Since transfected cells each receive multiple plasmids, the resulting viruses are unlikely to have a genotype-phenotype link.
-To establish this link and select for functional variants, we first passaged viral libraries for four days at a low initial multiplicity of infection (MOI) of 0.01 in SupT1 cells expressing CCR5.
-We then infected SupT1 cells expressing CCR5 with the passaged libraries at a high MOI ($>$1) and harvested reverse-transcribed unintegrated viral DNA 12 hours post-infection, before additional rounds of replication could occur.
-Essentially, this second infection imposes an additional round of selection for viruses with entry-competent Env variants.
-Finally, we deep sequenced the plasmid library before selection and the unintegrated proviral DNA after selection to quantify the change in frequency of each mutation.
-We estimated error rates from PCR and deep sequencing by sequencing wildtype plasmids.
-We also estimated error rates from viral replication by sequencing wildtype viruses passaged in parallel with the mutant viruses.
-We conducted the entire experiment in biological triplicate for each homolog (Fig \ref{fig:ch3_dms_schematic}B) (except that we only sequenced a single wildtype plasmid for the BF520 experiment).
+We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain of HIV~\citep{haddox2016experimental}.
+In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
+Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
+This noise is a major problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably identify differences that exceed the magnitude of the experimental noise.
 
@@
 
-For the deep mutational-scanning experiments to succeed, the selection step must efficiently purge deleterious mutations from the library.
-As a first measure of the strength of selection, we examined changes in per-codon mutation frequencies averaged across all sites in Env.
-Codon-level mutations can be classified as being nonsynonymous, synonymous, or leading to a stop codon.
-Deep sequencing of the mutant libraries allowed us to quantify the frequency each of these mutation types.
-In addition, deep sequencing of the wildtype controls allowed us to quantify mutational errors from PCR, deep sequencing, and viral replication.
-We corrected for these errors by subtracting mutation frequencies in the wildtype plasmids from the those in the mutant plasmids, and the mutation frequencies in the wildtype viruses from those in the mutant viruses.
-Figure \ref{fig:ch3_mutfreqs}A shows the resulting error-corrected per-codon mutation frequencies for each replicate before and after selection.
-The overall mutation frequency was higher in the BG505 libraries compared to the BF520 libraries, consistent with the higher mutation frequency estimated by Sanger sequencing Fig~\ref{suppfig:ch3_sanger_sequencing_supp}.
+We therefore modified the deep mutational scanning strategy into the approach illustrated in Figure~\ref{fig:dms_schematic}A.
+This approach had the following substantive modifications from that of \citet{haddox2016experimental}: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we use a larger number of virions ($\ge 3 \times 10^6$ infectious units per library in the current work versus $5\times 10^5$ in the prior work) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions (Figure~\ref{fig:dms_schematic}A).
+We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
 
@@
 \caption{\label{fig:mutfreqs}
-CAPTION
+The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
+{\bf (A)}
+The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), the plasmid mutant libraries (\emph{mutDNA}), the mutant viruses after the final infection (\emph{mutvirus}), and virus generated from the wildtype plasmids (\emph{virus}).
+Mutations are categorized by whether they are nonsynonymous, synonymous, or introduce a stop codon. 
+The \emph{DNA} samples show that the sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
+Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the appropriate background error rates given by the \emph{virus} and \emph{DNA} samples, respectively.
+The numerical data are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:mutfreqs}.
+More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
+{\bf (B)}
 }
-\figdata{The raw numerical data plotted in panel (A) are in the file \texttt{avgmutfreqs.csv}.}
+\figdata{\label{figdata:mutfreqs} The raw numerical data plotted in panel (A) are in the file \texttt{avgmutfreqs.csv}.}
 \figdata{The amino-acid preferences for each replicate and the replicate averages for each homolog are in \texttt{all\_prefs\_unscaled.zip}.}
 \end{figure}
+
+We estimated error rates from PCR and deep sequencing by sequencing wildtype plasmids.
+We also estimated error rates from viral replication by sequencing wildtype viruses passaged in parallel with the mutant viruses.
+We conducted the entire experiment in biological triplicate for each homolog (Fig \ref{fig:ch3_dms_schematic}B) (except that we only sequenced a single wildtype plasmid for the BF520 experiment).
+
+For the deep mutational-scanning experiments to succeed, the selection step must efficiently purge deleterious mutations from the library.
+As a first measure of the strength of selection, we examined changes in per-codon mutation frequencies averaged across all sites in Env.
+Codon-level mutations can be classified as being nonsynonymous, synonymous, or leading to a stop codon.
+Deep sequencing of the mutant libraries allowed us to quantify the frequency each of these mutation types.
+In addition, deep sequencing of the wildtype controls allowed us to quantify mutational errors from PCR, deep sequencing, and viral replication.
+We corrected for these errors by subtracting mutation frequencies in the wildtype plasmids from the those in the mutant plasmids, and the mutation frequencies in the wildtype viruses from those in the mutant viruses.
+Figure \ref{fig:ch3_mutfreqs}A shows the resulting error-corrected per-codon mutation frequencies for each replicate before and after selection.
+The overall mutation frequency was higher in the BG505 libraries compared to the BF520 libraries, consistent with the higher mutation frequency estimated by Sanger sequencing Fig~\ref{suppfig:ch3_sanger_sequencing_supp}.
 
@@
 \label{suppfile:code}
-The code to perform all steps in the analysis beginning with downloading the FASTQ files is in \texttt{analysis\_code.zip}.}
+The code to perform all steps in the analysis beginning with downloading the FASTQ files is in \texttt{analysis\_code.zip}.
+Specifically, this file contains a Jupyter notebook that performs the full analysis, as well as all required input data and all reasonably sized output files.}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:html}
+An HTML rendering of the Jupyter notebook that performs the computational analysis.
+The actual notebook is in Supplementary~file~\ref{suppfile:code}, but if you just want to look at the analysis rather than run it, then you may prefer to examine this file instead.
+In particular, the notebook contains detailed plots summarizing the deep sequencing data as generated using the \texttt{dms\_tools2} software~\citep{bloom2015software}.
+}
 \end{suppfile}

### 2017-12-02 paper/manuscript.tex
added table of average mutfreqs that includes error-corrected amounts and percent remaining

@@
 Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
-This noise is a major problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably identify differences that exceed the magnitude of the experimental noise.
+This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably identify differences that exceed the magnitude of the experimental noise.
 
@@
 We therefore modified the deep mutational scanning strategy into the approach illustrated in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive modifications from that of \citet{haddox2016experimental}: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we use a larger number of virions ($\ge 3 \times 10^6$ infectious units per library in the current work versus $5\times 10^5$ in the prior work) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions (Figure~\ref{fig:dms_schematic}A).
+This approach had the following substantive changes from that of \citet{haddox2016experimental}: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we use a larger number of virions ($\ge 3 \times 10^6$ infectious units per library in the current work versus $5\times 10^5$ in the prior work) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking diversity (Figure~\ref{fig:dms_schematic}A).
 We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
@@
 {\bf (A)}
-The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), the plasmid mutant libraries (\emph{mutDNA}), the mutant viruses after the final infection (\emph{mutvirus}), and virus generated from the wildtype plasmids (\emph{virus}).
-Mutations are categorized by whether they are nonsynonymous, synonymous, or introduce a stop codon. 
-The \emph{DNA} samples show that the sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
-Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the appropriate background error rates given by the \emph{virus} and \emph{DNA} samples, respectively.
-The numerical data are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:mutfreqs}.
+The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
+Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
+The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
+Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
 More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
 {\bf (B)}
+Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
+Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
+$R$ is the Pearson correlation coefficient. 
+The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
 }
-\figdata{\label{figdata:mutfreqs} The raw numerical data plotted in panel (A) are in the file \texttt{avgmutfreqs.csv}.}
-\figdata{The amino-acid preferences for each replicate and the replicate averages for each homolog are in \texttt{all\_prefs\_unscaled.zip}.}
+\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
+{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in Figure~\ref{fig:mutfreqs}.
+Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
+We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
+We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
+{\input{figures/avgmutfreqs}}
+\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
 \end{figure}

### 2017-12-03 paper/manuscript.tex
more progress on revising results (DMS section)

@@
 
-We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
-The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
-We used the T332N variant of BG505 Env because it has a glycosylation site that is targeted by many anti-HIV antibodies but is absent in wildtype BG505~\citep{sanders2013next}.
-The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
-We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.
-
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.9\textwidth]{figures/tree_plot.pdf}}
+\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
 \caption{\label{fig:tree}
@@
 \end{figure}
+
+We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
+The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
+We used the T332N variant of BG505 Env because it has a glycosylation site that is targeted by many anti-HIV antibodies but is absent in wildtype BG505~\citep{sanders2013next}.
+The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
+We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.
 
@@
 
-We therefore modified the deep mutational scanning strategy into the approach illustrated in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes from that of \citet{haddox2016experimental}: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we use a larger number of virions ($\ge 3 \times 10^6$ infectious units per library in the current work versus $5\times 10^5$ in the prior work) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking diversity (Figure~\ref{fig:dms_schematic}A).
+We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
 We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
+Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
 
@@
 
-We estimated error rates from PCR and deep sequencing by sequencing wildtype plasmids.
-We also estimated error rates from viral replication by sequencing wildtype viruses passaged in parallel with the mutant viruses.
-We conducted the entire experiment in biological triplicate for each homolog (Fig \ref{fig:ch3_dms_schematic}B) (except that we only sequenced a single wildtype plasmid for the BF520 experiment).
+The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
+Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
+The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
+This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
+The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
+The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
+These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
+Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
+Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed summaries of all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
-For the deep mutational-scanning experiments to succeed, the selection step must efficiently purge deleterious mutations from the library.
-As a first measure of the strength of selection, we examined changes in per-codon mutation frequencies averaged across all sites in Env.
-Codon-level mutations can be classified as being nonsynonymous, synonymous, or leading to a stop codon.
-Deep sequencing of the mutant libraries allowed us to quantify the frequency each of these mutation types.
-In addition, deep sequencing of the wildtype controls allowed us to quantify mutational errors from PCR, deep sequencing, and viral replication.
-We corrected for these errors by subtracting mutation frequencies in the wildtype plasmids from the those in the mutant plasmids, and the mutation frequencies in the wildtype viruses from those in the mutant viruses.
-Figure \ref{fig:ch3_mutfreqs}A shows the resulting error-corrected per-codon mutation frequencies for each replicate before and after selection.
-The overall mutation frequency was higher in the BG505 libraries compared to the BF520 libraries, consistent with the higher mutation frequency estimated by Sanger sequencing Fig~\ref{suppfig:ch3_sanger_sequencing_supp}.
-
-The observed changes in per-codon mutation frequencies are consistent with the experiments imposing strong purifying selection.
-Mutations leading to pre-mature stop codons are expected to be highly deleterious and thus efficiently purged by selection.
-Indeed, stop codons were purged to between 3-17\% their starting frequencies in each replicate (see the red numbers in Figure \ref{fig:ch3_mutfreqs}A).
-This purging was more efficient in the BG505 replicates than in the BF520 replicates, suggesting that the BG505 experiments imposed stronger selection.
-Nonsynonymous mutations can have more varied effects depending on the site and amino-acid change in question.
-A variety of studies indicate that many if not most amino-acid mutations tend to be deleterious to any protein~\cite{guo2004protein,shafikhani1997generation,bloom2005thermodynamic}.
-Consistent with this expectation, we found that nonsynonymous mutations were purged to 40-50\% their starting frequencies across all replicates.
-In contrast, the effects of synonymous mutations might be expected to be more neutral than the effects of nonsynonymous mutations.
-This expectation does not always hold~\cite{parmley2006evidence,cuevas2012fitness,subramaniam2013serine,zanini2013quantifying}, especially in regions where codon mutations have the potential to disrupt important RNA secondary structures, such as HIV's Rev-response element, which overlaps with the \textit{env} codon sequence~\cite{fernandes2012hiv,haddox2016experimental}.
-However, when averaging across all sites, we found that the synonymous mutations were mostly retained upon selection, only decreasing to 87-95\% their starting frequency in each replicate.
-Overall, these changes are broadly consistent with selection strongly purging deleterious variants in the libraries.
-
-Next, we used the deep-sequencing data to infer each homolog's site-specific amino-acid preferences.
-For each site ($r$), we inferred that site's relative preference ($\pi_{r,a}$) for each amino acid ($a$) as values that are proportional to that amino acid's enrichment or depletion upon selection.
-These preferences are defined to sum to one at each site (i.e., $\sum_{a} \pi_{r,a} = 1$).
-In making these inferences, we used the deep-sequencing data of the wildtype controls to statistically correct for errors.
-We mutagenized 670 sites in BG505 and and 662 sites BF520.
-Thus, we estimated 14,070 (= 670 $\times$ 21) and 13,902 (=662 $\times$ 21) site-specific amino-acid preferences for these homologs, respectively.
-Fig \ref{fig:ch3_bg505_prefs_logo} shows the preferences for BG505 averaged between replicates and rescaled to optimally reflect the strength of selection in nature (as described in the next section).
-Fig \ref{fig:ch3_bf520_prefs_logo} is the same as Fig \ref{fig:ch3_bg505_prefs_logo}, but shows the preferences for BF520 instead of BG505.
-Each homolog's preferences are highly heterogeneous among sites.
-For instance, although some sites strongly prefer just a single amino acid (e.g., site 54), other sites, such as the region of variable loop 4 between sites 396-413, prefer a large number of amino acids roughly equally.
-Of sites that tolerate multiple amino acids, some sites prefer amino acids with a wide variety of chemical properties (e.g., site 31), while other sites only prefer amino acids with similar properties (e.g., site 35 only prefers aromatic amino acids).
-Overall, these data quantify each homolog's ability to tolerate each amino acid at each site.
-
-Comparing the preferences between experimental replicates of the same homolog allowed us to quantify the level of experimental noise in our estimates (Figs \ref{fig:ch3_mutfreqs} B and C).
-Our estimates were largely reproducible between replicates for a give homolog, with Pearson correlation coefficients ranging from 0.59-0.75.
-The estimates were more reproducible for BG505 than for BF520.
-This trend might be explained by the fact that stop codons were less efficiently purged in the BF520 experiments (Fig \ref{fig:ch3_mutfreqs}A).
-Thus, a higher fraction of nonfunctional variants may have randomly survived the selection step in the BF520 experiments, which would be expected to result in increased noise.
-Overall, however, the above results indicate that our estimates were largely reproducible for both homologs.
+We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
+As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
+Our libraries mutagenized 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
+The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
+These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
+Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to make comparisons to the earlier inferior LAI data set. 
+Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across both Envs, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
+This fact is an initial hint that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
 
@@
 \label{suppfile:code}
-The code to perform all steps in the analysis beginning with downloading the FASTQ files is in \texttt{analysis\_code.zip}.
-Specifically, this file contains a Jupyter notebook that performs the full analysis, as well as all required input data and all reasonably sized output files.}
+The code to perform all steps in the analysis is in \texttt{analysis\_code.zip}.
+Specifically, this file contains a Jupyter notebook that performs the analysis, all required input data, and all reasonably sized output files.
+The Jupyter notebook downloads the deep sequencing data, processes it with the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, and also performs a variety of downstream analyses that generate most of the figures for this paper.
+}
 \end{suppfile}
@@
 An HTML rendering of the Jupyter notebook that performs the computational analysis.
-The actual notebook is in Supplementary~file~\ref{suppfile:code}, but if you just want to look at the analysis rather than run it, then you may prefer to examine this file instead.
-In particular, the notebook contains detailed plots summarizing the deep sequencing data as generated using the \texttt{dms\_tools2} software~\citep{bloom2015software}.
+The actual notebook is in Supplementary~file~\ref{suppfile:code}, but if you just want to look at the analysis rather than run it, then you may prefer this file instead.
+In particular, the notebook contains plots detailing the deep sequencing data analysis as generated using the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 }

### 2017-12-03 paper/manuscript.tex
Table 1 legend

@@
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
-However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by our group~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
+However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
 The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
@@
 {\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
-\figdata{\label{figdata:alignment} The sequence alignment of clade A \textit{env} coding sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
+\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
 \figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
@@
 \subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
-We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain of HIV~\citep{haddox2016experimental}.
+We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain~\citep{haddox2016experimental}.
 In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
 Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
-This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably identify differences that exceed the magnitude of the experimental noise.
+This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.
 
@@
 To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
-We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce the sequencing error rate.
-We also deep sequenced identically handled wildtype controls to estimate error and mutation rates.
+We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
+We also deep sequenced identically handled wildtype controls to estimate error rates.
 Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
@@
 We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
 We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
@@
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to make comparisons to the earlier inferior LAI data set. 
+Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to compare to the earlier inferior LAI data set. 
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across both Envs, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
@@
 
-\subsection*{Re-scaling the experimental measurements to optimally describe HIV evolution in nature}
+\subsection*{Amino-acid preferences of the BG505 and BF520 Envs and their relationship to HIV evolution in nature}
 Next, we sought to compare Env's amino-acid preferences between BG505 and BF520.
@@
 \caption{\label{tab:phydms}
-Phylogenetic models that incorporate Env's preferences indicate that selection was less stringent in the lab than in nature.}
-\tabledata{Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites are in \texttt{modelcomparison.md}.}
+Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature better than a standard substitution model.
+Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
+Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing either deep mutational scanning dataset describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
+Both ExpCM models have a stringency parameter $>$1, indicating that natural evolution prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
+All models draw $\omega$ (the dN/dS ratio for the Goldman-Yang model, and the relative dN/dS ratio after accounting for the measured constraints for the ExpCM) from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
+The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected, as determined using the approach in \citet{bloom2017identification}.
+Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
+Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results of analyses of the same tree with additional substitution models.
+}
+\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
 \end{fullwidth}
@@
 Since both homologs were rescaled with respect to the same sequences, we expect this rescaling to also normalize the preferences across experiments, helping to correct for differences in selection strength between experiments.
-
-In contrast to BG505 and BF520, for both alignments, the ExpCMs inferred a stringency parameter of $\sim$1 for LAI (Table \ref{tab:ch3_phydms}).
-Thus, both the stringency parameter and the overall model performance was lower for LAI than for the other homologs.
-One explanation for these differences is biological.
-Perhaps the preferences for LAI -- a lab-adapted virus isolated from chronic infection -- are simply less representative of the dominant selective pressures that shape Env's evolution in nature.
-An alternative explanation is experimental.
-It is possible that if we repeated the LAI deep mutational scan using the same method we used for BG505 and BF520, our estimates would change.
-We are currently repeating these experiments.
-Thus, for the remainder of this chapter, I will only focus on comparing the preferences for BG505 and BF520.
 

### 2017-12-03 paper/manuscript.tex
legends for prefs logo plots

@@
 
+\renewcommand{\topfraction}{1}
+\renewcommand{\bottomfraction}{1}
+\renewcommand{\textfraction}{0}
 \renewcommand{\floatpagefraction}{1}
+\setcounter{topnumber}{9}
+\setcounter{bottomnumber}{9}
+\setcounter{totalnumber}{20}
 
@@
 {\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
-\caption{\label{fig:bg505_prefs_logo}
-The rescaled averaged site-specific amino-acid preferences for BG505.
-This logo plot shows the site-specific amino-acid preferences for BG505 after averaging between replicates and then rescaling them using the stringency parameter from Table \ref{tab:phydms} inferred for BG505 from the group-M alignment.
-Each site has a stack of 20 letters corresponding to the 20 amino acids.
-Letter heights, which sum to one at each site, are proportional to the site's preference for each amino acid.
-The top bar (R) indicates gp120 variable loops, other regions in gp120, or gp41.
-The bottom bar (WT) shows the wildtype amino-acid sequence for BG505.
-Sites are numbered according to the HXB2 numbering scheme~\cite{korber1998numbering}).
+\caption{\label{fig:BG505prefs}
+Amino-acid preferences for the BG505 Env.
+At each site, the height of the letter is proportional for that site's preference for that amino acid.
+The upper overlaid color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
+The lower overlaid color bar indicates the evidence that the site is evolving faster or slower than expected in nature given the experiments~\citep[see][]{bloom2017identification}.
+The letters above the logos indicate the wildtype amino acid in BG505.
+Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
+This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in Table \ref{tab:phydms}.
+The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
 }
-\figdata{The re-scaled preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
-\figdata{Sequence of the BG505 Env and mapping from sequential 1, 2, ... numbering of this sequence (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $Q$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
+\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
+\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 \caption{\label{fig:bf520_prefs_logo}
-The re-scaled averaged site-specific amino-acid preferences for BF520.
-This figure is the same as Fig \ref{fig:bg505_prefs_logo}, but for BF520 instead of BG505.
+Amino-acid preferences for the BF520 Env.
+This figure is the same as Fig \ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
 }
-\figdata{The re-scaled preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
-\figdata{Sequence of the BF520 Env and mapping from sequential 1, 2, ... numbering of this sequence (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
+\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
+\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
 \figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}

### 2017-12-03 paper/manuscript.tex
Results section on phydms comparison and logo plots

@@
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to compare to the earlier inferior LAI data set. 
-Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across both Envs, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
+Because of the clear superiority of the current measurements, in this paper we focus on the new BG505 and BF520 datasets, and do not attempt to compare to the inferior LAI dataset. 
+Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
 This fact is an initial hint that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
@@
 \subsection*{Amino-acid preferences of the BG505 and BF520 Envs and their relationship to HIV evolution in nature}
-Next, we sought to compare Env's amino-acid preferences between BG505 and BF520.
-We also sought to compare these homologs to LAI~\cite{haddox2016experimental}.
-However, a concern is that differences we see between homologs may be due to experimental differences, rather than true biological differences.
-For instance, the LAI experiments involved passaging the viral libraries for a substantially longer amount of time in cell culture, which may have imposed increased selection.
-Additionally, even though we conducted the BG505 and BF520 experiments using very similar protocols, the experiments seemed to have exerted different levels of purifying selection, as indicated by differential purging of stop codons (Fig \ref{fig:ch3_mutfreqs}A).
-The strength of selection in our experiments may also differ from the strength of selection in nature.
-Ideally, however, we would like to compare the preferences after rescaling them to reflect the intensity of selection in a natural setting.
-
-To address these concerns, we estimated differences in selection strength between our experiments and nature using a phylogenetic approach.
-A software package called \texttt{phydms} can be used to incorporate site-specific amino-acid preferences into substitution models for maximum-likelihood phylogenetics~\cite{hilton2017phydms}.
-These models, known as experimentally informed codon models (ExpCMs), define the probability of a substitution at site $r$ to amino-acid $j$ from amino-acid $i$ as being proportional to the ratio of the site's preferences for these amino acids: $\frac{\pi_{r,j}}{\pi_{r,i}}$.
-Thus, if $j$ is more preferred than $i$, the probability of substituting from $j$ to $i$ is greater than substituting from $i$ to $j$.
-These models also define a stringency parameter ($\beta$) that rescales the preference for each amino acid $a$ at a site to be: $\pi_{r,a}^{\beta} / \lbrack \sum_{a} \pi_{r,a}^{\beta} \rbrack$.
-Given an alignment of natural sequences, ExpCMs can be used to infer the value of this parameter that rescales the preferences to optimally describe the evolution of these sequences.
-If the inferred stringency parameter is greater than one, the indication is that selection in nature tends to prefer the same amino acids as the experiments, but with greater stringency.
-
-We created ExpCMs with each homolog's averaged preferences.
-We then used these ExpCMs to analyze the \textit{env} sequences used to make the trees in Fig \ref{fig:ch3_tree}.
-To do so, we first used \texttt{RAxML} to infer the topology of each tree using the GTRCAT model.
-Fixing this topology, we then optimized the tree's branch lengths using either a standard codon-substitution model (YNGKP M5) or one of the ExpCMs.
-Table \ref{tab:ch3_phydms} compares the performance of each model with respect to each tree.
-For both trees, the ExpCMs describe Env's natural evolution far better than the standard model, as gauged by differences Akaike information criterion (AIC), which quantifies the likelihood of the data given the model while taking into account the number of model parameters, with higher AIC values indicating worse model performance.
-Notably, BG505 and BF520's preferences describe Env's evolution substantially better than LAI's preferences, though the ExpCM for LAI still outperforms the standard model.
-As a control, for each homolog, we averaged that homolog's preferences across all sites and made a ExpCM that modeled each site using these averaged preferences.
-As expected, these site-averaged models perform worse than the site-specific ones, indicating that the preferences recapitulate site-specific constraints in nature.
-The site-averaged models perform about as well as the standard model, which also lacks site-specific information.
-Tables \ref{tab:ch3_phydms_groupM_supp} and \ref{tab:ch3_phydms_subtypeA_supp} show the results of ExpCMs made using the preferences of individual replicates.
+The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
+Direct comparisons between experimentally measured amino-acid preferences and amino-acid frequencies in natural sequences are confounded by the fact that the natural sequences are evolutionarily related.
+This problem can be overcome by making the comparison in a phylogenetic context.
+Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms,bloom2017identification}.
+Importantly, since we expect many sites in Env to be under diversifying selection from immunity, we extended the ExpCM's to draw the relative dN/dS parameter ($\omega$) from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
 
@@
 \caption{\label{tab:phydms}
-Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature better than a standard substitution model.
+Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
 Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
-Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing either deep mutational scanning dataset describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
-Both ExpCM models have a stringency parameter $>$1, indicating that natural evolution prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
-All models draw $\omega$ (the dN/dS ratio for the Goldman-Yang model, and the relative dN/dS ratio after accounting for the measured constraints for the ExpCM) from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
+Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
+Both ExpCM models have a stringency parameter $>$1.
+All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
 The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected, as determined using the approach in \citet{bloom2017identification}.
@@
 
-For both alignments, the ExpCMs inferred stringency parameters of $>$1 for both BG505 and BF520, indicating that selection was weaker in our experiments for these homologs than in nature.
-The results are largely consistent between the two alignments, indicating that the rescaling is robust to the different levels of sequence diversity we examined.
-We decided to rescale the preferences for BG505 and BF520 using the stringency parameters from the group-M analysis (Table \ref{tab:ch3_phydms}).
-These resulting rescaled preferences are shown in Figs \ref{fig:ch3_bg505_prefs_logo} and \ref{fig:ch3_bf520_prefs_logo} for BG505 and BF520, respectively.
-Since both homologs were rescaled with respect to the same sequences, we expect this rescaling to also normalize the preferences across experiments, helping to correct for differences in selection strength between experiments.
+Table~\ref{tab:phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
+In addition to simply comparing model fits, we can interpret the model parameters.
+ExpCM's have a stringency parameter that relates selection in the experiments to that in nature.
+A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
+Both ExpCM's have stringency parameters $>$1 (Table~\ref{tab:phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
+We can also interpret the $\omega$ parameter.
+For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous one -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
+But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
+Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that immune selection drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
 
@@
 {\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
-\caption{\label{fig:bf520_prefs_logo}
+\caption{\label{fig:BF520prefs}
 Amino-acid preferences for the BF520 Env.
@@
 
-\begin{figure}
-%\includegraphics[width=0.5\textwidth]{figures/diversifying_selection_corr.pdf}
-\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
-\caption{\label{fig:divselcorr}
-Correlation between sites of diversifying and purifying selection among homologs.
-}
-\figsupp[Amino-acid preferences and alignment frequencies at sites of purifying selection.]{
-This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
-Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
-For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
-The $Q$-value indicated is the maximum of that for the two homologs.
-}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
-\figsupp[Distribution of relative solvent accessibilities among sites of diversifying and purifying selection.]{
-Sites are grouped by whether they are under diversifying or purifying selection at $Q < 0.05$ in \emph{both} homologs, or whether they do not fall into either of these categories.
-The absolute solvent accessibility of each site in the BG505 SOSIP trimer in PDB 5FYL~\citep{?} was calculated using DSSP~\citep{?}, and normalized to a relative solvent accessibility using the absolute accessibilities provided by \citet{?}.
-As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
-}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
-\figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}
-\end{figure}
+A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
+Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
+At sites of known functional importance, these preferences are usually consistent with prior knowledge.
+For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
+Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
+The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
+For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
+It is of course beyond the scope of this paper to catalog associations between our measurements and all prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
 
@@
 }
+\end{figure}
+
+\subsection{Diversifying selection}
+
+\begin{figure}
+\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
+\caption{\label{fig:divselcorr}
+Correlation between sites of diversifying and purifying selection among homologs.
+}
+\figsupp[Amino-acid preferences and alignment frequencies at sites of purifying selection.]{
+This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
+Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
+For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
+The $Q$-value indicated is the maximum of that for the two homologs.
+}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
+\figsupp[Distribution of relative solvent accessibilities among sites of diversifying and purifying selection.]{
+Sites are grouped by whether they are under diversifying or purifying selection at $Q < 0.05$ in \emph{both} homologs, or whether they do not fall into either of these categories.
+The absolute solvent accessibility of each site in the BG505 SOSIP trimer in PDB 5FYL~\citep{?} was calculated using DSSP~\citep{?}, and normalized to a relative solvent accessibility using the absolute accessibilities provided by \citet{?}.
+As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
+}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
+\figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}
 \end{figure}

### 2017-12-04 paper/manuscript.tex
Partial progress on section on calculating shifts

@@
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
-This fact is an initial hint that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
+This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
 
-\subsection*{Amino-acid preferences of the BG505 and BF520 Envs and their relationship to HIV evolution in nature}
+\subsection*{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
 The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
-It is of course beyond the scope of this paper to catalog associations between our measurements and all prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
+It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
 
-\subsection*{Differences between homologs}
-Having rescaled the preferences of each homolog to optimally describe Env's evolution in nature, we next sought to compare the preferences between homologs to determine how much Env's preferences have shifted over evolutionary time.
-We estimated preferences for 670 and 662 sites in BG505 and BF520, respectively.
-We compared the subset of 616 sites that are shared between homologs and are readily alignable in the group-M multiple-sequence alignment.
-Thus, in total, we compared 12,320 (=616$\times$20) amino-acid preferences between the homologs.
-
-As an initial gene-wide measure of these shifts, we simply quantified the correlation of the preferences between homologs.
-We expected some of the observed differences to be due to experimental noise rather than true biological differences.
-We estimated this noise by correlating the preferences between replicates for the same homolog after rescaling the preferences for each replicate using the stringency parameter from Table \ref{tab:ch3_phydms} for the appropriate homolog, as inferred from the group-M alignments.
-Fig \ref{fig:ch3_homologs_corr}A shows this correlation between a single pair of replicates for both BG505 and BF520.
-These comparisons provide an approximate ceiling for the expected correlation of replicates between homologs if both homologs have identical preferences.
-Conversely, if Env's preferences have substantially shifted over evolutionary time, the correlation between homologs is expected to be much lower.
-Fig \ref{fig:ch3_homologs_corr}B shows the correlation between homologs for a single pair of replicates.
-These replicates are nearly as well correlated as two replicates from the same homolog, suggesting that Env's preferences are substantially conserved between homologs.
-As a control, we correlated Env's preferences to the preferences of a non-homologous protein -- influenza hemagglutinin (HA) -- estimated in another study ~\cite{doud2016accurate} and rescaled with the appropriate stringency parameter from Table 1 of that study.
-HA's preferences were highly repeatable between the pair of replicates shown in Fig \ref{fig:ch3_homologs_corr}A.
-However, comparing replicates between BG505 and HA across all 480 sites that overlap in sequential numbering yielded a very low correlation, as would be expected when comparing preferences for non-homologous sites.
-Fig \ref{fig:ch3_homologs_corr} C shows that these trends are consistent across all three replicates for each protein.
-Thus, the correlations between Env homologs are much higher than the expected correlation between two non-homologous proteins.
+\subsection*{Shifts in amino-acid preferences between BG505 and BF520}
+The most fundamental question that we seek to address is if the effects of mutations are similar across Envs, or highly idiosyncratic to the particular Env variant under study.
+We can get some hint simply by examining Figure~\ref{fig:mutfreqs}B, which shows the correlations among the amino-acid preferences measured for all experimental replicates for both BG505 and BF520.
+All replicates are strongly correlated with each other regardless of which Env was used, indicating that many aspects of the preferences are conserved between BG505 and BF520.
+But the correlations between replicate measurements for the same Env are always greater than or equal to those for replicate measurements on different Envs, suggesting at least some shifts in preferences between BG505 and BF520.
+However, simply comparing correlation coefficients in this way cannot identify specific sites where mutational effects have shifted, nor can it quantify the magnitude of any shifts.
 
@@
 \includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
-\caption{\label{fig:diffprefs}
-Difference in preferences.
+\caption{\label{fig:prefsdist}
+Env sites with shifted amino-acid preferences between BG505 and BF520.
+{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at five example sites.
+We have triplicate measurements for each Env.
+We calculate the distance between each pair of replicate measurements, and group these into comparisons \emph{between} the two Envs and \emph{within} replicates for the same Env.
+We compute the root-mean-square distance (RMSD) for both sets of comparisons, which we denote as RMSD$_{\rm{between}}$ and RMSD$_{\rm{within}}$.
+The latter quantity is a measure of experimental noise.
+The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
+{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites.
+The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
+The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
+We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
+Using this approach, 30 of the 659 sites have significant shifts (corrected distance $\ge$0.22).
+{\bf (C)} All sites that have significantly shifted their amino-acid preferences at an FDR of 0.01.
+For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
 }
-\figdata{The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
+\figdata{\label{figdata:prefsdist}The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
 \end{fullwidth}
@@
 
-Next, we sought to quantify shifts in Env's preferences at a site-specific level.
-Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.
-Thus, we quantified shifts using an approach that corrects for noise, similar to the one used in a previous study that deep mutational-scanning data between homologs of influenza's nucleoprotein \cite{doud2015site}.
-Fig \ref{fig:ch3_distances}A shows how we quantified shifts for a few example sites.
-For a given site $r$, we define the distance in preferences between an arbitrary pair of replicates $i$ and $j$ as: $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
-We then use this metric to measure the distance between all pairwise combinations of replicates from both homologs.
-We quantified the un-corrected distance between homologs as the root mean square of all replicate-replicate distances between homologs ($RMSD_{between}$).
-Since $D_{r}^{i,j}$ can range between 0-1, $RMSD_{between}$ also ranges between 0-1, with 0 indicating no differences and 1 indicating extreme differences between the homologs.
-Next, we quantified experimental noise as the root mean square of all replicate-replicate distances where both replicates came from the \textit{same} homolog ($RMSD_{within}$).
-Finally, we computed the noise-corrected distance between homologs ($RMSD_{corrected}$) by subtracting $RMSD_{between}$ by $RMSD_{within}$.
-
-Sites where we repeatedly measured large differences between homologs are expected to have $RMSD_{corrected}$ values much greater than zero (e.g., sites 512 and 309; Fig \ref{fig:ch3_distances}A).
-In contrast, sites where we repeatedly measured small differences between homologs are expected to have $RMSD_{corrected}$ values close to zero (e.g., site 296).
-Sites where the noise completely overwhelmed the biological signal are also expected to have $RMSD_{corrected}$ values close to zero (e.g., site 607).
-Thus, a $RMSD_{corrected}$ value near zero could indicate that a site's preferences are conserved between homologs, or that the noise at a site is high.
+We therefore used a more statistically rigorous approach to identify sites where the amino-acid preferences had shifted between BG505 and BF520 by an amount that exceeds the noise in our experiments.
+We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from Table~\ref{tab:phydms} to uniformly calibrate all measurements to match the stringency of natural selection on clade A Envs.
+We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (Figure~\ref{fig:prefsdist}-source~data~\ref{figdata:prefsdist}).
+For each of these sites, we then calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in Figure~\ref{fig:prefsdist}A.
+This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates of the same Env.
 
@@
 
+\subsubsection*{Preference differences}Next, we sought to quantify shifts in Env's preferences at a site-specific level.
+Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.
+Thus, we quantified shifts using an approach that corrects for noise, similar to the one used in a previous study that deep mutational-scanning data between homologs of influenza's nucleoprotein \cite{doud2015site}.
+Fig \ref{fig:ch3_distances}A shows how we quantified shifts for a few example sites.
+For a given site $r$, we define the distance in preferences between an arbitrary pair of replicates $i$ and $j$ as: $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
+We then use this metric to measure the distance between all pairwise combinations of replicates from both homologs.
+We quantified the un-corrected distance between homologs as the root mean square of all replicate-replicate distances between homologs ($RMSD_{between}$).
+Since $D_{r}^{i,j}$ can range between 0-1, $RMSD_{between}$ also ranges between 0-1, with 0 indicating no differences and 1 indicating extreme differences between the homologs.
+Next, we quantified experimental noise as the root mean square of all replicate-replicate distances where both replicates came from the \textit{same} homolog ($RMSD_{within}$).
+Finally, we computed the noise-corrected distance between homologs ($RMSD_{corrected}$) by subtracting $RMSD_{between}$ by $RMSD_{within}$.
+
+Sites where we repeatedly measured large differences between homologs are expected to have $RMSD_{corrected}$ values much greater than zero (e.g., sites 512 and 309; Fig \ref{fig:ch3_distances}A).
+In contrast, sites where we repeatedly measured small differences between homologs are expected to have $RMSD_{corrected}$ values close to zero (e.g., site 296).
+Sites where the noise completely overwhelmed the biological signal are also expected to have $RMSD_{corrected}$ values close to zero (e.g., site 607).
+Thus, a $RMSD_{corrected}$ value near zero could indicate that a site's preferences are conserved between homologs, or that the noise at a site is high.
+
 \section{Acknowledgments}

### 2017-12-05 paper/manuscript.tex
more progress on shifts section, includng material about stability

@@
 The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
-By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
+By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
 Overall, our work provides across-strain genotype-to-phenotype maps that inform the analysis of Env's structure and evolution.
@@
 As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
-Our libraries mutagenized 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
+Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
 The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, in this paper we focus on the new BG505 and BF520 datasets, and do not attempt to compare to the inferior LAI dataset. 
+Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to the earlier inferior LAI dataset. 
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
@@
 Amino-acid preferences for the BF520 Env.
-This figure is the same as Fig \ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
+This figure is the same as Figure~\ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
 }
@@
 \subsection*{Shifts in amino-acid preferences between BG505 and BF520}
-The most fundamental question that we seek to address is if the effects of mutations are similar across Envs, or highly idiosyncratic to the particular Env variant under study.
+The most fundamental question that we seek to address is how similar the effects of mutations are between the two Envs.
 We can get some hint simply by examining Figure~\ref{fig:mutfreqs}B, which shows the correlations among the amino-acid preferences measured for all experimental replicates for both BG505 and BF520.
@@
 But the correlations between replicate measurements for the same Env are always greater than or equal to those for replicate measurements on different Envs, suggesting at least some shifts in preferences between BG505 and BF520.
-However, simply comparing correlation coefficients in this way cannot identify specific sites where mutational effects have shifted, nor can it quantify the magnitude of any shifts.
+However, simply comparing correlation coefficients in this way does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
 
@@
 The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
-The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
+The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin (HA) protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
 We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
@@
 For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
+The sites are sorted by the magnitude of the shift.
 }
@@
 
-We therefore used a more statistically rigorous approach to identify sites where the amino-acid preferences had shifted between BG505 and BF520 by an amount that exceeds the noise in our experiments.
-We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from Table~\ref{tab:phydms} to uniformly calibrate all measurements to match the stringency of natural selection on clade A Envs.
+We therefore used a more rigorous approach to identify sites where the amino-acid preferences differ between BG505 and BF520 by an amount that exceeds the noise in our experiments.
+We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from Table~\ref{tab:phydms} to calibrate all measurements to the stringency of natural selection.
 We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (Figure~\ref{fig:prefsdist}-source~data~\ref{figdata:prefsdist}).
-For each of these sites, we then calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in Figure~\ref{fig:prefsdist}A.
-This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates of the same Env.
+For each site, we calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in Figure~\ref{fig:prefsdist}A.
+This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates for the same Env.
+Figure~\ref{fig:prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and several other sites with more intermediate behaviors.
 
-The blue histogram in Fig \ref{fig:ch3_distances}B shows the distribution of site-specific $RMSD_{corrected}$ values between BG505 and BF520 for all 616 sites being compared (Fig \ref{fig:ch3_homologs_corr}).
-Most sites have distances greater than zero.
-Yet, only a small fraction of sites have large distances as extreme as sites 512 and 309 ($RMSD_{corrected}$ = 0.30-0.47; Fig \ref{fig:ch3_distances}A), as indicated by the small tail at the positive end of the distribution.
-Thus, it appears that few sites have dramatically shifted in preference.
-However, since $RMSD_{corrected}$ is influenced by noise, and since this noise is heterogeneous between sites, it was unclear exactly what the distribution would look like if Env's preferences had actually shifted at a large number of sites.
-We estimated the expected shape of this distribution by comparing BG505's preferences with the preferences from influenza HA, comparing all 480 sites that overlap in primary sequence.
-Since these proteins are non-homologous, we expected large site-specific differences in their preferences, mimicking large shifts during evolution.
-The green histogram in Fig \ref{fig:ch3_distances}B shows the resulting distribution of $RMSD_{corrected}$ values between Env and HA.
-As expected, the values in the Env-HA comparison tend to be much larger than the values in the Env-Env comparison.
-Thus, most differences between Env homologs are small-to-intermediate in effect size relative to the typical differences we observe for non-homologous sites.
+The overall distribution of shifts between BG505 and BF520 is shown in Figure~\ref{fig:prefsdist}B.
+Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
+This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
+How should we interpret this distribution -- have mutational effects have shifted a lot, or not very much?
+We can establish an upper-bound for how much sites can shift by comparing Env to a \emph{non}-homologous protein.
+Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
+Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
+We can also establish a lower-bound by creating a null distribution for the expected shifts if all differences are simply due to experimental noise.
+This null distribution is created by randomizing the experimental replicates among Envs.
+Figure~\ref{fig:prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
+The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.
 
-To examine the preference shifts in greater detail, we made a logo plot that shows the estimated shift for each amino acid at each of the 616 sites being compared (Fig \ref{fig:ch3_diffs_logo}).
-As described in the figure legend, this logo plot shows the results of subtracting the averaged preferences between homologs and then adjusting the total height of the letters at each site to correspond to that site's $RMSD_{corrected}$ value from Fig \ref{fig:ch3_distances}B.
-Most sites have small-to-intermediate stack heights, reflecting the fact that most site have relatively small-to-intermediate $RMSD_{corrected}$ values in Fig \ref{fig:ch3_distances}B.
-Sites where we detect large shifts between homologs are distributed throughout Env's primary sequence.
-The observed shifts follow a variety of patterns.
-At some sites, the preferences are strongly shifted for just a few amino acids (e.g., sites 309 and 288).
-At other sites, the preferences are strongly shifted in terms of the overall number of tolerated mutations (e.g., sites 512, 516, 599).
-Many of the largest shifts involved shifting from one hyrdrophobic amino acid to another (e.g., sites 165, 288, 307, 309).
-Overall, the shifts in the preferences are highly heterogeneous between sites.
-We are currently validating our findings for a subset of mutations by independently cloning each of these mutations into proviral plasmids, and then testing the ability of mutant viruses to replicate in cell culture.
+We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (Figure~\ref{fig:prefsdist}B).
+There are 30 such sites at a false discovery rate of 0.1.
+Figure~\ref{fig:prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
+For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
+Such broadening and narrowing of a site's mutational tolerance is often observed during protein evolution, and is frequently~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability} although not always~\citep{starr2017alternative} linked to changes in protein stability -- with a more stable protein typically being more mutationally tolerant.
+Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
+It is therefore interesting to note that the shifted sites often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) tolerate more mutations in the more stable BG505 Env.
+
+ An important caveat is that the foregoing analysis only identifies sites with shifts that exceed the magnitude of the experimental noise.
+ It is possible that other sites have undergone small shifts not detectable by our experiments -- indeed, a recent study by \citet{starr2017pervasive} found that small-effect mutational shifts are common during the evolution of the Hsp90 protein in yeast.
+ However, it is not technically feasible to experimentally measure the fitness effects of mutations to HIV with the same precision as for model organisms such as yeast.
+Therefore, we can only conclude that 30 Env sites have undergone \emph{detectable} shifts in amino-acid preferences between BG505 and BF520.
+It is possible that more sensitive experiments would find additional small shifts, but our approach should have identified all sites of large to moderate shifts.
 
@@
 \section{Acknowledgments}
-\jdbcomment{add this}
+We thank Michael Doud and Orr Ashenberg for computer code that helped form the basis for some of the analyses.
+We thank Kelly Lee for helpful input about the relative stabilities of the BG505 and BF520 Envs.
+\jdbcomment{add more}
 

### 2017-12-05 paper/manuscript.tex
added Results section on entrenchment

@@
 
-\subsection*{Two Env homologs from clade A transmitted-founder viruses}
+\subsection{Two Env homologs from clade A transmitted-founder viruses}
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
@@
 
-\subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
+\subsection{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
 We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain~\citep{haddox2016experimental}.
@@
 
-\subsection*{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
+\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
 The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
@@
 
-\subsection*{Shifts in amino-acid preferences between BG505 and BF520}
+\subsection{Shifts in amino-acid preferences between BG505 and BF520}
 The most fundamental question that we seek to address is how similar the effects of mutations are between the two Envs.
@@
 Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
-It is therefore interesting to note that the shifted sites often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) tolerate more mutations in the more stable BG505 Env.
+The fact that the shifted sites usually (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) tolerate more mutations in the more stable BG505 Env is therefore consistent with the view that increased protein stability is often associated with shifts towards higher mutational tolerance.
 
- An important caveat is that the foregoing analysis only identifies sites with shifts that exceed the magnitude of the experimental noise.
- It is possible that other sites have undergone small shifts not detectable by our experiments -- indeed, a recent study by \citet{starr2017pervasive} found that small-effect mutational shifts are common during the evolution of the Hsp90 protein in yeast.
+However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
+For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
+Several other sites show similar but somewhat milder flips in which amino acids they prefer.
+In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to fully rationalize many aspects of the observed sequence-function relationships.
+
+ An important caveat to the foregoing analysis is that it only identifies the 30 sites with shifts that exceed the experimental noise.
+ It is possible that other sites have undergone small shifts not detectable by our experiments -- indeed, a recent study by \citet{starr2017pervasive} found that small shifts are common during the evolution of the Hsp90 protein in yeast.
  However, it is not technically feasible to experimentally measure the fitness effects of mutations to HIV with the same precision as for model organisms such as yeast.
 Therefore, we can only conclude that 30 Env sites have undergone \emph{detectable} shifts in amino-acid preferences between BG505 and BF520.
-It is possible that more sensitive experiments would find additional small shifts, but our approach should have identified all sites of large to moderate shifts.
+It is possible that more sensitive experiments would find additional small shifts.
 
-\subsection*{Looking at how shifts relate to evolution and structure}
+\subsection{Entrenchment of substitutions contributes to mutational shifts}
+What is the evolutionary process by which sites in Env shift their amino-acid preferences?
+One idea that has developed in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution.
+Specifically, after \citet{pollock2012amino} suggested that amino-acid preferences systematically shift at sites of substitutions, \citet{shah2015contingency} formalized a definition of entrenchment which \citet{starr2017pervasive} subsequently reduced to experimental practice.
+Entrenchment is defined as the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
+Given two homologs, if there is no entrenchment then the effect of mutating a site in the first homolog to its identity in the second will simply be the opposite of mutating the site in the second homolog to its identity in the first.
+But if there is entrenchment, then both mutations will be unfavorable, since the site is entrenched at its preferred identity in each homolog.
+
+\begin{figure}
+\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.7\textwidth]{figures/entrenchment.pdf}}
+\caption{\label{fig:entrenchment}
+Entrenchment of substitutions during the divergence of BG505 and BF520.
+There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
+The blue densities show the effects of all these mutations to BG505 and BF520.
+The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
+In the absence of entrenchment, the effect of mutating a site in BG505 to its identity in BF520 should be the negative of the effect of mutating the site in BF520 to its identity in BG505.
+In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
+Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions tend to be entrenched in the protein variant in which they have fixed.
+The effect of a mutation is quantified as the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
+}
+\end{figure}
+
+Figure~\ref{fig:entrenchment} shows the actual distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
+As expected under entrenchment, the average effect of these mutations is negative -- although there are a substantial number of sites where the substitutions distinguishing BG505 and BF520 are not entrenched, since not all mutational flips are deleterious.
+We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (Figure~\ref{fig:entrenchment}).
+This comparison shows that reverting even highly entrenched substitutions is still more favorable than random amino-acid mutations.
+Therefore, entrenchment occurs for some but not all substitutions that distinguish BG505 and BF520, and the magnitude of entrenchment is less than the effect of a typical random mutation.
+
+\subsection{Looking at how shifts relate to evolution and structure}
 
@@
 \figdata{Statistical testing of whether significantly shifted sites are more likely to have substituted between homologs are in \texttt{shifts\_vs\_subs\_table.txt}.}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.7\textwidth]{figures/entrenchment.pdf}}
-\caption{\label{fig:entrenchment}
-CAPTION
-}
 \end{figure}
@@
 
-\subsubsection*{Sequence numbering}
+\subsection{Sequence numbering}
 We use the HXB2 numbering system \cite{korber1998numbering} for each Env homolog unless otherwise noted.
@@
 
-\subsubsection*{Codon-mutant libraries}
+\subsection{Codon-mutant libraries}
 For BF520 \textit{env}, we used the codon-mutant libraries generated from a previous study \cite{dingens2017comprehensive}.
@@
 
-\subsubsection*{Generation and passaging of viruses}
+\subsection{Generation and passaging of viruses}
 For BF520, we analyzed the viruses generated and passaged in our previous study \cite{dingens2017comprehensive}.
@@
 
-\subsubsection*{Generation of samples for Illumina sequencing}
+\subsection{Generation of samples for Illumina sequencing}
 We deeply sequenced each plasmid library before selection and each viral library after the two passages, as well as the wildtype plasmids and wildtype viruses that served as controls.
@@
 
-\subsubsection*{Analysis of deep-sequencing data}
+\subsection{Analysis of deep-sequencing data}
 
@@
 
-\subsubsection*{Alignments and phylogenetic analyses of HIV-1 {\it env} sequences}
+\subsection{Alignments and phylogenetic analyses of HIV-1 {\it env} sequences}
 We made the trees in Fig \ref{ref:ch3_tree} using multiple-sequence alignments downloaded from the HIV sequence database (\url{http://www.hiv.lanl.gov/}) that we then manually curated in the following ways:
@@
 
-\subsubsection*{Preference differences}Next, we sought to quantify shifts in Env's preferences at a site-specific level.
+\subsection{Preference differences}Next, we sought to quantify shifts in Env's preferences at a site-specific level.
 Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.

### 2017-12-06 paper/manuscript.tex
more work on Results

@@
 \usepackage{tikz}
+
+\usepackage{amsmath} % or simply amstext
+\newcommand{\angstrom}{\textup{\AA}}
 
@@
 
+\subsection{Structural and evolutionary properties of shifted sites}
+
+\begin{figure}
+{\bf \Large A} \\
+\includegraphics[width=0.8\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
+\\
+{\bf \Large B} \hspace{0.28\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
+\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
+\hspace{0.01\textwidth}
+\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
+\hspace{0.01\textwidth}
+\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
+\caption{\label{fig:shiftfeatures}
+Characteristics of significantly shifted sites.
+{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from gray to orange according to the magnitude of the mutational shift at each site (orange indicates large shift).
+Sites that are significantly shifted according to Figure~\ref{fig:prefsdist}B are in spheres, and all other sites are in cartoon representation.
+{\bf (B)} 
+There is no significant difference in relative solvent accessibility of sites that have and have not undergone significant shifts.
+The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
+{\bf (C)} 
+Sites of significant shifts are clustered in Env's structure.
+Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
+{\bf (D)}
+Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520. 
+The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
+All panels only show the 577 sites that are resolved in the crystal structure.
+Of the 30 significantly shifted sites identified in Figure~\ref{fig:prefsdist}B, only 28 are crystallographically resolved.
+Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
+$P$ values were calculated using the Mann-Whitney $U$ test.
+}
+\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
+{The sites of significant shifts as determined in Figure~\ref{fig:prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
+However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
+{\input{figures/shifts_vs_subs_table}}
+\end{figure}
+
+What properties distinguish the sites that have undergone substantial shifts in their amino-acid preferences?
+Figure~\ref{fig:shiftfeatures}A shows the locations of the significantly shifted sites on the crystal structure of Env.
+There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
+However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is indeed the case (Figure~\ref{fig:shiftfeatures}C).
+Therefore, whatever process drives shifts in Env's mutational tolerance appear to affect physically interacting clusters of residues in a coordinated fashion.
+
+An obvious hypothesis is that the strongly shifted sites are either sites that have substituted between BG505 and BF520, or are sites that physically contact such substitutions.
+According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site itself and its neighbors, thereby shifting the amino-acid preferences of all sites in this physical cluster.
+
+Next, we sought to explore the evolutionary basis of these shifts.
+One prediction is that the shifts would be concentrated at sites that differ between homologs.
+For instance, at site 309, Env's preferences shift with wildtype amino acid, where the BG505 wildtype amino acid (isoleucine) is more preferred in BG505 and the BF520 wildtype amino acid (leucine) is more preferred in BF520.
+This trend was observed in another study that compared shifts in mutational effects between homologs of influenza's nucleoprotein~\cite{doud2015site}.
+However, when we considered all sites, we did not find evidence that shifts at variable sites were stronger than shifts at conserved sites (Fig \ref{fig:ch3_shifts_at_variable_sites}).
+In the future, I plan to examine other evolutionary explanations for these shifts.
+I also plan to examine these shifts in context of Env's structure, which might lend insight into their functional basis.
+
 \subsection{Entrenchment of substitutions contributes to mutational shifts}
@@
 
-\subsection{Looking at how shifts relate to evolution and structure}
-
-Next, we sought to explore the evolutionary basis of these shifts.
-One prediction is that the shifts would be concentrated at sites that differ between homologs.
-For instance, at site 309, Env's preferences shift with wildtype amino acid, where the BG505 wildtype amino acid (isoleucine) is more preferred in BG505 and the BF520 wildtype amino acid (leucine) is more preferred in BF520.
-This trend was observed in another study that compared shifts in mutational effects between homologs of influenza's nucleoprotein~\cite{doud2015site}.
-However, when we considered all sites, we did not find evidence that shifts at variable sites were stronger than shifts at conserved sites (Fig \ref{fig:ch3_shifts_at_variable_sites}).
-In the future, I plan to examine other evolutionary explanations for these shifts.
-I also plan to examine these shifts in context of Env's structure, which might lend insight into their functional basis.
-
-\begin{figure}
-{\bf \Large A} \\
-\includegraphics[width=0.9\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
-\\
-{\bf \Large B} \hspace{0.38\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
-\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
-\hspace{0.01\textwidth}
-\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
-\hspace{0.01\textwidth}
-\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
-\caption{\label{fig:shiftfeatures}
-CAPTION.
-Note that the analyses in panels B-D are restricted to only those sites that are resolved in the Env crystal structure.
-}
-\figdata{Statistical testing of whether significantly shifted sites are more likely to have substituted between homologs are in \texttt{shifts\_vs\_subs\_table.txt}.}
-\end{figure}
-
 \subsection{Diversifying selection}
@@
 Sites are grouped by whether they are under diversifying or purifying selection at $Q < 0.05$ in \emph{both} homologs, or whether they do not fall into either of these categories.
-The absolute solvent accessibility of each site in the BG505 SOSIP trimer in PDB 5FYL~\citep{?} was calculated using DSSP~\citep{?}, and normalized to a relative solvent accessibility using the absolute accessibilities provided by \citet{?}.
+Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
 As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.

### 2017-12-07 paper/manuscript.tex
some progress on diversifying selection section

@@
 The immediate evolutionary space accessible to any given Env is defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
-Most studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to any given Env variant.
+Most mutational studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to an Env variant.
 
@@
 These shifted sites usually strongly prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
-The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
+The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
 By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
@@
 All panels only show the 577 sites that are resolved in the crystal structure.
-Of the 30 significantly shifted sites identified in Figure~\ref{fig:prefsdist}B, only 28 are crystallographically resolved.
+Of the 30 significantly shifted sites in Figure~\ref{fig:prefsdist}B, only 28 are crystallographically resolved.
 Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
@@
 
-What properties distinguish the sites that have undergone substantial shifts in their amino-acid preferences?
-Figure~\ref{fig:shiftfeatures}A shows the locations of the significantly shifted sites on the crystal structure of Env.
+What distinguishes the sites that have undergone significant shifts in their amino-acid preferences?
+Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
 There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
-However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is indeed the case (Figure~\ref{fig:shiftfeatures}C).
-Therefore, whatever process drives shifts in Env's mutational tolerance appear to affect physically interacting clusters of residues in a coordinated fashion.
+However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
+Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
 
-An obvious hypothesis is that the strongly shifted sites are either sites that have substituted between BG505 and BF520, or are sites that physically contact such substitutions.
-According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site itself and its neighbors, thereby shifting the amino-acid preferences of all sites in this physical cluster.
+An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
+According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
+But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
+There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most of the shifted sites have not substituted (only 8 of the 30 shifted sites differ between the two Envs).
+The lack of a strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which have found that shifts are dramatically more pronounced at sites of substitutions.
+The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
+On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
 
-Next, we sought to explore the evolutionary basis of these shifts.
-One prediction is that the shifts would be concentrated at sites that differ between homologs.
-For instance, at site 309, Env's preferences shift with wildtype amino acid, where the BG505 wildtype amino acid (isoleucine) is more preferred in BG505 and the BF520 wildtype amino acid (leucine) is more preferred in BF520.
-This trend was observed in another study that compared shifts in mutational effects between homologs of influenza's nucleoprotein~\cite{doud2015site}.
-However, when we considered all sites, we did not find evidence that shifts at variable sites were stronger than shifts at conserved sites (Fig \ref{fig:ch3_shifts_at_variable_sites}).
-In the future, I plan to examine other evolutionary explanations for these shifts.
-I also plan to examine these shifts in context of Env's structure, which might lend insight into their functional basis.
-
-\subsection{Entrenchment of substitutions contributes to mutational shifts}
-What is the evolutionary process by which sites in Env shift their amino-acid preferences?
-One idea that has developed in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution.
-Specifically, after \citet{pollock2012amino} suggested that amino-acid preferences systematically shift at sites of substitutions, \citet{shah2015contingency} formalized a definition of entrenchment which \citet{starr2017pervasive} subsequently reduced to experimental practice.
-Entrenchment is defined as the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
-Given two homologs, if there is no entrenchment then the effect of mutating a site in the first homolog to its identity in the second will simply be the opposite of mutating the site in the second homolog to its identity in the first.
-But if there is entrenchment, then both mutations will be unfavorable, since the site is entrenched at its preferred identity in each homolog.
+\subsection{Entrenchment of substitutions modestly contributes to mutational shifts}
 
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.7\textwidth]{figures/entrenchment.pdf}}
+\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.68\textwidth]{figures/entrenchment.pdf}}
 \caption{\label{fig:entrenchment}
-Entrenchment of substitutions during the divergence of BG505 and BF520.
+Entrenchment of substitutions during Env evolution.
 There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
-The blue densities show the effects of all these mutations to BG505 and BF520.
+The blue densities show the effects of all these mutations to each Env.
 The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
-In the absence of entrenchment, the effect of mutating a site in BG505 to its identity in BF520 should be the negative of the effect of mutating the site in BF520 to its identity in BG505.
+In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
 In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
-Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions tend to be entrenched in the protein variant in which they have fixed.
+Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
 The effect of a mutation is quantified as the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
@@
 
-Figure~\ref{fig:entrenchment} shows the actual distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
-As expected under entrenchment, the average effect of these mutations is negative -- although there are a substantial number of sites where the substitutions distinguishing BG505 and BF520 are not entrenched, since not all mutational flips are deleterious.
+One idea that has recently gained support in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution.
+Specifically, after \citet{pollock2012amino} suggested that amino-acid preferences systematically shift at sites of substitutions, \citet{shah2015contingency} formalized a definition of entrenchment which \citet{starr2017pervasive} subsequently reduced to experimental practice.
+Entrenchment is the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
+Given two homologs, if there is no entrenchment then the effect of mutating a site in the first homolog to its identity in the second will simply be the opposite of mutating the site in the second homolog to its identity in the first.
+But if there is entrenchment, then both mutations will be unfavorable, since the site is entrenched at its preferred identity in each homolog.
+
+Figure~\ref{fig:entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
+As expected under entrenchment, the average effect of these mutations is deleterious -- although there are a substantial number of sites where the mutational flips are not deleterious. 
 We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (Figure~\ref{fig:entrenchment}).
-This comparison shows that reverting even highly entrenched substitutions is still more favorable than random amino-acid mutations.
+This comparison shows that even unfavorable inter-Env mutational flips are generally more favorable than random amino-acid mutations.
 Therefore, entrenchment occurs for some but not all substitutions that distinguish BG505 and BF520, and the magnitude of entrenchment is less than the effect of a typical random mutation.
+Entrenchment of substitutions therefore contributes to some of the mutational shifts.
+But given that many of these shifts occur at sites that do not even differ between the Envs (Figure~\ref{fig:shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.
 
-\subsection{Diversifying selection}
+\subsection{Comparing selection in the lab to natural selection}
+All the foregoing experiments measure the effects of mutations on viral replication \emph{in a T-cell line in the lab}.
+But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
+For instance, antibody pressure might increase the rate of evolution at some sites~\citep{wei2003antibody}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites.
+Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
+In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar} -- something that is of less utility since we already know that Env is under strong functional constraint.
 
@@
 \includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
-\caption{\label{fig:divselcorr}
-Correlation between sites of diversifying and purifying selection among homologs.
+\caption{\label{fig:divsel}
+Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
+We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
+We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
+{\bf (A)}
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
+Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
+{\bf (B)}
+Sites of slower or faster than expected evolution given the BF520 experiments.
+For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+{\bf (C)}
+The results are similar regardless of whether the BG505 or BF520 experiments are used.
+Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
+All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
+{\bf (D)} 
+A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
+Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form the hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
 }
-\figsupp[Amino-acid preferences and alignment frequencies at sites of purifying selection.]{
+\figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
+Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they do fall into neither of these categories.
+Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
+As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
+}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
+\figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
 This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
@@
 }{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
-\figsupp[Distribution of relative solvent accessibilities among sites of diversifying and purifying selection.]{
-Sites are grouped by whether they are under diversifying or purifying selection at $Q < 0.05$ in \emph{both} homologs, or whether they do not fall into either of these categories.
-Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
-As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
-}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
-\figdata{The $\omega_r$ and associated $P$-values plotted in this figure are in \texttt{merged\_omegabysite.csv}.}
+\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
 \end{figure}
+
+We determined whether each site in Env was evolving faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
 
@@
 \section{Acknowledgments}
-We thank Michael Doud and Orr Ashenberg for computer code that helped form the basis for some of the analyses.
+We thank Michael Doud and Orr Ashenberg for computer code that formed the basis for some of the analyses.
+We thank Andrew Ward for pointing out to us that some of the sites with slower-than-expected rates of evolution are involved in Env's conformational changes upon receptor binding.
 We thank Kelly Lee for helpful input about the relative stabilities of the BG505 and BF520 Envs.

### 2017-12-08 paper/manuscript.tex
draft of diversifying selection section

@@
 All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
-The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected, as determined using the approach in \citet{bloom2017identification}.
+The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
 Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
-Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results of analyses of the same tree with additional substitution models.
+Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
 }
@@
 \subsection{Comparing selection in the lab to natural selection}
-All the foregoing experiments measure the effects of mutations on viral replication \emph{in a T-cell line in the lab}.
+All the foregoing experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
 But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
-For instance, antibody pressure might increase the rate of evolution at some sites~\citep{wei2003antibody}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites.
+For instance, antibody pressure might increase the rate of evolution at some sites~\citep{albert1990rapid,wei2003antibody,richman2003rapid}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites.
 Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
-In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar} -- something that is of less utility since we already know that Env is under strong functional constraint.
+%In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar}.
+% -- something that is of less utility since we already know that Env is under strong functional constraint.
 
@@
 Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form the hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
+All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
 }
@@
 
-We determined whether each site in Env was evolving faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
+We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
+The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
+As shown in Table~\ref{tab:phydms}, the standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution.
+This finding is only of modest interest, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
+In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
+
+The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B.
+The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
+The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
+
+What causes some sites to evolve faster or slower in nature than expected from the experiments?
+The answer in both cases is likely to be immune selection.
+Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
+It is known that Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
+Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
+Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
+In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
+These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
+Sites that contribute to such immune-evasion strategies will be under a constraint in nature that is not present in our experiments, since the experiments do not select for immune evasion.
+%Indeed, many of the sites that evolve slower than expected appear to be involved in these general immune-evasion strategies.
+For instance, sites of N-linked glycans tend to be more conserved in nature than the experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as are sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
 

### 2017-12-08 paper/manuscript.tex
edits to all sections up to results

@@
 The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
-We used the T332N variant of BG505 Env because it has a glycosylation site that is targeted by many anti-HIV antibodies but is absent in wildtype BG505~\citep{sanders2013next}.
+We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
 The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the clade A alignment (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
-We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
+We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
 Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
@@
 Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
-Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed summaries of all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
@@
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to the earlier inferior LAI dataset. 
+Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
@@
 This problem can be overcome by making the comparison in a phylogenetic context.
-Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms,bloom2017identification}.
+Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms}.
 Importantly, since we expect many sites in Env to be under diversifying selection from immunity, we extended the ExpCM's to draw the relative dN/dS parameter ($\omega$) from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
@@
 At each site, the height of the letter is proportional for that site's preference for that amino acid.
-The upper overlaid color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
-The lower overlaid color bar indicates the evidence that the site is evolving faster or slower than expected in nature given the experiments~\citep[see][]{bloom2017identification}.
+The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
+The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}.
 The letters above the logos indicate the wildtype amino acid in BG505.
@@
 \figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
-\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
+\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
 \figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
@@
 This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
-How should we interpret this distribution -- have mutational effects have shifted a lot, or not very much?
-We can establish an upper-bound for how much sites can shift by comparing Env to a \emph{non}-homologous protein.
+How should we interpret this distribution -- have mutational effects shifted a lot, or not very much?
+We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
 Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
@@
 Several other sites show similar but somewhat milder flips in which amino acids they prefer.
-In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to fully rationalize many aspects of the observed sequence-function relationships.
+In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to rationalize many aspects of the observed sequence-function relationships, including the cause of these shifts.
 
@@
 
-What distinguishes the sites that have undergone significant shifts in their amino-acid preferences?
+What distinguishes the sites that have undergone significant shifts?
 Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
@@
 But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
-There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most of the shifted sites have not substituted (only 8 of the 30 shifted sites differ between the two Envs).
-The lack of a strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which have found that shifts are dramatically more pronounced at sites of substitutions.
+There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
+The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
 The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
@@
 \subsection{Comparing selection in the lab to natural selection}
-All the foregoing experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
+Our experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
 But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
@@
 Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
-%In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar}.
-% -- something that is of less utility since we already know that Env is under strong functional constraint.
 
@@
 The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
-As shown in Table~\ref{tab:phydms}, the standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution.
-This finding is only of modest interest, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
+The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
+This finding is unsurprising, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
 In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
 
-The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B.
+The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B, and also overlaid on the logoplots in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} as the $\omega_r$ values.
 The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
@@
 Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
-It is known that Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
+Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
 Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
@@
 These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
-Sites that contribute to such immune-evasion strategies will be under a constraint in nature that is not present in our experiments, since the experiments do not select for immune evasion.
-%Indeed, many of the sites that evolve slower than expected appear to be involved in these general immune-evasion strategies.
-For instance, sites of N-linked glycans tend to be more conserved in nature than the experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as are sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
+For instance, sites of N-linked glycans change slower than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.

### 2017-12-09 paper/manuscript.tex
first 2 paragraphs of Results

@@
 Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
-Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the evolutionary capacity of most sites is conserved between moderately diverged viral strains.
+Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the amino-acid preferences of most sites are conserved between moderately diverged viral strains.
 \end{abstract}
@@
 \section{Discussion}
+We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral replication in cell culture.
+These sequence-function maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
+However, the major value of these maps comes not from comparing them with prior knowledge, but from the fact that prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
+Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive sequence-function maps enable this in a way that prior piecemeal studies of individual mutations cannot.
+
+But our maps come with a potentially serious caveat: each one is measured for just a single Env variant.
+The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
+This question is reminiscent of one that was grappled with in the early days of protein crystallography, when it first became possible to build maps of a protein's structure.
+Because it was not (and is still not) possible to crystallize every variant of every protein, it was necessary to determine whether sequence-structure maps could be usefully generalized among homologs.
+Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
+This rough generalizability of sequence-structure maps holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
+
 We experimentally estimated the effects of all single amino-acid changes to two Env homologs with 85\% amino-acid identity.

### 2017-12-10 paper/manuscript.tex
full draft of Discussion

--- 2017-12-08 paper/manuscript.tex
@@
 \section{Discussion}
-We experimentally estimated the effects of all single amino-acid changes to two Env homologs with 85\% amino-acid identity.
-We did so using a high-throughput technique called deep mutational scanning.
-For each homolog, this approach involved making libraries of \textit{env} with random codon mutations, selecting for mutations that supported viral replication in cell culture, and then deep sequencing the starting and ending libraries to quantify the effect of each mutation.
-These experiments imposed strong purifying selection, as indicated by the near-complete depletion of mutations leading to premature stop codons.
-For each homolog, the results allowed us to estimate each site's preference for each of the 20 amino acids.
-We found that these estimates were largely repeatable between experimental replicates of the same homolog.
+We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral replication in cell culture.
+These amino-acid preference maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
+However, the main value of these maps comes not from comparing them with prior knowledge, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
+Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in a way that prior piecemeal studies of individual mutations cannot.
 
-We then compared site-specific amino-acid preferences between homologs.
-In an initial gene-wide comparison, we simply correlated the preferences between homologs across all sites.
-We expected that the differences we observed would be due to a combination of true biological differences and experimental noise.
-We quantified experimental noise by correlating the preferences between experimental replicates of the same homolog.
-When we then correlated the preferences between replicates from different homologs, we found that the correlation was nearly as high as the correlation between replicates from the same homolog, suggesting that the true preferences of each homolog are fairly well conserved.
-As a control, we compared Env's preferences with the preferences from a non-homologous protein -- influenza HA -- among sites that overlap in primary sequence of these proteins.
-As expected, the correlation in preferences between these non-homologous proteins was very low.
-This finding provided the first indication that although Env's preferences differ to some extent between homologs, they are still much more conserved than expected for two non-homologous proteins.
+But our maps come with a serious caveat: each one is measured for just a single Env variant.
+The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
+This question is reminiscent of one that was grappled with in the early days of protein crystallography, when it first became possible to build maps of a protein's structure.
+Because it was not (and is still not) possible to crystallize every variant of a protein, it was necessary to determine whether protein-structure maps could be usefully generalized among homologs.
+Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
+This rough generalizability of protein-structure maps holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
 
-Next, we quantified shifts in Env's amino-acid preferences at a more site-specific level.
-We computed shifts using a metric that takes into account experimental noise.
-At each site, this metric estimates the expected shift due to the noise, as captured by experimental replicates, and then corrects for this noise when measuring the shifts between homologs.
-We found that for most sites, the shifts between homologs were small-to-intermediate in effect size, with only a few sites having large shifts.
-We used the same metric to compare Env and influenza HA at sites that overlap in primary sequence.
-As expected for two non-homologous proteins, we found that most sites had large shifts in mutational effects.
-Thus, this site-specific analysis provided additional support that the shifts observed between Env homologs were not nearly as large as shifts between non-homologous proteins.
+Our results show that amino-acid preference maps of Env also have a useful level of conservation for many purposes.
+From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site.
+From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral replication imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
 
-Overall, these results help elucidate the strength of epistasis in Env's long-term evolution.
-Although large shifts in mutational effects can occur during protein evolution, our results indicate that such events were rare for the Env homologs we analyzed.
-Deep mutational scanning has also been used to analyze mutational shifts between other homologs.
-One study compared two homologs of influenza's nucleoprotein with 94\% amino-acid identity~\cite{doud2015site}.
-This study found that most mutational effects were conserved.
-A lower-throughput study of nucleoprotein mutations with a variety of stability effects found that these effects were also largely conserved among homologs that ranged from 94\%-72\% amino-acid identity~\cite{ashenberg2013mutational}.
-Another deep mutational-scanning study compared three homologous TIM-barrel proteins, each with 30-40\% amino-acid identity to one another~\cite{chan2017correlation}.
-Even for these considerably divergent homologs, mutational effects were still found to be correlated between homologs at many sites.
-Thus, our findings with Env are qualitatively similar to the above studies in suggesting that although epistasis can cause mutational effects to diverge over time, it does not usually completely erase site-specific preferences.
+Of course, from the perspective of protein evolution, the most interesting sites are the exceptions to the general conservation of amino-acid preferences.
+Consistent with studies of other proteins~\citep{natarajan2013epistasis,harms2014historical,doud2015site,starr2017pervasive}, we find a subset of sites that change markedly in which mutations they tolerate.
+Some shifted sites simply accommodate more amino acids in the more stable BG505 Env -- a type of shift that has been well-documented for other proteins~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
+But interestingly, there is no strong trend for shifts to be enhanced at sites that differ between BG505 and BF520.
+Recent studies of protein evolution have focused on the idea that substitutions become ``entrenched'' as sites shift to accommodate new amino acids~\citep{pollock2012amino,shah2015contingency,bazykin2015changing,starr2017pervasive}.
+Indeed, a prior protein-wide comparison of amino-acid preferences across homologs found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
+But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: sites that differ between BG505 and BF520 are not more shifted than other sites.
+This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
 
-Our results also shed light on the evolutionary basis of shifts in mutational effects.
-The deep mutational scan comparing nucleoprotein homologs found that shifts were enriched at sites that differed in wildtype amino-acid sequence between homologs~\cite{doud2015site}.
-This trend might be expected if proteins quickly evolve to accommodate historical amino-acid changes, as suggested from computational modeling~\cite{pollock2012amino}.
-However, the median shift in Env's preferences was similar between sites that were the same vs. different in wildtype sequence.
-Additional comparisons will be required to determine which pattern is more typical.
+Going forward, it will be interesting to determine how the results of experiments such as ours are affected by defined modifications in viral genotype or selective pressure.
+A challenge of studying HIV is that almost everything must be simplified to make it tractable.
+Our studies are no different -- after all, we are measuring viral replication in a T-cell line in the lab, not in infected humans.
+But our experiments provide highly quantitative data that is amenable to rigorous analysis.
+Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between our experiments and nature to identify sites under different pressure in the lab and in humans.
+Future experiments that modulate selection pressures in other relevant ways should provide further insight into the forces that drive and constrain HIV's evolution.
 
-Future work could expand upon our findings to analyze other Env homologs and the effects of mutations on other Env phenotypes.
-In this study, we compared two homologs from subtype A.
-It is possible that mutational effects are more shifted between more divergent Envs.
-Deep mutational scanning of Env from a different subtype could address this question (our previous results for LAI are difficult to compare with BG505 and BF520, for reasons described above).
-The specific phenotype we selected for in our study was the ability of Env to support viral replication in cell culture in the \textit{absence} of immune selection (e.g., antibodies).
-However, it is of great interest to characterize the effects of mutations to Env on antibody escape, especially for antibodies being used in immunotherapies or as templates in vaccine design.
-We have previously used deep mutational scanning to comprehensively identify single amino-acid mutations that allow BF520 Env to escape a bNAb~\cite{dingens2017comprehensive}.
-This technique, when applied to two different Env homologs, could be used to comprehensively determine the extent to which antibody-escape mutations have shifted during Env evolution.
-
+\clearpage
 \section{Methods and Materials}

### 2017-12-10 paper/manuscript.tex
some minor revisions to Intro and Results

@@
 Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
-Most mutations similarly affect both Envs, but the amino-acid preferences of a substantial minority of sites have clearly shifted.
-These shifted sites usually prefer a specific amino acid in one Env, but can tolerate many amino acids in the other.
+Most mutations similarly affect both Envs, but the amino-acid preferences of a minority of sites have clearly shifted.
+These shifted sites usually prefer a specific amino acid in one Env, but tolerate many amino acids in the other.
 Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
@@
 Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
-By comparing these complete mutational maps, we identify sites where amino-acid preferences have shifted between the Envs.
-Most sites show no detectable shifts, but 30 sites have clearly altered their amino-acid preferences.
-These shifted sites usually strongly prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
-The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
+We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
+Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
+These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
+The shifted sites cluster in structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
 By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
-Overall, our work provides across-strain genotype-to-phenotype maps that inform the analysis of Env's structure and evolution.
+Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.
 
@@
 The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
-This finding is unsurprising, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
+This finding is unsurprising, since it is well known that Env is under functional constraint.
 In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
 
-The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B, and also overlaid on the logoplots in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} as the $\omega_r$ values.
+The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B, and overlaid on the logoplots in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} as the $\omega_r$ values.
 The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
@@
 However, the main value of these maps comes not from comparing them with prior knowledge, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
-Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in a way that prior piecemeal studies of individual mutations cannot.
+Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in ways that prior piecemeal studies of mutations cannot.
 
@@
 But our experiments provide highly quantitative data that is amenable to rigorous analysis.
-Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between our experiments and nature to identify sites under different pressure in the lab and in humans.
+Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
 Future experiments that modulate selection pressures in other relevant ways should provide further insight into the forces that drive and constrain HIV's evolution.

### 2017-12-10 paper/manuscript.tex
progress on Methods up thru deep sequencing analysis

@@
 \usepackage{tikz}
+\usepackage{seqsplit}
 
@@
 
-\subsection{Sequence numbering}
-We use the HXB2 numbering system \cite{korber1998numbering} for each Env homolog unless otherwise noted.
-We defined Env's ``variable loops'' according to \url{http://www.hiv.lanl.gov/}, but did not consider the disulfide-bonded cysteines as part of the loops.
+\subsection{Creation of codon-mutant libraries}
+Our codon mutant libraries mutagenized all sites in \textit{env} to all 64 codons, except that the signal peptide and cytoplasmic tail were not mutagenized.
+The rationale for excluding these regions is that they are not part of Env's ectodomain, and may be prone to mutations that strongly modulate Env's expression level.
 
-\subsection{Codon-mutant libraries}
-For BF520 \textit{env}, we used the codon-mutant libraries generated from a previous study \cite{dingens2017comprehensive}.
-For BG505 \textit{env}, we generated codon-mutant libraries using essentially the same methods as the previous study, but with a few modifications.
-We computationally generated the sequences of the codon-tiling primers for the PCR-mutagenesis step using the same approach as in~\cite{dingens2017comprehensive}.
-The sequences of these primers will be made available as a supplemental file upon publication of this work.
-The end primers for this mutagenesis step were: 5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3' and 5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'.
-As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \cite{poss1999variants} -- another subtype-A transmitted/founder virus, using the same high-efficiency cloning vector used previously \cite{dingens2017comprehensive}.
-To do so, we first digested the cloning vector with BsmBI.
-We then used PCR to elongate the amplicons to include on either end 30 basepairs that are identical in sequence to the ends of the BsmBI-digested vector.
-The primers we used for this PCR were:
+The codon-mutant libraries were generated using the approach originally described in \citet{bloom2014experimentally}, with the modification of \citet{dingens2017comprehensive} to ensure more uniform primer melting temperatures.
+The computer script used to design the mutagenesis primers (along with some detailed implementation notes) is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
+For BF520, the three libraries are the same ones described by \citet{dingens2017comprehensive}.
+For BG505, we created three libraries for this study.
+The wildtype BG505 sequence used for these libraries is in Supplemental~file~\ref{suppfile:BG505}.
+The BG505 mutagenesis primers are in Supplemental~file~\ref{suppfile:BG505primers}.
+
+The end primers for the BG505 mutagenesis were: 5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3' and 5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'.
+As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \citep[another subtype-A transmitted/founder virus;][]{poss1999variants} using the high-efficiency cloning vector described in \citet{dingens2017comprehensive}.
+For this cloning, we digested the cloning vector with BsmBI, and then used PCR to elongate the amplicons to include 30 nucleotides at each end that were identical in sequence to the ends of the BsmBI-digested vector.
+The primers for this PCR were:
 5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3' and 5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'.
@@
 The transformations yielded between 1.5-3.6 million unique clones for each of the three replicate libraries, as estimated by plating 1:2,000 dilutions of the transformations.
-As before, we scraped the plated colonies and maxiprepped the plasmid DNA, but this time we did not include the 4-hour outgrowth step after the scraping step.
-For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} in the \textit{env} locus of the full-length Q23 proviral plasmid.
+We scraped the plated colonies and maxiprepped the plasmid DNA; unlike in \citet{dingens2017comprehensive}, we did not include a 4-hour outgrowth step after the scraping step.
+For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} cloned into the same Q23 proviral plasmid.
+See Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger} and Figure~\ref{fig:mutfreqs}A for information on the average mutation rate in these libraries as estimated by Sanger sequencing and deep sequencing, respectively.
 
 \subsection{Generation and passaging of viruses}
-For BF520, we analyzed the viruses generated and passaged in our previous study \cite{dingens2017comprehensive}.
-For BG505, we generated and passaged viral libraries using essentially the same methods.
-First, for each replicate, we generated mutant viruses by transfecting 293T cells in three 6-well plates with a mixture of 2 ug mutant plasmid DNA and 6 uL FuGENE 6 Transfection Reagent (Promega, E269A) and 100 uL DMEM per well.
-293T cells were seeded in D10 media (DMEM supplemented with 10\% FBS, 1\% 200 mM L-glutamine, and 1\% of a solution of 10,000 units/mL penicillin and 10,000 $\mu$g/mL streptomycin) the day before transfection with 0.5 million cells per well, such that they were approximately 50\% confluent the next day.
-In parallel, we generated wildtype viruses by transfecting one 6-well plate of 293T cells with wildtype plasmid, using the same amount of DNA and transfection reagent as above.
-At 2days post-transfection, we harvested the transfection supernatant, passed it through a 0.2$\mu$m filter, treated the supernatant with DNAse to digest residual plasmid DNA (as in~\cite{haddox2016experimental}), and froze aliquots at -80$^{\circ}$C.
-We thawed and titered aliquots using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran as described in~\cite{dingens2017comprehensive}.
+For BG505, we generated mutant virus libraries from the proviral plasmid libraries by transfecting 293T cells in three 6-well plates (so 18 wells total per library) with a per-well mixture of 2 $\mu$g plasmid DNA, 6 $\mu$l FuGENE 6 Transfection Reagent (Promega, E269A), and 100 $\mu$l DMEM.
+The 293T cells were seeded at $5\times 10^5$ cells/well in D10 media (DMEM supplemented with 10\% FBS, 1\% 200 mM L-glutamine, and 1\% of a solution of 10,000 units/mL penicillin and 10,000 $\mu$g/mL streptomycin) the day before transfection, such that they were approximately 50\% confluent at the time of transfection.
+In parallel, we generated wildtype viruses by transfecting one 6-well plate of 293T cells with each wildtype plasmid replicate.
+At 2 days post-transfection, we harvested the transfection supernatant, passed it through a 0.2 $\mu$m filter to remove cells, treated the supernatant with DNAse to digest residual plasmid DNA as in~\citet{haddox2016experimental}, and froze aliquots at -80$^{\circ}$C.
+We thawed and titered aliquots using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran as described in~\citet{dingens2017comprehensive}.
 
-Next, we conducted the initial viral passage in SupT1.CCR5 cells (obtained from Dr. James Hoxie \cite{boyd2015mutations}).
-During this passage, cells were maintained in R10 media, which has the same composition as D10 (described above), but has RPMI-1640 (GE Healthcare Life Sciences, SH30255.01) in the place of DMEM, with 10 $\mu$g/mL DEAE-dextran to enhance viral infection.
+We conducted the low MOI viral passage illustrated in Figure~\ref{fig:dms_schematic}A in SupT1.CCR5 cells~\citep[obtained from Dr. James Hoxie;][]{boyd2015mutations}.
+During this passage, cells were maintained in R10 media, which has the same composition as the D10 described above, except RPMI-1640 (GE Healthcare Life Sciences, SH30255.01) is used in the place of DMEM.
+In addition, the media contained 10 $\mu$g/mL DEAE-dextran to enhance viral infection.
 We infected cells with 4 million (for replicate 1) or 5 million (for replicates 2 and 3) TZM-bl infectious units of mutant virus at an MOI of 0.01, with cells at a starting concentration of 1 million cells/mL in vented tissue-culture flasks (Fisher Scientific, 14-826-80).
-At 1 day post-infection, we pelleted cells, aspirated the supernatant, and resuspended cell pellets in fresh media including DEAE-dextran.
-At 2 days post-infection, we doubled the volume of each culture with fresh media including DEAE-dextran.
-At 4 days post-infection, we pelleted cells, passed the virus-containing supernatant through a 0.2 $\mu$m filter, concentrated the virus $\sim$30 fold using ultracentrifugation as described in~\cite{dingens2017comprehensive}, and then froze aliquots at -80$^{\circ}$C.
-In parallel, for each replicate, we also passaged 0.2 million (for replicate 1) or 0.5 million (for replicates 2 and 3) infectious units of wildtype virus with and using the same conditions.
-We thawed and titered these aliquots using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran.
+At day 1 post-infection, we pelleted cells, aspirated the supernatant, and resuspended cell pellets in the same volume of fresh media still including the DEAE-dextran.
+At 2 days post-infection, we doubled the volume of each culture with fresh media still including DEAE-dextran.
+At 4 days post-infection, we pelleted cells, passed the viral supernatant through a 0.2 $\mu$m filter, concentrated the virus $\sim$30 fold using ultracentrifugation as described in \citet{dingens2017comprehensive}, and froze aliquots at -80$^{\circ}$C.
+In parallel, for each replicate, we also passaged $2\times 10^5$ (for replicate 1) or $5 \times 10^5$ (for replicates 2 and 3) TZM-bl infectious units of wildtype virus using the same procedure.
+To obtain final titers for our concentrated virus, we thawed one of the aliquots stored at -80$^{\circ}$C and titered using the TZM-bl assay in the presence of 10 $\mu$g/mL DEAE-dextran.
 
-We conducted a second, much shorter viral passage by infecting cells with the passage-1 viruses.
-For each virus, we infected 1 million TMZ-bl infectious units into 1 million SupT1.CCR5 cells in the presence of 100 $\mu$g/mL DEAE-dextran (a 10-fold higher concentration than to concentration we used in the TZM-bl assay to titer the virus, meaning the actual number of infectious units in this passage was probably higher).
+For the final short-duration infection illustrated in Figure~\ref{fig:dms_schematic}A, for each replicate we infected $10^6$ TZM-bl infectious units into $10^6$ SupT1.CCR5 cells in the presence of 100 $\mu$g/mL DEAE-dextran (note that this is a 10-fold higher concentration of DEAE-dextran than for the other steps, meaning that the effective MOI of infection is higher if DEAE-dextran has the expected effect of enhancing viral infection).
 Three hours post-infection, we pelleted the cells and resuspended them in fresh media without any DEAE-dextran.
-At 12 hours post-infection, we pelleted cells, washed them once with PBS, and then used a miniprep kit to harvest reverse-transcribed unintegrated viral DNA.
+At 12 hours post-infection, we pelleted cells, washed them once with PBS, and then used a miniprep kit to harvest reverse-transcribed unintegrated viral DNA \citep{haddox2016experimental}.
 
-\subsection{Generation of samples for Illumina sequencing}
-We deeply sequenced each plasmid library before selection and each viral library after the two passages, as well as the wildtype plasmids and wildtype viruses that served as controls.
-First, we generated PCR amplicons of \textit{env} using the same approach as~\cite{dingens2017comprehensive} with the primers: 5'-GAAGACAGTGGCAATGAGAGTGATGG-3' and 5'-TTCCCTAACCTCAGGCCATCC-3'.
-Next, we sequenced these amplicons using a barcoded sub-amplicon sequencing strategy to reduce the sequencing error rate, as previously described~\cite{doud2016accurate,haddox2016experimental}, dividing \textit{env} into seven amplicons.
-The primer pairs used to generate these subamplicons are as follows, where N characters represent randomized sites in the barcode region of the primer:
+The generation, passaging and deep sequencing of BF520 was done in a highly similar fashion, except that we only had a single replicate of the wildtype control.
+Note that the final passaged BF520 mutant libraries analyzed here actually correspond to the ``no-antibody'' controls described in \citet{dingens2017comprehensive}, but that study did not analyze the initial plasmid mutant libraries relative to these passaged viruses, and so was not able to provide measurements of the amino-acid preferences.
 
-\begin{itemize}
-{\footnotesize
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGATCTTGGGGATGATAATAATCTGTAGTGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGGTGACATTGGTACACTGTAGAGTAAC-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNCCATGTGTAAAGTTAACCCCTCTCTGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGAACTTCTTATCCTTACACTTTAGGATCGC-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNCATACATTATTGTGCCCCAGCTGG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNCAATGTGCTTGTCTTATATCCCCTATTATGTC-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGGACAAGCATTCTATGCAACAGGG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNTGCTTTATTCTGCATGGGAGAGTTATAC-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGTCAAATAGCACGGGGTCAAATGAC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNCCAAGGAAGACAGCTCCTATTCCAAC-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGTGGTGGGGAGAGAAAAAAGAGC-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNGTTTCTATTACTCCAACTAGAGTTCCAGGG-3'
-\item 5'-CTTTCCCTACACGACGCTCTTCCGATCTNNNNNNNNGGAAAACTCATCTGCACCACTAATGTG-3' and 5'-GGAGTTCAGACGTGTGCTCTTCCGATCTNNNNNNNNTGAGTATCCCTGCCTAACTCTATGTATTACAG-3'
-}
-\end{itemize}
+\subsection{Illumina deep sequencing}
+We deep sequenced all of the samples shown in Figure~\ref{fig:mutfreqs}A: the plasmid mutant libraries and wildtype plasmid controls, and the cDNA from the final mutant viruses and wildtype virus controls.
+In order to increase the sequence accuracy, we used a barcoded-subamplicon sequencing strategy.
+This general strategy was originally applied in the context of deep mutational scanning by \citet{wu2014high}, and the specific protocol used in our work is described in \citet{doud2016accurate} (see also \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
 
-The resulting deep-sequencing data will be made available on the Sequence Read Archive upon the publication of this work.
+The primers used for BG505 are in Supplementary~file~\ref{suppfile:B505bcsubamp}.
+The primers used for BF520 are in \citet{dingens2017comprehensive}.
+The data generated by the Illumina deep sequencing are on the Sequence Read Archive under the accession numbers provided at the beginning of the Jupyter notebook in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
 
 \subsection{Analysis of deep-sequencing data}
+We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+A Jupyter notebook that performs the entire analysis including generation of most of the figures in this paper is in Supplementary~file~\ref{suppfile:code}, which also contains all necessary input data to fully reproduce the analysis.
+An HTML rendering of this notebook is in Supplementary~file~\ref{suppfile:html}.
+Citations are also due to \texttt{weblogo}~\citep{crooks2004weblogo} and \texttt{ggseqlogo}~\citep{?}, which were used in the generation of the logoplots. 
+\jdbcomment{also joypy?}
 
-We analyzed the deep-sequencing data using the \texttt{dms\_tools} software package~\cite{bloom2015software}.
-The specific code we used will be made available upon the publication of this work.
-A summary of our workflow is as follows, and involves multiple programs encoded in \texttt{dms\_tools}:
-First, we aligned sequencing reads to \textit{env} and computed site-specific codon mutation frequencies using \texttt{dms\_barcodedsubamplicons}, using \texttt{dms\_summarizealignments} to make plots summarizing these frequencies.
-Next, we inferred Env's preferences from the codon counts using \texttt{dms\_inferprefs}.
-We then used a custom script to compare these preferences between homologs.
+The Jupyter notebooks in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} also contain numerous plots that summarize relevant aspects of the deep sequencing such as read depth, per-codon mutation frequency, mutation types, etc.
+Supplementary~file~\ref{suppfile:code} also contains text files and CSV files with all of the numerical values shown in these plots.
 
@@
 We thank Kelly Lee for helpful input about the relative stabilities of the BG505 and BF520 Envs.
+We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
 \jdbcomment{add more}
@@
 
+\begin{suppfile}
+\caption{
+\label{suppfile:BG505}
+The sequence of the wildtype BG505 \textit{env} used in our study is in FASTA format in the file \texttt{BG505\_env.fasta}.
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:BG505primers}
+This file gives the primers used for the BG505 codon mutagenesis.
+\jdbcomment{Hugh, can you add this file?}
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:B505bcsubamp}
+The primers used for the BG505 barcoded-subamplicon sequencing are in the file \texttt{BG505\_bcsubamp\_primers.txt}.
+}
+\end{suppfile}
+
 \end{document}

### 2017-12-10 paper/manuscript.tex
full draft of methods

@@
 Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
-The effect of a mutation is quantified as the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
+The effect of a mutation is quantified as the log of the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
 }
@@
 
-The end primers for the BG505 mutagenesis were: 5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3' and 5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'.
+The end primers for the BG505 mutagenesis were: \seqsplit{5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3'} and \seqsplit{5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'}.
 As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \citep[another subtype-A transmitted/founder virus;][]{poss1999variants} using the high-efficiency cloning vector described in \citet{dingens2017comprehensive}.
@@
 The primers for this PCR were:
-5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3' and 5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'.
+\seqsplit{5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3'} and \seqsplit{5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'}.
 Next, we used NEBuilder HiFi DNA Assembly (NEB, E2621S) to clone the \textit{env} amplicons into the BsmBI-digested plasmids.
@@
 We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
-A Jupyter notebook that performs the entire analysis including generation of most of the figures in this paper is in Supplementary~file~\ref{suppfile:code}, which also contains all necessary input data to fully reproduce the analysis.
+The algorithm that goes from the deep-sequencing counts to the amino-acid preferences is that described in \citet{bloom2015software} (see also \url{https://jbloomlab.github.io/dms_tools2/prefs.html}).
+A Jupyter notebook that performs the entire analysis including generation of most of the figures in this paper is in Supplementary~file~\ref{suppfile:code}.
 An HTML rendering of this notebook is in Supplementary~file~\ref{suppfile:html}.
-Citations are also due to \texttt{weblogo}~\citep{crooks2004weblogo} and \texttt{ggseqlogo}~\citep{?}, which were used in the generation of the logoplots. 
-\jdbcomment{also joypy?}
 
 The Jupyter notebooks in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} also contain numerous plots that summarize relevant aspects of the deep sequencing such as read depth, per-codon mutation frequency, mutation types, etc.
-Supplementary~file~\ref{suppfile:code} also contains text files and CSV files with all of the numerical values shown in these plots.
+Supplementary~file~\ref{suppfile:code} also contains text files and CSV files with the numerical values shown in these plots.
 
-\subsection{Alignments and phylogenetic analyses of HIV-1 {\it env} sequences}
-We made the trees in Fig \ref{ref:ch3_tree} using multiple-sequence alignments downloaded from the HIV sequence database (\url{http://www.hiv.lanl.gov/}) that we then manually curated in the following ways:
-First, we removed sequences that differed in length from the HXB2 reference sequence (including gap characters), or which contained a premature stop codon, ambiguous residue, or frame-shift mutation.
-Next, we removed columns in the alignment for which we lack deep mutational scanning data or are not present in HXB2, columns that have $>$5\% gap characters, or columns in variable loops that looked poorly aligned by eye.
-The sequences used to make the trees in Fig \ref{ref:ch3_tree} were randomly selected from the larger alignment, with a defined number of sequences per clade, as described in the figure.
-The topologies of these trees were inferred by \texttt{RAxML} using the GTRCAT model.
-In the \texttt{phydms}~\cite{hilton2017phydms} analyses, we fixed the topologies of these trees, and then compared the ability of the different models to optimize the branch lengths.
-The trees shown in Fig \ref{ref:ch3_tree} are the results of optimizing the branch lengths using the YNGKP M5 model.
-The code we used to generate these alignments, build the trees, and conduct the \texttt{phydms} analyses will be made available upon publication of this work.
+Citations are also owed to \texttt{weblogo}~\citep[][\url{http://weblogo.threeplusone.com/}]{crooks2004weblogo} and \texttt{ggseqlogo}~\citep[][\url{https://omarwagih.github.io/ggseqlogo/}]{wagih2017ggseqlogo}, which were used in the generation of the logoplots.
 
-\subsection{Preference differences}Next, we sought to quantify shifts in Env's preferences at a site-specific level.
-Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.
-Thus, we quantified shifts using an approach that corrects for noise, similar to the one used in a previous study that deep mutational-scanning data between homologs of influenza's nucleoprotein \cite{doud2015site}.
-Fig \ref{fig:ch3_distances}A shows how we quantified shifts for a few example sites.
-For a given site $r$, we define the distance in preferences between an arbitrary pair of replicates $i$ and $j$ as: $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
-We then use this metric to measure the distance between all pairwise combinations of replicates from both homologs.
-We quantified the un-corrected distance between homologs as the root mean square of all replicate-replicate distances between homologs ($RMSD_{between}$).
-Since $D_{r}^{i,j}$ can range between 0-1, $RMSD_{between}$ also ranges between 0-1, with 0 indicating no differences and 1 indicating extreme differences between the homologs.
-Next, we quantified experimental noise as the root mean square of all replicate-replicate distances where both replicates came from the \textit{same} homolog ($RMSD_{within}$).
-Finally, we computed the noise-corrected distance between homologs ($RMSD_{corrected}$) by subtracting $RMSD_{between}$ by $RMSD_{within}$.
+\subsection{Alignments and phylogenetic analyses of Env sequences}
+A basic description of the process used to generate the clade A sequence alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, the alignment mask in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, and the phylogenetic tree in Figure~\ref{fig:tree} are provided in the legend to that figure.
+An algorithmic description of how the alignment and tree were generated are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
 
-Sites where we repeatedly measured large differences between homologs are expected to have $RMSD_{corrected}$ values much greater than zero (e.g., sites 512 and 309; Fig \ref{fig:ch3_distances}A).
-In contrast, sites where we repeatedly measured small differences between homologs are expected to have $RMSD_{corrected}$ values close to zero (e.g., site 296).
-Sites where the noise completely overwhelmed the biological signal are also expected to have $RMSD_{corrected}$ values close to zero (e.g., site 607).
-Thus, a $RMSD_{corrected}$ value near zero could indicate that a site's preferences are conserved between homologs, or that the noise at a site is high.
+For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in Figure~\ref{fig:tree}.
+The Goldman-Yang (or YNGKP) model used in Table~\ref{tab:phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
+For the ExpCM shown in Table~\ref{tab:phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
+These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
+For both the M5 model and the ExpCM with a gamma-distributed $\omega$, the number of categories in the discretized gamma distribution is four.
+
+Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for a wider set of models than those used in Table~\ref{tab:phydms}. 
+These include the M0 model of \citet{yang2000codon}, ExpCM without a gamma-distributed $\omega$, and ExpCM in which the amino-acid preferences are averaged across sites as a control to ensure that the improved performance of these models is due to their site-specificity.
+Note how for these Env alignments, using a gamma-distributed $\omega$ is very important in order for the ExpCMs to outperform the M5 model -- we suspect this is because there are many sites of strong diversifying selection.
+
+For detection of sites with faster or slower than expected evolution, we used the approach in \citet{bloom2017identification}, which is exactly modeled on the FEL approach of \citet{kosakovsky2005not} but extended to ExpCM.
+This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
+For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.
+
+Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contains the code that runs \texttt{phydms} to reproduce all of these analyses.
+
+\subsection{Identifying sites of shifted amino-acid preference}
+When identifying shifts in amino-acid preferences between the two Envs, we needed a way to quantify differences between the Envs while accounting for the fact that our measurements are noisy.
+The approach we use is based closely on that of \citet{doud2015site}, and is illustrated graphically in Figure~\ref{fig:prefsdist}A.
+The RMSD$_{\rm{corrected}}$ value is our measure of the magnitude of the shift.
+Figure~\ref{fig:prefsdist}A, its legend, and the associated text completely explains these calculations with the following exception: they do not detail how the ``distance'' between any two preference measurements was calculated.
+The distance between preferences at each site was simply defined as half of the sum of absolute value of the difference between preferences for each amino acid.
+Specifically, for a given site $r$, let $\pi_{r,a}^{i}$ be the preference for amino-acid $a$ in homolog $i$ (e.g., BG505) and let $\pi_{r,a}^{j}$ be the preference for that same amino acid in homolog $j$ (e.g., BF520).
+Then the distance between the homologs at this site is simply $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
+The factor of $\frac{1}{2}$ is used so that the maximum distance will always fall between zero and one.
+
+\subsection{Analysis of entrenchment}
+For the analysis in Figure~\ref{fig:entrenchment}, the results are presented in terms of the mutational effects rather than the amino-acid preferences. 
+If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in Table~\ref{tab:phydms}), then estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.
+
+\subsection{Data and code availability}
+All code and input data required to reproduce all analyses in this paper are in Supplementary~file~\ref{suppfile:code} (see also Supplementary~file~\ref{suppfile:html}).
+The deep sequencing data are on the Sequence Read Archive with the accession numbers listed in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
 
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-\jdbcomment{add more}
+\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.}
 

### 2017-12-10 paper/manuscript.tex
a few more tiny tweaks

@@
 This question is reminiscent of one that was grappled with in the early days of protein crystallography, when it first became possible to build maps of a protein's structure.
-Because it was not (and is still not) possible to crystallize every variant of a protein, it was necessary to determine whether protein-structure maps could be usefully generalized among homologs.
+Because it was not (and is still not) possible to crystallize every variant of a protein, it was necessary to determine whether protein structures could be usefully generalized among homologs.
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein-structure maps holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
 

### 2017-12-11 paper/manuscript_Adam.tex
added commented version of manuscript

--- 2017-12-10 paper/manuscript.tex
@@
 \newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+\newcommand{\asdcomment}[1]{\emph{\color{green} [#1]}}
 
@@
 
+\asdcomment{I might add one sentence relating to comparing selection in nature vs in lab/using these datasets in phylogenetic modeling to the abstract (alluding to identifying sites under immune selection).
+However, I had troubles adding anything meaningful here and still keeping the abstract clear.}
+
 \section{Introduction}
@@
 
-Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid}.
+Env's rapid evolution has dire consequences for anti-HIV immunity; it limits the breadth of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid} and presents a significant hurdle to developing a broadly protective vaccine.
 Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
@@
 
+\asdcomment{Here, it might also be nice to have a panel containing a tree with these viruses and all clades.
+ Hugh has a figure made already that include BG505 and BF520 on such a tree.}
+
 We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
+\asdcomment{Here, we should also cite Wu et al 2006 JVI (where the envs were isolated) and maybe Nduati et al 2000 JAMA (clinical trial where samples were collected).}
 The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
@@
 BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+\asdcomment{this language could be improved, I had a hard time with exactly what we are saying.}
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 
+\asdcomment{I think it might be nice to add one paragraph and a supplemental figure on one brief functional analysis - the mutational tolerance of glycans in cell culture. 
+Hugh has already done a lot of analyses on glycans that are and are not functionally conserved in both strains. 
+I think this is of a lot of interest (and has obvious immune evasion conclusions), and you could imagine these observations alone being a paper if assayed individually.}
 \subsection{Shifts in amino-acid preferences between BG505 and BF520}
@@
 Several other sites show similar but somewhat milder flips in which amino acids they prefer.
-In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to rationalize many aspects of the observed sequence-function relationships, including the cause of these shifts.
+In later sections, we investigate evolutionary and structural characteristics of some of these shifted sites.
 
@@
 Entrenchment of substitutions therefore contributes to some of the mutational shifts.
-But given that many of these shifts occur at sites that do not even differ between the Envs (Figure~\ref{fig:shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.
+But given that many of these shifts occur at sites that do not even differ between the Envs (Figure~\ref{fig:shiftfeatures}D), there is at least one additional unknown cause of the shifting of Env amino-acid preferences beyond entrenchment.
 
@@
 Our experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
-But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
+But HIV actually evolves in humans, where additional selection pressures on Env are rampant.
 For instance, antibody pressure might increase the rate of evolution at some sites~\citep{albert1990rapid,wei2003antibody,richman2003rapid}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites.
@@
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that Env structures are surprisingly similar between distantly related Env homologs (cite Stewart-Jones 2016, Verkerke 2016, Pugach et al 2015 JVI, Gristick et al 2017 Struc Bio, Julien et al 2015 PNAS).
 
 Our results show that amino-acid preference maps of Env also have a useful level of conservation for many purposes.
-From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site.
+From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse Env strains.
 From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral replication imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
+ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).
 \jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.}

### 2017-12-12 paper/manuscript.tex
added Adam comments to manuscript

@@
 
-We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
+We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{nduati2000effect,wu2006neutralization,goo2014early}.
 The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 Several other sites show similar but somewhat milder flips in which amino acids they prefer.
-In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to rationalize many aspects of the observed sequence-function relationships, including the cause of these shifts.
+In later sections, we investigate evolutionary and structural characteristics of these shifted sites.
 
@@
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter Env's structure and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, these alterations are sufficiently modest that a few structural maps have facilitated the interpretation of studies on a vast number of HIV strains.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, all characterized Env structures remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.
 
 Our results show that amino-acid preference maps of Env also have a useful level of conservation for many purposes.
-From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site.
+From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
 From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral replication imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.}
+\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.
+ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).}
 

### 2017-12-12 paper/manuscript_Sarah.tex
started adding Sarah comments

--- 2017-12-10 paper/manuscript.tex
@@
 \newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+\newcommand{\skhcomment}[1]{\emph{\color{magenta} [#1]}}
 
@@
 \end{abstract}
+\skhcomment{How ubiquitous is the term "shifted"? }
 
@@
 We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
-Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
+Most sites show no detectable shifts, but 30 sites have clearly altered preferences.
 These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
@@
 The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
+\skhcomment{Maybe change the color of the BG505 and BF520 labels so they stick out more? Also, add a label for some sort of landmark sequence (such as the oldest sequence) to contextualize the tree.}
 }
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+BG505 and BF520 have typical divergence \skhcomment{Do you mean they have roughly the mean divergence?
+I might change the wording or provide this value.} for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+\skhcomment{I am not sure I understand the difference between the denominator 836 and the denominator 616.}
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 These replicates therefore account for \emph{all} sources of noise and error in the experiments.
+\skhcomment{Three comments on the schematic.
+1.
+The genome strands might be too wide for the size of the figure.
+2.
+I think the grey plasmid should be changed to orange.
+3.
+The dark and light orange are difficult to tell apart.}
 }
@@
 The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
+\skhcomment{1.
+The axis text is a little bit blurry on my copy.
+2.
+I would swap the order of barplots so that BF520 is close to the BF520 correlation plots.
+3.
+Could the mutation type legend be horizontal?}
 }
@@
 This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
+\skhcomment{Could you discuss selection only in the context of error-corrected rates?}
 The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
@@
 These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
-Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
-Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data but supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
@@
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
+\skhcomment{Later on in the text this claim is more conservative and takes into account the fact that BF520 is basically as well correlated with itself as with BG505.}
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
@@
 Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
-Both ExpCM models have a stringency parameter $>$1.
+Both ExpCM models have a stringency parameter $>$1, indicating selection in nature prefers the same amino acids as the experiments, but with greater stringency.
 All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
@@
 Both ExpCM's have stringency parameters $>$1 (Table~\ref{tab:phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
+\skhcomment{Comment here that re-scaled preferences are used throughout the rest of the study?}
 We can also interpret the $\omega$ parameter.
@@
 Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
+\skhcomment{Is it common to report the mean $\omega$ when a model uses $\Gamma\omega$? I usually think of "gene-wide average $\omega$" to be $\omega$ from M0.}
 But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
@@
 The sites are sorted by the magnitude of the shift.
+\skhcomment{Could you bold the sites which are discussed in the text? Since they are ranked by shift magnitude it is difficult to "find" 288.}
 }
@@
 Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
-There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
-However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
+There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core and there is not a significant association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted (Figure~\ref{fig:shiftfeatures}B, Mann-Whitney $U$ Test, $p$ = 0.15) .
+However, Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure (Figure~\ref{fig:shiftfeatures}C, Mann-Whitney $U$ Test, $p$ = 0.01).
 Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
 
+\skhcomment{This paragraph was the hardest for me to understand.}
 An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
@@
 But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
+\skhcomment{I am having a hard time imagining "sites that contact a substitution" rather than "sites that contact sites which have experienced a substitution."}
 There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
@@
 One idea that has recently gained support in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution.
+Entrenchment is the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
 Specifically, after \citet{pollock2012amino} suggested that amino-acid preferences systematically shift at sites of substitutions, \citet{shah2015contingency} formalized a definition of entrenchment which \citet{starr2017pervasive} subsequently reduced to experimental practice.
-Entrenchment is the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
 Given two homologs, if there is no entrenchment then the effect of mutating a site in the first homolog to its identity in the second will simply be the opposite of mutating the site in the second homolog to its identity in the first.
@@
 We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
-We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
 {\bf (A)}
 One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
-Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
+Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 ($Q$ value < 0.05, ~\citep{storey2003statistical}) are shown in spheres.
 {\bf (B)}
@@
 The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
-The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
+The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
 This finding is unsurprising, since it is well known that Env is under functional constraint.
@@
 \subsection{Alignments and phylogenetic analyses of Env sequences}
+\skhcomment{Add in the formula for preference re-scaling?}
 A basic description of the process used to generate the clade A sequence alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, the alignment mask in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, and the phylogenetic tree in Figure~\ref{fig:tree} are provided in the legend to that figure.
@@
 These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
-For both the M5 model and the ExpCM with a gamma-distributed $\omega$, the number of categories in the discretized gamma distribution is four.
+For both the M5 model and the ExpCM with a gamma-distributed $\omega$, we used four categories for the discretized gamma distribution.
 

### 2017-12-12 paper/manuscript_Hugh.tex

--- 2017-12-12 paper/manuscript.tex
@@
 \newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+\newcommand{\hhcomment}[1]{\emph{\color{blue} [#1]}}
 
@@
 Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
-The immediate evolutionary space accessible to any given Env is defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
+The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
+\hhcomment{I added "largely" since if you're considering all double- and triple-nt mutations away from the parent this would not only include all single amino-acid mutations, but also include some double- and triple-aa mutations.
+In the end, I'm fine with the original sentence, but I just thought I'd voice my opinion since the original wording sounded a little off.}
 Most mutational studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to an Env variant.
@@
 But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
-During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
-Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
+During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+\hhcomment{When reading my thesis, Michael and Julie both asked me to define "shift" the first time I used it, as neither were certain exactly what it meant.
+That is why I changed "shift" to "change" in the above sentence, and then put "shift" in quotation marks in the below sentence, as I thought it would help clear things up.
+We could also define "shift" more explicitly by giving an example such as, a change at site A could cause another site to shift from preferring Alanine to Leucine.}
+Such "shifts" in mutational effects can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
 Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
-The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
+The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large, heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
@@
 {\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
-The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+The number of mutations per gene roughly followed a Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
@@
 
-The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
+The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
 Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
@@
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
+Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier, inferior LAI dataset. 
 Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
+Note that while there is a trend for amino acids with a high preference in the BG505 replicates to have a lower preference in the BF520 replicates, this trend does not necessarily reflect differences in underlying mutational effects.
+That is because this trend is also shaped by differences in the stringency of selection between our deep mutational scans (purifying selection was stronger in the BG505 experiments; Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
+Later in this paper, we correct for such experimental differences by calibrating each dataset to match the stringency of selection in nature.
 
@@
 We can also interpret the $\omega$ parameter.
-For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous one -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones.
+The gene-wide average of $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
 Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
-But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that immune selection drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
+But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
+Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
 
@@
 Env sites with shifted amino-acid preferences between BG505 and BF520.
-{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at five example sites.
+{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
 We have triplicate measurements for each Env.
@@
 This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates for the same Env.
-Figure~\ref{fig:prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and several other sites with more intermediate behaviors.
+Figure~\ref{fig:prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and two other sites with more intermediate behaviors.
 
@@
 Figure~\ref{fig:prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
-For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
+For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
 Such broadening and narrowing of a site's mutational tolerance is often observed during protein evolution, and is frequently~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability} although not always~\citep{starr2017alternative} linked to changes in protein stability -- with a more stable protein typically being more mutationally tolerant.
@@
 However, it is not technically feasible to experimentally measure the fitness effects of mutations to HIV with the same precision as for model organisms such as yeast.
+\hhcomment{I was confused, are you saying it's not technically feasible to obtain this precision with our DMS experiments, or just in general with HIV? If just with HIV DMS, perhaps clarify this?}
 Therefore, we can only conclude that 30 Env sites have undergone \emph{detectable} shifts in amino-acid preferences between BG505 and BF520.
@@
 {\bf (B)} 
-There is no significant difference in relative solvent accessibility of sites that have and have not undergone significant shifts.
+There is no significant difference in relative the solvent accessibility of sites that have and have not undergone significant shifts.
 The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
@@
 {\bf (D)}
-Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520. 
+Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520 or at sites that physically contact the substituted sites.
 The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
@@
 {\bf (A)}
-One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to grey to red based on the strength of evidence that sites are evolving slower than expected (blue), as fast as expected (grey), or faster than expected (red) given the BG505 experiments.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
@@
 A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
-Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form the hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
+Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
 All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
@@
 \figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
-Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they do fall into neither of these categories.
+Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they fall into neither of these categories.
 Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
@@
 Indeed, a prior protein-wide comparison of amino-acid preferences across homologs found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
-But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: sites that differ between BG505 and BF520 are not more shifted than other sites.
+But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted actually have the same wildtype amino acid.
+\hhcomment{I thought that the previous way it was worded failed to acknowledge that there was a boarder-line significant trend. And I thought it would be relevant to emphasize what I added.}
 This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
 
+\hhcomment{I thought the original version of this paragraph was a bit redundant, so I have restructured it. Just an idea...}
+In sum, we have shown how amino-acid preferences from two clade A Env variants can be compared to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
+However, a challenge of studying HIV is that almost everything must be simplified to make it tractable.
+Our studies are no different -- we are measuring viral replication in a T-cell line in the lab, not in infected humans.
 Going forward, it will be interesting to determine how the results of experiments such as ours are affected by defined modifications in viral genotype or selective pressure.
-A challenge of studying HIV is that almost everything must be simplified to make it tractable.
-Our studies are no different -- after all, we are measuring viral replication in a T-cell line in the lab, not in infected humans.
-But our experiments provide highly quantitative data that is amenable to rigorous analysis.
-Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
-Future experiments that modulate selection pressures in other relevant ways should provide further insight into the forces that drive and constrain HIV's evolution.
+For instance, although the Env variants we tested have 86\% protein identity, Env variants in nature can have as little as 65\% protein identity~\citep{lynch2009appreciating}.
+Repeating our experiment with more divergent Env variants, including those from different clades, could quantify shifts in mutational effects over an even wider swath of Env's evolution.
+Alternatively, adding additional selective pressures (e.g., antibody selection) to our experiment could help determine the extent that adaptive mutations differ in effects between Env variants.
+Overall, such experiments should provide further insight into the forces that drive and constrain HIV's evolution.
 
@@
 For the analysis in Figure~\ref{fig:entrenchment}, the results are presented in terms of the mutational effects rather than the amino-acid preferences. 
-If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in Table~\ref{tab:phydms}), then estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.
+If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in Table~\ref{tab:phydms}), then the estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.
 
@@
 \label{suppfile:BG505primers}
-This file gives the primers used for the BG505 codon mutagenesis.
-\jdbcomment{Hugh, can you add this file?}
+The sequences of the primers used for the BG505 codon mutagenesis are in the file \texttt{BG505\_codon\_mutagenesis\_primers.txt}.
 }

### 2017-12-12 paper/manuscript_Hugh.tex
Minor tweak to Hugh's edits

@@
 For instance, although the Env variants we tested have 86\% protein identity, Env variants in nature can have as little as 65\% protein identity~\citep{lynch2009appreciating}.
-Repeating our experiment with more divergent Env variants, including those from different clades, could quantify shifts in mutational effects over an even wider swath of Env's evolution.
+Repeating our experiment with more divergent Env variants from transmitted-founder viruses, including those from different clades, could quantify shifts in mutational effects over an even wider swath of Env's evolution.
 Alternatively, adding additional selective pressures (e.g., antibody selection) to our experiment could help determine the extent that adaptive mutations differ in effects between Env variants.

### 2017-12-12 paper/manuscript.tex
Add file with BG505 codon mutagenesis primers
added Sarah Hilton comments

@@
 We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
-Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
+Most sites show no detectable shifts, but 30 sites have clearly shifted preferences.
 These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
+In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
+In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
-Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
-Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
@@
 Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
+Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters, with the re-scaling done as described in \citet{hilton2017phydms}.
 At sites of known functional importance, these preferences are usually consistent with prior knowledge.
@@
 There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
-However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
+However, Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
 Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
@@
 According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
-But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
+But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (Figure~\ref{fig:shiftfeatures}C).
 There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
@@
 The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
-The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
+The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
 This finding is unsurprising, since it is well known that Env is under functional constraint.
@@
 These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
-For both the M5 model and the ExpCM with a gamma-distributed $\omega$, the number of categories in the discretized gamma distribution is four.
+The preferences were re-scaled by the stringency parameters in Table~\ref{tab:phydms} as described in \citet{hilton2017phydms}.
+For both the M5 model and the ExpCM with a gamma-distributed $\omega$, we used four categories for the discretized gamma distribution.
 

### 2017-12-12 paper/manuscript.tex
added Hugh comments

@@
 Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
-The immediate evolutionary space accessible to any given Env is defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
+The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
 Most mutational studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to an Env variant.
@@
 But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
-During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as ``shifts'' in a site's amino-acid preferences.
 Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
@@
 Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset.
-Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
+
+While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B).
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
+Note also that while there is a trend for the highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most of the high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal), this trend does not necessarily reflect differences between the two Envs.
+Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
+Later in this paper, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
 
@@
 Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
-But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that immune selection drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
+But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
+Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
 
@@
 {\bf (B)} 
-There is no significant difference in relative solvent accessibility of sites that have and have not undergone significant shifts.
+There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
 The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
@@
 {\bf (D)}
-Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520. 
+Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
 The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
@@
 {\bf (A)}
-One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites are evolving slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
@@
 Indeed, a prior protein-wide comparison of amino-acid preferences across homologs found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
-But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: sites that differ between BG505 and BF520 are not more shifted than other sites.
+But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs.
 This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
 
-Going forward, it will be interesting to determine how the results of experiments such as ours are affected by defined modifications in viral genotype or selective pressure.
+Going forward, it will be interesting to determine how the results of experiments such as ours are affected by other defined modifications in viral genotype or selective pressure.
 A challenge of studying HIV is that almost everything must be simplified to make it tractable.

### 2017-12-13 paper/manuscript.tex
added all funding acknowledgments

@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.
+This work was supported by the following grants from the NIH: DP1-DA039543 from the NIDA to JO and R01-AI127893 from the NIAID to JDB.
+The work was also supported in part by a Collaboration for AIDS Vaccine Discovery Grant (OPP1111923).
+The research of JDB is supported in part by a Faculty Scholar Grant from the Howard Hughes Medical Institute and the Simons Foundation.
+HKH was supported by a Cell and Molecular Biology Training Grant (T32GM007270) from the NIGMS of the NIH.
 ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).}

### 2017-12-14 paper/manuscript.tex
added Julie Overbaugh comments

@@
 \affil[3]{Department of Genome Sciences, University of Washington, Seattle, WA}
-\affil[4]{Human Biology Division, Fred Hutchinson Cancer Research Center, Seattle, WA}
+\affil[4]{Human Biology Division and Epidemiology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
 \contrib[\authfn{1}]{These authors contributed equally to this work}
@@
 These mutational effects can shift as the virus evolves.
-However, the prevalence of such shifts remains unclear.
+However, the prevalence of such shifts in mutational effects remains unclear.
 Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
@@
 
-Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid}.
+Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid,burton2005antibody}.
 Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
-The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
-Most mutational studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to an Env variant.
+The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all single amino-acid mutations to Env.
+Most mutational studies have measured how just a small number of these mutations affect viral replication in cell culture, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} single amino-acid mutations mutations to an Env variant.
 
@@
 However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
-The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
+The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic,ronen2015hiv}.
 
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
+\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
 \caption{\label{fig:tree}
@@
 The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
-Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
-The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
-The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
+Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
+The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
+The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
 }
-\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
+\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
 {The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
@@
 \subsection{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
-We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain~\citep{haddox2016experimental}.
-In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
-Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
-This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.
+We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
+Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
+This last consideration is especially important when comparing mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.
+Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
+We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
+Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
 
@@
 \end{figure}
-
-We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
-We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
-Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
 
@@
 The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
-These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset.
+These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.
 
-While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B).
+While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
-Note also that while there is a trend for the highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most of the high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal), this trend does not necessarily reflect differences between the two Envs.
+Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the two Envs.
 Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
-Later in this paper, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
+In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
 
-\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
+\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution}
 The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
@@
 \subsection{Shifts in amino-acid preferences between BG505 and BF520}
-The most fundamental question that we seek to address is how similar the effects of mutations are between the two Envs.
-We can get some hint simply by examining Figure~\ref{fig:mutfreqs}B, which shows the correlations among the amino-acid preferences measured for all experimental replicates for both BG505 and BF520.
-All replicates are strongly correlated with each other regardless of which Env was used, indicating that many aspects of the preferences are conserved between BG505 and BF520.
-But the correlations between replicate measurements for the same Env are always greater than or equal to those for replicate measurements on different Envs, suggesting at least some shifts in preferences between BG505 and BF520.
-However, simply comparing correlation coefficients in this way does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
+The most fundamental question that we seek to address is how similar the amino-acid preferences are between the two Envs.
+We have already noted that Figure~\ref{fig:mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
+However, simply comparing correlation coefficients does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
 
@@
 For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
-Such broadening and narrowing of a site's mutational tolerance is often observed during protein evolution, and is frequently~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability} although not always~\citep{starr2017alternative} linked to changes in protein stability -- with a more stable protein typically being more mutationally tolerant.
+Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
 Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
-The fact that the shifted sites usually (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) tolerate more mutations in the more stable BG505 Env is therefore consistent with the view that increased protein stability is often associated with shifts towards higher mutational tolerance.
+Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) more mutationally tolerant in BG505.
 
@@
 Several other sites show similar but somewhat milder flips in which amino acids they prefer.
-In later sections, we investigate evolutionary and structural characteristics of these shifted sites.
-
- An important caveat to the foregoing analysis is that it only identifies the 30 sites with shifts that exceed the experimental noise.
- It is possible that other sites have undergone small shifts not detectable by our experiments -- indeed, a recent study by \citet{starr2017pervasive} found that small shifts are common during the evolution of the Hsp90 protein in yeast.
- However, it is not technically feasible to experimentally measure the fitness effects of mutations to HIV with the same precision as for model organisms such as yeast.
-Therefore, we can only conclude that 30 Env sites have undergone \emph{detectable} shifts in amino-acid preferences between BG505 and BF520.
-It is possible that more sensitive experiments would find additional small shifts.
 
@@
 \subsection{Comparing selection in the lab to natural selection}
-Our experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
+Our experiments measure the effects of mutations on multi-cycle viral replication in a T-cell line in the lab.
 But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
@@
 \section{Discussion}
-We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral replication in cell culture.
+We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for multi-cycle viral replication in cell culture.
 These amino-acid preference maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
-However, the main value of these maps comes not from comparing them with prior knowledge, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
+However, the main value of these maps comes not from comparing them with prior knowledge, which serves to simply validate the approach.
+The main value comes from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
 Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in ways that prior piecemeal studies of mutations cannot.
 
-But our maps come with a serious caveat: each one is measured for just a single Env variant.
+But these maps come with a potentially serious caveat: each one is measured for just a single Env variant.
 The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
@@
 Recent studies of protein evolution have focused on the idea that substitutions become ``entrenched'' as sites shift to accommodate new amino acids~\citep{pollock2012amino,shah2015contingency,bazykin2015changing,starr2017pervasive}.
-Indeed, a prior protein-wide comparison of amino-acid preferences across homologs found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
+Indeed, a prior protein-wide comparison of amino-acid preferences across homologs of influenza nucleoprotein found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
 But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs.
@@
 HKH was supported by a Cell and Molecular Biology Training Grant (T32GM007270) from the NIGMS of the NIH.
-ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).}
+ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).
 

### 2017-12-14 paper/manuscript.tex
a few small edits

@@
 During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
-We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as ``shifts'' in a site's amino-acid preferences.
+We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
 Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
@@
 
-\subsection{Two Env homologs from clade A transmitted-founder viruses}
+\subsection{Two Envs from clade A transmitted-founder viruses}
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
@@
 
-\subsection{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
+\subsection{Deep mutational scanning of each Env}
 We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
 Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
-This last consideration is especially important when comparing mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.
+This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
 Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
@@
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
-Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the two Envs.
+Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
 Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
@@
 Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
-Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters, with the re-scaling done as described in \citet{hilton2017phydms}.
+(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
 At sites of known functional importance, these preferences are usually consistent with prior knowledge.
@@
 \figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
-Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they fall into neither of these categories.
+Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} Envs, or whether they fall into neither of these categories.
 Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
@@
 \figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
-This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
+This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both Envs.
 Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
-For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
-The $Q$-value indicated is the maximum of that for the two homologs.
+For each site, the plots show the preferences averaged across replicates and re-scaled for each Env, as well as the frequencies of amino acids in the clade A Env alignment. 
+The $Q$-value indicated is the maximum of that for the two Envs.
 }{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}

### 2017-12-14 paper/manuscript.tex
added N-glycan supp figure

@@
 Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
-We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
-We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
+We calculated the statistical evidence that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
 {\bf (A)}
@@
 Sites of slower or faster than expected evolution given the BF520 experiments.
-For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+For both Envs, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
 {\bf (C)}
 The results are similar regardless of whether the BG505 or BF520 experiments are used.
-Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
+Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
 All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
@@
 }{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
+\figsupp[\label{figsupp:glycans}Amino-acid preferences and alignment frequencies for glycosylation motifs.]
+{This plot shows all sites that are N-linked glycosylation motifs in both BG505 and BF520.
+Each motif is named by the number of the asparagine, and the preferences (averaged across replicates) for each Env are shown as well as the frequencies of amino acids across the clade A alignment at each of the three positions in the motif (N-X-S/T).
+As can be seen from this plot, at many of the sites the experiments measures relatively broad mutational tolerance, but the natural Env sequences have a strongly conserved motif.
+We suspect this is because many glycans serve as as shield against immunity in nature -- a function that is not required in cell culture.}
+{\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
 \figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
@@
 The answer in both cases is likely to be immune selection.
-Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
+Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
 Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
 Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
+
 Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
@@
 Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
-For instance, sites of N-linked glycans change slower than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+For instance, sites of N-linked glycans change slower in nature than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.

### 2017-12-14 paper/manuscript.tex
replaced "viral replication" with "viral growth" as some people apparently think of "replication" as just transcription and not multi-cycle

@@
 However, the prevalence of such shifts in mutational effects remains unclear.
-Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
+Here we quantify the effects on viral growth of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
 Most mutations similarly affect both Envs, but the amino-acid preferences of a minority of sites have clearly shifted.
@@
 The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all single amino-acid mutations to Env.
-Most mutational studies have measured how just a small number of these mutations affect viral replication in cell culture, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} single amino-acid mutations mutations to an Env variant.
+Most mutational studies have measured how just a small number of these mutations affect viral growth in cell culture, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} single amino-acid mutations mutations to an Env variant.
 
@@
 
-Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
+Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral growth of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
 We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
@@
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
-However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
+However, the only prior work that has measured how all Env amino-acid mutations affect HIV growth is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
 The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic,ronen2015hiv}.
@@
 The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
-We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.
+We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral growth.
 
@@
 \subsection{Deep mutational scanning of each Env}
-We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
+We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV growth in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
 Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
@@
 Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
 We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
@@
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
-To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
+To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
 To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
@@
 The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
-These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
+These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
 Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
@@
 \subsection{Comparing selection in the lab to natural selection}
-Our experiments measure the effects of mutations on multi-cycle viral replication in a T-cell line in the lab.
+Our experiments measure the effects of mutations on viral growth in a T-cell line in the lab.
 But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
@@
 \section{Discussion}
-We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for multi-cycle viral replication in cell culture.
+We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral growth in cell culture.
 These amino-acid preference maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
@@
 From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
-From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral replication imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
+From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
 
@@
 A challenge of studying HIV is that almost everything must be simplified to make it tractable.
-Our studies are no different -- after all, we are measuring viral replication in a T-cell line in the lab, not in infected humans.
+Our studies are no different -- after all, we are measuring viral growth in a T-cell line in the lab, not in infected humans.
 But our experiments provide highly quantitative data that is amenable to rigorous analysis.

### 2017-12-14 paper/manuscript.tex
a few very minor edits

@@
 By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
-Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.
+Overall, our work provides complete across-strain maps of mutational effects that inform analyses of Env's evolution and function.
 
@@
 This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
-Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
+Our modified deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
 This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
@@
 The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
-On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
+On the other hand, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
 

### 2017-12-14 paper/manuscript.tex
added reference

@@
 But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
-During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis,klink2017parallel}.
 We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.

### 2017-12-15 paper/manuscript_Adam2.tex
added second round oof Adam's comments

--- 2017-12-14 paper/manuscript.tex
@@
 Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
+
 Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
+Intringuingly, we find very little selection to maintain glycosylation in our cell culutre experiments.
+Of the 21 N-linked glycosylations sites shared between BG505 and BF52, only 4 are under strong selection to maintain the glycan in our experiments while they are largely conserved in nature (Figure 9S2).
+This highlights the role that these glycans play in immune selection, which is absent in our experiments, and agree with/adds to previous literature (cite). 
 It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
@@
 Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
-For instance, sites of N-linked glycans change slower in nature than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+For instance, many sites of N-linked glycans change slower in nature than our would be expected based on our experimental measurements of cell culture replication (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
+Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D) also evolve slower in nature than expected.
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
@@
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, all characterized Env structures remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate},  SOSIP trimer structures from diverse Env strains remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.
 
@@
 
-Going forward, it will be interesting to determine how the results of experiments such as ours are affected by other defined modifications in viral genotype or selective pressure.
-A challenge of studying HIV is that almost everything must be simplified to make it tractable.
-Our studies are no different -- after all, we are measuring viral growth in a T-cell line in the lab, not in infected humans.
-But our experiments provide highly quantitative data that is amenable to rigorous analysis.
+These experiments provide highly quantitative datasets on the functional tolerance of Env that is amenable to rigorous functional and evolutionary analyses.
 Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.

### 2017-12-15 paper/manuscript.tex
merged Adam changes into manuscript

@@
 Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
-Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
-The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
+
+Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, shifts in amino-acid preferences could make the outcome of any study highly dependent on the Env used.
+The extent to which this is actually the case is unclear, since the few protein-wide studies of such shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
@@
 We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{nduati2000effect,wu2006neutralization,goo2014early}.
-The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
+The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric,gristick2016natively}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
 We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
-It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
 
@@
 
-One idea that has recently gained support in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution.
-Specifically, after \citet{pollock2012amino} suggested that amino-acid preferences systematically shift at sites of substitutions, \citet{shah2015contingency} formalized a definition of entrenchment which \citet{starr2017pervasive} subsequently reduced to experimental practice.
+One idea that has recently gained support in the protein-evolution field is that substitutions become ``entrenched'' by subsequent evolution~\citep{pollock2012amino,shah2015contingency,starr2017pervasive}.
 Entrenchment is the tendency of a mutational reversion to become increasingly unfavorable as a sequence evolves.
@@
 In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
-These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
+These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric,gristick2016natively} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
 Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
-For instance, sites of N-linked glycans change slower in nature than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+For instance, we find very little selection to maintain most glycans in our cell-culture experiments.
+Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments -- despite the fact that most are conserved in nature (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
+This finding concords with prior literature suggesting that these glycans are selected primarily for their role in immune evasion~\citep{pugach2004prolonged,wang2013systematic,rathore2017glycosylation}.
+Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D) also evolve slower in nature than expected from our experiments.
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
@@
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, all characterized Env structures remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, SOSIP trimer structures from diverse Env strains remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2016natively}.
 
@@
 
-Going forward, it will be interesting to determine how the results of experiments such as ours are affected by other defined modifications in viral genotype or selective pressure.
-A challenge of studying HIV is that almost everything must be simplified to make it tractable.
-Our studies are no different -- after all, we are measuring viral growth in a T-cell line in the lab, not in infected humans.
-But our experiments provide highly quantitative data that is amenable to rigorous analysis.
+Our experiments provide highly quantitative data on the mutational tolerance of Env under selection for viral growth in cell culture.
+These data are amenable to rigorous functional and evolutionary analyses.
 Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.

### 2017-12-15 paper/manuscript_Hugh.tex
Add Hugh's second round of edits in separate file

--- 2017-12-12 paper/manuscript.tex
@@
 These mutational effects can shift as the virus evolves.
-However, the prevalence of such shifts remains unclear.
-Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
+However, the prevalence of such shifts in mutational effects remains unclear.
+Here we quantify the effects on viral growth of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
 Most mutations similarly affect both Envs, but the amino-acid preferences of a minority of sites have clearly shifted.
@@
 
-Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid}.
+Env's rapid evolution has dire consequences for anti-HIV immunity, since it erodes the efficacy of most neutralizing antibodies~\citep{albert1990rapid,wei2003antibody,richman2003rapid,burton2005antibody}.
 Because of this public-health importance, numerous studies have experimentally characterized aspects of the ``evolutionary landscape'' that Env traverses.
-The immediate evolutionary space accessible to any given Env is defined by the effects on viral fitness of all $19 \times 856 \sim 1.6 \times 10^4$ single amino-acid mutations that can be made to a typical-length 856-residue Env.
-Most mutational studies have measured how just a small fraction of these mutations affect viral replication in cell culture~\citep{freed1989mutational,cordonnier1989single,olshevsky1990identification,lu2001structural,basmaciogullari2002identification,jacobs2005alanine,zwick2005anti,lynch2015hiv}, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} mutations to an Env variant.
+The immediate evolutionary space accessible to any given Env is largely defined by the effects on viral fitness of all single amino-acid mutations to Env.
+Most mutational studies have measured how just a small number of these mutations affect viral growth in cell culture, although it has recently become possible to use deep mutational scanning to measure the effects of many~\citep{al2014high,duenas2016saturation} or even all~\citep{haddox2016experimental} single amino-acid mutations mutations to an Env variant.
 
 But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
-During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
+During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis,klink2017parallel}.
+We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
 Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
-Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
-The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
-Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
+Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, shifts in amino-acid preferences could make the outcome of any study highly dependent on the Env used.
+The extent to which this is actually the case is unclear, since the few protein-wide studies of such shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
+
+Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral growth of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
 We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
-Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
+Most sites show no detectable shifts, but 30 sites have clearly shifted preferences.
 These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
@@
 
-\subsection{Two Env homologs from clade A transmitted-founder viruses}
+\subsection{Two Envs from clade A transmitted-founder viruses}
 The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
-However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
-The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.
+However, the only prior work that has measured how all Env amino-acid mutations affect HIV growth is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
+The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic,ronen2015hiv}.
 
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
+\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
 \caption{\label{fig:tree}
@@
 We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{nduati2000effect,wu2006neutralization,goo2014early}.
-The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
+The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric,gristick2016natively}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
 We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
 The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
-We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.
+We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral growth.
 
@@
 Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
-BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
+In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
+In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
 
-\subsection{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
-We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain~\citep{haddox2016experimental}.
-In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
-Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
-This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.
+\subsection{Deep mutational scanning of each Env}
+We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV growth in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
+Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
+This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
+Our modified deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
+We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
+Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
 
@@
 {\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
-The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+The number of mutations per gene followed a Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+\hhcomment{I still don't think "roughly" fits in here since it's not an adjective.}
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
-To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
+To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
 To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
@@
 
-We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
-We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
-Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
-
 \begin{figure}
@@
 Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
-The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
+The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
+\hhaddox{I would say that reverse-transcription errors fall under the category of viral-replication errors.}
 Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
@@
 The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
-These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
-Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
-Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
+Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
@@
 The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
-These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
-Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
-Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
+These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.
+
+While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
+Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
+Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
+In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
 
-\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
+\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution}
 The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
@@
 We can also interpret the $\omega$ parameter.
-For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous one -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+\hhcomment{I still think it should be "ones" instead of "one", since it's standing in place for "mutations" not "mutation", right?}
 Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
-But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that immune selection drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
+But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
+Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
 
@@
 Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
+(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
 At sites of known functional importance, these preferences are usually consistent with prior knowledge.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
-It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
+It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the similarity of our findings with the above mutational studies, and the ability of our data to improve phylogenetic models of Env's natural evolution, both suggest that our measurements authentically describe the selection on Env.
+\hhcomment{It looks like Adam suggested removing the above sentence.
+But, I liked it as a concluding sentence.
+I think it's much weaker to just end on the second-to-last sentence about the variable loops.
+I've re-added this sentence and have tweaked it a bit in ways I think better sum up the entire section.
+I also moved one of the parts I deleted to the Discussion, where I thought it was a better fit.}
 
 \subsection{Shifts in amino-acid preferences between BG505 and BF520}
-The most fundamental question that we seek to address is how similar the effects of mutations are between the two Envs.
-We can get some hint simply by examining Figure~\ref{fig:mutfreqs}B, which shows the correlations among the amino-acid preferences measured for all experimental replicates for both BG505 and BF520.
-All replicates are strongly correlated with each other regardless of which Env was used, indicating that many aspects of the preferences are conserved between BG505 and BF520.
-But the correlations between replicate measurements for the same Env are always greater than or equal to those for replicate measurements on different Envs, suggesting at least some shifts in preferences between BG505 and BF520.
-However, simply comparing correlation coefficients in this way does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
+The most fundamental question that we seek to address is how similar the amino-acid preferences are between the two Envs.
+We have already noted that Figure~\ref{fig:mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
+However, simply comparing correlation coefficients does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
 
@@
 Figure~\ref{fig:prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
-For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
-Such broadening and narrowing of a site's mutational tolerance is often observed during protein evolution, and is frequently~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability} although not always~\citep{starr2017alternative} linked to changes in protein stability -- with a more stable protein typically being more mutationally tolerant.
+For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
+Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
 Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
-The fact that the shifted sites usually (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) tolerate more mutations in the more stable BG505 Env is therefore consistent with the view that increased protein stability is often associated with shifts towards higher mutational tolerance.
+Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) more mutationally tolerant in BG505.
 
@@
 For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
-Several other sites show similar but somewhat milder flips in which amino acids they prefer.
-In later sections, we investigate evolutionary and structural characteristics of these shifted sites.
-
- An important caveat to the foregoing analysis is that it only identifies the 30 sites with shifts that exceed the experimental noise.
- It is possible that other sites have undergone small shifts not detectable by our experiments -- indeed, a recent study by \citet{starr2017pervasive} found that small shifts are common during the evolution of the Hsp90 protein in yeast.
- However, it is not technically feasible to experimentally measure the fitness effects of mutations to HIV with the same precision as for model organisms such as yeast.
-Therefore, we can only conclude that 30 Env sites have undergone \emph{detectable} shifts in amino-acid preferences between BG505 and BF520.
-It is possible that more sensitive experiments would find additional small shifts.
+Thus, not only do sites vary in terms of the degree that they have shifted, but also in terms of the type of shift observed.
 
@@
 {\bf (D)}
-Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520. 
+Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
 The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+\hhcomment{On a second reading of this figure, I'm a bit confused.
+It seems like sites that fall into the "substitution" category would also fall into either the "contacts substitution" or the "not near substitution" category.
+I'm not sure I understand the rationale for why the three categories would all be mutually exclusive.
+It seems to me like a more straightforward comparison would be for panel D to \emph{only} compare sites based on whether or not they contact a substituted site, i.e., removing the "substitution" box plot.
+Since the supplemental table to this figure already tests for a difference in shifted vs.
+non-shifted sites, it doesn't seem like removing the "substitution" box plot would leave us with much less information.
+I would be interested to hear your thoughts on this if you have time to talk over the phone or in person.}
 Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
@@
 \subsection{Comparing selection in the lab to natural selection}
-Our experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
+Our experiments measure the effects of mutations on viral growth in a T-cell line in the lab.
 But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
@@
 Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
-We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
-We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
+We calculated the statistical evidence that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
 {\bf (A)}
-One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites are evolving slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
 {\bf (B)}
-Sites of slower or faster than expected evolution given the BF520 experiments.
-For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+Sites of slower- or faster-than-expected evolution given the BF520 experiments.
+For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
 {\bf (C)}
 The results are similar regardless of whether the BG505 or BF520 experiments are used.
-Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
+Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
 All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
@@
 \figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
-Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they do fall into neither of these categories.
+Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} Envs, or whether they fall into neither of these categories.
 Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
@@
 }{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
+\figsupp[\label{figsupp:glycans}Amino-acid preferences and alignment frequencies for glycosylation motifs.]
+{This plot shows all sites that are N-linked glycosylation motifs in both BG505 and BF520.
+Each motif is named by the residue number of the asparagine, and the preferences (averaged across replicates) for each Env are shown as well as the frequencies of amino acids across the clade A alignment at each of the three positions in the motif (N-X-S/T).
+As can be seen from this plot, at many of the sites the experiments measure relatively broad mutational tolerance, but the natural Env sequences have a strongly conserved motif.
+We suspect this is because many glycans serve as a shield against immunity in nature -- a function that is not required in cell culture.}
+{\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
 \figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
-This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
+This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both Envs.
 Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
-For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
-The $Q$-value indicated is the maximum of that for the two homologs.
+For each site, the plots show the preferences averaged across replicates and re-scaled for each Env, as well as the frequencies of amino acids in the clade A Env alignment. 
+The $Q$-value indicated is the maximum of that for the two Envs.
 }{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
@@
 We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
-The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
-The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
+The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM's that incorporate experimental measurements as described by \citet{bloom2017identification}.
+The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
 This finding is unsurprising, since it is well known that Env is under functional constraint.
-In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
+In contrast, ExpCM's that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
 
@@
 The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
-The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
+The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, suggesting that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
+\hhcomment{The word "meaning" here sounds too strong to me.}
 
@@
 The answer in both cases is likely to be immune selection.
-Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
+Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
 Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
 Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
+
 Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
 In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
-These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
+These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric,gristick2016natively} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
 Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
-For instance, sites of N-linked glycans change slower than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
+For instance, we find very little selection to maintain most glycans in our cell-culture experiments.
+Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments -- despite the fact that most are conserved in nature (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
+This finding concords with prior literature suggesting that these glycans are selected primarily for their role in immune evasion~\citep{pugach2004prolonged,wang2013systematic,rathore2017glycosylation}.
+Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D) also evolve slower in nature than expected from our experiments.
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
@@
 \section{Discussion}
-We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral replication in cell culture.
+We have experimentally measured the preference for each amino acid at each site in the ectodomain and transmembrane domain of two Envs under selection for viral growth in cell culture.
 These amino-acid preference maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
-However, the main value of these maps comes not from comparing them with prior knowledge, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
+However, the main value of these maps comes not from comparing them with prior knowledge, which serves to simply validate the approach, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
 Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in ways that prior piecemeal studies of mutations cannot.
 
-But our maps come with a serious caveat: each one is measured for just a single Env variant.
+But these maps come with a potentially serious caveat: each one is measured for just a single Env variant.
 The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
@@
 From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
-From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral replication imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
+Indeed, we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
+From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
 
@@
 Recent studies of protein evolution have focused on the idea that substitutions become ``entrenched'' as sites shift to accommodate new amino acids~\citep{pollock2012amino,shah2015contingency,bazykin2015changing,starr2017pervasive}.
-Indeed, a prior protein-wide comparison of amino-acid preferences across homologs found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
-But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: sites that differ between BG505 and BF520 are not more shifted than other sites.
+Indeed, a prior protein-wide comparison of amino-acid preferences across homologs of influenza nucleoprotein found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
+But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs.
 This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
 
-Going forward, it will be interesting to determine how the results of experiments such as ours are affected by defined modifications in viral genotype or selective pressure.
-A challenge of studying HIV is that almost everything must be simplified to make it tractable.
-Our studies are no different -- after all, we are measuring viral replication in a T-cell line in the lab, not in infected humans.
-But our experiments provide highly quantitative data that is amenable to rigorous analysis.
+Our experiments provide highly quantitative data on the mutational tolerance of Env under selection for viral growth in cell culture.
+These data are amenable to rigorous functional and evolutionary analyses.
 Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.
-ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).}
+This work was supported by the following grants from the NIH: DP1-DA039543 from the NIDA to JO and R01-AI127893 from the NIAID to JDB.
+The work was also supported in part by a Collaboration for AIDS Vaccine Discovery Grant (OPP1111923).
+The research of JDB is supported in part by a Faculty Scholar Grant from the Howard Hughes Medical Institute and the Simons Foundation.
+HKH was supported by a Cell and Molecular Biology Training Grant (T32GM007270) from the NIGMS of the NIH.
+ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).
 

### 2017-12-15 paper/manuscript_Hugh.tex
Minor tweak to one of Hugh's comments

@@
 The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
-\hhaddox{I would say that reverse-transcription errors fall under the category of viral-replication errors.}
+\hhcomment{I would say that reverse-transcription errors fall under the category of viral-replication errors.}
 Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).

### 2017-12-15 paper/manuscript_Hugh.tex
Another minor tweak to Hugh's comments

@@
 It seems to me like a more straightforward comparison would be for panel D to \emph{only} compare sites based on whether or not they contact a substituted site, i.e., removing the "substitution" box plot.
-Since the supplemental table to this figure already tests for a difference in shifted vs.
-non-shifted sites, it doesn't seem like removing the "substitution" box plot would leave us with much less information.
+Since the supplemental table to this figure already tests for a difference in shifts at substituted vs.
+non-substituted sites, it doesn't seem like removing the "substitution" box plot would leave us with much less information.
 I would be interested to hear your thoughts on this if you have time to talk over the phone or in person.}

### 2017-12-16 paper/manuscript.tex
merged Hugh comments

@@
 {\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
-The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
@@
 Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
-The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
+The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
 Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
@@
 We can also interpret the $\omega$ parameter.
-For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous one -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
 Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
+It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.
 
@@
 For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
-Several other sites show similar but somewhat milder flips in which amino acids they prefer.
+Thus, there is variation in both the extent and types of shifts observed.
 
@@
 Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
-The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
 Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
@@
 {\bf (A)}
-One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites are evolving slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
 {\bf (B)}
-Sites of slower or faster than expected evolution given the BF520 experiments.
-For both Envs, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+Like (A) but using the data from the BF520 experiments.
+For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
 {\bf (C)}
@@
 {This plot shows all sites that are N-linked glycosylation motifs in both BG505 and BF520.
-Each motif is named by the number of the asparagine, and the preferences (averaged across replicates) for each Env are shown as well as the frequencies of amino acids across the clade A alignment at each of the three positions in the motif (N-X-S/T).
-As can be seen from this plot, at many of the sites the experiments measures relatively broad mutational tolerance, but the natural Env sequences have a strongly conserved motif.
-We suspect this is because many glycans serve as as shield against immunity in nature -- a function that is not required in cell culture.}
+Each motif is named by the residue number of the asparagine, and the preferences (averaged across replicates) for each Env are shown as well as the frequencies of amino acids across the clade A alignment at each of the three positions in the motif (N-X-S/T).
+As can be seen from this plot, the experiments measure relatively broad mutational tolerance at many sites where the natural Env sequences have a strongly conserved motif.
+We suspect this is because many glycans serve as a shield against immunity in nature -- a function that is not required in cell culture.}
 {\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
@@
 We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
-The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
+The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM's that incorporate experimental measurements as described by \citet{bloom2017identification}.
 The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
 This finding is unsurprising, since it is well known that Env is under functional constraint.
-In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
+In contrast, ExpCM's that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
 
@@
 The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
-The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
+The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, suggesting that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
 
@@
 \section{Discussion}
-We have experimentally measured the preference for each amino acid at each site in the ectodomain of two Envs under selection for viral growth in cell culture.
+We have experimentally measured the preference for each amino acid at each site in the ectodomain and transmembrane domain of two Envs under selection for viral growth in cell culture.
 These amino-acid preference maps are generally consistent with prior knowledge about sites that are important for protein properties such as receptor binding or disulfide-mediated stability.
-However, the main value of these maps comes not from comparing them with prior knowledge, which serves to simply validate the approach.
-The main value comes from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
+However, the main value of these maps comes not from comparing them with prior knowledge, but from the fact that such prior knowledge encompasses just a small fraction of the vast mutational space available to Env.
 Because Env evolves so rapidly, every study of this protein must be placed in an evolutionary context, and our comprehensive amino-acid preference maps potentially enable this in ways that prior piecemeal studies of mutations cannot.
@@
 From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
+Indeed, we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
 From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.

### 2018-01-07 paper/manuscript.tex
tweaks for eLife full submission

@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-This work was supported by the following grants from the NIH: DP1-DA039543 from the NIDA to JO and R01-AI127893 from the NIAID to JDB.
-The work was also supported in part by a Collaboration for AIDS Vaccine Discovery Grant (OPP1111923).
-The research of JDB is supported in part by a Faculty Scholar Grant from the Howard Hughes Medical Institute and the Simons Foundation.
-HKH was supported by a Cell and Molecular Biology Training Grant (T32GM007270) from the NIGMS of the NIH.
-ASD was supported by an NSF Graduate Research Fellowship (DGE-1256082).
 

### 2018-02-09 paper/ResponseToReviews.tex
added response to reviews outline

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\title{Response to reviews of ``Mapping mutational effects along the evolutionary landscape of HIV envelope'' for \textit{eLife}}
\author{Hugh K. Haddox, Adam S. Dingens, Sarah K. Hilton, Julie Overbaugh, and Jesse D. Bloom}

\begin{document}
\maketitle

\emph{Below, the original comments {\color{blue} are in blue}, and our responses are in black.}

\color{blue}

\section*{SUMMARY}

Information on the possible amino acid trajectories of the rapidly evolving HIV envelope protein has long been limited. In this manuscript, you extend previous work from your lab analyzing the mutational landscape of the HIV-1 Env gene during in vitro viral replication. The work focuses on two clade A transmitter founder viruses, BG505 and BF520, which differ by more than 100 mutations. A deep mutational scanning approach is used to identify the permitted mutations at each residue of Env. Of particular interest is how amino acid preferences differ for these two proteins. A set of 30 sites with clear shifts in amino acid preference are identified. These data are compared to known mutational constraints on Env and to one another, and sources for the shifts in mutant affects are proposed by eliminating many scenarios. The results offer a body of knowledge useful for identifying i) residues and networks that are critical for basic function and ii) hot spots for evolvability or immune pressure. The analysis has been performed to a high standard (though some clarifications are necessary; see minor comments), and the openness of the data is a great service to the community. The paper represents a significant contribution towards understanding the structure-function relationship of HIV envelope. Therefore, the paper is likely to be accepted for publication if the following points are adequately addressed.

{\color{black}
Thanks for the excellent summary of our work.
We appreciate the favorable evaluation of the rigor of the work and the importance of the data.
The comments below are extremely helpful, and we have revised the manuscript in line with these suggestions.
These revisions have further strengthened the manuscript.
}

\section*{ESSENTIAL REVISIONS}

1. How are mutations that strongly modulate Env's expression level taken into account?

{\color{red}
The best I can think of here is just to explain our rationale for excluding signal peptide and cytoplasmic tail. 
I don't think we have a way of addressing this comment beyond that?
Hugh and Adam, do you have additional ideas?
}

2. Only the native, pre-fusion Env trimer is used to calculate proximity of substituted and shifted residues. It is possible that substitution-shift colocalization may be detected in other conformations of Env as it undergoes structural rearrangement to mediate fusion, as opposed to acting through longer-range allostery in the pre-fusion conformation. Another well-characterized structure that should be considered is the post-fusion six-helix bundle of gp41.

{\color{red}
This is a really good suggestion.
Hugh, are you able to add analysis with this and any other relevant conformations that are substantially different?
We should think about whether these would be added just as analyses, or whether we'd also show such a structure.
}

3. Overall, it is found that the shifted sites tend to cluster together, but they are not necessarily the same as the sites with substitutions between the two Envs or ones that contact them. This analysis is not pursued in great detail, however, leaving the reasons why shifted sites tend to cluster together unexplained. Given that the authors also observe some evidence of entrenchment of substitutions between the Envs, is it possible that interactions between the shifted sites themselves are important? This question could be explored through a more detailed examination of the structure.

{\color{red}
Seems like this might be best examined by looking carefully at some of the clusters?
I'm not sure how to do it with a statistical analysis.
Hugh and Adam---do you have any thoughts on this?
}

\section*{MINOR POINTS}

1. Readers may miss the rationale for excluding the signal peptide and cytoplasmic tail if it is only found in the Methods.

{\color{red}Yes, we should move this up. Will be done in context of answering essential revision 1, and then we can describe change here.}

2. Page 3: When referring to the modifications made in this paper versus the previous publication, it is not clear whether the cell type difference was expression of CCR5 or use of a new cell line altogether.

{\color{red}
Easy to fix.
}

3. Page 6: The dN/dS parameter is introduced, but its meaning in the two different models is only described in the next paragraph. Noting that a description follows may help readers.

{\color{red}
Easy to fix.}

4. The color scheme of structures in Fig 7a, 9a is difficult to discern. Higher contrast colors please.

{\color{red}
Hugh, can you do this?}

5. The methodology used here was previously developed in your group, which would benefit from a short description in the main text. For example, how exactly is $\omega$ defined, what is the stringency parameter, can the use of $\omega$ be motivated, how is $\omega_r$ defined and motivated etc? What are the assumptions in using these measures? We think this is necessary for a stand-alone paper aimed at a broad audience. All the details of the method should not be repeated, but an intuitive summary in the main text would be helpful.

{\color{red}
This is easy to do. Maybe Jesse should do this?}

6. With the current version 2.2.0 of dms\_tools2 installed via pip today (January 19, 2018), the included iPython notebook throws an error. Error text is below: 
\begin{verbatim}
--------------------------------------------------------------------------- 
ModuleNotFoundError Traceback (most recent call last) 
in <module>() 
36 import dms_tools2.dssp 
37 import dms_tools2.compareprefs 
---> 38 import dms_tools2.protstruct 
39 import dms_tools2.rplot 
40 print(dms_tools2.rplot.versionInfo())

ModuleNotFoundError: No module named 'dms_tools2.protstruct'
\end{verbatim}

{\color{red}
This is now fixed by the latest version.}

7. It might be helpful to increase the transparency on the points in Figure 3B, because at the current level it's impossible to resolve any detail in the blob near zero.

{\color{red}
Not sure if this will work in practice because then other points become to sparse. 
Jesse can also make contour plots to show as supplement.}

8. In Figure 4, it's not clear why the p-value for $\omega$ greater/less than one is used for shading rather than the estimated $\omega$ itself. 
{\color{red}
Jesse can add explanation here.}

9. What is the rationale for rescaling amino acid preferences by the stringency parameter before comparing them? Based on (Hilton et al., 2017), my understanding is that only the amino acid preferences are measured directly. The stringency parameter appears to be based on a phylogenetic model fit. This rescaling would emphasize differences in the basic ordering of amino acid preferences over differences of intensity, but the latter would seem to also be important. 
{\color{red}
Jesse can respond to this. I suppose we could also repeat all with un-rescaled if we wanted.}

10. In the text (p8, bottom) it is stated that the amino acid preferences are rescaled for comparison. But this is not mentioned in Figure 6 or described clearly in the Methods (p18, bottom). For the sake of clarity, it would be helpful to introduce new notation for rescaled preferences and indicate clearly where they are used. 
{\color{red}
We can do this.}

11. In the discussion it is noted that: "But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs." Does wildtype refer to the amino acids in the actual BG505 and BF520 sequences? The most preferred amino acid is the same in only 12/30 cases, based on Figure 6C. 
{\color{red}
Yes, it refers to wildtype. I suppose we could modify to show wildtype in plot, but not sure if that is needed.}

12. A body of computational work that has been done to infer the effects of epistasis on HIV proteins (including ENV) is not adequately referenced. (Posted 5th Feb 2018) 
{\color{red}
This includes the new Arup Chakroborty paper. Others?}

\end{document}

### 2018-02-09 paper/eLife_original_submission.tex
added response to reviews outline

--- 2018-01-07 paper/manuscript.tex
@@
 Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
-The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
-Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
-The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
+The tree shows the 69 Envs in the alignment in \FIG{tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
+Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in \FIG{tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
+The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in \FIGSUPP[tree]{identity}.
 The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
@@
 \figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
-{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
+{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in \FIG{tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in \FIG{tree}-source~data~\ref{figdata:mask}.
 There are 616 non-masked sites.
@@
 Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
-Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
+\FIG{tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
 BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
 In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
-In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
+In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
-Our modified deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
-We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
+Our modified deep mutational scanning strategy is in \FIG{dms_schematic}A.
+This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
+We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (\FIG{dms_schematic}B).
 Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.
@@
 {\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
-The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
+The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{fsanger}).
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
@@
 The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
-Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
+Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIGSUPP[mutfreqs]{mutfreqs}).
 More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
@@
 $R$ is the Pearson correlation coefficient. 
-The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
+The numerical values for the preferences are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
 }
 \figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
-{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in Figure~\ref{fig:mutfreqs}.
+{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs}.
 Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
@@
 The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
-Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
-The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
-This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
-The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
+\FIG{mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
+The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (\FIG{mutfreqs}A).
+This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIGSUPP[mutfreqs]{mutfreqs}).
+The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIGSUPP[mutfreqs]{mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
 The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
 These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
-Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+\FIG{mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
 
 We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
-As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
+As graphically illustrated in \FIG{dms_schematic}A, the preferences for each site are normalized to sum to one.
 Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
-The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
+The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B.
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.
 
-While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
+While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (\FIG{mutfreqs}B, compare red and blue versus gray plots).
 This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
-Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
-Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
+Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
+Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIGSUPP[mutfreqs]{mutfreqs} shows that purifying selection better purged stop codons in BG505).
 In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
@@
 Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
-Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
+Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in \FIG{tree}.
 Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
@@
 Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
-Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
+\TABLE{phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
 }
@@
 
-Table~\ref{tab:phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
+\TABLE{phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
 In addition to simply comparing model fits, we can interpret the model parameters.
@@
 A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
-Both ExpCM's have stringency parameters $>$1 (Table~\ref{tab:phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
+Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
 We can also interpret the $\omega$ parameter.
 For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
-Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
+Indeed, \TABLE{phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
 But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
+\TABLE{phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
 
@@
 Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
-This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in Table \ref{tab:phydms}.
+This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
 The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
@@
 Amino-acid preferences for the BF520 Env.
-This figure is the same as Figure~\ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
+This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
 }
@@
 A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
-Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
+\FIG{BG505prefs} and \FIG{BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in \TABLE{phydms}.
 (Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
 At sites of known functional importance, these preferences are usually consistent with prior knowledge.
-For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
+For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (\FIG{BG505prefs} and \FIG{BF520prefs}).
 Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
 The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
-For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
+For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in \FIG{BG505prefs} and \FIG{BF520prefs}, such as 132 to 195).
 It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.
@@
 The most fundamental question that we seek to address is how similar the amino-acid preferences are between the two Envs.
-We have already noted that Figure~\ref{fig:mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
+We have already noted that \FIG{mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
 However, simply comparing correlation coefficients does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.
@@
 We therefore used a more rigorous approach to identify sites where the amino-acid preferences differ between BG505 and BF520 by an amount that exceeds the noise in our experiments.
-We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from Table~\ref{tab:phydms} to calibrate all measurements to the stringency of natural selection.
-We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (Figure~\ref{fig:prefsdist}-source~data~\ref{figdata:prefsdist}).
-For each site, we calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in Figure~\ref{fig:prefsdist}A.
+We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from \TABLE{phydms} to calibrate all measurements to the stringency of natural selection.
+We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (\FIG{prefsdist}-source~data~\ref{figdata:prefsdist}).
+For each site, we calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in \FIG{prefsdist}A.
 This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates for the same Env.
-Figure~\ref{fig:prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and two other sites with more intermediate behaviors.
+\FIG{prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and two other sites with more intermediate behaviors.
 
-The overall distribution of shifts between BG505 and BF520 is shown in Figure~\ref{fig:prefsdist}B.
+The overall distribution of shifts between BG505 and BF520 is shown in \FIG{prefsdist}B.
 Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
@@
 We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
-Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
+\FIG{prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
 Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
@@
 This null distribution is created by randomizing the experimental replicates among Envs.
-Figure~\ref{fig:prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
+\FIG{prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
 The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.
 
-We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (Figure~\ref{fig:prefsdist}B).
+We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (\FIG{prefsdist}B).
 There are 30 such sites at a false discovery rate of 0.1.
-Figure~\ref{fig:prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
-For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
+\FIG{prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
+For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in \FIG{prefsdist}C.
 Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
 Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
-Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) more mutationally tolerant in BG505.
+Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in \FIG{prefsdist}C) more mutationally tolerant in BG505.
 
 However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
-For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
+For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (\FIG{prefsdist}C).
 Thus, there is variation in both the extent and types of shifts observed.
@@
 {\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from gray to orange according to the magnitude of the mutational shift at each site (orange indicates large shift).
-Sites that are significantly shifted according to Figure~\ref{fig:prefsdist}B are in spheres, and all other sites are in cartoon representation.
+Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
 {\bf (B)} 
@@
 The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
-Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
+\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
 All panels only show the 577 sites that are resolved in the crystal structure.
-Of the 30 significantly shifted sites in Figure~\ref{fig:prefsdist}B, only 28 are crystallographically resolved.
+Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 are crystallographically resolved.
 Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
@@
 \figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
-{The sites of significant shifts as determined in Figure~\ref{fig:prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
+{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
 However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
@@
 What distinguishes the sites that have undergone significant shifts?
-Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
-There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
-However, Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
+\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
+There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
+However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (\FIG{shiftfeatures}C).
 Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
@@
 According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
-But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (Figure~\ref{fig:shiftfeatures}C).
-There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
+But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (\FIG{shiftfeatures}C).
+There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIGSUPP[shiftfeatures]{subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
 The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
@@
 
-Figure~\ref{fig:entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
+\FIG{entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
 As expected under entrenchment, the average effect of these mutations is deleterious -- although there are a substantial number of sites where the mutational flips are not deleterious. 
-We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (Figure~\ref{fig:entrenchment}).
+We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (\FIG{entrenchment}).
 This comparison shows that even unfavorable inter-Env mutational flips are generally more favorable than random amino-acid mutations.
@@
 Entrenchment of substitutions therefore contributes to some of the mutational shifts.
-But given that many of these shifts occur at sites that do not even differ between the Envs (Figure~\ref{fig:shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.
+But given that many of these shifts occur at sites that do not even differ between the Envs (\FIG{shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.
 
@@
 Like (A) but using the data from the BF520 experiments.
-For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (\FIGSUPP[divsel]{divselRSA}).
 {\bf (C)}
 The results are similar regardless of whether the BG505 or BF520 experiments are used.
-Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
-All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
+Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (\FIGSUPP[divsel]{glycans}).
+All sites that evolve slower than expected for both experimental datasets are in \FIGSUPP[divsel]{slower}.
 {\bf (D)} 
@@
 Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
-All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
+All analyses used the phylogenetic tree in \FIG{tree}.
 }
@@
 Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} Envs, or whether they fall into neither of these categories.
-Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
+Relative solvent accessibilities were calculated as in \FIG{shiftfeatures}.
 As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
@@
 The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM's that incorporate experimental measurements as described by \citet{bloom2017identification}.
-The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
+The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (\TABLE{phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (\TABLE{phydms}, $\omega_r > 1$).
 This finding is unsurprising, since it is well known that Env is under functional constraint.
-In contrast, ExpCM's that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).
+In contrast, ExpCM's that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (\TABLE{phydms}).
 
-The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B, and overlaid on the logoplots in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} as the $\omega_r$ values.
-The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
-The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, suggesting that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.
+The sites that evolve slower or faster than expected from the experiments are shown in \FIG{divsel}A,B, and overlaid on the logoplots in \FIG{BG505prefs} and \FIG{BF520prefs} as the $\omega_r$ values.
+The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (\FIG{divsel}C).
+The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, suggesting that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in \FIG{tree}.
 
@@
 The answer in both cases is likely to be immune selection.
-Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
+Most of the sites of faster-than-expected evolution are on the surface of Env (\FIG{divsel}A,B and \FIGSUPP[divsel]{divselRSA}).
 Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
@@
 For instance, we find very little selection to maintain most glycans in our cell-culture experiments.
-Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments -- despite the fact that most are conserved in nature (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
+Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments -- despite the fact that most are conserved in nature (\FIG{divsel}C and \FIGSUPP[divsel]{glycans}).
 This finding concords with prior literature suggesting that these glycans are selected primarily for their role in immune evasion~\citep{pugach2004prolonged,wang2013systematic,rathore2017glycosylation}.
-Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D) also evolve slower in nature than expected from our experiments.
+Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (\FIG{divsel}D) also evolve slower in nature than expected from our experiments.
 Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
@@
 From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
-Indeed, we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
+Indeed, we anticipate that the complete maps of mutational effects in \FIG{BG505prefs} and \FIG{BF520prefs} will be useful for future sequence-structure-function studies.
 From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
@@
 For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} cloned into the same Q23 proviral plasmid.
-See Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger} and Figure~\ref{fig:mutfreqs}A for information on the average mutation rate in these libraries as estimated by Sanger sequencing and deep sequencing, respectively.
+See \FIGSUPP[dms_schematic]{sanger} and \FIG{mutfreqs}A for information on the average mutation rate in these libraries as estimated by Sanger sequencing and deep sequencing, respectively.
 
@@
 
-We conducted the low MOI viral passage illustrated in Figure~\ref{fig:dms_schematic}A in SupT1.CCR5 cells~\citep[obtained from Dr. James Hoxie;][]{boyd2015mutations}.
+We conducted the low MOI viral passage illustrated in \FIG{dms_schematic}A in SupT1.CCR5 cells~\citep[obtained from Dr. James Hoxie;][]{boyd2015mutations}.
 During this passage, cells were maintained in R10 media, which has the same composition as the D10 described above, except RPMI-1640 (GE Healthcare Life Sciences, SH30255.01) is used in the place of DMEM.
@@
 
-For the final short-duration infection illustrated in Figure~\ref{fig:dms_schematic}A, for each replicate we infected $10^6$ TZM-bl infectious units into $10^6$ SupT1.CCR5 cells in the presence of 100 $\mu$g/mL DEAE-dextran (note that this is a 10-fold higher concentration of DEAE-dextran than for the other steps, meaning that the effective MOI of infection is higher if DEAE-dextran has the expected effect of enhancing viral infection).
+For the final short-duration infection illustrated in \FIG{dms_schematic}A, for each replicate we infected $10^6$ TZM-bl infectious units into $10^6$ SupT1.CCR5 cells in the presence of 100 $\mu$g/mL DEAE-dextran (note that this is a 10-fold higher concentration of DEAE-dextran than for the other steps, meaning that the effective MOI of infection is higher if DEAE-dextran has the expected effect of enhancing viral infection).
 Three hours post-infection, we pelleted the cells and resuspended them in fresh media without any DEAE-dextran.
@@
 \subsection{Illumina deep sequencing}
-We deep sequenced all of the samples shown in Figure~\ref{fig:mutfreqs}A: the plasmid mutant libraries and wildtype plasmid controls, and the cDNA from the final mutant viruses and wildtype virus controls.
+We deep sequenced all of the samples shown in \FIG{mutfreqs}A: the plasmid mutant libraries and wildtype plasmid controls, and the cDNA from the final mutant viruses and wildtype virus controls.
 In order to increase the sequence accuracy, we used a barcoded-subamplicon sequencing strategy.
@@
 \subsection{Alignments and phylogenetic analyses of Env sequences}
-A basic description of the process used to generate the clade A sequence alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, the alignment mask in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, and the phylogenetic tree in Figure~\ref{fig:tree} are provided in the legend to that figure.
+A basic description of the process used to generate the clade A sequence alignment in \FIG{tree}-source~data~\ref{figdata:alignment}, the alignment mask in \FIG{tree}-source~data~\ref{figdata:mask}, and the phylogenetic tree in \FIG{tree} are provided in the legend to that figure.
 An algorithmic description of how the alignment and tree were generated are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
 
-For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in Figure~\ref{fig:tree}.
-The Goldman-Yang (or YNGKP) model used in Table~\ref{tab:phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
-For the ExpCM shown in Table~\ref{tab:phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
+For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in \FIG{tree}.
+The Goldman-Yang (or YNGKP) model used in \TABLE{phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
+For the ExpCM shown in \TABLE{phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
 These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
-The preferences were re-scaled by the stringency parameters in Table~\ref{tab:phydms} as described in \citet{hilton2017phydms}.
+The preferences were re-scaled by the stringency parameters in \TABLE{phydms} as described in \citet{hilton2017phydms}.
 For both the M5 model and the ExpCM with a gamma-distributed $\omega$, we used four categories for the discretized gamma distribution.
 
-Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for a wider set of models than those used in Table~\ref{tab:phydms}. 
+\TABLE{phydms}-source-data~\ref{tabledata:phydms} shows the results for a wider set of models than those used in \TABLE{phydms}. 
 These include the M0 model of \citet{yang2000codon}, ExpCM without a gamma-distributed $\omega$, and ExpCM in which the amino-acid preferences are averaged across sites as a control to ensure that the improved performance of these models is due to their site-specificity.
@@
 When identifying shifts in amino-acid preferences between the two Envs, we needed a way to quantify differences between the Envs while accounting for the fact that our measurements are noisy.
-The approach we use is based closely on that of \citet{doud2015site}, and is illustrated graphically in Figure~\ref{fig:prefsdist}A.
+The approach we use is based closely on that of \citet{doud2015site}, and is illustrated graphically in \FIG{prefsdist}A.
 The RMSD$_{\rm{corrected}}$ value is our measure of the magnitude of the shift.
-Figure~\ref{fig:prefsdist}A, its legend, and the associated text completely explains these calculations with the following exception: they do not detail how the ``distance'' between any two preference measurements was calculated.
+\FIG{prefsdist}A, its legend, and the associated text completely explains these calculations with the following exception: they do not detail how the ``distance'' between any two preference measurements was calculated.
 The distance between preferences at each site was simply defined as half of the sum of absolute value of the difference between preferences for each amino acid.
@@
 \subsection{Analysis of entrenchment}
-For the analysis in Figure~\ref{fig:entrenchment}, the results are presented in terms of the mutational effects rather than the amino-acid preferences. 
-If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in Table~\ref{tab:phydms}), then the estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.
+For the analysis in \FIG{entrenchment}, the results are presented in terms of the mutational effects rather than the amino-acid preferences. 
+If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in \TABLE{phydms}), then the estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.
 

### 2018-02-15 paper/ResponseToReviews.tex
tiny tweak to revisions
added contour version of correlation scatter plots

@@
 
-{\color{red}
-Not sure if this will work in practice because then other points become to sparse. 
-Jesse can also make contour plots to show as supplement.}
+{\color{black}
+We have added a new figure supplement (Figure 3--Figure supplement 2) that shows the correlations as a KDE contour plot rather than a scatter plot.
+As is apparent from this new supplement, the vast majority of points are near the origin (low preference in both replicates). 
+This is expected since most mutations are highly deleterious.
+Therefore, the correlation signal is due to relatively small fraction of amino acids that have non-negligible preference---a fact that makes sesne, since most mutations are so deleterious as to be evolutionarily irrelevant.
+Unfortunately, given the skew of the data it is not possible to find a transparency that strikes a good balance between showing the high-preference points clearly and not over-plotting the low preference points.
+Therefore, we have elected to keep the original Figure 3B (scatter plot) while also adding the new contour plot in Figure 3--Figure supplement 2) to show both elements of the data.
+}
 
@@
 
-12. A body of computational work that has been done to infer the effects of epistasis on HIV proteins (including ENV) is not adequately referenced. (Posted 5th Feb 2018) 
+12. A body of computational work that has been done to infer the effects of epistasis on HIV proteins (including ENV) is not adequately referenced. 
 {\color{red}

### 2018-02-15 paper/manuscript.tex
added contour version of correlation scatter plots

@@
 The numerical values for the preferences are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
+\FIGSUPP[mutfreqs]{contour} shows the correlations using contour rather than scatter plots.
 }
@@
 {\input{figures/avgmutfreqs}}
+\figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
+{
+This figure shows the same data as in \FIG{mutfreqs}B except that it shows KDE contour plots rather than points.
+As is obvious from this representation, the vast majority of the points fall near the origin (very low preference in both replicates), and most of the correlation signal is therefore due to the relatively modest number of amino acids that have high preference at any given site.
+This is expected, since most amino acid mutations at most sites will be strongly deleterious, and so have low preference.
+}{\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
 \figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}

### 2018-02-17 paper/manuscript.tex
re-ran with newer version of `phydms`
Slightly changes LnL for YNGKP models since
old `phydms` had a minor bug for these models.

@@
 
-For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in \FIG{tree}.
+For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}, version 2.2.1]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in \FIG{tree}.
 The Goldman-Yang (or YNGKP) model used in \TABLE{phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.

### 2018-02-17 paper/ResponseToReviews.tex
add WT to plots for shifted sites, re-run notebook
We added the wildtype amino acid above the Env for
each site (this is panel Figure 6C).

Also, the notebook was cleanly run from the top
and paper was modified to indicate version
2.2.4 of `dms_tools2` was used.

@@
 								% TeX will automatically convert eps --> pdf in pdflatex
+
+\usepackage{hyperref}
+
 \usepackage{amssymb}
@@
 
-{\color{red}
-This is now fixed by the latest version.}
+{\color{black}
+This is now fixed.
+Previously we were running the analysis with a developmental version of \texttt{dms\_tools2} (version 2.2.dev1) that was not yet on \texttt{PyPI}.
+Now the analysis uses a stable version (2.2.4) that is on \texttt{PyPI} and installable with \texttt{pip}.
+Therefore, the analysis should now work if \texttt{dms\_tools2} is installed with \texttt{rplot} as described here: \url{https://jbloomlab.github.io/dms_tools2/installation.html#installing-the-rplot-module}.
+Note that some other external Python packages are also needed as indicated by the imports at the beginning of the iPython notebook.
+}
 
@@
 The most preferred amino acid is the same in only 12/30 cases, based on Figure 6C.
-{\color{red}
-Yes, it refers to wildtype. I suppose we could modify to show wildtype in plot, but not sure if that is needed.}
+
+{\color{black}
+This was unclear in the original manuscript.
+We have updated Figure 6C to also show the wildtype amino acid at each site.
+This makes it clear that even though the most preferred amino acid is only the same in 12/30 cases, the wildtype amino acid is the same in 18/30 sequences.
+This is because although the wildtype is always among the most preferred amino acid, it is not always the most preferred one.
+We have updated the sentence in the Discussion to explain this, and also to say the ``majority of sites'' rather than ``most sites'' since the former probably is a more accurate description of 18/30.}
 

### 2018-02-17 paper/manuscript.tex
add WT to plots for shifted sites, re-run notebook
We added the wildtype amino acid above the Env for
each site (this is panel Figure 6C).

Also, the notebook was cleanly run from the top
and paper was modified to indicate version
2.2.4 of `dms_tools2` was used.

@@
 For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
+The wildtype amino acid for that Env is indicated using the small black letters above each logo plot; note how the wildtype amino acid is frequently but not always the most preferred one.
 The sites are sorted by the magnitude of the shift.
@@
 Indeed, a prior protein-wide comparison of amino-acid preferences across homologs of influenza nucleoprotein found a significant enrichment of shifts at sites of substitutions~\citep{doud2015site}.
-But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs.
+But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: the majority of sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs even though the preferences have shifted.
 This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
@@
 \subsection{Analysis of deep-sequencing data}
-We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
+We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}, version 2.2.4]{bloom2015software}.
 The algorithm that goes from the deep-sequencing counts to the amino-acid preferences is that described in \citet{bloom2015software} (see also \url{https://jbloomlab.github.io/dms_tools2/prefs.html}).

### 2018-02-17 paper/ResponseToReviews.tex
explain why we report P rather than omega_r

@@
 8. In Figure 4, it's not clear why the p-value for $\omega$ greater/less than one is used for shading rather than the estimated $\omega$ itself.
-{\color{red}
-Jesse can add explanation here.}
+
+{\color{black}
+We have added text in the legend of Figure 4 to explain this.
+Briefly, for individual sites, the point estimates of dN/dS values (or $\omega_r$) are not reliable when using a FEL framework for the reasons described in Pond et al (2015) and Murrell et al (2012).
+
+Here is the basic intuition. 
+Imagine two sites: one has 3 nonsynonymous substitutions and 0 synonymous substitutions, while the other has 20 nonsynonymous substitutions and 2 synonymous substitutons.
+If we just calculated $\omega_r$, we would get a value of infinity for the first site and 10 for the second site.
+But in reality, we think there is substantially stronger evidence for diversifying selection at the second site, since there are many more nonsynonymous substitutions.
+In an FEL framework, we assess the statistical evidence that $\omega_r \ne 1$, and find much more evidence for the second site.
+This increased evidence is manifested by the larger $P$-value rather than the value of $\omega_r$ itself.}
 

### 2018-02-17 paper/manuscript.tex
explain why we report P rather than omega_r

@@
 The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
-The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}.
+The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
 The letters above the logos indicate the wildtype amino acid in BG505.

### 2018-02-17 paper/ResponseToReviews.tex
more explanation about re-scaling of preferences

@@
 This rescaling would emphasize differences in the basic ordering of amino acid preferences over differences of intensity, but the latter would seem to also be important.
-{\color{red}
-Jesse can respond to this. I suppose we could also repeat all with un-rescaled if we wanted.}
+
+{\color{black}
+In this comment and the one immediately below (\#10), the reviewers point out the need to both justify why the preferences were re-scaled, and to clearly explain that this was done.
+
+In the Results section where we first discuss the re-scaling, we have now added an entire paragraph that discusses this fact:
+\begin{quote}
+\textsl{For the entire rest of the paper, we use the experimentally measured preferences re-scaled by the stringency parameters in Table 1.
+The reason we do this is to distinguish genuine differences between the two Envs from mere variation in the strength of selection between the two sets of experiments.
+Re-scaling both sets of preferences to optimally describe Env evolution in nature is a principled way to standardize the measurements; see Hilton et al (2017) and the Methods section entitled ``Re-scaling the preferences'' for a more detailed explanation.}
+\end{quote}
+At other key points (such as Figure 6 and the bottom of page 18 in the original manuscript) we now also emphasize the fact that the re-scaled preferences are used.
+
+The detailed justification for re-scaling is provided in the new Methods section entitled ``Re-scaling the preferences.''
+Briefly, the preferences themselves are calculated from the enrichment of mutations after selection.
+But the extent of enrichment is a function of how the experiment was performed (e.g., how long the selection was performed) as well as the inherent effects of the mutations. 
+For instance, let's say a given mutation is depleted 3-fold after 2 rounds of viral replication. 
+Then if we instead allowed $2^2 = 4$ rounds of viral replication, then we would expect the mutation to be depleted $3^2 = 9$-fold.
+Although we did our best to standardize the experiments with the two different Envs (BG505 and BF520), it is impossible to completely standardize experiments with different viruses since they grow at different inherent rates.
+Therefore, we need some method to ensure that we have put the measurements for the two Envs on the same ``scale.''
+Re-scaling by the stringency parameter is a principled way to do this.
+See the new Methods section ``Re-scaling the preferences'' for a justification of why this is a principled approach.
+}
 
@@
 For the sake of clarity, it would be helpful to introduce new notation for rescaled preferences and indicate clearly where they are used.
-{\color{red}
-We can do this.}
+
+{\color{black}
+This is an important point that we have addressed.
+Se our response to comment \#9 immediately above for a description of the changes that we have made.}
 

### 2018-02-17 paper/manuscript.tex
more explanation about re-scaling of preferences

@@
 
-A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
-\FIG{BG505prefs} and \FIG{BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in \TABLE{phydms}.
-(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
+For the entire rest of the paper, we use the experimentally measured preferences re-scaled by the stringency parameters in \TABLE{phydms}.
+The reason we do this is to distinguish genuine differences between the two Envs from mere variation in the strength of selection between the two sets of experiments.
+Re-scaling both sets of preferences to optimally describe Env evolution in nature is a principled way to standardize the measurements; see \citet{hilton2017phydms} and the Methods section entitled ``Re-scaling the preferences'' for a more detailed explanation.
+
+A qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
+\FIG{BG505prefs} and \FIG{BF520prefs} show the re-scaled across-replicate average of the amino-acid preferences for each Env.
 At sites of known functional importance, these preferences are usually consistent with prior knowledge.
@@
 Env sites with shifted amino-acid preferences between BG505 and BF520.
+Note that the preferences have been re-scaled using the stringency parameters in \TABLE{phydms} to enable direct comparison across Envs.
 {\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
@@
 
+\subsection{Re-scaling the preferences}
+The amino-acid preferences that are directly extracted from the deep sequencing data essentially give the enrichment / depletion of each mutation, normalized to sum to one at each site~\cite[][\url{https://jbloomlab.github.io/dms_tools2/prefs.html}]{bloom2015software}.
+However, the extent that any mutation is enriched or depleted is a combination of two factors: the inherent effect of that mutation, and the ``stringency'' of the experimental selection.
+For instance, if the selection is weak, then deleterious mutations will only be slightly depleted; conversely, if selection is strong, then deleterious mutations will be greatly depleted.
+The fact that the preferences depend on the stringency of the experimental selection is important if we want to compare results between Envs.
+The reason is that our goal is to identify differences in the inherent effects of mutations between Envs, not simply find differences due to variation in experimental stringency.
+Of course, we have done our best to perform the experiments for BG505 and BF520 equivalently, but because these are different viruses with different growth rates, it is impossible to exactly match the experimental stringencies.
+This can be seen in \FIGSUPP[mutfreqs]{mutfreqs}, which shows that stop codons were more depleted for BG505 than BF520, indicating that selection in our experiments was more stringent for BG505.
+
+How should we best re-scale the preferences?
+It turns out that raising them to a power is the best approach.
+To see that, imagine a mutation that is depleted 3-fold after 2 rounds of viral growth. 
+If our experiment instead allowed $2^2 = 4$ rounds of viral growth, then the mutation would be depleted $3^2 = 9$-fold.
+More generally, if a mutation is enriched in frequency by $\phi$-fold after $n$ rounds of viral growth, then it will be enriched in frequency by $\phi^{\beta}$-fold after $\beta n$ rounds of viral growth.
+Since the amino-acid preferences are conceptually equivalent to the re-normalized enrichments of mutations~\cite[][\url{https://jbloomlab.github.io/dms_tools2/prefs.html}]{bloom2015software}, it therefore makes sense that the re-scaled preference $\pi_{r,a}$ for amino-acid $a$ at site $r$ should be related to the directly measured preference $\hat{\pi}_{r,a}$ by $\pi_{r,a} \propto \left(\hat{\pi}_{r,a}\right)^{\beta}$.
+And indeed, this is exactly the re-scaling scheme described in \citet{hilton2017phydms} that we use to re-scale our preferences for BG505 and BF520.
+
+The last point is how to choose the re-scaling parameter $\beta$ for each Env.
+It turns out that the features that we have described above for our experiments are also a feature of natural evolution: the expected frequency of a substitution during evolution depends not only on the inherent fitness effect of that mutation, but also on the effective population size, which is conceptually somewhat similar to the stringency of selection.
+It turns out that in a mutation-selection phylogenetic model of evolution, if the amino-acid preferences are taken to represent the ``fitness effects'' of mutations, then the exponential scaling parameter $\beta$ is proportional to the effective population size~\citep{halpern1998evolutionary,mccandlish2014modeling,bloom2014informed}.
+Therefore, fitting the $\beta$ parameter using a phylogenetic approach enables standardization of the preferences for the two Envs, and re-scales the preferences so that they best match with the actual stringency of selection observed in nature~\citep{hilton2017phydms}.
+
+Note that in practice this re-scaling scheme is roughly equivalent to a more heuristic approach that has been used by \citet{gray2017analysis} and others.
+In this heuristic approach, the log-transformed enrichment ratios from different experiments are adjusted so that the distributions have equal spreads.
+Since multiplying log-transformed enrichment ratios is equivalent to exponentiating amino-acid preferences, these two re-scaling procedures apply the same mathematical transformation.
+
 \subsection{Identifying sites of shifted amino-acid preference}
@@
 The distance between preferences at each site was simply defined as half of the sum of absolute value of the difference between preferences for each amino acid.
-Specifically, for a given site $r$, let $\pi_{r,a}^{i}$ be the preference for amino-acid $a$ in homolog $i$ (e.g., BG505) and let $\pi_{r,a}^{j}$ be the preference for that same amino acid in homolog $j$ (e.g., BF520).
+Specifically, for a given site $r$, let $\pi_{r,a}^{i}$ be the re-scaled preference for amino-acid $a$ in homolog $i$ (e.g., BG505) and let $\pi_{r,a}^{j}$ be the re-scaled preference for that same amino acid in homolog $j$ (e.g., BF520).
 Then the distance between the homologs at this site is simply $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.

### 2018-02-17 paper/manuscript.tex
adjusted explanation of re-scaling based on Sarah Hilton comments

@@
 How should we best re-scale the preferences?
-It turns out that raising them to a power is the best approach.
-To see that, imagine a mutation that is depleted 3-fold after 2 rounds of viral growth. 
+Raising them to a power is a sensible approach.
+To see why, imagine a mutation that is depleted 3-fold after 2 rounds of viral growth. 
 If our experiment instead allowed $2^2 = 4$ rounds of viral growth, then the mutation would be depleted $3^2 = 9$-fold.
-More generally, if a mutation is enriched in frequency by $\phi$-fold after $n$ rounds of viral growth, then it will be enriched in frequency by $\phi^{\beta}$-fold after $\beta n$ rounds of viral growth.
+More generally, if a mutation is enriched in frequency by $\phi$-fold after $n$ rounds of viral growth, then it will be enriched in frequency by $\phi^{\beta}$-fold after $\beta \times n$ rounds of viral growth.
 Since the amino-acid preferences are conceptually equivalent to the re-normalized enrichments of mutations~\cite[][\url{https://jbloomlab.github.io/dms_tools2/prefs.html}]{bloom2015software}, it therefore makes sense that the re-scaled preference $\pi_{r,a}$ for amino-acid $a$ at site $r$ should be related to the directly measured preference $\hat{\pi}_{r,a}$ by $\pi_{r,a} \propto \left(\hat{\pi}_{r,a}\right)^{\beta}$.

### 2018-02-18 paper/figures/omegabysite_structural_analysis/analysis_code_conformational_dynamics_network.py
Add PyMOL scripts for analyzing shifts on other structures

@@ -18,9 +18,11 @@ cmd.delete('all')
 cmd.fetch(structure) #, type='pdb1')
 
 # Remove non-Env chains
-# gp41 = chain A
-# gp120 = chain G
-cmd.remove ('c;C,H,I,J,K,L,M,N')
+# gp41 = chain A, B, D
+# gp120 = chain G, I, J
+# CD4 = chains C, E, F
+# 17b = chains H, K, M and L, N, O
+cmd.remove ('c;C,E,F,H,K,M,L,N,O,I,J,B,D')
 #cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
 
 # Tweak initial display and color of Env monomers

### 2018-02-18 paper/figures/shifts_on_structure/pymol_analysis_1AIK.py

@@ -1,19 +1,13 @@
 """
-This script is for analyzing RMSD_corrected values in PyMOL
+This is a PyMOL script for analyzing shifts in preferences on the structure 1AIK
 
-It colors the structure of Env according to site-specific
-RMSD_corrected values on a color scale from grey to orange,
-with sites with significant RMSD_corrected values shown in
-spheres.
-
-When `take_pictures` is set to true, this script generates pictures
-of Env: `Env_pymol_face1.png` and `Env_pymol_face2.png`, which are
-rotated 120 degrees relative to one another.
+This script is similar to `pymol_analysis_5FYL.py`, but performs the analysis on
+the structure 1AIK instead of 5FYL.
 
 This script intended to be run in the PyMOL terminal as:
-    run pymol_analysis_code.py
+    run pymol_analysis_1AIK.py
 
-Hugh Haddox, November-27-2017
+Hugh Haddox, February-17-2018
 """
 
 # Imports
@@ -21,24 +15,21 @@ import pymol
 from pymol import cmd
 
 # Fetch structure
-structure = '5fyl'
+structure = '1AIK'
 cmd.reinitialize()
 cmd.delete('all')
 cmd.fetch(structure) #, type='pdb1')
 
-# Remove non-Env chains
-# gp41 = chain B
-# gp120 = chain G
-# 35O22 = chains D and E
-# PGT122 = chains H and L
-cmd.remove ('c;D,E,H,L,U,V')
-cmd.symexp(structure, structure, structure, 3) # generate symmetry partners based on crystal structure
+# Generate symmetry partners based on crystal structure, using a distance of
+# 1.9 since this reproduces the timer without segments from any other adjacent
+# monomers in the crystal structure
+cmd.symexp(structure, structure, structure, 1.9)
 
 # Tweak initial display and color of Env monomers
 cmd.hide('everything')
 cmd.bg_color('white')
 cmd.show('cartoon')
-cmd.color('grey90')
+cmd.color('grey40')
 #cmd.color('grey20', structure)
 cmd.set('cartoon_transparency', '0.5')
 cmd.set('cartoon_transparency', '0', structure)
@@ -76,22 +67,24 @@ print "max: {0}".format(max_RMSD)
 sites_with_data = RMSD_dict.keys()
 sites_not_in_structure = []
 for site in sites_with_data:
-    print(site, RMSD_dict[site])
+    #print(site, RMSD_dict[site])
     cmd.alter("{0} and resi {1}".format(structure, site),
[…]

### 2018-02-18 paper/figures/shifts_on_structure/pymol_analysis_5FYL.py

@@ -1,0 +1,158 @@
+"""
+This is a PyMOL script for analyzing shifts in preferences on the structure 5FYL
+
+It colors the structure of Env according to site-specific RMSD_corrected values
+on a color scale from white (no shift) to red (large shift), with sites with
+significant RMSD_corrected values shown as spheres.
+
+When `take_pictures` is set to true, this script generates pictures of Env:
+`5FYL_pymol_face1.png` and `5FYL_pymol_face2.png`, which are rotated 120 degrees
+relative to one another.
+
+This script intended to be run in the PyMOL terminal as:
+    run pymol_analysis_5FYL.py
+
+Hugh Haddox, November-27-2017
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '5FYL'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Remove non-Env chains
+# gp41 = chain B
+# gp120 = chain G
+# 35O22 = chains D and E
+# PGT122 = chains H and L
+cmd.remove ('c;D,E,H,L,U,V')
+
+# Generate symmetry partners based on crystal structure
+cmd.symexp(structure, structure, structure, 3)
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey40')
+#cmd.color('grey20', structure)
+cmd.set('cartoon_transparency', '0.5')
+cmd.set('cartoon_transparency', '0', structure)
+#cmd.hide('hetatm')
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+    lines = f.readlines()[1:]
+    for line in lines:
+        elements = line.split(',')
+        site = elements[0]
+        RMSDcorrected = elements[1]
+        significant_shift = elements[6]
+        RMSD_dict[site] = float(RMSDcorrected)
+        sig_dict[site] = significant_shift
[…]

### 2018-02-18 paper/figures/shifts_on_structure/pymol_analysis_5VN3.py

@@ -1,0 +1,160 @@
+"""
+This is a PyMOL script for analyzing shifts in preferences on the structure 5VN3
+
+This script is similar to `pymol_analysis_5FYL.py`, but performs the analysis on
+the structure 5VN3 instead of 5FYL.
+
+This script intended to be run in the PyMOL terminal as:
+    run pymol_analysis_5VN3.py
+
+Hugh Haddox, February-17-2018
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '5VN3'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# In the original script `pymol_analysis_5FYL.py`, the `structure` variable was
+# the name of a `PyMOL` object of the gp160 monomer. For this structure, it is
+# the name of an object of the entire trimer. So that the rest of the code is
+# compatible with this new naming system, I will rename the trimer object and
+# then change the monomer object to equal `structure`.
+cmd.set_name(structure, '5VN3_trimer')
+cmd.select('gp160', 'c;A,G')
+cmd.create(structure, 'gp160')
+
+# Remove non-Env chains
+# gp41 = chain A, B, D
+# gp120 = chain G, I, J
+# CD4 = chains C, E, F
+# 17b = chains H, K, M and L, N, O
+cmd.remove ('c;C,E,F,H,K,M,L,N,O')
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey40')
+#cmd.color('grey20', structure)
+cmd.set('cartoon_transparency', '0.5')
+cmd.set('cartoon_transparency', '0', structure)
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+    lines = f.readlines()[1:]
+    for line in lines:
+        elements = line.split(',')
+        site = elements[0]
+        RMSDcorrected = elements[1]
[…]

### 2018-02-18 paper/figures/curate_5VN3_structure/color_Env_RSA.py
Progress on curating other structures

@@ -1,0 +1,78 @@
+"""
+This script is for coloring Env's structure by RSA
+
+It is intended to be run in the PyMOL terminal as:
+    run color_Env_entropy.py
+
+Hugh Haddox, January-28-2016
+"""
+
+import pymol
+from pymol import cmd
+
+# Input variables
+structure = '5VN3'
+
+# Clear pymol, get structure, and remove non-Env chains
+# gp41 = chain A, B, D
+# gp120 = chain G, I, J
+# CD4 = chains C, E, F
+# 17b = chains H, K, M and L, N, O
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+cmd.remove ('c;C,E,F,H,K,M,L,N,O')
+cmd.set_name(structure, '5VN3_trimer')
+cmd.select('gp160', 'c;A,G')
+cmd.create(structure, 'gp160')
+
+# Read in RSA values from file
+d = {}
+with open ('5VN3_RSA_and_SS.txt') as f:
+    for (i, line) in enumerate(f):
+        line = line.strip().split(',')
+        site = line[0]
+        RSA = line[3]
+        if i == 0:
+            assert site == 'site'
+            assert RSA == 'RSA'
+        else:
+            site = int(site)
+            d[site] = float(RSA)
+
+# Read in entropies from a file of amino-acid preferences (HXB2 numbering)
+minRSA = min(d.values())
+maxRSA = max(d.values())
+
+print "The minimum and maximum RSA values are:"
+print "min: %s"%minRSA
+print "max: %s"%maxRSA
+
+# Replace residue b-factors with RSA values
+for site in d:
+    cmd.alter("resi %d" % site, "b = %g" % d[site])
+
+# cmd.hide('everything')
+# cmd.bg_color('white')
+cmd.show('surface', '5VN3')
+cmd.select(None)
+cmd.spectrum('b', 'blue_red', '5VN3', minimum=minRSA, maximum=maxRSA)
+
+# Make sure all sites in the structure have an RSA value. Not all sites in the
+# structure have RSA values, but it looks like the sites lacking them are all
+# unresolved, but are still incorporated in the PDB file somehow, and so counted
+# in the list of total sites.
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+sites_without_RSA_values = [
[…]

### 2018-02-18 paper/figures/curate_5VN3_structure/make_pdb_file_for_dssp.py

@@ -1,0 +1,24 @@
+"""
+This script is for loading 5VN3 and removing non-Env chains
+
+It is intended to be run in the PyMOL terminal as:
+    run make_pdb_file_for_dssp.py
+
+Hugh Haddox, February-17-2018
+"""
+import pymol
+from pymol import cmd, stored
+
+# Input variables
+structure = '5VN3'
+
+# Clear pymol, get structure, and remove non-Env chains
+# gp41 = chain A, B, D
+# gp120 = chain G, I, J
+# CD4 = chains C, E, F
+# 17b = chains H, K, M and L, N, O
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+cmd.remove ('c;C,E,F,H,K,M,L,N,O')
+
+cmd.save('5VN3_Env_trimer.pdb')

### 2018-02-19 paper/figures/shifts_on_structure/pymol_analysis_1ENV.py
Additional analysis of other structures

@@ -1,0 +1,34 @@
+"""
+This is a PyMOL script for analyzing shifts in preferences on the structure 1ENV
+
+This script is similar to `pymol_analysis_5FYL.py`, but performs the analysis on
+the structure 1ENV instead of 5FYL.
+
+This script intended to be run in the PyMOL terminal as:
+    run pymol_analysis_1ENV.py
+
+Hugh Haddox, February-18-2018
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '1ENV'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Generate symmetry partners based on crystal structure, using a distance of
+# 1.9 since this reproduces the timer without segments from any other adjacent
+# monomers in the crystal structure
+cmd.symexp(structure, structure, structure, 2.5)
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey40')
+#cmd.set('cartoon_transparency', '0.5')
+#cmd.set('cartoon_transparency', '0', structure)

### 2018-02-20 paper/manuscript.tex
Change color spectra

@@
 Characteristics of significantly shifted sites.
-{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from gray to orange according to the magnitude of the mutational shift at each site (orange indicates large shift).
+{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from white to red according to the magnitude of the mutational shift at each site (red indicates large shift).
 Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
@@
 {\bf (A)}
-One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
+One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to white to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (white) or faster than expected (red) given the BG505 experiments.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
@@
 Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
+The color spectrum is similar to the one in (A) and (B), but is colored from blue to \textit{grey} to red for increased visibility of the grey residues in the hydrophobic network.
 All analyses used the phylogenetic tree in \FIG{tree}.

### 2018-02-20 paper/figures/omegabysite_structural_analysis/analysis_code_conformational_dynamics_network.py

@@ -29,7 +29,7 @@ cmd.remove ('c;C,E,F,H,K,M,L,N,O,I,J,B,D')
 cmd.hide('everything')
 cmd.bg_color('white')
 cmd.show('cartoon')
-cmd.color('grey90')
+cmd.color('grey40')
 #cmd.color('grey20', structure)
 cmd.set('cartoon_transparency', '0.5')
 cmd.set('cartoon_transparency', '0', structure)
@@ -54,7 +54,7 @@ for h in homologs:
 	with open('../merged_omegabysite.csv') as f:
 		lines = f.readlines()[1:]
 		for line in lines:
-			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir) = line.strip().split(',')
+			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir, N_glycans) = line.strip().split(',')
 			omega_d[Env][site] = float(log10Pdir)
 			if float(Q) < Q_cutoff and float(omega) < 1.0:
 				sites_sig_slow[Env].append(site)
@@ -77,7 +77,7 @@ for site in sites_with_data:
     cmd.alter("{0} and resi {1}".format(structure, site), "b = {0}".format(avg_log10P_dir))
     if site not in unique_sites_in_structure:
     	sites_not_in_structure.append(site)
-cmd.spectrum('b', 'red_grey_blue', structure, minimum = min_omega, maximum = max_omega)
+cmd.spectrum('b', 'blue_white_red', structure, minimum = min_omega, maximum = max_omega)
 print ("\nSites with data not in structure: {0}".format(', '.join(sites_not_in_structure)))
 
 # Color residues lacking data white

### 2018-02-20 paper/figures/omegabysite_structural_analysis/analysis_code_omega_by_site.py

@@ -2,7 +2,7 @@
 This script is for analyzing omega-by-site values in PyMOL
 
 It is intended to be run in the PyMOL terminal as:
-    run analysis_code_omega_by_site.py
+	run analysis_code_omega_by_site.py
 
 Hugh Haddox, November-2-2017
 """
@@ -32,7 +32,7 @@ for h in homologs:
 	cmd.hide('everything')
 	cmd.bg_color('white')
 	cmd.show('cartoon')
-	cmd.color('grey90')
+	cmd.color('grey40')
 	#cmd.color('grey20', structure)
 	cmd.set('cartoon_transparency', '0.5')
 	cmd.set('cartoon_transparency', '0', structure)
@@ -55,7 +55,7 @@ for h in homologs:
 	with open('../merged_omegabysite.csv') as f:
 		lines = f.readlines()[1:]
 		for line in lines:
-			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir) = line.strip().split(',')
+			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir, N_glycans) = line.strip().split(',')
 			omega_d[Env][site] = float(log10Pdir)
 			if float(Q) < Q_cutoff and float(omega) < 1.0:
 				sites_sig_slow[Env].append(site)
@@ -75,16 +75,16 @@ for h in homologs:
 	sites_with_data = BG505_sites
 	sites_not_in_structure = []
 	for site in sites_with_data:
-	    cmd.alter("{0} and resi {1}".format(structure, site), "b = {0}".format(omega_d[h][site]))
-	    if site not in unique_sites_in_structure:
-	    	sites_not_in_structure.append(site)
-	cmd.spectrum('b', 'red_grey_blue', structure, minimum = min_omega, maximum = max_omega)
+		cmd.alter("{0} and resi {1}".format(structure, site), "b = {0}".format(omega_d[h][site]))
+		if site not in unique_sites_in_structure:
+			sites_not_in_structure.append(site)
+	cmd.spectrum('b', 'blue_white_red', structure, minimum = min_omega, maximum = max_omega)
 	print ("\nSites with data not in structure: {0}".format(', '.join(sites_not_in_structure)))
 
-	# Color residues lacking data white
+	# Color residues lacking data black
 	sites_lacking_data = [site for site in unique_sites_in_structure if site not in sites_with_data]
[…]

### 2018-02-20 paper/figures/shifts_on_structure/pymol_analysis_5FYL.py

@@ -107,18 +107,32 @@ print ("\nOf the significant sites, {0} of them are in the structure".format(len
 
 # Take a pictures of Env rotated 120 degrees relative to one another
 cmd.set_view ("""\
-     0.413237274,    0.018612767,   -0.910427451,\
-     0.910516560,    0.006436472,    0.413409144,\
-     0.013554142,   -0.999800861,   -0.014289021,\
-    -0.000450239,   -0.000266090, -364.028167725,\
-    64.596366882,   39.767127991,   -5.709826946,\
-   224.197448730,  503.894958496,  -20.000000000""")
+     0.606518745,   -0.016123615,    0.794892550,\
+     0.760999739,    0.301236391,   -0.574545264,\
+    -0.230190367,    0.953389049,    0.194983304,\
+     0.000000000,    0.000000000, -364.900909424,\
+   165.407760620,  160.255493164,  155.138549805,\
+   225.052154541,  504.749664307,  -20.000000000 """)
+
 take_pictures = False
 if take_pictures:
 	cmd.bg_color('white')
 	cmd.png('{0}_pymol_face1.png'.format(structure), width=1000, dpi=1000, ray=1)
 	cmd.rotate('y', '120')
 	cmd.png('{0}_pymol_face2.png'.format(structure), width=1000, dpi=1000, ray=1)
+
+# Identify sites that have substituted
+substituted_sites = []
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines())[1:]
+	for line in lines:
+		 (site, RMSDcorrected, RMSDbetween,RMSDwithin,
+		 BG505, BF520, significant_shift, substituted) = line.split(',')[:8]
+		 if substituted == 'TRUE':
+			 substituted_sites.append(site)
+print("There are {0} substituted sites:")
+print(", ".join(substituted_sites))
+cmd.select('subs', structure + ' and resi ' + '+'.join(substituted_sites))
 
 # Annotations of structural features
 # gp120 and gp41

### 2018-02-20 paper/figures/shifts_on_structure/pymol_analysis_1ENV.py
Add progress on mapping substitutions on structure

@@ -32,3 +32,16 @@ cmd.show('cartoon')
 cmd.color('grey40')
 #cmd.set('cartoon_transparency', '0.5')
 #cmd.set('cartoon_transparency', '0', structure)
+
+# Identify sites that have substituted
+substituted_sites = []
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		(site, RMSDcorrected, RMSDbetween,RMSDwithin, BG505, BF520, significant_shift, substituted) = line.split(',')[:8]
+		if substituted == 'True':
+			substituted_sites.append(site)
+substituted_sites = list(set(substituted_sites)) # for some reason, this is needed to handle duplicates
+cmd.select('subs', structure + ' and resi ' + '+'.join(substituted_sites))
+print("\nThere are {0} substituted sites:".format(len(substituted_sites)))
+print(', '.join(substituted_sites))

### 2018-02-20 paper/figures/shifts_on_structure/pymol_analysis_5FYL.py

@@ -10,7 +10,7 @@ When `take_pictures` is set to true, this script generates pictures of Env:
 relative to one another.
 
 This script intended to be run in the PyMOL terminal as:
-    run pymol_analysis_5FYL.py
+	run pymol_analysis_5FYL.py
 
 Hugh Haddox, November-27-2017
 """
@@ -58,14 +58,14 @@ for site in sites_in_structure:
 RMSD_dict = {}
 sig_dict = {}
 with open('../BG505_to_BF520_prefs_dist.csv') as f:
-    lines = f.readlines()[1:]
-    for line in lines:
-        elements = line.split(',')
-        site = elements[0]
-        RMSDcorrected = elements[1]
-        significant_shift = elements[6]
-        RMSD_dict[site] = float(RMSDcorrected)
-        sig_dict[site] = significant_shift
+	lines = f.readlines()[1:]
+	for line in lines:
+		elements = line.split(',')
+		site = elements[0]
+		RMSDcorrected = elements[1]
+		significant_shift = elements[6]
+		RMSD_dict[site] = float(RMSDcorrected)
+		sig_dict[site] = significant_shift
 sig_sites = [site for site in sig_dict if sig_dict[site]=='True']
 min_RMSD = min(RMSD_dict.values())
 max_RMSD = max(RMSD_dict.values())
@@ -77,18 +77,18 @@ print "max: {0}".format(max_RMSD)
 sites_with_data = RMSD_dict.keys()
 sites_not_in_structure = []
 for site in sites_with_data:
-    #print(site, RMSD_dict[site])
-    cmd.alter("{0} and resi {1}".format(structure, site),
-                "b = {0}".format(RMSD_dict[site]))
-    if site not in unique_sites_in_structure:
-    	sites_not_in_structure.append(site)
+	#print(site, RMSD_dict[site])
+	cmd.alter("{0} and resi {1}".format(structure, site),
+				"b = {0}".format(RMSD_dict[site]))
+	if site not in unique_sites_in_structure:
+		sites_not_in_structure.append(site)
 color_scheme = 'white_red'
 cmd.spectrum('b', color_scheme, structure, minimum=min_RMSD, maximum=max_RMSD)
 print ("\nSites with data not in structure: {0}".format(
-                                ', '.join(sites_not_in_structure)))
+								', '.join(sites_not_in_structure)))
 print ("Sites with significant shifts not in structure: {0}".format(
-                                ', '.join(s for s in sig_sites
[…]

### 2018-02-20 paper/figures/shifts_on_structure/pymol_analysis_5VN3.py

@@ -122,6 +122,21 @@ if take_pictures:
 	cmd.rotate('y', '120')
 	cmd.png('{0}_pymol_face2.png'.format(structure), width=1000, dpi=1000, ray=1)
 
+# Identify sites that have substituted
+substituted_sites = []
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		(site, RMSDcorrected, RMSDbetween,RMSDwithin, BG505, BF520, significant_shift, substituted) = line.split(',')[:8]
+		if substituted == 'True':
+			substituted_sites.append(site)
+substituted_sites = list(set(substituted_sites)) # for some reason, this is needed to handle duplicates
+cmd.select('subs', structure + ' and resi ' + '+'.join(substituted_sites))
+print("\nThere are {0} substituted sites:".format(len(substituted_sites)))
+print(', '.join(substituted_sites))
+subs_in_structure = [site for site in substituted_sites if site in unique_sites_in_structure]
+print ("\nOf the substituted sites, {0} of them are in the structure".format(len(subs_in_structure)))
+
 # Annotations of structural features
 # gp120 and gp41
 cmd.select('gp120', structure + ' and resi 31-511')

### 2018-02-20 paper/manuscript.tex
Addressed minor point 2.

@@
 Our modified deep mutational scanning strategy is in \FIG{dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
+This approach had the following substantive changes: we used SupT1.CCR5 cells (a SupT1 cell line that was lentiviral transduced to express CCR5 in addition to CXCR4) to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
 We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (\FIG{dms_schematic}B).

### 2018-02-20 paper/ResponseToReviews.tex
Addressed minor point 2.
Expanded on addressing minor point 2.

@@
 
-{\color{red}
-Easy to fix.
+{\color{black}
+We have clarified that we used the SupT1.CCR5 cell line to allow for infection with viruses bearing CCR5-tropic Envs.
+This cell line is a SupT1 clone that was lentiviral transduced to also express CCR5 in addition to CXCR4 and CD4, which SupT1 cells express naturally.
+In our original work with CXCR4-tropic LAI Env, we used SupT1 cells.
 }

### 2018-02-20 paper/manuscript.tex
Expanded on addressing minor point 2.

--- 2018-02-20 paper/manuscript.tex
@@
 Our modified deep mutational scanning strategy is in \FIG{dms_schematic}A.
-This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
+This approach had the following substantive changes: instead of SupT1 cells, we used SupT1.CCR5 cells (SupT1 cells lentiviral transduced to express CCR5 in addition to CXCR4 \citep{boyd2015mutations}) to support growth of viruses with transmitted-founder, CCR5-tropic Envs; we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity; and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
 We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (\FIG{dms_schematic}B).

### 2018-02-20 paper/manuscript.tex
fully incorporated figs with 5FYL, 5VN3, and min

@@
 \hspace{0.03\textwidth}
-\includegraphics[clip=true, trim=0in 0in 0in 1.2in, width=0.68\textwidth]{figures/distance_distribution.pdf}
+\includegraphics[clip=true, trim=0in 0in 0in 0.4in, width=0.65\textwidth]{figures/distance_distribution.pdf}
 \vspace{0.15in}
@@
 \begin{figure}
-{\bf \Large A} \\
-\includegraphics[width=0.8\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
+{\bf \Large A} \hspace{0.6\textwidth} {\bf \Large B}  \\
+\includegraphics[width=0.57\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
+\hspace{0.06\textwidth}
+\includegraphics[width=0.32\textwidth]{figures/rsa_vs_shifts.pdf}
 \\
-{\bf \Large B} \hspace{0.28\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
-\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
-\hspace{0.01\textwidth}
-\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
-\hspace{0.01\textwidth}
-\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
+ {\bf \Large C} \hspace{0.45\textwidth} {\bf \Large D} \\
+\includegraphics[width=0.45\textwidth]{figures/shifts_proximity.pdf}
+\hspace{0.02\textwidth}
+\includegraphics[width=0.53\textwidth]{figures/conservation_vs_shifts.pdf} 
 \caption{\label{fig:shiftfeatures}
@@
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.68\textwidth]{figures/entrenchment.pdf}}
+\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
 \caption{\label{fig:entrenchment}
@@
 As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
-}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
+}{\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
 \figsupp[\label{figsupp:glycans}Amino-acid preferences and alignment frequencies for glycosylation motifs.]

### 2018-02-22 paper/figures/shifts_on_structure/pymol_analysis_1AIK.py
Progress in responding to major comment 2

@@ -105,6 +105,22 @@ if take_pictures:
 	cmd.rotate('y', '120')
 	cmd.png('{0}_pymol_face2.png'.format(structure), width=1000, dpi=1000, ray=1)
 
+# Identify sites that have substituted
+substituted_sites = []
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		(site, RMSDcorrected, RMSDbetween,RMSDwithin, BG505, BF520, significant_shift, substituted) = line.split(',')[:8]
+		if substituted == 'True':
+			substituted_sites.append(site)
+substituted_sites = list(set(substituted_sites)) # for some reason, this is needed to handle duplicates
+cmd.select('subs', structure + ' and resi ' + '+'.join(substituted_sites))
+print("\nThere are {0} substituted sites:".format(len(substituted_sites)))
+print(', '.join(substituted_sites))
+subs_in_structure = [site for site in substituted_sites if site in unique_sites_in_structure]
+print ("\nOf the substituted sites, {0} of them are in the structure".format(len(subs_in_structure)))
+print(' ,'.join(subs_in_structure))
+
 # Annotations of structural features
 # gp120 and gp41
 cmd.select('gp120', structure + ' and resi 31-511')

### 2018-02-22 paper/figures/shifts_on_structure/pymol_analysis_1ENV.py

@@ -5,7 +5,7 @@ This script is similar to `pymol_analysis_5FYL.py`, but performs the analysis on
 the structure 1ENV instead of 5FYL.
 
 This script intended to be run in the PyMOL terminal as:
-    run pymol_analysis_1ENV.py
+	run pymol_analysis_1ENV.py
 
 Hugh Haddox, February-18-2018
 """
@@ -33,12 +33,87 @@ cmd.color('grey40')
 #cmd.set('cartoon_transparency', '0.5')
 #cmd.set('cartoon_transparency', '0', structure)
 
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		elements = line.split(',')
+		site = elements[0]
+		# The numbering for this structure follows HXB2 numbering, but is
+		# indexed at 1 where HBX2 is indexed at 512. Also, the first 29 resiudes
+		# are not derived from Env, so those should be skipped
+		try:
+			site = int(site)
+			if site >= 512 + 29:
+				site = str(site-511)
+			else:
+				continue
+		except ValueError:
+			continue
+		RMSDcorrected = elements[1]
+		significant_shift = elements[6]
+		RMSD_dict[site] = float(RMSDcorrected)
+		sig_dict[site] = significant_shift
+sig_sites = [site for site in sig_dict if sig_dict[site]=='True']
+min_RMSD = min(RMSD_dict.values())
+max_RMSD = max(RMSD_dict.values())
+print "\nThe minimum and maximum RMSDcorrected values are:"
+print "min: {0}".format(min_RMSD)
+print "max: {0}".format(max_RMSD)
+
+# Color residues by omega-by-site log10P values
+sites_with_data = RMSD_dict.keys()
+sites_not_in_structure = []
+for site in sites_with_data:
+	#print(site, RMSD_dict[site])
+	cmd.alter("{0} and resi {1}".format(structure, site),
+				"b = {0}".format(RMSD_dict[site]))
+	if site not in unique_sites_in_structure:
+		sites_not_in_structure.append(site)
+color_scheme = 'white_red'
+cmd.spectrum('b', color_scheme, structure, minimum=min_RMSD, maximum=max_RMSD)
+print ("\nSites with data not in structure: {0}".format(
+								', '.join(sites_not_in_structure)))
[…]

### 2018-02-22 paper/ResponseToReviews.tex
Progress on response to major comment 2

@@
 {\color{red}
-This is a really good suggestion.
-Hugh, are you able to add analysis with this and any other relevant conformations that are substantially different?
-We should think about whether these would be added just as analyses, or whether we'd also show such a structure.
+This is an excellent suggestion.
+We investigated substitution-shift co-localization in two other conformations: the CD4-bound conformation and the post-fusion six-helix bundle.
+In the end, our basic conclusions about co-localization remained the same.
+But, we believe these other conformations were extremely important to consider, and that the results substantially strengthen the conclusion that epistasis may be acting via long-range interactions.
+
+First, we investigated the CD4-bound conformation.
+
+Note: In the process of making these revisions, we discovered a minor bug that affected our calculations of residue-residue distances.
+We have fixed that bug in the revised version.
+None of the qualitative results change (the trends that were previously significant are still significant, and the ones that previously were not significant are still not significant.
+However, some of the quantitative values in the box plots and the corresponding P-values have changed slightly.
 }

### 2018-02-22 paper/manuscript.tex
Progress on response to major comment 2

@@
 Characteristics of significantly shifted sites.
-{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from white to red according to the magnitude of the mutational shift at each site (red indicates large shift).
+{\bf (A)} One monomer of the closed pre-fusion Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from white to red according to the magnitude of the mutational shift at each site (red indicates large shift).
 Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
@@
 There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
+This observation holds for both the pre-fusion Env trimer in (A) and the CD4-bound Env trimer~\citep[PDB 5VN3;][]{ozorowski2017open}.
 The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
 {\bf (C)} 
-Sites of significant shifts are clustered in Env's structure.
-Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
+Sites of significant shifts are clustered in the structures of both the closed pre-fusion Env trimer and the CD4-bound Env trimer.
+The left two plots show the distance of each significantly shifted and not-shifted site to the closest other shifted site in the indicated structure.
+The right-most plot shows the minimum distance across both structures.
+The trend for shifts to cluster becomes stronger when considering the minimum distance, indicating that multiple conformations may shape this trend.
 {\bf (D)}
 Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
-The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+The left-most plot shows the magnitudes of the shifts in context of the closed pre-fusion state among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 204 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+The center plot shows these magnitudes in context of the CD4-bound state, while the right-most plot shows these magnitudes when considering both structures, with the relative number of sites in each category shown in parentheses.
+In the right-most plot, non-substituted sites are considered to be contacting a substitution if they do so in at least one structure.
 \FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.

### 2018-02-23 paper/manuscript.tex
Add suplemental figure on six-helix bundle

@@
 {\input{figures/shifts_vs_subs_table}}
+\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
+{This figure shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
+The $\sim$80 residues per monomer that are resolved in this structure include four significantly shifted sites and five substituted sites.
+Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle, where we highlight the side chain of each site in a different color.
+As shown in \FIG{prefsdist}D, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
+For the two shifted sites that point inward towards the bundles core (583 and 587), this importance may involve in packing.
+Based on this structure, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby, in part since the arginine side chain is not fully resolved.
+}
+{\includegraphics[width=0.4\textwidth]{figures/shifts_on_structure/shifts_on_six_helix_bundle.pdf}}
 \end{figure}

### 2018-02-23 paper/figures/shifts_on_structure/pymol_analysis_1ENV.py

@@ -82,10 +82,10 @@ print ("Sites with significant shifts not in structure: {0}".format(
 											if s in sites_not_in_structure)))
 
 # Color residues lacking data black
-sites_lacking_data = [site for site in unique_sites_in_structure if site not in sites_with_data]
-if len(sites_lacking_data) > 0:
-	print ("\nThe following sites in the structure, but lacking data will be colored white: {0}".format(', '.join(sites_lacking_data)))
-	cmd.color('black', '{0} and resi {1}'.format(structure, '+'.join(sites_lacking_data)))
+#sites_lacking_data = [site for site in unique_sites_in_structure if site not in sites_with_data]
+#if len(sites_lacking_data) > 0:
+#	print ("\nThe following sites in the structure, but lacking data will be colored black: {0}".format(', '.join(sites_lacking_data)))
+#	cmd.color('black', '{0} and resi {1}'.format(structure, '+'.join(sites_lacking_data)))
 
 # Show significant sites as spheres
 print ("\nThere are {0} sites with significant RMSD corrected values".format(len(sig_sites)))

### 2018-02-23 paper/ResponseToReviews.tex
better explanation of ExpCM, omega, stringency parameters

@@
 
-{\color{red}
-Easy to fix.}
+{\color{black}
+This is a good suggestion.
+We have re-structured the relevant text (second and third paragraph of third subsection of Results) to expand the explanation of the dN/dS model parameters and put them in a more logical order.}
 
@@
 
-{\color{red}
-This is easy to do. Maybe Jesse should do this?}
+{\color{black}
+This is good suggestion, and related to Minor Point \#3 above.
+We have re-structured the relevant text (first few paragraphs of third subsection of Results) to substantially expand discussion of these models and parameters.
+We have also expanded the discussion around $\omega_r$ in the last subsection of the Results and in the figure / table legends.}
 

### 2018-02-23 paper/manuscript.tex
better explanation of ExpCM, omega, stringency parameters

@@
 Direct comparisons between experimentally measured amino-acid preferences and amino-acid frequencies in natural sequences are confounded by the fact that the natural sequences are evolutionarily related.
-This problem can be overcome by making the comparison in a phylogenetic context.
-Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms}.
-Importantly, since we expect many sites in Env to be under diversifying selection from immunity, we extended the ExpCM's to draw the relative dN/dS parameter ($\omega$) from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
+This problem can be overcome by making the comparison in a phylogenetic context to account for the evolutionary relationships among sequences.
+
+Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env's evolution.
+An ExpCM is a phylogenetic substitution model that incorporates the functional constraints measured in a deep mutational scanning experiment~\citep{hilton2017phydms}.
+If the experiment captures much of the actual evolutionary constraint on a gene, then an ExpCM will describe the gene's natural evolution better than a standard phylogenetic codon substitution model.
+The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
+In contrast, an ExpCM accounts for the site-specific functional constraints measured in the deep mutational scan, and then additionally adds an $\omega$ parameter that now represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
+Importantly, since we expect some sites in Env to be under diversifying selection from immunity, we extended the ExpCM's described in \citet{hilton2017phydms} to draw $\omega$ from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
 
@@
 All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
-The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
+The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification} (see also the last section of the Results).
 Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
@@
 
-\TABLE{phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
-In addition to simply comparing model fits, we can interpret the model parameters.
-ExpCM's have a stringency parameter that relates selection in the experiments to that in nature.
-A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
-Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
-We can also interpret the $\omega$ parameter.
-For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
+\TABLE{phydms} shows that ExpCM's informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
+In addition to the improved fit of the ExpCM's, we can also interpret the $\omega$ parameter.
+Recall that for standard codon substitution models, $\omega$ is simply the rate of fixation of nonsynonymous mutations relative to synonymous ones.
+For such models, the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
 Indeed, \TABLE{phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
 But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-\TABLE{phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
+But for ExpCM's, the gene-wide average $\omega$ is $>$1 (\TABLE{phydms}), indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints on the protein.
+
+ExpCM's also have a stringency parameter that relates selection in the experiments to that in nature.
+Essentially, this experiment indicates how strongly natural selection prefers the amino acids that are preferred in the deep mutational scanning~\citep{hilton2017phydms}
+A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency.
+Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
 
@@
 The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
-The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
+The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
 The letters above the logos indicate the wildtype amino acid in BG505.
@@
 We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.
-The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM's that incorporate experimental measurements as described by \citet{bloom2017identification}.
+The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models are conceptually similar but account for the experimentally measured amino-acid preferences as described by \citet{bloom2017identification}.
+All three models test if individual sites evolve faster or slower than expected, but they ``expect'' different things: the dN/dS model expects nonsynonymous and synonymous mutations fix at the same rate, while the ExpCM expects the rate at a site to depend on the experimentally measured functional constraints.
+In all cases, the evidence that a site $r$ evolves differently than expected is statistically summarized by the $P$-value that $\omega_r$ is $>$ or $<$ 1.
 The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (\TABLE{phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (\TABLE{phydms}, $\omega_r > 1$).

### 2018-02-23 paper/ResponseToReviews.tex
clarification of ExpCM / omega based on @skhilton comments

@@
 {\color{black}
-We have added text in the legend of Figure 4 to explain this.
+We have added text in the legend of Figure 4 and the ``Alignments and phylogenetic analyses of Env sequences'' subsection to explain this.
 Briefly, for individual sites, the point estimates of dN/dS values (or $\omega_r$) are not reliable when using a FEL framework for the reasons described in Pond et al (2015) and Murrell et al (2012).
+Therefore, the $P$-value is a better indication of the strength of the statistical evidence than the value of $\omega_r$ itself.
 

### 2018-02-23 paper/manuscript.tex
clarification of ExpCM / omega based on @skhilton comments

@@
 If the experiment captures much of the actual evolutionary constraint on a gene, then an ExpCM will describe the gene's natural evolution better than a standard phylogenetic codon substitution model.
-The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
-In contrast, an ExpCM accounts for the site-specific functional constraints measured in the deep mutational scan, and then additionally adds an $\omega$ parameter that now represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
+The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a single parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
+In contrast, an ExpCM accounts for the preference of each site for each of the 20 amino acids under the functional selection in the deep mutational scan, and then additionally adds an $\omega$ parameter that represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
 Importantly, since we expect some sites in Env to be under diversifying selection from immunity, we extended the ExpCM's described in \citet{hilton2017phydms} to draw $\omega$ from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
@@
 The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
-The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
+The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}.
+We report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of $\omega_r$ are unreliable for individual sites due to low numbers of observations, making the $P$-value a better indicator of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
 The letters above the logos indicate the wildtype amino acid in BG505.
-Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
+Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}.
 This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
@@
 This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
+The actual point estimates of $\omega_r$ are unreliable for individual sites due to the limited number of observations, so we report the $P$-value that $\omega_r$ is not equal to one, which is a better indication of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
 For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.

### 2018-02-23 paper/ResponseToReviews.tex
cited literature on HIV epistasis and fitness landscapes

@@
 12. A body of computational work that has been done to infer the effects of epistasis on HIV proteins (including ENV) is not adequately referenced.
-{\color{red}
-This includes the new Arup Chakroborty paper. Others?}
+
+{\color{black}
+This is a good suggestion. 
+Indeed, the first paper to make an in-depth attempt to parameterize an epistatic fitness landscape of Env was just published as our manuscript was under review (Louie et al, 2018).
+We now cite this paper as well as some related ones.
+
+In the Introduction, we note this body of work:
+\begin{quote}
+\textsl{
+Epistasis among a few combinations of Env mutations has been experimentally demonstrated (da Silva et al., 2010), and epistatic fitness landscapes have been computationally inferred for a variety of HIV proteins (Kouyos et al., 2012; Ferguson et al., 2013; Mann et al., 2014; Barton et al., 2015) including Env (Louie et al., 2018).
+}
+\end{quote}
+
+Then in the Discussion, we raise the intriguing possibility that our data might help inform these computational approaches:
+\begin{quote}
+\textsl{Our large-scale datasets of mutational effects in multiple viral strains should be useful for efforts to computationally parameterize ``fitness landscapes'' of Env (Kouyos et al., 2012; Ferguson et al., 2013; Mann et al., 2014; Barton et al., 2015; Louie et al., 2018).}
+\end{quote}}
 

### 2018-02-23 paper/manuscript.tex
cited literature on HIV epistasis and fitness landscapes

@@
 Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, shifts in amino-acid preferences could make the outcome of any study highly dependent on the Env used.
-The extent to which this is actually the case is unclear, since the few protein-wide studies of such shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
+Indeed, epistasis among a few combinations of Env mutations has been experimentally demonstrated~\citep{da2010fitness}, and epistatic fitness landscapes have been computationally inferred for a variety of HIV proteins~\citep{kouyos2012exploring,ferguson2013translating,mann2014fitness,barton2015scaling} including Env~\citep{louie2018fitness}.
+However, the only protein-wide experimental studies of how amino-acid preferences shift during evolution have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.
 
@@
 This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
+Regardless of the exact mechanism, our large-scale datasets of mutational effects in multiple viral strains should be useful for efforts to computationally parameterize ``fitness landscapes'' of Env~\citep{kouyos2012exploring,ferguson2013translating,mann2014fitness,barton2015scaling,louie2018fitness}.
 

### 2018-02-23 paper/manuscript.tex
discussed expression level and limiting to ectodomain and transmembrane domain

@@
 BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
-In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
-In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
+However, in our experiments we mutagenized only the ectodomain and transmembrane domain of Env, and excluded the signal peptide and cytoplasmic tail.
+The reason is that we measure how Env mutations affect viral growth, which is influenced both by the functionality of Env protein molecules and their expression level.
+Mutations in the signal peptide and cytoplasmic tail commonly affect Env expression level~\citep{chakrabarti1989cytoplasmic,yuste2004modulation,li1994control}, so we excluded these regions with the goal of reducing the degree to which we simply identified mutations that affected Env expression.
+In the ectodomain and transmembrane domains of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
 The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.
@@
 Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in \FIG{prefsdist}C) more mutationally tolerant in BG505.
+Differences in Env's expression level might also contribute to a general broadening or narrowing of tolerance to subsequent mutations.
+Because our experiments select for viral growth---which is affected by both Env function and expression---it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.
 
@@
 Our codon mutant libraries mutagenized all sites in \textit{env} to all 64 codons, except that the signal peptide and cytoplasmic tail were not mutagenized.
-The rationale for excluding these regions is that they are not part of Env's ectodomain, and may be prone to mutations that strongly modulate Env's expression level.
+The rationale for excluding these regions is that they are not part of Env's ectodomain, and are prone to mutations that strongly modulate Env's expression level~\citep{chakrabarti1989cytoplasmic,yuste2004modulation,li1994control}.
 

### 2018-02-25 paper/manuscript.tex
Add supp fig on clusters of shifted sites

--- 2018-02-23 paper/manuscript.tex
@@
 If the experiment captures much of the actual evolutionary constraint on a gene, then an ExpCM will describe the gene's natural evolution better than a standard phylogenetic codon substitution model.
-The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a single parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
-In contrast, an ExpCM accounts for the preference of each site for each of the 20 amino acids under the functional selection in the deep mutational scan, and then additionally adds an $\omega$ parameter that represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
+The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
+In contrast, an ExpCM accounts for the site-specific functional constraints measured in the deep mutational scan, and then additionally adds an $\omega$ parameter that now represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
 Importantly, since we expect some sites in Env to be under diversifying selection from immunity, we extended the ExpCM's described in \citet{hilton2017phydms} to draw $\omega$ from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.
@@
 The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
-The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}.
-We report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of $\omega_r$ are unreliable for individual sites due to low numbers of observations, making the $P$-value a better indicator of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
+The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
 The letters above the logos indicate the wildtype amino acid in BG505.
-Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}.
+Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
 This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
@@
 {\input{figures/shifts_vs_subs_table}}
-\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
-{This figure shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
-The $\sim$80 residues per monomer that are resolved in this structure include four significantly shifted sites and five substituted sites.
-Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle, where we highlight the side chain of each site in a different color.
-As shown in \FIG{prefsdist}D, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
-For the two shifted sites that point inward towards the bundles core (583 and 587), this importance may involve in packing.
-Based on this structure, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby, in part since the arginine side chain is not fully resolved.
+\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Clusters of shifted and substituted sites in context of different conformations of Env.]
+{This figure gives a fine-grained view of three clusters of shifted sites in context of different conformations of Env.
+In each structure, we only show side chains for shifted sites or other sites of interest, such as substituted sites.
+Side chains are colored differently according to the key.
+{\bf (A)}
+This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}.
+Each of these shifted sites has also substituted.
+Further, this cluster is adjacent to other substituted sites, raising the possibility of epistasis among these sites.
+This region of the structure involves packing between variable loops (V1/V2 and V3) and between these loops and an adjacent Env protomer.
+This packing is known to control the conformational dynamics of these loops~\cite{guttman2014cd4,powell2017plasticity}.
+Notably, a recent study found that mutations at two of the shifted sites (165 and 307) resulted in Env trimers with aberrant conformations~\cite{lee2017broadly}.
+Thus, it is possible that this cluster of shifted sites is involved in controlling Env's dynamics.
+Indeed, although each of these shifted sites prefers hydrophobic amino acids in both homologs, one homolog often has a higher preference for bulkier aromatic amino acids than the other.
+This difference could make structural sense if these residues pack in different arrangements in the different homologs, possibly due to the substitutions in this region.
+However, this hypothesis is difficult to test since there is not a high-resolution structure for BF520.
+{\bf (B)}
+This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
+These sites are shown in context of both states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
+Nearly one third of the shifted residues cluster within this structural region (9/30).
+In contrast to panel (A), only a few of the shifted sites are adjacent to substitutions or have substituted themselves.
+This trend is found in both the closed and open conformations, suggesting that the shifts may be primarily due to long-range epistatic interactions.
+Such long-range interactions seem especially plausible in this region given that it is dynamic over a large swath or residues.
+Indeed, one of the shifted sites (69) overlaps with the hydrophobic network, and mutations at another (64) have been shown to modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
+{\bf (C)}
+This panel shows a cluster of shifted sites in context of a three-helix bundle in Env, each helix of which is derived from a different monomer and which pack together to form an inter-monomer core.
+These helices form the core of the post-fusion six-helix bundle shown on the right~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, but are also present in a similar conformation in the closed pre-fusion state and the open CD4-bound state.
+These sites are symmetric around the axis of the bundle in an arrangement that is largely similar among the different structures.
+In the open conformation, one of these sites (582) is in close proximity to another site that has both shifted and substituted (520), which could account for site 582's shift in preference.
+Two of the shifted sites (583 and 587) pack in the bundle's core, suggesting that the packing of residues in this region may differ between homologs.
+However, unlike panel (A), where we observed a similar phenomenon, there are no substituted sites that also pack in a nearby location.
+Thus, aside from site 582, it is difficult to say from a structural standpoint why these sites may have shifted.
 }
-{\includegraphics[width=0.4\textwidth]{figures/shifts_on_structure/shifts_on_six_helix_bundle.pdf}}
+{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
 \end{figure}
@@
 This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
-The actual point estimates of $\omega_r$ are unreliable for individual sites due to the limited number of observations, so we report the $P$-value that $\omega_r$ is not equal to one, which is a better indication of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
 For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.

### 2018-02-25 paper/figures/shifts_on_structure/pymol_analysis_5FYL_hydrophobic_network.py

@@ -1,0 +1,183 @@
+"""
+This is a PyMOL script for analyzing shifts in preferences on the structure 5FYL
+
+It colors the structure of Env according to site-specific RMSD_corrected values
+on a color scale from white (no shift) to red (large shift), with sites with
+significant RMSD_corrected values shown as spheres.
+
+When `take_pictures` is set to true, this script generates pictures of Env:
+`5FYL_pymol_face1.png` and `5FYL_pymol_face2.png`, which are rotated 120 degrees
+relative to one another.
+
+This script intended to be run in the PyMOL terminal as:
+	run pymol_analysis_5FYL.py
+
+Hugh Haddox, November-27-2017
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '5FYL'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# Remove non-Env chains
+# gp41 = chain B
+# gp120 = chain G
+# 35O22 = chains D and E
+# PGT122 = chains H and L
+cmd.remove ('c;D,E,H,L,U,V')
+
+# Generate symmetry partners based on crystal structure
+cmd.symexp(structure, structure, structure, 3)
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey40')
+#cmd.color('grey20', structure)
+#cmd.set('cartoon_transparency', '0.5')
+#cmd.set('cartoon_transparency', '0', structure)
+#cmd.hide('hetatm')
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+	lines = f.readlines()[1:]
+	for line in lines:
+		elements = line.split(',')
+		site = elements[0]
+		RMSDcorrected = elements[1]
+		significant_shift = elements[6]
+		RMSD_dict[site] = float(RMSDcorrected)
+		sig_dict[site] = significant_shift
[…]

### 2018-02-25 paper/figures/shifts_on_structure/pymol_analysis_5VN3.py

@@ -83,7 +83,7 @@ for site in sites_with_data:
     if site not in unique_sites_in_structure:
     	sites_not_in_structure.append(site)
 color_scheme = 'white_red'
-cmd.spectrum('b', color_scheme, structure, minimum=min_RMSD, maximum=max_RMSD)
+###cmd.spectrum('b', color_scheme, structure, minimum=min_RMSD, maximum=max_RMSD)
 print ("\nSites with data not in structure: {0}".format(
                                 ', '.join(sites_not_in_structure)))
 print ("Sites with significant shifts not in structure: {0}".format(
@@ -100,7 +100,7 @@ if len(sites_lacking_data) > 0:
 print ("\nThere are {0} sites with significant RMSD corrected values".format(len(sig_sites)))
 print (', '.join(sig_sites))
 cmd.select('sig_RMSDcorrected', '{0} and resi {1}'.format(structure, '+'.join(sig_sites)))
-cmd.show('spheres', 'sig_RMSDcorrected')
+###cmd.show('spheres', 'sig_RMSDcorrected')
 cmd.hide('spheres', 'het')
 
 # Report how many significant sites are in the structure
@@ -138,6 +138,11 @@ subs_in_structure = [site for site in substituted_sites if site in unique_sites_
 print ("\nOf the substituted sites, {0} of them are in the structure".format(len(subs_in_structure)))
 
 # Annotations of structural features
+# a0 helix
+cmd.select('a0 helix', structure +' and resi 63-72')
+# network of hydrophobic residues from Ozorowski, 2017, Nature
+cmd.select('hydrophobic network', structure +' and resi 435+427+111+112+69+72+564')
+
 # gp120 and gp41
 cmd.select('gp120', structure + ' and resi 31-511')
 cmd.select('gp41', structure +' and resi 512-664')

### 2018-02-25 paper/figures/shifts_on_structure/pymol_analysis_5VN3_hydrophobic_network.py

@@ -1,0 +1,184 @@
+"""
+This is a PyMOL script for analyzing shifts in preferences on the structure 5VN3
+
+This script is similar to `pymol_analysis_5FYL.py`, but performs the analysis on
+the structure 5VN3 instead of 5FYL.
+
+This script intended to be run in the PyMOL terminal as:
+    run pymol_analysis_5VN3.py
+
+Hugh Haddox, February-17-2018
+"""
+
+# Imports
+import pymol
+from pymol import cmd
+
+# Fetch structure
+structure = '5VN3'
+cmd.reinitialize()
+cmd.delete('all')
+cmd.fetch(structure) #, type='pdb1')
+
+# In the original script `pymol_analysis_5FYL.py`, the `structure` variable was
+# the name of a `PyMOL` object of the gp160 monomer. For this structure, it is
+# the name of an object of the entire trimer. So that the rest of the code is
+# compatible with this new naming system, I will rename the trimer object and
+# then change the monomer object to equal `structure`.
+cmd.set_name(structure, '5VN3_trimer')
+cmd.select('gp160', 'c;A,G')
+cmd.create(structure, 'gp160')
+
+# Remove non-Env chains
+# gp41 = chain A, B, D
+# gp120 = chain G, I, J
+# CD4 = chains C, E, F
+# 17b = chains H, K, M and L, N, O
+cmd.remove ('c;C,E,F,H,K,M,L,N,O')
+
+# Tweak initial display and color of Env monomers
+cmd.hide('everything')
+cmd.bg_color('white')
+cmd.show('cartoon')
+cmd.color('grey40')
+#cmd.color('grey20', structure)
+#cmd.set('cartoon_transparency', '0.5')
+#cmd.set('cartoon_transparency', '0', structure)
+
+# Identify unique sites in structure
+sites_in_structure = []
+cmd.iterate("(name ca)","sites_in_structure.append(resi)")
+unique_sites_in_structure = []
+for site in sites_in_structure:
+	if site not in unique_sites_in_structure:
+		unique_sites_in_structure.append(site)
+
+# Get site-specific RMSDcorrected values and whether the shift at a site is
+# significant
+RMSD_dict = {}
+sig_dict = {}
+with open('../BG505_to_BF520_prefs_dist.csv') as f:
+    lines = f.readlines()[1:]
+    for line in lines:
+        elements = line.split(',')
+        site = elements[0]
+        RMSDcorrected = elements[1]
+        significant_shift = elements[6]
[…]

### 2018-02-25 paper/ResponseToReviews.tex

@@
 
-{\color{red}
-The best I can think of here is just to explain our rationale for excluding signal peptide and cytoplasmic tail. 
-I don't think we have a way of addressing this comment beyond that?
-Hugh and Adam, do you have additional ideas?
+{\color{black}
+This is an important point.
+Our experiments directly measure how mutations affect viral growth.
+This approach has many advantages over simple surface display approaches, since it is more relevant to the true function of Env.
+However, a caveat is that viral growth is a convolution of Env function and expression, so our experiments cannot separate these two phenotypes.
+
+We have taken steps to limit simply finding mutations affect expression.
+We have done this by \emph{not} mutagenizing the signal peptide and cytoplasmic tail, which are the most common sites of mutations that affect expression levels.
+
+We have added text to make this point clearly, and emphasize the associated caveats.
+When we first describe the deep mutational scanning approach, we now say:
+\begin{quote}
+\textsl{
+However, in our experiments we mutagenized only the ectodomain and transmembrane domain of Env, and excluded the signal peptide and cytoplasmic tail.
+The reason is that we measure how Env mutations affect viral growth, which is influenced both by the functionality of Env protein molecules and their expression level.
+Mutations in the signal peptide and cytoplasmic tail commonly affect Env expression level (Chakrabarti et al, 1989; Yuste et al, 2004; Li et al, 1994), so we excluded these regions with the goal of reducing the degree to which we simply identified mutations that affected Env expression.
+}
+\end{quote}
+Then when we describe the results, we clearly emphasize how expression level (like Env protein stability, which was already discussed) could have general effects on the results:
+\begin{quote}
+\textsl{
+Differences in Env's expression level might also contribute to a general broadening or narrowing of tolerance to subsequent mutations.
+Because our experiments select for viral growth---which is affected by both Env function and expression---it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.
+}
+\end{quote}
 }
@@
 
-{\color{red}Yes, we should move this up. Will be done in context of answering essential revision 1, and then we can describe change here.}
+{\color{black}
+This is a good suggestion.
+We now also explain this rationale in the first part of the Results as discussed in the response to Major Point \#1 above.}
 

### 2018-02-25 paper/ResponseToReviews.tex
Address major comment 2

@@
 
-{\color{red}
+{\color{black}
 This is an excellent suggestion.
-We investigated substitution-shift co-localization in two other conformations: the CD4-bound conformation and the post-fusion six-helix bundle.
-In the end, our basic conclusions about co-localization remained the same.
-But, we believe these other conformations were extremely important to consider, and that the results substantially strengthen the conclusion that epistasis may be acting via long-range interactions.
+We investigated substitution-shift co-localization in two other conformations: the open CD4-bound conformation and the post-fusion six-helix bundle.
+In the end, our basic conclusions about co-localization remained the same, adding greater support to the hypothesis that epistasis is acting via long-range interactions.
 
-First, we investigated the CD4-bound conformation.
+First, we investigated the open CD4-bound conformation.
+Figure 7D statistically tests the hypothesis that shifts are larger at sites that have substituted or sites that contacts substitutions.
+Initially, this figure only showed the results of this test for the closed pre-fusion conformation.
+Now, it shows the results for both the closed an open conformations (see the first two panels from the left in Figure 7D).
+Both conformations gave the same general result: in neither case was there statistical support for the above hypothesis.
+We also tested this hypothesis when considering both structures at once, classifying sites as contacting substitutions if they do so in at least one structure.
+Even then, we did not find support for the above hypothesis (see the right-most panel in Figure 7D).
+Thus, our investigation of the open conformation has re-enforced our initial conclusion that many of the shifts are likely due to long-range, rather than short-range, epistatic interactions.
+
+Of note, we also repeated the statistical analyses in Figure 7B and Figure 7C with the open conformation.
+In both cases, the results were qualitatively similar for both conformations.
+Thus, our initial conclusions about the solvent accessibility (Figure 7B) and clustering of shifted sites (Figure 7B) for the closed state also hold for the open state.
+
+Next, we investigated substitution-shift co-localization in the post-fusion six-helix bundle.
+Only $\sim$80 residues per Env monomer are resolved in available structures of the six-helix bundle.
+Of these structures, the one with largest number of shifted sites (PDB: 1ENV) only had a total of four shifted sites and five substituted sites.
+These numbers are too low to perform the same quantitative statistical tests from Figure 7 with the six-helix bundle.
+But, we have added a new figure (Figure 7 - Figure supplement 2) that qualitatively analyzes substitution-shift co-localization.
+We found that three of the four shifted sites cluster at one end of the helix, along with one of the five substitutions.
+We wondered whether this cluster was unique to the six-helix bundle, but found that it was also present in the closed and open states as well (see the first two panels from the right for this figure).
+Thus, although we agree that it was valuable to analyze the six-helix bundle in addition to the closed and open states, we did not observe any new patterns of co-localization that were not also present in the other states.
+
+We have added the results of the new analyses described above to the first two paragraphs of the section in the Results entitled "Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.
+
+In total, we have expanded our structural analyses to include not only the closed state, but also the open state and the six-helix bundle.
+This had made our analysis much more comprehensive.
+Further, the agreement of results between the different structures, particularly the closed and open states, has substantially strengthened our initial conclusion about substitution-shift co-localization.
 

### 2018-02-25 paper/manuscript.tex
Address major comment 2

@@
 {\bf (C)} 
-Sites of significant shifts are clustered in the structures of both the closed pre-fusion Env trimer and the CD4-bound Env trimer.
+Sites of significant shifts are clustered in the structures of both the closed and open Env trimers.
 The left two plots show the distance of each significantly shifted and not-shifted site to the closest other shifted site in the indicated structure.
-The right-most plot shows the minimum distance across both structures.
+The right-most plot shows the minimum distance across both the closed and open conformations.
 The trend for shifts to cluster becomes stronger when considering the minimum distance, indicating that multiple conformations may shape this trend.
@@
 Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
-The left-most plot shows the magnitudes of the shifts in context of the closed pre-fusion state among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 204 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
-The center plot shows these magnitudes in context of the CD4-bound state, while the right-most plot shows these magnitudes when considering both structures, with the relative number of sites in each category shown in parentheses.
-In the right-most plot, non-substituted sites are considered to be contacting a substitution if they do so in at least one structure.
+The left-most plot shows the magnitudes of the shifts in context of the closed conformation of Env among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 204 non-substituted sites that physically contact a substitution in this structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+The center plot shows these magnitudes in context of the open conformation, while the right-most plot shows these magnitudes when considering both the closed and open conformations, with the relative number of sites in each category shown in parentheses.
+In the right-most plot, non-substituted sites are considered to be contacting a substitution if they do so in at least one structural context.
 \FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
-All panels only show the 577 sites that are resolved in the crystal structure.
-Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 are crystallographically resolved.
+All panels only show the 577 and 520 sites that are resolved in the structures of the closed and open conformations, repectively.
+Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 and 27 are structurally resolved in the closed and open conformations, respectively.
 Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
@@
 {\input{figures/shifts_vs_subs_table}}
-\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Clusters of shifted and substituted sites in context of different conformations of Env.]
+
+\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
+{
+The left-most panel shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
+There are $\sim$80 residues per monomer that are resolved in this structure, including four significantly shifted sites and five substituted sites.
+Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle.
+We show the side chains for each of these sites colored according to whether the site has shifted (red) or substituted (yellow).
+We wondered whether this cluster was unique to the six-helix bundle, but found that it is also present in roughly the same structural arrangement in both the closed pre-fusion conformation (center panel) and the open CD4-bound conformation (right panel; note that some of the surrounding residues in both the closed and open structures have been hidden for the sake of clarity).
+As shown in \FIG{prefsdist}C, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
+The two shifted sites that point inward towards the bundle's core (583 and 587) may be important for packing.
+Based on these structures, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby.
+}
+{\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}
+
+\figsupp[\label{figsupp:shifted_site_clusters}Clusters of shifted and substituted sites in context of different conformations of Env.]
 {This figure gives a fine-grained view of three clusters of shifted sites in context of different conformations of Env.
@@
 Indeed, one of the shifted sites (69) overlaps with the hydrophobic network, and mutations at another (64) have been shown to modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
-{\bf (C)}
-This panel shows a cluster of shifted sites in context of a three-helix bundle in Env, each helix of which is derived from a different monomer and which pack together to form an inter-monomer core.
-These helices form the core of the post-fusion six-helix bundle shown on the right~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, but are also present in a similar conformation in the closed pre-fusion state and the open CD4-bound state.
-These sites are symmetric around the axis of the bundle in an arrangement that is largely similar among the different structures.
-In the open conformation, one of these sites (582) is in close proximity to another site that has both shifted and substituted (520), which could account for site 582's shift in preference.
-Two of the shifted sites (583 and 587) pack in the bundle's core, suggesting that the packing of residues in this region may differ between homologs.
-However, unlike panel (A), where we observed a similar phenomenon, there are no substituted sites that also pack in a nearby location.
-Thus, aside from site 582, it is difficult to say from a structural standpoint why these sites may have shifted.
 }
@@
 What distinguishes the sites that have undergone significant shifts?
-\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
-There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
-However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (\FIG{shiftfeatures}C).
+First, we analyzed the distribution of shifted sites in context of Env's three-dimensional structure.
+Env's structure is highly conformationally dynamic and undergoes large changes upon receptor binding and membrane fusion.
+In an effort to account for these dynamics, we examined multiple conformational states of Env, including: the closed pre-fusion state~\cite{julien2013crystal,pancera2014structure,stewart2016trimeric}, the open CD4-bound state~\cite{pancera2010structure,ozorowski2017open}, and the post-fusion six-helix bundle~\cite{weissenhorn1997atomic,chan1997core,tan1997atomic}.
+\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env in the closed pre-fusion state.
+There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and statistical analysis of both the closed and open states of Env (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
+However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case for both the closed and open states (\FIG{shiftfeatures}C).
 Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
+Moreover, these factors are present in multiple conformations of Env.
+We also investigated the distribution of shifted sites in context of the post-fusion six-helix bundle.
+However, only $\sim$80 residues are resolved in this structure, and of these sites, only four have shifted amino-acid preferences.
+This low number of sites does not lend itself to statistical analysis, but a qualitative analysis of these sites on the structure revealed that three of the four shifted sites cluster at one end of the helical bundle (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
 
@@
 According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
-But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (\FIG{shiftfeatures}C).
+But surprisingly, for both the closed and open conformations, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (\FIG{shiftfeatures}D).
+In the case of the six-helix bundle, there are a total of five structurally resolved substituted sites, one of which is adjacent to the cluster of shifted sites (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
+The number of resolved shifted and substituted sites in this structure is too low for to perform the same statistical analysis as in \FIG{shiftfeatures}D.
+However, we found that the cluster of shifted and substituted sites in the six-helix bundle is also present in the closed and open states (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
+Thus, this cluster is already well-accounted for by other states.
+
 There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIGSUPP[shiftfeatures]{subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
@@
 The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
-On the other hand, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
+On the other hand, as stated above, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
 

### 2018-02-26 paper/ResponseToReviews.tex
Add response to major revision 3

@@
 
-We have added the results of the new analyses described above to the first two paragraphs of the section in the Results entitled "Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.
+We have added the results of the new analyses described above to the first two paragraphs of the section in the Results entitled ``Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.
 
@@
 
-{\color{red}
-Seems like this might be best examined by looking carefully at some of the clusters?
-I'm not sure how to do it with a statistical analysis.
-Hugh and Adam---do you have any thoughts on this?
+{\color{black}
+This comment raises a very interesting question.
+To answer it, we set about examining clusters of shifted sites in greater detail.
+We found that two clusters occur within highly dynamic regions of Env.
+One hypothesis as to why shifted sites cluster in these regions is that their highly dynamic nature allows long-range epistatic interactions to be readily propagated between shifted and substituted sites that are distant from one another.
+
+The reviewers asked whether it is possible that the shifted sites interact and whether this could help explain the entrenchment we observe.
+If the above hypothesis is true, then shifted sites could interact by virtue of all being linked in a dynamic network.
+Though, because Env's dynamics are so complex, the exact mechanism of epistasis for entrenched residues is difficult to discern.
+
+We have added a new supplemental figure (Figure 7 - figure supplement 3) that provides a fine-grained structural view of each of the above clusterss.
+We have also added an entire paragraph detailing these observations at the end of the section in the Results called ``Structural and evolutionary properties of shifted sites".
+Overall, these additions strengthen the manuscript by providing a better structural intuition for how long-range epistatic interactions might occur in Env.
 }

### 2018-02-26 paper/manuscript.tex
Add response to major revision 3

@@
 
-\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
+\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted sites in the post-fusion six-helix bundle of Env.]
 {
 The left-most panel shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
+The other three helices that make up this bundle are located in another part of the structure.
 There are $\sim$80 residues per monomer that are resolved in this structure, including four significantly shifted sites and five substituted sites.
@@
 
-\figsupp[\label{figsupp:shifted_site_clusters}Clusters of shifted and substituted sites in context of different conformations of Env.]
-{This figure gives a fine-grained view of three clusters of shifted sites in context of different conformations of Env.
-In each structure, we only show side chains for shifted sites or other sites of interest, such as substituted sites.
-Side chains are colored differently according to the key.
+\figsupp[\label{figsupp:dynamic_clusters}Clusters of shifted sites in highly dynamic regions of Env.]
+{This figure gives a fine-grained view of two clusters of shifted sites, both of which reside in highly dynamic regions of Env.
+In each structure, we only show side chains for shifted sites or other sites of interest, which include substituted sites and sites that are proposed to take part in a hydrophobic network that helps control Env's dynamics~\cite{ozorowski2017open}.
+These side chains are colored differently according to the key.
 {\bf (A)}
-This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}.
-Each of these shifted sites has also substituted.
-Further, this cluster is adjacent to other substituted sites, raising the possibility of epistasis among these sites.
-This region of the structure involves packing between variable loops (V1/V2 and V3) and between these loops and an adjacent Env protomer.
-This packing is known to control the conformational dynamics of these loops~\cite{guttman2014cd4,powell2017plasticity}.
-Notably, a recent study found that mutations at two of the shifted sites (165 and 307) resulted in Env trimers with aberrant conformations~\cite{lee2017broadly}.
-Thus, it is possible that this cluster of shifted sites is involved in controlling Env's dynamics.
-Indeed, although each of these shifted sites prefers hydrophobic amino acids in both homologs, one homolog often has a higher preference for bulkier aromatic amino acids than the other.
-This difference could make structural sense if these residues pack in different arrangements in the different homologs, possibly due to the substitutions in this region.
-However, this hypothesis is difficult to test since there is not a high-resolution structure for BF520.
+This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}, a region that is not resolved in the open CD4-bound conformation.
+These sites are in Env's first/second variable loop (164 and 165) and its third variable loop (307 and 309), which pack against one another and against an adjacent Env protomer (the inter-protomer boundary is indicated by a dashed line).
+Each of the shifted sites has substituted.
+And these sites are in the immediate vicinity of other non-shifted, but substituted sites.
+Thus, it seems likely that these shifts at least partially arise from short-range epistatic interactions within this cluster.
+However, longer-range epistatic interactions with more distant sites also seem plausible given the highly dynamic nature of this region.
 {\bf (B)}
 This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
-These sites are shown in context of both states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
+One of the shifted sites (purple) is in this network.
+These sites are shown in context of both structural states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
 Nearly one third of the shifted residues cluster within this structural region (9/30).
@@
 This trend is found in both the closed and open conformations, suggesting that the shifts may be primarily due to long-range epistatic interactions.
-Such long-range interactions seem especially plausible in this region given that it is dynamic over a large swath or residues.
-Indeed, one of the shifted sites (69) overlaps with the hydrophobic network, and mutations at another (64) have been shown to modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
 }
@@
 There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and statistical analysis of both the closed and open states of Env (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
+
 However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case for both the closed and open states (\FIG{shiftfeatures}C).
@@
 On the other hand, as stated above, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
+
+Indeed, many of the shifted sites cluster within regions of Env that are highly dynamic.
+\FIGSUPP[shiftfeatures]{dynamic_clusters} shows the structural context of these clusters in finer detail.
+One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
+Upon CD4 binding, these loops become highly disordered~\cite{guttman2014cd4,ozorowski2017open}.
+Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\cite{lee2017broadly}.
+The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\cite{ozorowski2017open}.
+One of the shifted sites (69) overlaps with this network, and mutations at another (64) have been shown to strongly modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
+In total, these two clusters consist of nearly half (13/30) of the shifted sites.
+One hypothesis as to why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
+Indeed, as shown in \FIGSUPP[shiftfeatures]{dynamic_clusters}, these dynamic regions also contain several substituted sites.
+However, due to the complexity of Env's dynamics, it is difficult to discern exactly how these interactions might occur.
 

### 2018-02-26 paper/manuscript.tex
Add response to minor comment 4

@@
 One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to white to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (white) or faster than expected (red) given the BG505 experiments.
+We used a color scale that only ranges between significance values of $-6 < \log_{10}Q < 6$, which decreases the resolution between a small number of sites with extreme values, but provides greater resolution among other sites.
+Sites for which we lack $\omega_r$ estimates are colored black.
 Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
@@
 Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
-The color spectrum is similar to the one in (A) and (B), but is colored from blue to \textit{grey} to red for increased visibility of the grey residues in the hydrophobic network.
 All analyses used the phylogenetic tree in \FIG{tree}.

### 2018-02-26 paper/figures/omegabysite_structural_analysis/analysis_code_omega_by_site.py

@@ -10,6 +10,7 @@ Hugh Haddox, November-2-2017
 # Imports
 import pymol
 from pymol import cmd
+import math
 
 homologs = ['BG505', 'BF520']
 for h in homologs:
@@ -55,8 +56,8 @@ for h in homologs:
 	with open('../merged_omegabysite.csv') as f:
 		lines = f.readlines()[1:]
 		for line in lines:
-			(site, omega, P, dLnL, Q, Env, log10P, log10Pdir, N_glycans) = line.strip().split(',')
-			omega_d[Env][site] = float(log10Pdir)
+			(site, omega, P, dLnL, Q, Env, log10Q, log10Qdir, N_glycans) = line.strip().split(',')
+			omega_d[Env][site] = float(log10Qdir)
 			if float(Q) < Q_cutoff and float(omega) < 1.0:
 				sites_sig_slow[Env].append(site)
 			if float(Q) < Q_cutoff and float(omega) > 1.0:
@@ -67,7 +68,7 @@ for h in homologs:
 	print "min: {0}".format(min_omega)
 	print "max: {0}".format(max_omega)
 
-	# Color residues by omega-by-site log10P values
+	# Color residues by omega-by-site log10Q values
 
 	BG505_sites = sorted(omega_d['BG505'].keys())
 	BF520_sites = sorted(omega_d['BF520'].keys())
@@ -78,7 +79,14 @@ for h in homologs:
 		cmd.alter("{0} and resi {1}".format(structure, site), "b = {0}".format(omega_d[h][site]))
 		if site not in unique_sites_in_structure:
 			sites_not_in_structure.append(site)
-	cmd.spectrum('b', 'blue_white_red', structure, minimum = min_omega, maximum = max_omega)
+	cutoff_min_omega = -6
+	cutoff_max_omega = 6
+	if cutoff_min_omega:
+		print("Using manually specified min and max omega values of {0} and {1}".format(
+			cutoff_min_omega, cutoff_max_omega
+		))
+	cmd.spectrum('b', 'blue_white_red', structure,
+					minimum = cutoff_min_omega, maximum = cutoff_max_omega)
 	print ("\nSites with data not in structure: {0}".format(', '.join(sites_not_in_structure)))
 
 	# Color residues lacking data black
@@ -103,25 +111,12 @@ for h in homologs:
 		-0.000450239,   -0.000266090, -364.028167725,\
 		64.596366882,   39.767127991,   -5.709826946,\
[…]

### 2018-02-26 paper/manuscript.tex
fix dashes: these should be `---` rather than ` -- `

@@
 These shifted sites usually prefer a specific amino acid in one Env, but tolerate many amino acids in the other.
-Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
+Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs---and many occur at residues that do not even contact substitutions.
 Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the amino-acid preferences of most sites are conserved between moderately diverged viral strains.
@@
 We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
-Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
+Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing}---although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
 
@@
 {\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
-The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{fsanger}).
+The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{sanger}).
 We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
@@
 While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (\FIG{mutfreqs}B, compare red and blue versus gray plots).
-This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
+This fact hints that there are some shifts in amino-acid preferences between the two Envs---something that is investigated with more statistical rigor later in this paper.
 Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
@@
 A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency.
-Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
+Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms})---a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
 
@@
 For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (\FIG{BG505prefs} and \FIG{BF520prefs}).
-Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
+Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only}---and the cysteines at these sites are highly preferred in our deep mutational scanning.
 The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
-For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in \FIG{BG505prefs} and \FIG{BF520prefs}, such as 132 to 195).
+For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure}---and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in \FIG{BG505prefs} and \FIG{BF520prefs}, such as 132 to 195).
 It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.
@@
 This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
-How should we interpret this distribution -- have mutational effects shifted a lot, or not very much?
+How should we interpret this distribution---have mutational effects shifted a lot, or not very much?
 We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
@@
 In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
-In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
+In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution---and both distributions should be centered around zero if the two Envs are equivalently functional.
 Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
@@
 \FIG{entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
-As expected under entrenchment, the average effect of these mutations is deleterious -- although there are a substantial number of sites where the mutational flips are not deleterious. 
+As expected under entrenchment, the average effect of these mutations is deleterious---although there are a substantial number of sites where the mutational flips are not deleterious. 
 We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (\FIG{entrenchment}).
@@
 As can be seen from this plot, the experiments measure relatively broad mutational tolerance at many sites where the natural Env sequences have a strongly conserved motif.
-We suspect this is because many glycans serve as a shield against immunity in nature -- a function that is not required in cell culture.}
+We suspect this is because many glycans serve as a shield against immunity in nature---a function that is not required in cell culture.}
 {\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
@@
 These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric,gristick2016natively} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
-Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
+Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments---and indeed, such sites evolve more slowly than expected from our experiments.
 For instance, we find very little selection to maintain most glycans in our cell-culture experiments.
-Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments -- despite the fact that most are conserved in nature (\FIG{divsel}C and \FIGSUPP[divsel]{glycans}).
+Of the 21 N-linked glycosylation sites shared between BG505 and BF520, only four are under strong selection to maintain the glycan in our experiments---despite the fact that most are conserved in nature (\FIG{divsel}C and \FIGSUPP[divsel]{glycans}).
 This finding concords with prior literature suggesting that these glycans are selected primarily for their role in immune evasion~\citep{pugach2004prolonged,wang2013systematic,rathore2017glycosylation}.
@@
 Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
-This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, SOSIP trimer structures from diverse Env strains remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2016natively}.
+This rough generalizability of protein structures holds even for a protein as conformationally complex as Env---for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, SOSIP trimer structures from diverse Env strains remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2016natively}.
 
@@
 Indeed, we anticipate that the complete maps of mutational effects in \FIG{BG505prefs} and \FIG{BF520prefs} will be useful for future sequence-structure-function studies.
-From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
+From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments---and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
 
@@
 Consistent with studies of other proteins~\citep{natarajan2013epistasis,harms2014historical,doud2015site,starr2017pervasive}, we find a subset of sites that change markedly in which mutations they tolerate.
-Some shifted sites simply accommodate more amino acids in the more stable BG505 Env -- a type of shift that has been well-documented for other proteins~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
+Some shifted sites simply accommodate more amino acids in the more stable BG505 Env---a type of shift that has been well-documented for other proteins~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
 But interestingly, there is no strong trend for shifts to be enhanced at sites that differ between BG505 and BF520.
@@
 But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: the majority of sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs even though the preferences have shifted.
-This rather surprising result might be due to Env's exceptional conformational complexity -- mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
+This rather surprising result might be due to Env's exceptional conformational complexity---mutations can cause long-range alterations in Env's conformation~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, so it seems plausible that they might also shift mutational tolerance at distant sites.
 Regardless of the exact mechanism, our large-scale datasets of mutational effects in multiple viral strains should be useful for efforts to computationally parameterize ``fitness landscapes'' of Env~\citep{kouyos2012exploring,ferguson2013translating,mann2014fitness,barton2015scaling,louie2018fitness}.
@@
 These include the M0 model of \citet{yang2000codon}, ExpCM without a gamma-distributed $\omega$, and ExpCM in which the amino-acid preferences are averaged across sites as a control to ensure that the improved performance of these models is due to their site-specificity.
-Note how for these Env alignments, using a gamma-distributed $\omega$ is very important in order for the ExpCMs to outperform the M5 model -- we suspect this is because there are many sites of strong diversifying selection.
+Note how for these Env alignments, using a gamma-distributed $\omega$ is very important in order for the ExpCMs to outperform the M5 model---we suspect this is because there are many sites of strong diversifying selection.
 

### 2018-02-26 paper/manuscript.tex
some minor text editing / shortening up to *Entrenchment* subsection of Results

@@
 Our modified deep mutational scanning strategy is in \FIG{dms_schematic}A.
-This approach had the following substantive changes: instead of SupT1 cells, we used SupT1.CCR5 cells (SupT1 cells lentiviral transduced to express CCR5 in addition to CXCR4 \citep{boyd2015mutations}) to support growth of viruses with transmitted-founder, CCR5-tropic Envs; we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity; and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
+This approach had the following substantive changes: instead of SupT1 cells, we used SupT1.CCR5 cells (SupT1 cells that express CCR5 in addition to CXCR4 \citep{boyd2015mutations}) to support growth of viruses with transmitted-founder, CCR5-tropic Envs; we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity; and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
 We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (\FIG{dms_schematic}B).
@@
 Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
-Both ExpCM models have a stringency parameter $>$1.
+Both ExpCM's have a stringency parameter $>$1.
 All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
 The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification} (see also the last section of the Results).
-Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
+Analyses were performed using \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms}.
 \TABLE{phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
@@
 But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
-But for ExpCM's, the gene-wide average $\omega$ is $>$1 (\TABLE{phydms}), indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints on the protein.
+For the ExpCM's, the gene-wide average $\omega$ is $>$1 (\TABLE{phydms}), indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints on the protein.
 
 ExpCM's also have a stringency parameter that relates selection in the experiments to that in nature.
-Essentially, this experiment indicates how strongly natural selection prefers the amino acids that are preferred in the deep mutational scanning~\citep{hilton2017phydms}
+Essentially, this parameter indicates how strongly natural selection prefers the amino acids that are preferred in the deep mutational scanning~\citep{hilton2017phydms}.
 A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency.
@@
 For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure}---and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in \FIG{BG505prefs} and \FIG{BF520prefs}, such as 132 to 195).
-It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.
+It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our experiments do a reasonable job of authentically measuring functional selection on Env.
 
@@
 Differences in Env's expression level might also contribute to a general broadening or narrowing of tolerance to subsequent mutations.
-Because our experiments select for viral growth---which is affected by both Env function and expression---it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.
+The reason is that our experiments select for viral growth (which is affected by both Env function and expression), so it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.
 
@@
 There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
-This observation holds for both the pre-fusion Env trimer in (A) and the CD4-bound Env trimer~\citep[PDB 5VN3;][]{ozorowski2017open}.
+This observation holds for both the closed trimer conformation in (A) and the CD4-bound trimer conformation~\citep[PDB 5VN3;][]{ozorowski2017open}.
 The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
@@
 The left two plots show the distance of each significantly shifted and not-shifted site to the closest other shifted site in the indicated structure.
-The right-most plot shows the minimum distance across both the closed and open conformations.
-The trend for shifts to cluster becomes stronger when considering the minimum distance, indicating that multiple conformations may shape this trend.
+The right-most plot shows the minimum distance across both conformations.
+The trend for shifts to cluster becomes stronger when considering the minimum distance, suggesting multiple conformations contribute to this trend.
 {\bf (D)}
 Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
-The left-most plot shows the magnitudes of the shifts in context of the closed conformation of Env among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 204 non-substituted sites that physically contact a substitution in this structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
-The center plot shows these magnitudes in context of the open conformation, while the right-most plot shows these magnitudes when considering both the closed and open conformations, with the relative number of sites in each category shown in parentheses.
-In the right-most plot, non-substituted sites are considered to be contacting a substitution if they do so in at least one structural context.
-\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
-All panels only show the 577 and 520 sites that are resolved in the structures of the closed and open conformations, repectively.
-Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 and 27 are structurally resolved in the closed and open conformations, respectively.
+The plots show the magnitudes of the shifts among crystallographically resolved sites that have substituted between BG505 and BF520, the non-substituted sites that physically contact a substitution in the indicated structure(s) (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a borderline-significant tendency of significantly shifted sites to have substituted.
+All plots only show sites that are crystallographically resolved in the indicated structure(s). 
 Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
@@
 
-\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted sites in the post-fusion six-helix bundle of Env.]
+\figsupp[\label{figsupp:shifts_on_six_helix_bundle}Cluster of shifted sites in the post-fusion six-helix bundle of Env.]
 {
@@
 Env's structure is highly conformationally dynamic and undergoes large changes upon receptor binding and membrane fusion.
-In an effort to account for these dynamics, we examined multiple conformational states of Env, including: the closed pre-fusion state~\cite{julien2013crystal,pancera2014structure,stewart2016trimeric}, the open CD4-bound state~\cite{pancera2010structure,ozorowski2017open}, and the post-fusion six-helix bundle~\cite{weissenhorn1997atomic,chan1997core,tan1997atomic}.
+In an effort to account for these dynamics, we examined multiple conformational states of Env: the closed pre-fusion state~\citep{stewart2016trimeric}, the open CD4-bound state~\citep{ozorowski2017open}, and the post-fusion six-helix bundle~\citep{weissenhorn1997atomic}.
 \FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env in the closed pre-fusion state.
 There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and statistical analysis of both the closed and open states of Env (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
+We did not attempt to analyze the association between solvent accessibility and shift for the post-fusion six-helix bundle because crystal structures of this conformation only contain $\sim$80 Env residues~\citep{weissenhorn1997atomic,chan1997core,tan1997atomic}.
 
-However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case for both the closed and open states (\FIG{shiftfeatures}C).
+However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure.
+A statistical analysis confirms that there is clustering of shifted sites for the closed and open conformations, with the effect being strongest when we define contacts based on the closest intra-residue distance across these two conformations  (\FIG{shiftfeatures}C).
 Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
-Moreover, these factors are present in multiple conformations of Env.
-We also investigated the distribution of shifted sites in context of the post-fusion six-helix bundle.
-However, only $\sim$80 residues are resolved in this structure, and of these sites, only four have shifted amino-acid preferences.
-This low number of sites does not lend itself to statistical analysis, but a qualitative analysis of these sites on the structure revealed that three of the four shifted sites cluster at one end of the helical bundle (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
+We also investigated clustering of shifted sites in the post-fusion six-helix bundle.
+Because structures of this conformation only resolve the coordinates of $\sim$80 residues, we did not perform a statistical analysis.
+However, a qualitative analysis revealed that three of the four shifted sites that are resolved in the post-fusion conformation cluster at one end of the helical bundle (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundle}).
 
@@
 But surprisingly, for both the closed and open conformations, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (\FIG{shiftfeatures}D).
-In the case of the six-helix bundle, there are a total of five structurally resolved substituted sites, one of which is adjacent to the cluster of shifted sites (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
-The number of resolved shifted and substituted sites in this structure is too low for to perform the same statistical analysis as in \FIG{shiftfeatures}D.
-However, we found that the cluster of shifted and substituted sites in the six-helix bundle is also present in the closed and open states (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).
-Thus, this cluster is already well-accounted for by other states.
+For the six-helix bundle, there are five structurally resolved substituted sites, one of which is adjacent to the cluster of shifted sites (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundle}).
+The number of resolved shifted and substituted sites in this structure is too small for a meaningful statistical analysis of the type in \FIG{shiftfeatures}D.
+However, the cluster of shifted and substituted sites in the six-helix bundle is also present in the closed and open states (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundle}), and so is included in the statistical analyses in \FIG{shiftfeatures}D.
 
@@
 The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
-On the other hand, as stated above, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
+That fact that Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open} may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.
 
-Indeed, many of the shifted sites cluster within regions of Env that are highly dynamic.
+Indeed, many of the shifted sites cluster within regions of Env that are highly conformationally dynamic.
 \FIGSUPP[shiftfeatures]{dynamic_clusters} shows the structural context of these clusters in finer detail.
 One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
-Upon CD4 binding, these loops become highly disordered~\cite{guttman2014cd4,ozorowski2017open}.
-Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\cite{lee2017broadly}.
-The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\cite{ozorowski2017open}.
-One of the shifted sites (69) overlaps with this network, and mutations at another (64) have been shown to strongly modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
-In total, these two clusters consist of nearly half (13/30) of the shifted sites.
-One hypothesis as to why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
-Indeed, as shown in \FIGSUPP[shiftfeatures]{dynamic_clusters}, these dynamic regions also contain several substituted sites.
-However, due to the complexity of Env's dynamics, it is difficult to discern exactly how these interactions might occur.
+Upon CD4 binding, these loops become highly disordered~\citep{guttman2014cd4,ozorowski2017open}.
+Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\citep{lee2017broadly}.
+The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\citep{ozorowski2017open}.
+One of the shifted sites (site 69) overlaps with this network, and mutations at another (site 64) have been shown to strongly modulate the relative stability of the open and closed conformations~\citep{de2015immunogenicity}.
+In total, these two clusters consist of nearly half of the shifted sites (13 out of 30).
+One hypothesis why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
+Due to the complexity of Env's dynamics, it is difficult to discern exactly how these interactions might occur.
+But there is certainly a trend for sites that are conformationally dynamic to also be sites that show shifts in their amino-acid preferences during evolution.
 

### 2018-02-26 paper/manuscript.tex
more minor revisions to text

@@
 We wondered whether this cluster was unique to the six-helix bundle, but found that it is also present in roughly the same structural arrangement in both the closed pre-fusion conformation (center panel) and the open CD4-bound conformation (right panel; note that some of the surrounding residues in both the closed and open structures have been hidden for the sake of clarity).
-As shown in \FIG{prefsdist}C, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
+As shown in \FIG{prefsdist}C, each of the shifted sites are more tolerant of mutations in BG505 than in BF520.
 The two shifted sites that point inward towards the bundle's core (583 and 587) may be important for packing.
@@
 \figsupp[\label{figsupp:dynamic_clusters}Clusters of shifted sites in highly dynamic regions of Env.]
-{This figure gives a fine-grained view of two clusters of shifted sites, both of which reside in highly dynamic regions of Env.
-In each structure, we only show side chains for shifted sites or other sites of interest, which include substituted sites and sites that are proposed to take part in a hydrophobic network that helps control Env's dynamics~\cite{ozorowski2017open}.
-These side chains are colored differently according to the key.
+{A fine-grained view of two clusters of shifted sites, both of which are in conformationally dynamic regions of Env.
+In each structure, we only show side chains for shifted sites or other sites of interest, which include substituted sites and sites that are proposed to take part in a hydrophobic network that helps control Env's dynamics~\citep{ozorowski2017open}.
+These side chains are colored according to the key in the upper right.
 {\bf (A)}
-This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}, a region that is not resolved in the open CD4-bound conformation.
+A cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}, a region that is not resolved in the open CD4-bound conformation.
 These sites are in Env's first/second variable loop (164 and 165) and its third variable loop (307 and 309), which pack against one another and against an adjacent Env protomer (the inter-protomer boundary is indicated by a dashed line).
@@
 {\bf (B)}
-This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
+A cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\citep{ozorowski2017open}.
 One of the shifted sites (purple) is in this network.
@@
 One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to white to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (white) or faster than expected (red) given the BG505 experiments.
-We used a color scale that only ranges between significance values of $-6 < \log_{10}Q < 6$, which decreases the resolution between a small number of sites with extreme values, but provides greater resolution among other sites.
+%We used a color scale that only ranges between significance values of $-6 < \log_{10}Q < 6$, which decreases the resolution between a small number of sites with extreme values, but provides greater resolution among other sites.
 Sites for which we lack $\omega_r$ estimates are colored black.
@@
 An HTML rendering of this notebook is in Supplementary~file~\ref{suppfile:html}.
+A repository containing all of this code is also available at \url{https://github.com/jbloomlab/EnvMutationalShiftsPaper}~\citep[][copy archived at \url{https://github.com/elifesciences-publications/EnvMutationalShiftsPaper}]{haddox2018github}.
 
@@
 All code and input data required to reproduce all analyses in this paper are in Supplementary~file~\ref{suppfile:code} (see also Supplementary~file~\ref{suppfile:html}).
+A repository containing all of this code is also available at \url{https://github.com/jbloomlab/EnvMutationalShiftsPaper}~\citep[][copy archived at \url{https://github.com/elifesciences-publications/EnvMutationalShiftsPaper}]{haddox2018github}.
 The deep sequencing data are on the Sequence Read Archive with the accession numbers listed in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.

### 2018-02-26 paper/manuscript.tex

@@
 Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
-The plots show the magnitudes of the shifts among crystallographically resolved sites that have substituted between BG505 and BF520, the non-substituted sites that physically contact a substitution in the indicated structure(s) (any non-hydrogen atom within 3.5\angstrom), and all other sites.
+The plots show the magnitudes of the shifts among structurally resolved sites that have substituted between BG505 and BF520, the non-substituted sites that physically contact a substitution in the indicated structure(s) (any non-hydrogen atom within 3.5\angstrom), and all other sites.
 \FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a borderline-significant tendency of significantly shifted sites to have substituted.
-All plots only show sites that are crystallographically resolved in the indicated structure(s). 
+All plots only show sites that are structurally resolved in the indicated structure(s). 
 Structural distances and solvent accessibilities were calculated using all monomers in the trimer.

### 2018-02-26 paper/manuscript.tex
small tweaks to wording on clusters of shifted sites

--- 2018-02-26 paper/manuscript.tex
@@
 One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
-Upon CD4 binding, these loops become highly disordered~\citep{guttman2014cd4,ozorowski2017open}.
+These interactions are likely involved in regulating the transition between conformational states, and upon CD4 binding, these loops become highly disordered~\citep{guttman2014cd4,ozorowski2017open}.
 Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\citep{lee2017broadly}.
+Strikingly, this cluster of shifted sites may may reflect previously observed differences in the conformational dynamics of this regions between these two Envs; the V2 region of BF520 SOSIP trimer is more accessible to deuterium exchange than the BG505 SOSIP trimer~\citep{verkerke2016epitope}.
 The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\citep{ozorowski2017open}.
@@
 One hypothesis why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
-Due to the complexity of Env's dynamics, it is difficult to discern exactly how these interactions might occur.
-But there is certainly a trend for sites that are conformationally dynamic to also be sites that show shifts in their amino-acid preferences during evolution.
+It is difficult to discern exactly how these interactions might occur, but there is certainly a trend for sites that are conformationally dynamic to also be sites that show shifts in their amino-acid preferences during evolution.
 

### 2018-02-26 paper/ResponseToReviews.tex

--- 2018-02-25 paper/ResponseToReviews.tex
@@
 The results offer a body of knowledge useful for identifying i) residues and networks that are critical for basic function and ii) hot spots for evolvability or immune pressure.
+
 The analysis has been performed to a high standard (though some clarifications are necessary; see minor comments), and the openness of the data is a great service to the community.
@@
 
-{\color{red}
+{\color{black}
 This is an excellent suggestion.
-We investigated substitution-shift co-localization in two other conformations: the CD4-bound conformation and the post-fusion six-helix bundle.
-In the end, our basic conclusions about co-localization remained the same.
-But, we believe these other conformations were extremely important to consider, and that the results substantially strengthen the conclusion that epistasis may be acting via long-range interactions.
+In the initial paper, we only analyzed the closed pre-fusion conformation of the Env trimer.
+We now analyze substitution-shift co-localization in two other conformations: the open CD4-bound conformation and the post-fusion six-helix bundle.
+In the end, our basic conclusions remain the same, adding greater support to the hypothesis that epistasis is acting via long-range interactions.
 
-First, we investigated the CD4-bound conformation.
+First, we investigated the open CD4-bound conformation.
+Figure 7D statistically tests the hypothesis that shifts are larger at sites that have substituted or sites that contacts substitutions.
+Initially, this figure only showed the results of this test for the closed pre-fusion conformation.
+Now, it shows the results for both the closed and open conformations.
+Both conformations gave the same general result: there is no significant trend for shifted sites to contact sites of substitutions.
+We also considered both structures at once, classifying sites as contacting substitutions if they do so in at least one structure.
+Even then, we did not find an association (see the new right-most panel in Figure 7D).
+Thus, our investigation of the open conformation re-enforced our initial conclusion that many of the shifts are likely due to long-range, rather than short-range, epistatic interactions.
 
-Note: In the process of making these revisions, we discovered a minor bug that affected our calculations of residue-residue distances.
-We have fixed that bug in the revised version.
-None of the qualitative results change (the trends that were previously significant are still significant, and the ones that previously were not significant are still not significant.
-However, some of the quantitative values in the box plots and the corresponding P-values have changed slightly.
+We also repeated the analyses in Figure 7B and Figure 7C with the open conformation.
+Our initial conclusions about the solvent accessibility (Figure 7B) and clustering of shifted sites (Figure 7B) also held for the open conformation.
+
+Next, we investigated substitution-shift co-localization in the post-fusion six-helix bundle.
+Only $\sim$80 residues per Env monomer are resolved in available structures of the six-helix bundle.
+The structure with largest number of shifted sites (PDB: 1ENV) only contains four shifted sites and five substituted sites.
+These numbers are too low for the statistical tests used in Figure 7.
+But we have added a new figure (Figure 7 - Figure supplement 2) that qualitatively analyzes substitution-shift co-localization in the post-fusion six-helix bundle.
+Three of the four shifted sites cluster at one end of the helix, along with one of the five substitutions.
+However, this cluster is also present in the closed and open states as well (Figure 7 - Figure supplement 2).
+Thus, the post-fusion six-helix bundle did not reveal any new patterns of co-localization that were not also present in the other states.
+
+We now describe the results of the analyses in the first two paragraphs of the Results subsection ``Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.
+
+In the process of making these revisions, we discovered a minor bug that affected our calculations of residue-residue distances. 
+We fixed that bug in the revised version. 
+None of the main results change, but some of the quantitative values in the box plots and the corresponding P-values have changed slightly.
 }
@@
 
-{\color{red}
-Seems like this might be best examined by looking carefully at some of the clusters?
-I'm not sure how to do it with a statistical analysis.
-Hugh and Adam---do you have any thoughts on this?
+{\color{black}
+To address this comment, we examined the clusters of shifted sites in greater detail.
+Two clusters occur within conformationally dynamic regions of Env.
+One hypothesis as to why shifted sites cluster in these regions is that their conformationally dynamic nature allows long-range epistatic interactions to be propagated between shifted and substituted sites that are distant from one another.
+
+We have added a new supplemental figure (Figure 7 - figure supplement 3) that provides a fine-grained structural view of each of the clusters.
+We have also added an entire paragraph detailing these observations at the end of the section in the Results called ``Structural and evolutionary properties of shifted sites".
+Overall, these additions strengthen the manuscript by providing a better structural intuition for how long-range epistatic interactions might occur in Env.
 }
@@
 
-{\color{red}
-Hugh, can you do this?}
+{\color{black}
+We have now changed the color schemes to be higher contrast in both of these figures.
+The one in Fig 7a now goes from white-to-red instead of grey-to-red.
+And the one in Fig 9a now goes from blue-to-white-to-red instead of blue-to-grey-to-red.}
 

### 2018-02-26 paper/manuscript.tex
fixed typo on Adam commit (`may` was repeated twice in a row)`

@@
 Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\citep{lee2017broadly}.
-Strikingly, this cluster of shifted sites may may reflect previously observed differences in the conformational dynamics of this regions between these two Envs; the V2 region of BF520 SOSIP trimer is more accessible to deuterium exchange than the BG505 SOSIP trimer~\citep{verkerke2016epitope}.
+Strikingly, this cluster of shifted sites may reflect previously observed differences in the conformational dynamics of this regions between these two Envs; the V2 region of BF520 SOSIP trimer is more accessible to deuterium exchange than the BG505 SOSIP trimer~\citep{verkerke2016epitope}.
 The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\citep{ozorowski2017open}.

### 2018-02-27 paper/manuscript_for_final_submission/manuscript_nofigs_embeddedtabsrefs.tex
submitted _eLife_ revision

--- 2018-02-26 paper/manuscript.tex
@@
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
 \caption{\label{fig:tree}
@@
 The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
-{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
+{}
 \figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
+%\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
 \caption{\label{fig:dms_schematic}
@@
 Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
-}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
+}{}%\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
 \end{figure}
@@
 \begin{figure}
-\begin{minipage}[t]{0.34\textwidth}
-{\bf \Large A} \\ 
-\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
-\centerline{\small BG505}
-\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
-\centerline{\small BF520}
-\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
-\end{minipage}
-\begin{minipage}[t]{0.01\textwidth}
-\end{minipage}
-\begin{minipage}[t]{0.65\textwidth}
-{\bf \Large B} \\
-\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
-\end{minipage}
 \caption{\label{fig:mutfreqs}
@@
 We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
-{\input{figures/avgmutfreqs}}
+{}
 \figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
@@
 This is expected, since most amino acid mutations at most sites will be strongly deleterious, and so have low preference.
-}{\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
+}{}%\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
 \figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
@@
 \begin{fullwidth}
-{\centering \include{figures/gamma_omega_modelcomparison}}
+{\centering 
+\begin{tabular}{lrrrrrrrrr}
+\toprule
+           Model &  $\Delta$AIC &  LogLikelihood &  nParams &  stringency &  $\overline{\omega}$ &  $\omega_{\alpha}$ &  $\omega_{\beta}$ &  nsites $\omega_r > 1$ &  nsites $\omega_r < 1$ \\
+\midrule
+     ExpCM BF520 &          0.0 &       -35218.8 &        7 &         2.8 &                  1.4 &                1.0 &               0.7 &                     66 &                     35 \\
+     ExpCM BG505 &        269.0 &       -35353.3 &        7 &         2.1 &                  1.3 &                0.9 &               0.7 &                     65 &                     53 \\
+ Goldman-Yang M5 &       3455.1 &       -36941.4 &       12 &         nan &                  0.8 &                0.6 &               0.7 &                     14 &                    211 \\
+\bottomrule
+\end{tabular}
+}
 \caption{\label{tab:phydms}
@@
 \begin{fullwidth}
-{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
+%{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
 \caption{\label{fig:BG505prefs}
@@
 \begin{fullwidth}
-{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
+%{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
 \caption{\label{fig:BF520prefs}
@@
 \figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 \begin{fullwidth}
-{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
-\hspace{0.01\textwidth} 
-\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
-\hspace{0.03\textwidth}
-\includegraphics[clip=true, trim=0in 0in 0in 0.4in, width=0.65\textwidth]{figures/distance_distribution.pdf}
-\vspace{0.15in}
-\\ 
-{\bf \Large C} \\
-\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
 \caption{\label{fig:prefsdist}
@@
 \begin{figure}
-{\bf \Large A} \hspace{0.6\textwidth} {\bf \Large B}  \\
-\includegraphics[width=0.57\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
-\hspace{0.06\textwidth}
-\includegraphics[width=0.32\textwidth]{figures/rsa_vs_shifts.pdf}
-\\
- {\bf \Large C} \hspace{0.45\textwidth} {\bf \Large D} \\
-\includegraphics[width=0.45\textwidth]{figures/shifts_proximity.pdf}
-\hspace{0.02\textwidth}
-\includegraphics[width=0.53\textwidth]{figures/conservation_vs_shifts.pdf} 
 \caption{\label{fig:shiftfeatures}
@@
 However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
-{\input{figures/shifts_vs_subs_table}}
+{}
 
@@
 }
-{\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}
+{}%\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}
 
@@
 }
-{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
+{}%\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
 \end{figure}
@@
 \begin{figure}
-\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
+%\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
 \caption{\label{fig:entrenchment}
@@
 \begin{figure}
-\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
+%\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
 \caption{\label{fig:divsel}
@@
 As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
-}{\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
+}{}%\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
 \figsupp[\label{figsupp:glycans}Amino-acid preferences and alignment frequencies for glycosylation motifs.]
@@
 We suspect this is because many glycans serve as a shield against immunity in nature---a function that is not required in cell culture.}
-{\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
+{}%\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
 \figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
@@
 The $Q$-value indicated is the maximum of that for the two Envs.
-}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
+}{}%\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
 \figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
@@
 
-\bibliography{references.bib}
+\begin{thebibliography}{114}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\urlprefix}{}
+\providecommand{\doiprefix}{doi: }
+
+\bibitem[{Al-Mawsawi et~al.(2014)Al-Mawsawi, Laith Q and Wu, Nicholas C and
+  Olson, C Anders and Shi, Vivian Cai and Qi, Hangfei and Zheng, Xiaojuan and
+  Wu, Ting-Ting and Sun, Ren}]{al2014high}
+\textbf{\color{eLifeMediumGrey} Al-Mawsawi LQ}, Wu NC, Olson CA, Shi VC, Qi H,
+  Zheng X, Wu TT, Sun R.
+\newblock High-throughput profiling of point mutations across the HIV-1 genome.
+\newblock Retrovirology.  2014; 11(1):124.
+
+\bibitem[{Albert et~al.(1990)Albert, Jan and Abrahamsson, Bengt and Nagy,
+  Karoly and Aurelius, Elisabeth and Gaines, Hans and Nystr{\"o}m, Gunnel and
+  Feny{\"o}, Eva Maria}]{albert1990rapid}
+\textbf{\color{eLifeMediumGrey} Albert J}, Abrahamsson B, Nagy K, Aurelius E,
+  Gaines H, Nystr{\"o}m G, Feny{\"o} EM.
+\newblock Rapid development of isolate-specific neutralizing antibodies after
+  primary HIV-1 infection and consequent emergence of virus variants which
+  resist neutralization by autologous sera.
+\newblock AIDS.  1990; 4(2):107--112.
+
+\bibitem[{Almond et~al.(2010)Almond, David and Kimura, Tetsuya and Kong,
+  XiangPeng and Swetnam, James and Zolla-Pazner, Susan and Cardozo,
+  Timothy}]{almond2010structural}
+\textbf{\color{eLifeMediumGrey} Almond D}, Kimura T, Kong X, Swetnam J,
+  Zolla-Pazner S, Cardozo T.
+\newblock Structural conservation predominates over sequence variability in the
+  crown of {HIV} type 1's {V3} loop.
+\newblock AIDS research and human retroviruses.  2010; 26(6):717--723.
+
+\bibitem[{van Anken et~al.(2008)van Anken, Eelco and Sanders, Rogier W and
+  Liscaljet, I Marije and Land, Aafke and Bontjer, Ilja and Tillemans, Sonja
+  and Nabatov, Alexey A and Paxton, William A and Berkhout, Ben and Braakman,
+  Ineke}]{van2008only}
+\textbf{\color{eLifeMediumGrey} van Anken E}, Sanders RW, Liscaljet IM, Land A,
+  Bontjer I, Tillemans S, Nabatov AA, Paxton WA, Berkhout B, Braakman I.
+\newblock Only five of 10 strictly conserved disulfide bonds are essential for
+  folding and eight for function of the HIV-1 envelope glycoprotein.
+\newblock Molecular Biology of the Cell.  2008; 19(10):4298--4309.
+
+\bibitem[{Ashenberg et~al.(2013)Ashenberg, Orr and Gong, L Ian and Bloom, Jesse
+  D}]{ashenberg2013mutational}
+\textbf{\color{eLifeMediumGrey} Ashenberg O}, Gong LI, Bloom JD.
+\newblock Mutational effects on stability are largely conserved during protein
+  evolution.
+\newblock Proceedings of the National Academy of Sciences.  2013;
+  110(52):21071--21076.
+
+\bibitem[{Barton et~al.(2015)Barton, John P and Kardar, Mehran and Chakraborty,
+  Arup K}]{barton2015scaling}
+\textbf{\color{eLifeMediumGrey} Barton JP}, Kardar M, Chakraborty AK.
+\newblock Scaling laws describe memories of host--pathogen riposte in the {HIV}
+  population.
+\newblock Proceedings of the National Academy of Sciences.  2015;
+  112(7):1965--1970.
+
+\bibitem[{Bazykin(2015)Bazykin, Georgii A}]{bazykin2015changing}
+\textbf{\color{eLifeMediumGrey} Bazykin GA}.
+\newblock Changing preferences: deformation of single position amino acid
+  fitness landscapes and evolution of proteins.
+\newblock Biology Letters.  2015; 11(10):20150315.
+
+\bibitem[{Benjamini and Hochberg(1995)Benjamini, Yoav and Hochberg,
+  Yosef}]{benjamini1995controlling}
+\textbf{\color{eLifeMediumGrey} Benjamini Y}, Hochberg Y.
+\newblock Controlling the false discovery rate: a practical and powerful
+  approach to multiple testing.
+\newblock Journal of the Royal Statistical Society, Series B.  1995; p.
+  289--300.
+
+\bibitem[{Binley et~al.(2000)Binley, James M and Sanders, Rogier W and Clas,
+  Brian and Schuelke, Norbert and Master, Aditi and Guo, Yong and Kajumo,
+  Francis and Anselma, Deborah J and Maddon, Paul J and Olson, William C and
+  others}]{binley2000recombinant}
+\textbf{\color{eLifeMediumGrey} Binley JM}, Sanders RW, Clas B, Schuelke N,
+  Master A, Guo Y, Kajumo F, Anselma DJ, Maddon PJ, Olson WC, et~al.
+\newblock A recombinant human immunodeficiency virus type 1 envelope
+  glycoprotein complex stabilized by an intermolecular disulfide bond between
+  the gp120 and gp41 subunits is an antigenic mimic of the trimeric
+  virion-associated structure.
+\newblock Journal of Virology.  2000; 74(2):627--643.
+
+\bibitem[{Bloom(2014{\natexlab{a}})Bloom, Jesse D}]{bloom2014experimentally}
+\textbf{\color{eLifeMediumGrey} Bloom JD}.
+\newblock An experimentally determined evolutionary model dramatically improves
+  phylogenetic fit.
+\newblock Mol Biol Evol.  2014; 31(8):1956--1978.
+
+\bibitem[{Bloom(2014{\natexlab{b}})Bloom, Jesse D}]{bloom2014informed}
+\textbf{\color{eLifeMediumGrey} Bloom JD}.
+\newblock An experimentally informed evolutionary model improves phylogenetic
+  fit to divergent lactamase homologs.
+\newblock Molecular Biology and Evolution.  2014; 31(10):2753--2769.
+
+\bibitem[{Bloom(2015)Bloom, Jesse D}]{bloom2015software}
+\textbf{\color{eLifeMediumGrey} Bloom JD}.
+\newblock Software for the analysis and visualization of deep mutational
+  scanning data.
+\newblock BMC Bioinformatics.  2015; 16:168.
+
+\bibitem[{Bloom(2017)Bloom, Jesse D}]{bloom2017identification}
+\textbf{\color{eLifeMediumGrey} Bloom JD}.
+\newblock Identification of positive selection in genes is greatly improved by
+  using experimentally informed site-specific models.
+\newblock Biology Direct.  2017; 12(1):1.
+
+\bibitem[{Bloom et~al.(2006)Bloom, Jesse D and Labthavikul, Sy T and Otey,
+  Christopher R and Arnold, Frances H}]{bloom2006protein}
+\textbf{\color{eLifeMediumGrey} Bloom JD}, Labthavikul ST, Otey CR, Arnold FH.
+\newblock Protein stability promotes evolvability.
+\newblock Proceedings of the National Academy of Sciences.  2006;
+  103(15):5869--5874.
+
+\bibitem[{Boyd et~al.(2015)Boyd, David F and Peterson, Dylan and Haggarty, Beth
+  S and Jordan, Andrea PO and Hogan, Michael J and Goo, Leslie and Hoxie, James
+  A and Overbaugh, Julie}]{boyd2015mutations}
+\textbf{\color{eLifeMediumGrey} Boyd DF}, Peterson D, Haggarty BS, Jordan AP,
+  Hogan MJ, Goo L, Hoxie JA, Overbaugh J.
+\newblock Mutations in HIV-1 envelope that enhance entry with the macaque CD4
+  receptor alter antibody recognition by disrupting quaternary interactions
+  within the trimer.
+\newblock Journal of Virology.  2015; 89(2):894--907.
+
+\bibitem[{Burton et~al.(2005)Burton, Dennis R and Stanfield, Robyn L and
+  Wilson, Ian A}]{burton2005antibody}
+\textbf{\color{eLifeMediumGrey} Burton DR}, Stanfield RL, Wilson IA.
+\newblock Antibody vs. HIV in a clash of evolutionary titans.
+\newblock Proceedings of the National Academy of Sciences of the United States
+  of America.  2005; 102(42):14943--14948.
+
+\bibitem[{Chakrabarti et~al.(1989)Chakrabarti, L and Emerman, M and Tiollais, P
+  and Sonigo, P}]{chakrabarti1989cytoplasmic}
+\textbf{\color{eLifeMediumGrey} Chakrabarti L}, Emerman M, Tiollais P, Sonigo
+  P.
+\newblock The cytoplasmic domain of simian immunodeficiency virus transmembrane
+  protein modulates infectivity.
+\newblock Journal of Virology.  1989; 63(10):4395--4403.
+
+\bibitem[{Chan et~al.(1997)Chan, David C and Fass, Deborah and Berger, James M
+  and Kim, Peter S}]{chan1997core}
+\textbf{\color{eLifeMediumGrey} Chan DC}, Fass D, Berger JM, Kim PS.
+\newblock Core structure of gp41 from the HIV envelope glycoprotein.
+\newblock Cell.  1997; 89(2):263--273.
+
+\bibitem[{Chan et~al.(2017)Chan, Yvonne H and Venev, Sergey V and Zeldovich,
+  Konstantin B and Matthews, C Robert}]{chan2017correlation}
+\textbf{\color{eLifeMediumGrey} Chan YH}, Venev SV, Zeldovich KB, Matthews CR.
+\newblock Correlation of fitness landscapes from three orthologous TIM barrels
+  originates from sequence and structure constraints.
+\newblock Nature Communications.  2017; 8:14614.
+
+\bibitem[{Chothia and Lesk(1986)Chothia, Cyrus and Lesk, Arthur
+  M}]{chothia1986relation}
+\textbf{\color{eLifeMediumGrey} Chothia C}, Lesk AM.
+\newblock The relation between the divergence of sequence and structure in
+  proteins.
+\newblock The EMBO Journal.  1986; 5(4):823.
+
+\bibitem[{Crooks et~al.(2004)Crooks, Gavin E and Hon, Gary and Chandonia,
+  John-Marc and Brenner, Steven E}]{crooks2004weblogo}
+\textbf{\color{eLifeMediumGrey} Crooks GE}, Hon G, Chandonia JM, Brenner SE.
+\newblock WebLogo: a sequence logo generator.
+\newblock Genome Research.  2004; 14(6):1188--1190.
+
+\bibitem[{Davenport et~al.(2013)Davenport, Thaddeus M and Guttman, Miklos and
+  Guo, Wenjin and Cleveland, Brad and Kahn, Maria and Hu, Shiu-Lok and Lee,
+  Kelly K}]{davenport2013isolate}
+\textbf{\color{eLifeMediumGrey} Davenport TM}, Guttman M, Guo W, Cleveland B,
+  Kahn M, Hu SL, Lee KK.
+\newblock Isolate-specific differences in the conformational dynamics and
+  antigenicity of HIV-1 gp120.
+\newblock Journal of Virology.  2013; 87(19):10855--10873.
+
+\bibitem[{Dingens et~al.(2017)Dingens, Adam S and Haddox, Hugh K and Overbaugh,
+  Julie and Bloom, Jesse D}]{dingens2017comprehensive}
+\textbf{\color{eLifeMediumGrey} Dingens AS}, Haddox HK, Overbaugh J, Bloom JD.
+\newblock Comprehensive mapping of HIV-1 escape from a broadly neutralizing
+  antibody.
+\newblock Cell Host \& Microbe.  2017; 21:777--787.
+
+\bibitem[{Doud et~al.(2015)Doud, Michael B and Ashenberg, O and Bloom, Jesse
+  D}]{doud2015site}
+\textbf{\color{eLifeMediumGrey} Doud MB}, Ashenberg O, Bloom JD.
+\newblock Site-specific amino acid preferences are mostly conserved in two
+  closely related protein homologs.
+\newblock Molecular Biology and Evolution.  2015; 32(11):2944--2960.
+
+\bibitem[{Doud and Bloom(2016)Doud, Michael B and Bloom, Jesse
+  D}]{doud2016accurate}
+\textbf{\color{eLifeMediumGrey} Doud MB}, Bloom JD.
+\newblock Accurate measurement of the effects of all amino-acid mutations to
+  influenza hemagglutinin.
+\newblock Viruses.  2016; 8:155.
+
+\bibitem[{Duenas-Decamp et~al.(2016)Duenas-Decamp, Maria and Jiang, Li and
+  Bolon, Daniel and Clapham, Paul R}]{duenas2016saturation}
+\textbf{\color{eLifeMediumGrey} Duenas-Decamp M}, Jiang L, Bolon D, Clapham PR.
+\newblock Saturation Mutagenesis of the {HIV-1} Envelope {CD4} Binding Loop
+  Reveals Residues Controlling Distinct Trimer Conformations.
+\newblock PLoS Pathogens.  2016; 12(11):e1005988.
+
+\bibitem[{Faria et~al.(2014)Faria, Nuno R and Rambaut, Andrew and Suchard, Marc
+  A and Baele, Guy and Bedford, Trevor and Ward, Melissa J and Tatem, Andrew J
+  and Sousa, Jo{\~a}o D and Arinaminpathy, Nimalan and P{\'e}pin, Jacques and
+  others}]{faria2014early}
+\textbf{\color{eLifeMediumGrey} Faria NR}, Rambaut A, Suchard MA, Baele G,
+  Bedford T, Ward MJ, Tatem AJ, Sousa JD, Arinaminpathy N, P{\'e}pin J, et~al.
+\newblock The early spread and epidemic ignition of {HIV-1} in human
+  populations.
+\newblock Science.  2014; 346(6205):56--61.
+
+\bibitem[{Ferguson et~al.(2013)Ferguson, Andrew L and Mann, Jaclyn K and
+  Omarjee, Saleha and Ndung’u, Thumbi and Walker, Bruce D and Chakraborty,
+  Arup K}]{ferguson2013translating}
+\textbf{\color{eLifeMediumGrey} Ferguson AL}, Mann JK, Omarjee S, Ndung’u T,
+  Walker BD, Chakraborty AK.
+\newblock Translating {HIV} sequences into quantitative fitness landscapes
+  predicts viral vulnerabilities for rational immunogen design.
+\newblock Immunity.  2013; 38(3):606--617.
+
+\bibitem[{Gong et~al.(2013)Gong, Lizhi Ian and Suchard, Marc A and Bloom, Jesse
+  D}]{gong2013stability}
+\textbf{\color{eLifeMediumGrey} Gong LI}, Suchard MA, Bloom JD.
+\newblock Stability-mediated epistasis constrains the evolution of an influenza
+  protein.
+\newblock eLife.  2013; 2:e00631.
+
+\bibitem[{Goo et~al.(2014)Goo, Leslie and Chohan, Vrasha and Nduati, Ruth and
+  Overbaugh, Julie}]{goo2014early}
+\textbf{\color{eLifeMediumGrey} Goo L}, Chohan V, Nduati R, Overbaugh J.
+\newblock Early development of broadly neutralizing antibodies in
+  HIV-1-infected infants.
+\newblock Nature Medicine.  2014; 20(6):655--658.
+
+\bibitem[{Gray et~al.(2017)Gray, Vanessa E and Hause, Ronald J and Fowler,
+  Douglas M}]{gray2017analysis}
+\textbf{\color{eLifeMediumGrey} Gray VE}, Hause RJ, Fowler DM.
+\newblock Analysis of Large-Scale Mutagenesis Data To Assess the Impact of
+  Single Amino Acid Substitutions.
+\newblock Genetics.  2017; 207(1):53.
+
+\bibitem[{Gristick et~al.(2016)Gristick, Harry B and von Boehmer, Lotta and
+  West Jr, Anthony P and Schamber, Michael and Gazumyan, Anna and Golijanin,
+  Jovana and Seaman, Michael S and F{\"a}tkenheuer, Gerd and Klein, Florian and
+  Nussenzweig, Michel C and others}]{gristick2016natively}
+\textbf{\color{eLifeMediumGrey} Gristick HB}, von Boehmer L, West~Jr AP,
+  Schamber M, Gazumyan A, Golijanin J, Seaman MS, F{\"a}tkenheuer G, Klein F,
+  Nussenzweig MC, et~al.
+\newblock Natively glycosylated HIV-1 Env structure reveals new mode for
+  antibody recognition of the CD4-binding site.
+\newblock Nature Structural \& Molecular Biology.  2016; 23(10):906--915.
+
+\bibitem[{Guttman et~al.(2015)Guttman, M and Cupo, A and Julien, JP and
+  Sanders, RW and Wilson, IA and Moore, JP and Lee, KK}]{guttman2015antibody}
+\textbf{\color{eLifeMediumGrey} Guttman M}, Cupo A, Julien J, Sanders R, Wilson
+  I, Moore J, Lee K.
+\newblock Antibody potency relates to the ability to recognize the closed,
+  pre-fusion form of {HIV Env}.
+\newblock Nature Communications.  2015; 6:6144--6144.
+
+\bibitem[{Guttman et~al.(2014)Guttman, Miklos and Garcia, Natalie K and Cupo,
+  Albert and Matsui, Tsutomu and Julien, Jean-Philippe and Sanders, Rogier W
+  and Wilson, Ian A and Moore, John P and Lee, Kelly K}]{guttman2014cd4}
+\textbf{\color{eLifeMediumGrey} Guttman M}, Garcia NK, Cupo A, Matsui T, Julien
+  JP, Sanders RW, Wilson IA, Moore JP, Lee KK.
+\newblock CD4-induced activation in a soluble HIV-1 Env trimer.
+\newblock Structure.  2014; 22(7):974--984.
+
+\bibitem[{Haddox et~al.(2016)Haddox, Hugh K and Dingens, Adam S and Bloom,
+  Jesse D}]{haddox2016experimental}
+\textbf{\color{eLifeMediumGrey} Haddox HK}, Dingens AS, Bloom JD.
+\newblock Experimental estimation of the effects of all amino-acid mutations to
+  {HIV's} envelope protein on viral replication in cell culture.
+\newblock PLoS Pathogens.  2016; 12(12):e1006114.
+
+\bibitem[{Haddox et~al.(2018)Haddox, Hugh K and Dingens, Adam S and Hilton,
+  Sarah K and Overbaugh, Julie and Bloom, Jesse D}]{haddox2018github}
+\textbf{\color{eLifeMediumGrey} Haddox HK}, Dingens AS, Hilton SK, Overbaugh J,
+  Bloom JD.
+\newblock Computer code for ``Mapping mutational effects along the evolutionary
+  landscape of HIV envelope''.
+\newblock Github.  2018; commit ec358fd.
+\newblock
+  \urlprefix\url{https://github.com/jbloomlab/EnvMutationalShiftsPaper}.
+
+\bibitem[{Halpern and Bruno(1998)Halpern, Aaron L and Bruno, William
+  J}]{halpern1998evolutionary}
+\textbf{\color{eLifeMediumGrey} Halpern AL}, Bruno WJ.
+\newblock Evolutionary distances for protein-coding sequences: modeling
+  site-specific residue frequencies.
+\newblock Molecular Biology and Evolution.  1998; 15(7):910--917.
+
+\bibitem[{Harms and Thornton(2014)Harms, Michael J and Thornton, Joseph
+  W}]{harms2014historical}
+\textbf{\color{eLifeMediumGrey} Harms MJ}, Thornton JW.
+\newblock Historical contingency and its biophysical basis in glucocorticoid
+  receptor evolution.
+\newblock Nature.  2014; 512(7513):203.
+
+\bibitem[{Hedges et~al.(2006)Hedges, S Blair and Dudley, Joel and Kumar,
+  Sudhir}]{hedges2006timetree}
+\textbf{\color{eLifeMediumGrey} Hedges SB}, Dudley J, Kumar S.
+\newblock {TimeTree}: a public knowledge-base of divergence times among
+  organisms.
+\newblock Bioinformatics.  2006; 22(23):2971--2972.
+
+\bibitem[{Hilton et~al.(2017)Hilton, Sarah K and Doud, Michael B and Bloom,
+  Jesse D}]{hilton2017phydms}
+\textbf{\color{eLifeMediumGrey} Hilton SK}, Doud MB, Bloom JD.
+\newblock phydms: Software for phylogenetic analyses informed by deep
+  mutational scanning.
+\newblock PeerJ.  2017; 5:e3657.
+
+\bibitem[{Huang et~al.(2014)Huang, Jinghe and Kang, Byong H and Pancera, Marie
+  and Lee, Jeong Hyun and Tong, Tommy and Feng, Yu and Imamichi, Hiromi and
+  Georgiev, Ivelin S and Chuang, Gwo-Yu and Druz, Aliaksandr and
+  others}]{huang2014broad}
+\textbf{\color{eLifeMediumGrey} Huang J}, Kang BH, Pancera M, Lee JH, Tong T,
+  Feng Y, Imamichi H, Georgiev IS, Chuang GY, Druz A, et~al.
+\newblock Broad and potent HIV-1 neutralization by a human antibody that binds
+  the gp41-gp120 interface.
+\newblock Nature.  2014; 515(7525):138--142.
+
+\bibitem[{Julien et~al.(2013)Julien, Jean-Philippe and Cupo, Albert and Sok,
+  Devin and Stanfield, Robyn L and Lyumkis, Dmitry and Deller, Marc C and
+  Klasse, Per-Johan and Burton, Dennis R and Sanders, Rogier W and Moore, John
+  P and others}]{julien2013crystal}
+\textbf{\color{eLifeMediumGrey} Julien JP}, Cupo A, Sok D, Stanfield RL,
+  Lyumkis D, Deller MC, Klasse PJ, Burton DR, Sanders RW, Moore JP, et~al.
+\newblock Crystal structure of a soluble cleaved HIV-1 envelope trimer.
+\newblock Science.  2013; 342(6165):1477--1483.
+
+\bibitem[{Julien et~al.(2015)Julien, Jean-Philippe and Lee, Jeong Hyun and
+  Ozorowski, Gabriel and Hua, Yuanzi and de la Pe{\~n}a, Alba Torrents and de
+  Taeye, Steven W and Nieusma, Travis and Cupo, Albert and Yasmeen, Anila and
+  Golabek, Michael and others}]{julien2015design}
+\textbf{\color{eLifeMediumGrey} Julien JP}, Lee JH, Ozorowski G, Hua Y, de~la
+  Pe{\~n}a AT, de~Taeye SW, Nieusma T, Cupo A, Yasmeen A, Golabek M, et~al.
+\newblock Design and structure of two HIV-1 clade C SOSIP. 664 trimers that
+  increase the arsenal of native-like Env immunogens.
+\newblock Proceedings of the National Academy of Sciences.  2015;
+  112(38):11947--11952.
+
+\bibitem[{Kabsch and Sander(1983)Kabsch, Wolfgang and Sander,
+  Christian}]{kabsch1983dictionary}
+\textbf{\color{eLifeMediumGrey} Kabsch W}, Sander C.
+\newblock Dictionary of protein secondary structure: pattern recognition of
+  hydrogen-bonded and geometrical features.
+\newblock Biopolymers.  1983; 22(12):2577--2637.
+
+\bibitem[{Klink and Bazykin(2017)Klink, GV and Bazykin, GA}]{klink2017parallel}
+\textbf{\color{eLifeMediumGrey} Klink G}, Bazykin G.
+\newblock Parallel evolution of metazoan mitochondrial proteins.
+\newblock Genome Biology and Evolution.  2017; 9:1341--1350.
+
+\bibitem[{Korber et~al.(1998)Korber, Bette and Foley, Brian T and Kuiken, C and
+  Pillai, Satish K and Sodroski, Joseph G and others}]{korber1998numbering}
+\textbf{\color{eLifeMediumGrey} Korber B}, Foley BT, Kuiken C, Pillai SK,
+  Sodroski JG, et~al.
+\newblock Numbering positions in HIV relative to HXB2CG.
+\newblock Human Retroviruses and AIDS.  1998; 3:102--111.
+
+\bibitem[{Kosakovsky~Pond and Frost(2005)Kosakovsky Pond, Sergei L and Frost,
+  Simon DW}]{kosakovsky2005not}
+\textbf{\color{eLifeMediumGrey} Kosakovsky~Pond SL}, Frost SD.
+\newblock Not so different after all: a comparison of methods for detecting
+  amino acid sites under selection.
+\newblock Molecular Biology and Evolution.  2005; 22(5):1208--1222.
+
+\bibitem[{Kouyos et~al.(2012)Kouyos, Roger D and Leventhal, Gabriel E and
+  Hinkley, Trevor and Haddad, Mojgan and Whitcomb, Jeannette M and Petropoulos,
+  Christos J and Bonhoeffer, Sebastian}]{kouyos2012exploring}
+\textbf{\color{eLifeMediumGrey} Kouyos RD}, Leventhal GE, Hinkley T, Haddad M,
+  Whitcomb JM, Petropoulos CJ, Bonhoeffer S.
+\newblock Exploring the complexity of the {HIV-1} fitness landscape.
+\newblock PLoS Genetics.  2012; 8(3):e1002551.
+
+\bibitem[{Kumar et~al.(2017)Kumar, Amit and Natarajan, Chandrasekhar and
+  Moriyama, Hideaki and Witt, Christopher C and Weber, Roy E and Fago, Angela
+  and Storz, Jay F}]{kumar2017stability}
+\textbf{\color{eLifeMediumGrey} Kumar A}, Natarajan C, Moriyama H, Witt CC,
+  Weber RE, Fago A, Storz JF.
+\newblock Stability-mediated epistasis restricts accessible mutational pathways
+  in the functional evolution of avian hemoglobin.
+\newblock Molecular Biology and Evolution.  2017; 34(5):1240--1251.
+
+\bibitem[{Kwong et~al.(2002)Kwong, Peter D and Doyle, Michael L and Casper,
+  David J and Cicala, Claudia and Leavitt, Stephanie A and Majeed, Shahzad and
+  Steenbeke, Tavis D and Venturi, Miro and Chaiken, Irwin and Fung, Michael and
+  others}]{kwong2002hiv}
+\textbf{\color{eLifeMediumGrey} Kwong PD}, Doyle ML, Casper DJ, Cicala C,
+  Leavitt SA, Majeed S, Steenbeke TD, Venturi M, Chaiken I, Fung M, et~al.
+\newblock HIV-1 evades antibody-mediated neutralization through conformational
+  masking of receptor-binding sites.
+\newblock Nature.  2002; 420(6916):678--682.
+
+\bibitem[{Kwong et~al.(2000)Kwong, Peter D and Wyatt, Richard and Majeed,
+  Shahzad and Robinson, James and Sweet, Raymond W and Sodroski, Joseph and
+  Hendrickson, Wayne A}]{kwong2000structures}
+\textbf{\color{eLifeMediumGrey} Kwong PD}, Wyatt R, Majeed S, Robinson J, Sweet
+  RW, Sodroski J, Hendrickson WA.
+\newblock Structures of HIV-1 gp120 envelope glycoproteins from
+  laboratory-adapted and primary isolates.
+\newblock Structure.  2000; 8(12):1329--1339.
+
+\bibitem[{Lee et~al.(2017)Lee, Jeong Hyun and Andrabi, Raiees and Su, Ching-Yao
+  and Yasmeen, Anila and Julien, Jean-Philippe and Kong, Leopold and Wu,
+  Nicholas C and McBride, Ryan and Sok, Devin and Pauthner, Matthias and
+  others}]{lee2017broadly}
+\textbf{\color{eLifeMediumGrey} Lee JH}, Andrabi R, Su CY, Yasmeen A, Julien
+  JP, Kong L, Wu NC, McBride R, Sok D, Pauthner M, et~al.
+\newblock A broadly neutralizing antibody targets the dynamic HIV envelope
+  trimer apex via a long, rigidified, and anionic $\beta$-hairpin structure.
+\newblock Immunity.  2017; 46(4):690--702.
+
+\bibitem[{Li et~al.(1994)Li, Yan and Luo, Lizhong and Thomas, David Y and Kang,
+  O Yong}]{li1994control}
+\textbf{\color{eLifeMediumGrey} Li Y}, Luo L, Thomas DY, Kang OY.
+\newblock Control of expression, glycosylation, and secretion of HIV-1 gp120 by
+  homologous and heterologous signal sequences.
+\newblock Virology.  1994; 204(1):266--278.
+
+\bibitem[{Louie et~al.(2018)Louie, Raymond HY and Kaczorowski, Kevin J and
+  Barton, John P and Chakraborty, Arup K and McKay, Matthew
+  R}]{louie2018fitness}
+\textbf{\color{eLifeMediumGrey} Louie RH}, Kaczorowski KJ, Barton JP,
+  Chakraborty AK, McKay MR.
+\newblock Fitness landscape of the human immunodeficiency virus envelope
+  protein that is targeted by antibodies.
+\newblock Proceedings of the National Academy of Sciences.  2018; p. 201717765.
+
+\bibitem[{Lynch et~al.(2009)Lynch, Rebecca M and Shen, Tongye and Gnanakaran, S
+  and Derdeyn, Cynthia A}]{lynch2009appreciating}
+\textbf{\color{eLifeMediumGrey} Lynch RM}, Shen T, Gnanakaran S, Derdeyn CA.
+\newblock Appreciating {HIV} type 1 diversity: subtype differences in {Env}.
+\newblock AIDS Research and Human Retroviruses.  2009; 25(3):237--248.
+
+\bibitem[{Lyumkis et~al.(2013)Lyumkis, Dmitry and Julien, Jean-Philippe and de
+  Val, Natalia and Cupo, Albert and Potter, Clinton S and Klasse, Per-Johan and
+  Burton, Dennis R and Sanders, Rogier W and Moore, John P and Carragher,
+  Bridget and others}]{lyumkis2013cryo}
+\textbf{\color{eLifeMediumGrey} Lyumkis D}, Julien JP, de~Val N, Cupo A, Potter
+  CS, Klasse PJ, Burton DR, Sanders RW, Moore JP, Carragher B, et~al.
+\newblock Cryo-EM structure of a fully glycosylated soluble cleaved HIV-1
+  envelope trimer.
+\newblock Science.  2013; 342(6165):1484--1490.
+
+\bibitem[{Mann et~al.(2014)Mann, Jaclyn K and Barton, John P and Ferguson,
+  Andrew L and Omarjee, Saleha and Walker, Bruce D and Chakraborty, Arup and
+  Ndung'u, Thumbi}]{mann2014fitness}
+\textbf{\color{eLifeMediumGrey} Mann JK}, Barton JP, Ferguson AL, Omarjee S,
+  Walker BD, Chakraborty A, Ndung'u T.
+\newblock The fitness landscape of {HIV-1} gag: advanced modeling approaches
+  and validation of model predictions by in vitro testing.
+\newblock PLoS Computational Biology.  2014; 10(8):e1003776.
+
+\bibitem[{McCandlish and Stoltzfus(2014)McCandlish, David M and Stoltzfus,
+  Arlin}]{mccandlish2014modeling}
+\textbf{\color{eLifeMediumGrey} McCandlish DM}, Stoltzfus A.
+\newblock Modeling evolution using the probability of fixation: history and
+  implications.
+\newblock The Quarterly Review of Biology.  2014; 89(3):225--252.
+
+\bibitem[{Moore et~al.(2009)Moore, Penny L and Gray, Elin S and Morris,
+  Lynn}]{moore2009specificity}
+\textbf{\color{eLifeMediumGrey} Moore PL}, Gray ES, Morris L.
+\newblock Specificity of the autologous neutralizing antibody response.
+\newblock Current Opinion in HIV and AIDS.  2009; 4(5):358.
+
+\bibitem[{Munro et~al.(2014)Munro, James B and Gorman, Jason and Ma, Xiaochu
+  and Zhou, Zhou and Arthos, James and Burton, Dennis R and Koff, Wayne C and
+  Courter, Joel R and Smith, Amos B and Kwong, Peter D and
+  others}]{munro2014conformational}
+\textbf{\color{eLifeMediumGrey} Munro JB}, Gorman J, Ma X, Zhou Z, Arthos J,
+  Burton DR, Koff WC, Courter JR, Smith AB, Kwong PD, et~al.
+\newblock Conformational dynamics of single HIV-1 envelope trimers on the
+  surface of native virions.
+\newblock Science.  2014; 346(6210):759--763.
+
+\bibitem[{Murrell et~al.(2015)Murrell, Ben and Weaver, Steven and Smith, Martin
+  D and Wertheim, Joel O and Murrell, Sasha and Aylward, Anthony and Eren,
+  Kemal and Pollner, Tristan and Martin, Darren P and Smith, Davey M and
+  others}]{murrell2015gene}
+\textbf{\color{eLifeMediumGrey} Murrell B}, Weaver S, Smith MD, Wertheim JO,
+  Murrell S, Aylward A, Eren K, Pollner T, Martin DP, Smith DM, et~al.
+\newblock Gene-wide identification of episodic selection.
+\newblock Molecular Biology and Evolution.  2015; 32(5):1365--1371.
+
+\bibitem[{Murrell et~al.(2012)Murrell, Ben and Wertheim, Joel O and Moola,
+  Sasha and Weighill, Thomas and Scheffler, Konrad and Pond, Sergei L
+  Kosakovsky}]{murrell2012detecting}
+\textbf{\color{eLifeMediumGrey} Murrell B}, Wertheim JO, Moola S, Weighill T,
+  Scheffler K, Pond SLK.
+\newblock Detecting individual sites subject to episodic diversifying
+  selection.
+\newblock PLoS Genetics.  2012; 8(7):e1002764.
+
+\bibitem[{Natarajan et~al.(2013)Natarajan, Chandrasekhar and Inoguchi, Noriko
+  and Weber, Roy E and Fago, Angela and Moriyama, Hideaki and Storz, Jay
+  F}]{natarajan2013epistasis}
+\textbf{\color{eLifeMediumGrey} Natarajan C}, Inoguchi N, Weber RE, Fago A,
+  Moriyama H, Storz JF.
+\newblock Epistasis among adaptive mutations in deer mouse hemoglobin.
+\newblock Science.  2013; 340(6138):1324--1327.
+
+\bibitem[{Nduati et~al.(2000)Nduati, Ruth and John, Grace and Mbori-Ngacha,
+  Dorothy and Richardson, Barbra and Overbaugh, Julie and Mwatha, Anthony and
+  Ndinya-Achola, Jeckoniah and Bwayo, Job and Onyango, Francis E and Hughes,
+  James and others}]{nduati2000effect}
+\textbf{\color{eLifeMediumGrey} Nduati R}, John G, Mbori-Ngacha D, Richardson
+  B, Overbaugh J, Mwatha A, Ndinya-Achola J, Bwayo J, Onyango FE, Hughes J,
+  et~al.
+\newblock Effect of breastfeeding and formula feeding on transmission of HIV-1:
+  a randomized clinical trial.
+\newblock JAMA.  2000; 283(9):1167--1174.
+
+\bibitem[{Olshevsky et~al.(1990)Olshevsky, U and Helseth, E and Furman, C and
+  Li, J and Haseltine, W and Sodroski, J}]{olshevsky1990identification}
+\textbf{\color{eLifeMediumGrey} Olshevsky U}, Helseth E, Furman C, Li J,
+  Haseltine W, Sodroski J.
+\newblock Identification of individual human immunodeficiency virus type 1
+  gp120 amino acids important for CD4 receptor binding.
+\newblock Journal of Virology.  1990; 64(12):5701--5707.
+
+\bibitem[{Ozorowski et~al.(2017)Ozorowski, Gabriel and Pallesen, Jesper and de
+  Val, Natalia and Lyumkis, Dmitry and Cottrell, Christopher A and Torres,
+  Jonathan L and Copps, Jeffrey and Stanfield, Robyn L and Cupo, Albert and
+  Pugach, Pavel and others}]{ozorowski2017open}
+\textbf{\color{eLifeMediumGrey} Ozorowski G}, Pallesen J, de~Val N, Lyumkis D,
+  Cottrell CA, Torres JL, Copps J, Stanfield RL, Cupo A, Pugach P, et~al.
+\newblock Open and closed structures reveal allostery and pliability in the
+  HIV-1 envelope spike.
+\newblock Nature.  2017; 547(7663):360--363.
+
+\bibitem[{Pancera et~al.(2014)Pancera, Marie and Zhou, Tongqing and Druz,
+  Aliaksandr and Georgiev, Ivelin S and Soto, Cinque and Gorman, Jason and
+  Huang, Jinghe and Acharya, Priyamvada and Chuang, Gwo-Yu and Ofek, Gilad and
+  others}]{pancera2014structure}
+\textbf{\color{eLifeMediumGrey} Pancera M}, Zhou T, Druz A, Georgiev IS, Soto
+  C, Gorman J, Huang J, Acharya P, Chuang GY, Ofek G, et~al.
+\newblock Structure and immune recognition of trimeric pre-fusion HIV-1 Env.
+\newblock Nature.  2014; 514(7523):455--461.
+
+\bibitem[{Parrish et~al.(2013)Parrish, Nicholas F and Gao, Feng and Li, Hui and
+  Giorgi, Elena E and Barbian, Hannah J and Parrish, Erica H and Zajic, Lara
+  and Iyer, Shilpa S and Decker, Julie M and Kumar, Amit and
+  others}]{parrish2013phenotypic}
+\textbf{\color{eLifeMediumGrey} Parrish NF}, Gao F, Li H, Giorgi EE, Barbian
+  HJ, Parrish EH, Zajic L, Iyer SS, Decker JM, Kumar A, et~al.
+\newblock Phenotypic properties of transmitted founder {HIV-1}.
+\newblock Proceedings of the National Academy of Sciences.  2013;
+  110(17):6626--6633.
+
+\bibitem[{Peden et~al.(1991)Peden, Keith and Emerman, Michael and Montagnier,
+  Luc}]{peden1991changes}
+\textbf{\color{eLifeMediumGrey} Peden K}, Emerman M, Montagnier L.
+\newblock Changes in growth properties on passage in tissue culture of viruses
+  derived from infectious molecular clones of HIV-1 LAI, HIV-1 MAL, and HIV-1
+  ELI.
+\newblock Virology.  1991; 185(2):661--672.
+
+\bibitem[{Podgornaia and Laub(2015)Podgornaia, Anna I and Laub, Michael
+  T}]{podgornaia2015pervasive}
+\textbf{\color{eLifeMediumGrey} Podgornaia AI}, Laub MT.
+\newblock Pervasive degeneracy and epistasis in a protein-protein interface.
+\newblock Science.  2015; 347(6222):673--677.
+
+\bibitem[{Pollock et~al.(2012)Pollock, David D and Thiltgen, Grant and
+  Goldstein, Richard A}]{pollock2012amino}
+\textbf{\color{eLifeMediumGrey} Pollock DD}, Thiltgen G, Goldstein RA.
+\newblock Amino acid coevolution induces an evolutionary Stokes shift.
+\newblock Proceedings of the National Academy of Sciences.  2012;
+  109(21):E1352--E1359.
+
+\bibitem[{Pond et~al.(2010)Pond, Sergei Kosakovsky and Delport, Wayne and Muse,
+  Spencer V and Scheffler, Konrad}]{pond2010correcting}
+\textbf{\color{eLifeMediumGrey} Pond SK}, Delport W, Muse SV, Scheffler K.
+\newblock Correcting the bias of empirical frequency parameter estimators in
+  codon models.
+\newblock PLoS One.  2010; 5(7):e11230.
+
+\bibitem[{Posada and Buckley(2004)Posada, David and Buckley, Thomas
+  R}]{posada2004model}
+\textbf{\color{eLifeMediumGrey} Posada D}, Buckley TR.
+\newblock Model selection and model averaging in phylogenetics: advantages of
+  Akaike information criterion and Bayesian approaches over likelihood ratio
+  tests.
+\newblock Systematic Biology.  2004; 53(5):793--808.
+
+\bibitem[{Poss and Overbaugh(1999)Poss, Mary and Overbaugh,
+  Julie}]{poss1999variants}
+\textbf{\color{eLifeMediumGrey} Poss M}, Overbaugh J.
+\newblock Variants from the diverse virus population identified at
+  seroconversion of a clade A human immunodeficiency virus type 1-infected
+  woman have distinct biological properties.
+\newblock Journal of Virology.  1999; 73(7):5255--5264.
+
+\bibitem[{Pugach et~al.(2004)Pugach, Pavel and Kuhmann, Shawn E and Taylor,
+  Joann and Marozsan, Andre J and Snyder, Amy and Ketas, Thomas and Wolinsky,
+  Steven M and Korber, Bette T and Moore, John P}]{pugach2004prolonged}
+\textbf{\color{eLifeMediumGrey} Pugach P}, Kuhmann SE, Taylor J, Marozsan AJ,
+  Snyder A, Ketas T, Wolinsky SM, Korber BT, Moore JP.
+\newblock The prolonged culture of human immunodeficiency virus type 1 in
+  primary lymphocytes increases its sensitivity to neutralization by soluble
+  CD4.
+\newblock Virology.  2004; 321(1):8--22.
+
+\bibitem[{Pugach et~al.(2015)Pugach, Pavel and Ozorowski, Gabriel and Cupo,
+  Albert and Ringe, Rajesh and Yasmeen, Anila and de Val, Natalia and Derking,
+  Ronald and Kim, Helen J and Korzun, Jacob and Golabek, Michael and
+  others}]{pugach2015native}
+\textbf{\color{eLifeMediumGrey} Pugach P}, Ozorowski G, Cupo A, Ringe R,
+  Yasmeen A, de~Val N, Derking R, Kim HJ, Korzun J, Golabek M, et~al.
+\newblock A native-like SOSIP. 664 trimer based on an HIV-1 subtype B env gene.
+\newblock Journal of Virology.  2015; 89(6):3380--3395.
+
+\bibitem[{Rathore et~al.(2017)Rathore, Ujjwal and Saha, Piyali and
+  Kesavardhana, Sannula and Kumar, Aditya Arun and Datta, Rohini and
+  Devanarayanan, Sivasankar and Das, Raksha and Mascola, John R and
+  Varadarajan, Raghavan}]{rathore2017glycosylation}
+\textbf{\color{eLifeMediumGrey} Rathore U}, Saha P, Kesavardhana S, Kumar AA,
+  Datta R, Devanarayanan S, Das R, Mascola JR, Varadarajan R.
+\newblock Glycosylation of the core of the HIV-1 envelope subunit protein gp120
+  is not required for native trimer formation or viral infectivity.
+\newblock Journal of Biological Chemistry.  2017; 292(24):10197--10219.
+
+\bibitem[{Richman et~al.(2003)Richman, Douglas D and Wrin, Terri and Little,
+  Susan J and Petropoulos, Christos J}]{richman2003rapid}
+\textbf{\color{eLifeMediumGrey} Richman DD}, Wrin T, Little SJ, Petropoulos CJ.
+\newblock Rapid evolution of the neutralizing antibody response to HIV type 1
+  infection.
+\newblock Proceedings of the National Academy of Sciences.  2003;
+  100(7):4144--4149.
+
+\bibitem[{Risso et~al.(2014)Risso, Valeria A and Manssour-Triedo, Fadia and
+  Delgado-Delgado, Asunci{\'o}n and Arco, Rocio and Barroso-delJesus, Alicia
+  and Ingles-Prieto, Alvaro and Godoy-Ruiz, Raquel and Gavira, Jose A and
+  Gaucher, Eric A and Ibarra-Molero, Beatriz and others}]{risso2014mutational}
+\textbf{\color{eLifeMediumGrey} Risso VA}, Manssour-Triedo F, Delgado-Delgado
+  A, Arco R, Barroso-delJesus A, Ingles-Prieto A, Godoy-Ruiz R, Gavira JA,
+  Gaucher EA, Ibarra-Molero B, et~al.
+\newblock Mutational studies on resurrected ancestral proteins reveal
+  conservation of site-specific amino acid preferences throughout evolutionary
+  history.
+\newblock Molecular Biology and Evolution.  2014; 32(2):440--455.
+
+\bibitem[{Ronen et~al.(2015)Ronen, Keshet and Sharma, Amit and Overbaugh,
+  Julie}]{ronen2015hiv}
+\textbf{\color{eLifeMediumGrey} Ronen K}, Sharma A, Overbaugh J.
+\newblock HIV transmission biology: translation for HIV prevention.
+\newblock AIDS.  2015; 29:2219--2227.
+
+\bibitem[{Sagar et~al.(2006)Sagar, Manish and Wu, Xueling and Lee, Sandra and
+  Overbaugh, Julie}]{sagar2006human}
+\textbf{\color{eLifeMediumGrey} Sagar M}, Wu X, Lee S, Overbaugh J.
+\newblock Human immunodeficiency virus type 1 {V1-V2} envelope loop sequences
+  expand and add glycosylation sites over the course of infection, and these
+  modifications affect antibody neutralization sensitivity.
+\newblock Journal of Virology.  2006; 80(19):9586--9598.
+
+\bibitem[{Sander and Schneider(1991)Sander, Chris and Schneider,
+  Reinhard}]{sander1991database}
+\textbf{\color{eLifeMediumGrey} Sander C}, Schneider R.
+\newblock Database of homology-derived protein structures and the structural
+  meaning of sequence alignment.
+\newblock Proteins: Structure, Function, and Bioinformatics.  1991;
+  9(1):56--68.
+
+\bibitem[{Sanders et~al.(2013)Sanders, Rogier W and Derking, Ronald and Cupo,
+  Albert and Julien, Jean-Philippe and Yasmeen, Anila and de Val, Natalia and
+  Kim, Helen J and Blattner, Claudia and de la Pe{\~n}a, Alba Torrents and
+  Korzun, Jacob and others}]{sanders2013next}
+\textbf{\color{eLifeMediumGrey} Sanders RW}, Derking R, Cupo A, Julien JP,
+  Yasmeen A, de~Val N, Kim HJ, Blattner C, de~la Pe{\~n}a AT, Korzun J, et~al.
+\newblock A next-generation cleaved, soluble {HIV-1} Env trimer, {BG505
+  SOSIP.664 gp140}, expresses multiple epitopes for broadly neutralizing but
+  not non-neutralizing antibodies.
+\newblock PLoS Pathogens.  2013; 9(9):e1003618.
+
+\bibitem[{Sanders et~al.(2015)Sanders, Rogier W and Van Gils, Marit J and
+  Derking, Ronald and Sok, Devin and Ketas, Thomas J and Burger, Judith A and
+  Ozorowski, Gabriel and Cupo, Albert and Simonich, Cassandra and Goo, Leslie
+  and others}]{sanders2015hiv}
+\textbf{\color{eLifeMediumGrey} Sanders RW}, Van~Gils MJ, Derking R, Sok D,
+  Ketas TJ, Burger JA, Ozorowski G, Cupo A, Simonich C, Goo L, et~al.
+\newblock HIV-1 neutralizing antibodies induced by native-like envelope
+  trimers.
+\newblock Science.  2015; 349(6244):aac4223.
+
+\bibitem[{Sanders et~al.(2002)Sanders, Rogier W and Vesanen, Mika and Schuelke,
+  Norbert and Master, Aditi and Schiffner, Linnea and Kalyanaraman, Roopa and
+  Paluch, Maciej and Berkhout, Ben and Maddon, Paul J and Olson, William C and
+  others}]{sanders2002stabilization}
+\textbf{\color{eLifeMediumGrey} Sanders RW}, Vesanen M, Schuelke N, Master A,
+  Schiffner L, Kalyanaraman R, Paluch M, Berkhout B, Maddon PJ, Olson WC,
+  et~al.
+\newblock Stabilization of the soluble, cleaved, trimeric form of the envelope
+  glycoprotein complex of human immunodeficiency virus type 1.
+\newblock Journal of Virology.  2002; 76(17):8875--8889.
+
+\bibitem[{Shah et~al.(2015)Shah, Premal and McCandlish, David M and Plotkin,
+  Joshua B}]{shah2015contingency}
+\textbf{\color{eLifeMediumGrey} Shah P}, McCandlish DM, Plotkin JB.
+\newblock Contingency and entrenchment in protein evolution under purifying
+  selection.
+\newblock Proceedings of the National Academy of Sciences.  2015;
+  112(25):E3226--E3235.
+
+\bibitem[{Sharp and Hahn(2011)Sharp, Paul M and Hahn, Beatrice
+  H}]{sharp2011origins}
+\textbf{\color{eLifeMediumGrey} Sharp PM}, Hahn BH.
+\newblock Origins of {HIV} and the {AIDS} pandemic.
+\newblock Cold Spring Harbor Perspectives in Medicine.  2011; 1(1):a006841.
+
+\bibitem[{da~Silva et~al.(2010)da Silva, Jack and Coetzer, Mia and Nedellec,
+  Rebecca and Pastore, Cristina and Mosier, Donald E}]{da2010fitness}
+\textbf{\color{eLifeMediumGrey} da~Silva J}, Coetzer M, Nedellec R, Pastore C,
+  Mosier DE.
+\newblock Fitness epistasis and constraints on adaptation in a human
+  immunodeficiency virus type 1 protein region.
+\newblock Genetics.  2010; 185(1):293--303.
+
+\bibitem[{Simonich et~al.(2016)Simonich, Cassandra A and Williams, Katherine L
+  and Verkerke, Hans P and Williams, James A and Nduati, Ruth and Lee, Kelly K
+  and Overbaugh, Julie}]{simonich2016hiv}
+\textbf{\color{eLifeMediumGrey} Simonich CA}, Williams KL, Verkerke HP,
+  Williams JA, Nduati R, Lee KK, Overbaugh J.
+\newblock HIV-1 neutralizing antibodies with limited hypermutation from an
+  infant.
+\newblock Cell.  2016; 166(1):77--87.
+
+\bibitem[{Stamatakis(2014)Stamatakis, Alexandros}]{stamatakis2014raxml}
+\textbf{\color{eLifeMediumGrey} Stamatakis A}.
+\newblock RAxML version 8: a tool for phylogenetic analysis and post-analysis
+  of large phylogenies.
+\newblock Bioinformatics.  2014; 30:1312--1313.
+
+\bibitem[{Starcich et~al.(1986)Starcich, Bruno R and Hahn, Beatrice H and Shaw,
+  George M and McNeely, Paul D and Modrow, Susanne and Wolf, Hans and Parks,
+  Elizabeth S and Parks, Wade P and Josephs, Steven F and Gallo, Robert C and
+  others}]{starcich1986identification}
+\textbf{\color{eLifeMediumGrey} Starcich BR}, Hahn BH, Shaw GM, McNeely PD,
+  Modrow S, Wolf H, Parks ES, Parks WP, Josephs SF, Gallo RC, et~al.
+\newblock Identification and characterization of conserved and variable regions
+  in the envelope gene of HTLV-III/LAV, the retrovirus of AIDS.
+\newblock Cell.  1986; 45(5):637--648.
+
+\bibitem[{Starr et~al.(2017)Starr, Tyler N and Flynn, Julia M and Mishra, Parul
+  and Bolon, Daniel NA and Thornton, Joseph W}]{starr2017pervasive}
+\textbf{\color{eLifeMediumGrey} Starr TN}, Flynn JM, Mishra P, Bolon DN,
+  Thornton JW.
+\newblock Pervasive contingency and entrenchment in a billion years of Hsp90
+  evolution.
+\newblock bioRxiv.  2017; p. 189803.
+
+\bibitem[{Starr and Thornton(2016)Starr, Tyler N and Thornton, Joseph
+  W}]{starr2016epistasis}
+\textbf{\color{eLifeMediumGrey} Starr TN}, Thornton JW.
+\newblock Epistasis in protein evolution.
+\newblock Protein Science.  2016; 25(7):1204--1218.
+
+\bibitem[{Stewart-Jones et~al.(2016)Stewart-Jones, Guillaume BE and Soto,
+  Cinque and Lemmin, Thomas and Chuang, Gwo-Yu and Druz, Aliaksandr and Kong,
+  Rui and Thomas, Paul V and Wagh, Kshitij and Zhou, Tongqing and Behrens,
+  Anna-Janina and others}]{stewart2016trimeric}
+\textbf{\color{eLifeMediumGrey} Stewart-Jones GB}, Soto C, Lemmin T, Chuang GY,
+  Druz A, Kong R, Thomas PV, Wagh K, Zhou T, Behrens AJ, et~al.
+\newblock Trimeric {HIV-1-Env} Structures Define Glycan Shields from Clades {A,
+  B, and G}.
+\newblock Cell.  2016; 165(4):813--826.
+
+\bibitem[{de~Taeye et~al.(2015)de Taeye, Steven W and Ozorowski, Gabriel and de
+  la Pe{\~n}a, Alba Torrents and Guttman, Miklos and Julien, Jean-Philippe and
+  van den Kerkhof, Tom LGM and Burger, Judith A and Pritchard, Laura K and
+  Pugach, Pavel and Yasmeen, Anila and others}]{de2015immunogenicity}
+\textbf{\color{eLifeMediumGrey} de~Taeye SW}, Ozorowski G, de~la Pe{\~n}a AT,
+  Guttman M, Julien JP, van~den Kerkhof TL, Burger JA, Pritchard LK, Pugach P,
+  Yasmeen A, et~al.
+\newblock Immunogenicity of Stabilized HIV-1 Envelope Trimers with Reduced
+  Exposure of Non-neutralizing Epitopes.
+\newblock Cell.  2015; 163(7):1702--1715.
+
+\bibitem[{Tan et~al.(1997)Tan, Kemin and Liu, Jin-huan and Wang, Jia-huai and
+  Shen, Steven and Lu, Min}]{tan1997atomic}
+\textbf{\color{eLifeMediumGrey} Tan K}, Liu Jh, Wang Jh, Shen S, Lu M.
+\newblock Atomic structure of a thermostable subdomain of HIV-1 gp41.
+\newblock Proceedings of the National Academy of Sciences.  1997;
+  94(23):12303--12308.
+
+\bibitem[{Tien et~al.(2013)Tien, Matthew Z and Meyer, Austin G and Sydykova,
+  Dariya K and Spielman, Stephanie J and Wilke, Claus O}]{tien2013maximum}
+\textbf{\color{eLifeMediumGrey} Tien MZ}, Meyer AG, Sydykova DK, Spielman SJ,
+  Wilke CO.
+\newblock Maximum allowed solvent accessibilites of residues in proteins.
+\newblock PloS One.  2013; 8(11):e80635.
+
+\bibitem[{Verkerke et~al.(2016)Verkerke, Hans P and Williams, James A and
+  Guttman, Miklos and Simonich, Cassandra A and Liang, Yu and Filipavicius,
+  Modestas and Hu, Shiu-Lok and Overbaugh, Julie and Lee, Kelly
+  K}]{verkerke2016epitope}
+\textbf{\color{eLifeMediumGrey} Verkerke HP}, Williams JA, Guttman M, Simonich
+  CA, Liang Y, Filipavicius M, Hu SL, Overbaugh J, Lee KK.
+\newblock Epitope-independent purification of native-like envelope trimers from
+  diverse {HIV-1} isolates.
+\newblock Journal of Virology.  2016; 90(20):9471--9482.
+
+\bibitem[{Wagih(2017)Wagih, Omar}]{wagih2017ggseqlogo}
+\textbf{\color{eLifeMediumGrey} Wagih O}.
+\newblock ggseqlogo: a versatile {R} package for drawing sequence logos.
+\newblock Bioinformatics.  2017; 33(22):3645--3647.
+
+\bibitem[{Wang et~al.(2013)Wang, Wenbo and Nie, Jianhui and Prochnow, Courtney
+  and Truong, Carolyn and Jia, Zheng and Wang, Suting and Chen, Xiaojiang S and
+  Wang, Youchun}]{wang2013systematic}
+\textbf{\color{eLifeMediumGrey} Wang W}, Nie J, Prochnow C, Truong C, Jia Z,
+  Wang S, Chen XS, Wang Y.
+\newblock A systematic study of the N-glycosylation sites of HIV-1 envelope
+  protein on infectivity and antibody-mediated neutralization.
+\newblock Retrovirology.  2013; 10(1):1.
+
+\bibitem[{Wang et~al.(2002)Wang, Xiaojun and Minasov, George and Shoichet,
+  Brian K}]{wang2002evolution}
+\textbf{\color{eLifeMediumGrey} Wang X}, Minasov G, Shoichet BK.
+\newblock Evolution of an antibiotic resistance enzyme constrained by stability
+  and activity trade-offs.
+\newblock Journal of molecular biology.  2002; 320(1):85--95.
+
+\bibitem[{Waterston et~al.(2002)Waterston, RH and Lindblad-Toh, K and Birney, E
+  and Rogers, J and Abril, JF and Agarwal, P and Agarwala, R and Ainscough, R
+  and Alexandersson, M and An, P and others}]{waterston2002initial}
+\textbf{\color{eLifeMediumGrey} Waterston R}, Lindblad-Toh K, Birney E, Rogers
+  J, Abril J, Agarwal P, Agarwala R, Ainscough R, Alexandersson M, An P, et~al.
+\newblock Initial sequencing and comparative analysis of the mouse genome.
+\newblock Nature.  2002; 420(6915):520--562.
+
+\bibitem[{Wei et~al.(2003)Wei, Xiping and Decker, Julie M and Wang, Shuyi and
+  Hui, Huxiong and Kappes, John C and Wu, Xiaoyun and Salazar-Gonzalez, Jesus F
+  and Salazar, Maria G and Kilby, J Michael and Saag, Michael S and
+  others}]{wei2003antibody}
+\textbf{\color{eLifeMediumGrey} Wei X}, Decker JM, Wang S, Hui H, Kappes JC, Wu
+  X, Salazar-Gonzalez JF, Salazar MG, Kilby JM, Saag MS, et~al.
+\newblock Antibody neutralization and escape by HIV-1.
+\newblock Nature.  2003; 422(6929):307--312.
+
+\bibitem[{Weissenhorn et~al.(1997)Weissenhorn, Winfried and Dessen, A and
+  Harrison, SC and Skehel, JJ and Wiley, DC}]{weissenhorn1997atomic}
+\textbf{\color{eLifeMediumGrey} Weissenhorn W}, Dessen A, Harrison S, Skehel J,
+  Wiley D.
+\newblock Atomic structure of the ectodomain from HIV-1 gp41.
+\newblock Nature.  1997; 387(6631):426.
+
+\bibitem[{White et~al.(2010)White, Tommi A and Bartesaghi, Alberto and Borgnia,
+  Mario J and Meyerson, Joel R and de la Cruz, M Jason V and Bess, Julian W and
+  Nandwani, Rachna and Hoxie, James A and Lifson, Jeffrey D and Milne,
+  Jacqueline LS and others}]{white2010molecular}
+\textbf{\color{eLifeMediumGrey} White TA}, Bartesaghi A, Borgnia MJ, Meyerson
+  JR, de~la Cruz MJV, Bess JW, Nandwani R, Hoxie JA, Lifson JD, Milne JL,
+  et~al.
+\newblock Molecular architectures of trimeric SIV and HIV-1 envelope
+  glycoproteins on intact viruses: strain-dependent variation in quaternary
+  structure.
+\newblock PLoS Pathogens.  2010; 6(12):e1001249.
+
+\bibitem[{Wilen et~al.(2011)Wilen, Craig B and Parrish, Nicholas F and Pfaff,
+  Jennifer M and Decker, Julie M and Henning, Elizabeth A and Haim, Hillel and
+  Petersen, Josiah E and Wojcechowskyj, Jason A and Sodroski, Joseph and
+  Haynes, Barton F and others}]{wilen2011phenotypic}
+\textbf{\color{eLifeMediumGrey} Wilen CB}, Parrish NF, Pfaff JM, Decker JM,
+  Henning EA, Haim H, Petersen JE, Wojcechowskyj JA, Sodroski J, Haynes BF,
+  et~al.
+\newblock Phenotypic and immunologic comparison of clade {B}
+  transmitted/founder and chronic {HIV-1} envelope glycoproteins.
+\newblock Journal of Virology.  2011; 85(17):8514--8527.
+
+\bibitem[{Worobey et~al.(2008)Worobey, Michael and Gemmel, Marlea and Teuwen,
+  Dirk E and Haselkorn, Tamara and Kunstman, Kevin and Bunce, Michael and
+  Muyembe, Jean-Jacques and Kabongo, Jean-Marie M and Kalengayi, Rapha{\"e}l M
+  and Van Marck, Eric and others}]{worobey2008direct}
+\textbf{\color{eLifeMediumGrey} Worobey M}, Gemmel M, Teuwen DE, Haselkorn T,
+  Kunstman K, Bunce M, Muyembe JJ, Kabongo JMM, Kalengayi RM, Van~Marck E,
+  et~al.
+\newblock Direct evidence of extensive diversity of {HIV-1} in {Kinshasa} by
+  1960.
+\newblock Nature.  2008; 455(7213):661--664.
+
+\bibitem[{Wu et~al.(2014)Wu, Nicholas C and Young, Arthur P and Al-Mawsawi,
+  Laith Q and Olson, C Anders and Feng, Jun and Qi, Hangfei and Chen, Shu-Hwa
+  and Lu, I-Hsuan and Lin, Chung-Yen and Chin, Robert G and
+  others}]{wu2014high}
+\textbf{\color{eLifeMediumGrey} Wu NC}, Young AP, Al-Mawsawi LQ, Olson CA, Feng
+  J, Qi H, Chen SH, Lu IH, Lin CY, Chin RG, et~al.
+\newblock High-throughput profiling of influenza A virus hemagglutinin gene at
+  single-nucleotide resolution.
+\newblock Scientific Reports.  2014; 4:4942.
+
+\bibitem[{Wu et~al.(2006)Wu, Xueling and Parast, Adam B and Richardson, Barbra
+  A and Nduati, Ruth and John-Stewart, Grace and Mbori-Ngacha, Dorothy and
+  Rainwater, Stephanie MJ and Overbaugh, Julie}]{wu2006neutralization}
+\textbf{\color{eLifeMediumGrey} Wu X}, Parast AB, Richardson BA, Nduati R,
+  John-Stewart G, Mbori-Ngacha D, Rainwater SM, Overbaugh J.
+\newblock Neutralization escape variants of human immunodeficiency virus type 1
+  are transmitted from mother to infant.
+\newblock Journal of Virology.  2006; 80(2):835--844.
+
+\bibitem[{Yang(1994)Yang, Ziheng}]{yang1994maximum}
+\textbf{\color{eLifeMediumGrey} Yang Z}.
+\newblock Maximum likelihood phylogenetic estimation from DNA sequences with
+  variable rates over sites: approximate methods.
+\newblock Journal of Molecular Evolution.  1994; 39(3):306--314.
+
+\bibitem[{Yang et~al.(2000)Yang, Ziheng and Nielsen, Rasmus and Goldman, Nick
+  and Pedersen, Anne-Mette Krabbe}]{yang2000codon}
+\textbf{\color{eLifeMediumGrey} Yang Z}, Nielsen R, Goldman N, Pedersen AMK.
+\newblock Codon-substitution models for heterogeneous selection pressure at
+  amino acid sites.
+\newblock Genetics.  2000; 155(1):431--449.
+
+\bibitem[{Yuste et~al.(2004)Yuste, Elo{\'\i}sa and Reeves, Jacqueline D and
+  Doms, Robert W and Desrosiers, Ronald C}]{yuste2004modulation}
+\textbf{\color{eLifeMediumGrey} Yuste E}, Reeves JD, Doms RW, Desrosiers RC.
+\newblock Modulation of Env content in virions of simian immunodeficiency
+  virus: correlation with cell surface expression and virion infectivity.
+\newblock Journal of Virology.  2004; 78(13):6775--6785.
+
+\bibitem[{Zanini and Neher(2013)Zanini, Fabio and Neher, Richard
+  A}]{zanini2013quantifying}
+\textbf{\color{eLifeMediumGrey} Zanini F}, Neher RA.
+\newblock Quantifying selection against synonymous mutations in HIV-1 env
+  evolution.
+\newblock Journal of Virology.  2013; 87(21):11843--11850.
+
+\bibitem[{Zolla-Pazner and Cardozo(2010)Zolla-Pazner, Susan and Cardozo,
+  Timothy}]{zolla2010structure}
+\textbf{\color{eLifeMediumGrey} Zolla-Pazner S}, Cardozo T.
+\newblock Structure--function relationships of HIV-1 envelope sequence-variable
+  regions refocus vaccine design.
+\newblock Nature Reviews Immunology.  2010; 10(7):527--535.
+
+\end{thebibliography}
 

### 2018-02-28 paper/manuscript.tex
submitted _eLife_ revision
changes requested by _eLife_ editorial staff
Convert Figure supplements with tabular data
into source data.

Add references to all supplements in figure legends.

@@
 The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
-Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIGSUPP[mutfreqs]{mutfreqs}).
+Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}).
 More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
@@
 }
-\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
-{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs}.
-Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
-We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
-We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
-{\input{figures/avgmutfreqs}}
 \figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
@@
 \figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
+\figdata{\label{figdata:mutfreqs} Average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs} are in \texttt{avgmutfreqs.csv}.
+There is only one \emph{DNA} sample for BF520 which is listed three times with each BF520 replicate.
+We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} minus \emph{DNA}, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} minus the \emph{virus}.
+We use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection.}
 \end{figure}
@@
 The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (\FIG{mutfreqs}A).
-This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIGSUPP[mutfreqs]{mutfreqs}).
-The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIGSUPP[mutfreqs]{mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
+This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}).
+The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions.
 The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
@@
 Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
-The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B.
+The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B, and the preferences themselves are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
 These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.
@@
 Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
-Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIGSUPP[mutfreqs]{mutfreqs} shows that purifying selection better purged stop codons in BG505).
+Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs} shows that purifying selection better purged stop codons in BG505).
 In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.
@@
 The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
+The numerical values of the preferences are in \FIG{BG505prefs}-source~data~\ref{figdata:BG505prefs}, the mapping from sequential to HXB2 numbering is in \FIG{BG505prefs}-source~data~\ref{figdata:BG505numbering}, and the $\omega_r$ values are in \FIG{BG505prefs}-source~data~\ref{figdata:BG505omega}.
 }
-\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
-\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\figdata{\label{figdata:BG505prefs}The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
+\figdata{\label{figdata:BG505numbering}The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
+\figdata{\label{figdata:BG505omega}The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
+The numerical values of the preferences are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520prefs}, the mapping from sequential to HXB2 numbering is in \FIG{BF520prefs}-source~data~\ref{figdata:BF520numbering}, and the $\omega_r$ values are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520omega}.
 }
-\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
-\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
+\figdata{\label{figdata:BF520prefs}The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
+\figdata{\label{figdata:BF520numbering}The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
+\figdata{\label{figdata:BF520omega}The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
-{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites.
+{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites (see \FIG{prefsdist}-source~data~\ref{figdata:prefsdist} for numerical values).
 The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
@@
 The plots show the magnitudes of the shifts among structurally resolved sites that have substituted between BG505 and BF520, the non-substituted sites that physically contact a substitution in the indicated structure(s) (any non-hydrogen atom within 3.5\angstrom), and all other sites.
-\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a borderline-significant tendency of significantly shifted sites to have substituted.
+\FIG{shiftfeatures}-source~data~\ref{figdata:subs_shifts} shows that there is a borderline-significant tendency of significantly shifted sites to have substituted.
 All plots only show sites that are structurally resolved in the indicated structure(s). 
@@
 $P$ values were calculated using the Mann-Whitney $U$ test.
+\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundle} and \FIGSUPP[shiftfeatures]{dynamic_clusters} zoom in on some relevant clusters of sites.
 }
-\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
-{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
-However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
-{\input{figures/shifts_vs_subs_table}}
 
@@
 {\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
+
+\figdata{\label{figdata:subs_shifts} The sites of significant shifts in \FIG{prefsdist}B are somewhat more likely to have substituted between BG505 and BF520.
+This association is borderline statistically significant, with $P = 0.055$ using a Fisher's exact test on the contingency table in \texttt{shifts\_vs\_subs\_table.csv}.}
+
 \end{figure}
@@
 
-There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIGSUPP[shiftfeatures]{subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
+There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIG{shiftfeatures}-source~data~\ref{figdata:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
 The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
@@
 One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to white to red based on the strength of evidence that sites evolve slower than expected (blue), as expected (white) or faster than expected (red) given the BG505 experiments.
-%We used a color scale that only ranges between significance values of $-6 < \log_{10}Q < 6$, which decreases the resolution between a small number of sites with extreme values, but provides greater resolution among other sites.
 Sites for which we lack $\omega_r$ estimates are colored black.
@@
 All analyses used the phylogenetic tree in \FIG{tree}.
+The $\omega_r$ and $Q$-values are in \FIG{divsel}-source~data~\ref{figdata:omega}.
 }
@@
 }{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
-\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
+\figdata{\label{figdata:omega}The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
 \end{figure}
@@
 Of course, we have done our best to perform the experiments for BG505 and BF520 equivalently, but because these are different viruses with different growth rates, it is impossible to exactly match the experimental stringencies.
-This can be seen in \FIGSUPP[mutfreqs]{mutfreqs}, which shows that stop codons were more depleted for BG505 than BF520, indicating that selection in our experiments was more stringent for BG505.
+This can be seen in \FIG{mutfreqs}-source~data~\ref{figdata:mutfreqs}, which shows that stop codons were more depleted for BG505 than BF520, indicating that selection in our experiments was more stringent for BG505.
 

### 2018-02-28 paper/manuscript_for_final_submission/manuscript_nofigs_embeddedtabsrefs.tex
changes requested by _eLife_ editorial staff
Convert Figure supplements with tabular data
into source data.

Add references to all supplements in figure legends.

@@
 
+% do not show figures
+\renewcommand\includegraphics[2][]{}
+
 \usepackage{lipsum} % Required to insert dummy text
@@
 \begin{figure}
+\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
 \caption{\label{fig:tree}
@@
 The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
-{}
+{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
 \figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
@@
 \begin{figure}
-%\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
+\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
 \caption{\label{fig:dms_schematic}
@@
 Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
-}{}%\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
+}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
 \end{figure}
@@
 \begin{figure}
+%\begin{minipage}[t]{0.34\textwidth}
+%{\bf \Large A} \\ 
+%\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
+%\centerline{\small BG505}
+%\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
+%\centerline{\small BF520}
+%\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
+%\end{minipage}
+%\begin{minipage}[t]{0.01\textwidth}
+%\end{minipage}
+%\begin{minipage}[t]{0.65\textwidth}
+%{\bf \Large B} \\
+%\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
+%\end{minipage}
 \caption{\label{fig:mutfreqs}
@@
 }
-\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
-{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs}.
-Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
-We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
-We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
-{}
 \figsupp[\label{figsupp:contour} Correlations plotted on a contour rather than a scatter plot.]
@@
 This is expected, since most amino acid mutations at most sites will be strongly deleterious, and so have low preference.
-}{}%\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
+}{\includegraphics[width=\textwidth]{figures/allprefscorr_contour.pdf}}
 \figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
+\figdata{\label{figdata:mutfreqs} Average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs} are in \texttt{avgmutfreqs.csv}.
+There is only one \emph{DNA} sample for BF520 which is listed three times with each BF520 replicate.
+We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} minus \emph{DNA}, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} minus the \emph{virus}.
+We use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection.}
 \end{figure}
@@
 \begin{fullwidth}
-%{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
+{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
 \caption{\label{fig:BG505prefs}
@@
 \begin{fullwidth}
-%{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
+{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
 \caption{\label{fig:BF520prefs}
@@
 This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
+The numerical values of the preferences are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520prefs}, the mapping from sequential to HXB2 numbering is in \FIG{BF520prefs}-source~data~\ref{figdata:BF520numbering}, and the $\omega_r$ values are in \FIG{BF520prefs}-source~data~\ref{figdata:BF520omega}.
 }
-\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
-\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
-\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
+\figdata{\label{figdata:BF520prefs}The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
+\figdata{\label{figdata:BF520numbering}The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
+\figdata{\label{figdata:BF520omega}The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BF520\_omegabysite.tsv}.}
 \end{fullwidth}
@@
 \begin{fullwidth}
+%{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
+%\hspace{0.01\textwidth} 
+%\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
+%\hspace{0.03\textwidth}
+%\includegraphics[clip=true, trim=0in 0in 0in 0.4in, width=0.65\textwidth]{figures/distance_distribution.pdf}
+%\vspace{0.15in}
+%\\ 
+%{\bf \Large C} \\
+%\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
 \caption{\label{fig:prefsdist}
@@
 \begin{figure}
+%{\bf \Large A} \hspace{0.6\textwidth} {\bf \Large B}  \\
+%\includegraphics[width=0.57\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
+%\hspace{0.06\textwidth}
+%\includegraphics[width=0.32\textwidth]{figures/rsa_vs_shifts.pdf}
+%\\
+ %{\bf \Large C} \hspace{0.45\textwidth} {\bf \Large D} \\
+%\includegraphics[width=0.45\textwidth]{figures/shifts_proximity.pdf}
+%\hspace{0.02\textwidth}
+%\includegraphics[width=0.53\textwidth]{figures/conservation_vs_shifts.pdf} 
 \caption{\label{fig:shiftfeatures}
@@
 $P$ values were calculated using the Mann-Whitney $U$ test.
+\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundle} and \FIGSUPP[shiftfeatures]{dynamic_clusters} zoom in on some relevant clusters of sites.
 }
-\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
-{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
-However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
-{}
 
@@
 }
-{}%\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}
+{\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}
 
@@
 }
-{}%\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
+{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
+
+\figdata{\label{figdata:subs_shifts} The sites of significant shifts in \FIG{prefsdist}B are somewhat more likely to have substituted between BG505 and BF520.
+This association is borderline statistically significant, with $P = 0.055$ using a Fisher's exact test on the contingency table in \texttt{shifts\_vs\_subs\_table.csv}.}
+
 \end{figure}
@@
 \begin{figure}
-%\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
+\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
 \caption{\label{fig:entrenchment}
@@
 \begin{figure}
-%\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
+\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
 \caption{\label{fig:divsel}
@@
 As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
-}{}%\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
+}{\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
 \figsupp[\label{figsupp:glycans}Amino-acid preferences and alignment frequencies for glycosylation motifs.]
@@
 We suspect this is because many glycans serve as a shield against immunity in nature---a function that is not required in cell culture.}
-{}%\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
+{\includegraphics[width=\textwidth]{figures/glycan_prefs.pdf}}
 \figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
@@
 The $Q$-value indicated is the maximum of that for the two Envs.
-}{}%\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
-\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
+}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
+\figdata{\label{figdata:omega}The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
 \end{figure}

### 2018-03-22 paper/manuscript.tex
added results on cell-line validation

@@
 We conducted the low MOI viral passage illustrated in \FIG{dms_schematic}A in SupT1.CCR5 cells~\citep[obtained from Dr. James Hoxie;][]{boyd2015mutations}.
+The SupT1.CCR5 cells tested negative for mycoplasma. 
+The SupT1.CCR5 cell line was previously created by engineering the parental SupT1 cell line to express CCR5~\citep{boyd2015mutations}. 
+We used antibody staining followed by flow cytometry to validate that our stock of SupT1.CCR5 cells expressed CCR5, CXCR4, and CD4. 
+There is no validated STR profile for SupT1.CCR5 cells. 
+However, we performed STR profiling on our stock of cells and compared the results to the ATCC SupT1 (ATCC \#CRL-1942) reference profile. 
+We found that 11 of 14 alleles plus both amelogenin alleles matched the reference, with no additional mismatched alleles ins the SupT1.CCR5 profile. 
+Given the known instability of lymphoma cell lines~\citep{inoue2000frequent}, this level of identity suggests that the SupT1.CCR5 cells are indeed related to the parental SupT1 cells~\citep{capes2013match}.
+
 During this passage, cells were maintained in R10 media, which has the same composition as the D10 described above, except RPMI-1640 (GE Healthcare Life Sciences, SH30255.01) is used in the place of DMEM.

### 2018-03-22 paper/manuscript_for_final_submission/manuscript_nofigs_embeddedtabsrefs.tex
reference later GitHub commit
added results on cell-line validation

@@
 
+\bibitem[{Capes-Davis et~al.(2013)Capes-Davis, Amanda and Reid, Yvonne A and
+  Kline, Margaret C and Storts, Douglas R and Strauss, Ethan and Dirks, Wilhelm
+  G and Drexler, Hans G and MacLeod, Roderick AF and Sykes, Gregory and Kohara,
+  Arihiro and others}]{capes2013match}
+\textbf{\color{eLifeMediumGrey} Capes-Davis A}, Reid YA, Kline MC, Storts DR,
+  Strauss E, Dirks WG, Drexler HG, MacLeod RA, Sykes G, Kohara A, et~al.
+\newblock Match criteria for human cell line authentication: where do we draw
+  the line?
+\newblock International Journal of Cancer.  2013; 132(11):2510--2519.
+
 \bibitem[{Chakrabarti et~al.(1989)Chakrabarti, L and Emerman, M and Tiollais, P
@@
   landscape of HIV envelope''.
-\newblock Github.  2018; commit ec358fd.
+\newblock Github.  2018; commit 2355e76.
 \newblock
@@
 \newblock Nature.  2014; 515(7525):138--142.
+
+\bibitem[{Inoue et~al.(2000)Inoue, Kaoru and Kohno, Takashi and Takakura,
+  Satoshi and Hayashi, Yasuhide and Mizoguchi, Hideaki and Yokota,
+  Jun}]{inoue2000frequent}
+\textbf{\color{eLifeMediumGrey} Inoue K}, Kohno T, Takakura S, Hayashi Y,
+  Mizoguchi H, Yokota J.
+\newblock Frequent microsatellite instability and BAX mutations in T cell acute
+  lymphoblastic leukemia cell lines.
+\newblock Leukemia Research.  2000; 24(3):255--262.
 
